### 2020-03-25 ms/SI.tex
first commit

\documentclass[a4paper,11pt]{scrartcl}
\usepackage[utf8x]{inputenc}			
\usepackage[T1]{fontenc}				
\usepackage[english]{babel}			
%\usepackage{ulem}
\usepackage{color}						
\usepackage{amsmath,amstext,amssymb,amsfonts,amsthm,mathrsfs,chemarrow,stackrel,dsfont}   % Math packages
\usepackage{graphicx} 					
\usepackage{subcaption}					
%\usepackage{cite}	
\usepackage[round]{natbib} % Bio citation format

\usepackage{microtype}
\usepackage{array}
\usepackage{booktabs}

\usepackage{lineno}

\usepackage{xr-hyper}
\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}

%%% END HELPER CODE ------------------------------------
\myexternaldocument{main} %% add link to the maintext file of your revised manuscript in the file list (new file, from another project).

\usepackage[pdftex, breaklinks, colorlinks, citecolor=black, urlcolor=black, linkcolor=black]{hyperref}

%\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
%\usepackage{FiraSans}
% \florence Ah you do not like my typefaces ;-)
\usepackage{soul} % FD to strike out, highlight, ...

% Packages to comment out before submitting!
\usepackage[textsize = scriptsize, backgroundcolor = white, linecolor = black]{todonotes}
\usepackage{showlabels}

%\numberwithin{equation}{section}
%\setlength{\parindent}{0pt}
\allowdisplaybreaks[2]

\newcolumntype{P}[1]{>{\centering\arraybackslash}m{#1}}     % Table formatting
\renewcommand{\arraystretch}{1.1}                           % Table formatting

\definecolor{darkgreen}{rgb}{0.0, 0.42, 0.24}
\definecolor{darkblue}{rgb}{0.0, 0.0, 0.55}
\definecolor{change}{rgb}{0.5,0.,0.6}

%%%%%% our commenting commands
\newcommand{\florence}[1]{\textcolor{purple}{(#1)}} % Changed red to purple
\newcommand{\francois}[1]{\textcolor{darkblue}{(#1)}}
\newcommand{\hildegard}[1]{\textcolor{darkgreen}{(#1)}}
\newcommand{\pete}[1]{\textcolor{orange}{(#1)}}
\newcommand{\chg}[1]{\textcolor{change}{#1}}

%%%%%% Putting an S before the equation, section and figure numbering
\renewcommand\theequation{S\arabic{equation}}
\renewcommand\thesection{S\arabic{section}}
\renewcommand\thefigure{S\arabic{figure}}

%opening
\title{Supplementary Information: ``The effect of habitat choice on evolutionary rescue in subdivided populations''}
\author{Peter Czuppon, Fran\c{c}ois Blanquart, Hildegard Uecker,\\ Florence D\'{e}barre}
\date{\today}

\begin{document}

\maketitle

\tableofcontents
\newpage

%Adjusted line numbers (with S before number) but does not work with the cross-referencing... Edit: cross-referencing does not work because of transitivity! It seems you can only refer to ONE document anyway and not to multiple documents. This means that we need to include the linenumbers of the Supplement by hand anyway. Therefore, I changed the linenumbers to S\linenumber. Edit2: I found an easier solution
%\renewcommand\makeLineNumber{\ifodd\value{linenumber} \else\hss\linenumberfont S\thelinenumber\hskip\linenumbersep \fi}
\renewcommand\thelinenumber{S\arabic{linenumber}}
\linenumbers{}
\modulolinenumbers[2]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Model dynamics %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Deriving the model dynamics}
\linelabel{test}In this section we provide the mathematical \chg{details} of the model that is verbally described in the main text. We start by deriving the population dynamics of the wild type \chg{when alone} in both patch types. This will allow us to compute the local growth rate of the mutant \chg{when rare} in old-habitat patches, $a_{\text{old}}$.

Before we go into the details of the computation, we recall the form of the dispersal rates. For a wild-type individual to disperse to a new-habitat patch, this probability is given by

\begin{equation}\label{Seq:dispersal_rates_old}
    m_w^{\text{new}} = m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, ,
\end{equation}
%
with $m$ denoting the emigration probability, $f_{\text{old}}$ the frequency of old-habitat patches and $\widehat{\pi}_w$ the transformed wild-type bias towards old-habitat patches.
Analogously, the probability for a wild-type individual \chg{to emigrate and} to move to an old-habitat patch reads
%
\begin{equation}\label{Seq:dispersal_rate_new}
    m_w^{\text{old}} = m\, \frac{\widehat{\pi}_w f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, .
\end{equation}
\\

All the subsequent computations can be checked with a symbolic programming language (e.g. \textit{Mathematica}). A \textit{Mathematica} notebook is deposited on Gitlab\footnote{\url{https://gitlab.com/pczuppon/evolutionary\_rescue\_and\_dispersal}}.

\subsection*{Stationary wild-type population sizes}
We derive the (deterministic) stationary population sizes of the wild-type in both patch types. These values are denoted by $\widehat{N}_w^{k}$, where $k$ corresponds to the patch type (old or new). In old-patch habitats we assume that the population is always at its carrying capacity. Therefore, we have $\widehat{N}_w^{\text{old}} = K_{\text{old}}$.

In new-habitat patches, the stationary value is given by the solution of the following equation:

 \begin{equation}\label{Seq:wt_deme2}
	\begin{aligned}
		\widehat{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)(1-r)\widehat{N}_w^{\text{new}} \\
		&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} (1-r)  \widehat{N}_w^{\text{old}}\\
		 &= \left(1-m+m_w^{\text{new}} \right)(1-r)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \left(\frac{f_{\text{old}}}{1-f_{\text{old}}}\right) (1-r)  K_{\text{old}},
	\end{aligned}
\end{equation}
%
which simplifies to
%
\begin{equation}
	\widehat{N}_w^{\text{new}} = \frac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\ .
\end{equation}
%
Since we assume density regulation, this value cannot be larger than $K_{\text{new}}$, the carrying capacity of new-habitat patches. Hence, we find
\begin{equation}
	\widehat{N}_w^{\text{new}} =\left\{ \begin{array}{ll} 
		K_{\text{new}}, & \text{if }\frac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\geq K_{\text{new}}; \\
		\frac{m (1-r) f_{\text{old}} K_{\text{old}}
		}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}, & \text{otherwise}.
	\end{array}\right.
\end{equation}

\subsection*{Wild-type population sizes after dispersal}
In order to explicitly compute the growth rate of the mutant in old-habitat patches, we need an analytical expression for the number of wild-type individuals after the dispersal step. Later on, in the approximation of the probability of adaptation, we also use the number of wild-type individuals in new-habitat patches before reproduction. We denote these values by $\widetilde{N}_w^k$, where $k$ corresponds to the habitat type (old or new). These are given as the solutions to the following equations

\begin{equation}
	\begin{aligned}
		\widetilde{N}_w^{\text{old}} &= \left(1-m + m\, \frac{\widehat{\pi}_w f_{\text{old}}}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})}\,  \frac{f_{\text{old}}M}{f_{\text{old}}M} \right) \widehat{N}_w^{\text{old}} + m_w^{\text{old}} \frac{(1-f_{\text{old}})M}{f_{\text{old}}M}\widehat{N}_w^{\text{new}}\ , \\
		\widetilde{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{1-f_{\text{old}}+\pi f_{\text{old}}} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \widehat{N}_w^{\text{old}}\ .
	\end{aligned}
\end{equation}
Using $\widehat{N}_w^{\text{old}}=K_{\text{old}}$ and in the case of $\widehat{N}_w^{\text{new}} = K_{\text{new}}$ this reduces to
\begin{equation}\label{Seq:wt_eq1}
	\begin{aligned}
		\widetilde{N}_w^{\text{old}} &= \frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \\ 
		\widetilde{N}_w^{\text{new}}&= \frac{(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w (1-m))K_{\text{new}} + m f_{\text{old}} K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\ .
	\end{aligned}
\end{equation}
%
If on the other hand $\widehat{N}_w^{\text{new}}<K_{\text{new}}$ holds, we obtain
\begin{equation}\label{Seq:wt_eq2}
	\begin{aligned}
		\widetilde{N}_w^{\text{old}} &= \left(1-m+m_w^{\text{old}} \right)K_{\text{old}}\\
		&\qquad \qquad + \left(m_w^{\text{old}} \frac{m (1-r) f_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\left(\frac{1-f_{\text{old}}}{f_{\text{old}}}\right)\right) K_{\text{old}} \\
		%&= \left(1-m_1+\frac{m_1 m_2 (1-r)}{(1-(1-m_2)(1-r))}\right) K \\ 
		&= \frac{m \widehat{\pi}_w f_{\text{old}} + r (1-m) (1-f_{\text{old}} (1-\widehat{\pi}_w))}{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} K_{\text{old}}\ \\
		&= \left(1 - \frac{r m (1-f_{\text{old}}) }{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} \right) K_{\text{old}}\, ,\\
		\widetilde{N}_w^{\text{new}} &= \left(1-m+m_w^{\text{new}}\right) \frac{m (1-r) f_{\text{old}} }{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\, K_{\text{old}}\\
		&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}}}{(1-f_{\text{old}})} K_{\text{old}}\\
		%&= \left(\frac{(1-m_2) m_1 (1-r) f }{(1-f)(1-\left(1-m_2 \right)(1-r))} + \frac{m_1 f}{(1-f)}\right) K\\
		&= \frac{m f_{\text{old}}}{m \widehat{\pi}_w f_{\text{old}} + r(1- f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}  (1-m)) }\, K_{\text{old}}\, .
	\end{aligned}
\end{equation}

\subsection*{Wild-type population sizes during the environmental change}
\chg{Lastly, compute the (deterministic) wild-type population size over time during the environmental change. This value is used in the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue} in the main text, more precisely to estimate the number of rescue mutants that appear during the deterioration of patches.}

\chg{If a patch deteriorates its current population size is given by the carrying capacity of the old habitat, $K_{\text{old}}$. Subsequently, it is reduced by $(1-r)$, the fecundity of a single wild-type individual in a new-habitat patch. That is, neglecting dispersal for the moment, in generation $j$ after the degradation of a patch, we would have 
\begin{equation}
    N_w^{\text{new}}(j) = K_{\text{old}}(1-r)^j
\end{equation}
wild-type individuals in this patch. Including dispersal between the patches then results in the following number of wild-type individuals in a patch $i$ at time $j$, given that there are $k-1$ other new-habitat patches
\begin{equation}
    N_w^{i,k}(j) = (1-r)\left((1-m)N_w^{i,k}(j-1) + \frac{m_w^{\text{new}}}{k} (M-k) K_{\text{old}} + \frac{m_w^{\text{new}}}{k} \sum_{l=1}^k N_w^{l,k}(j-1) \right)\ .
\end{equation}
The first term represents the remaining individuals after emigration, $(1-m)$, the second and third term are immigrants from old- and new-habitat patches (distributed equally among the $k$ new-habitat patches), respectively. 
}

\subsection*{The local growth rate $a_{\text{old}}$}
As stated in eq.~\eqref{eq:sold} in the main text, we define the growth rate of a single mutant in the old habitat by
\begin{equation}
    1 + a_{\text{old}} = K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}_w^{\text{old}}}\, .    
\end{equation}
%
Having computed the number of wild-type individuals after dispersal in an old-habitat patch, $\widetilde{N}_w^{\text{old}}$, we can write the growth rate as follows

\begin{equation}\label{Seq:s_old}
    a_{\text{old}} = \left\{ \begin{array}{ll}
        \frac{\omega_m K_{\text{old}}}{\omega_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
        \frac{\omega_m}{\omega_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}. 
    \end{array}
    \right.
\end{equation}

\subsection*{The local growth rate $a_{\text{new}}$}
\chg{In new-habitat patches we assume that during the relevant phase of establishment of a rare mutant the carrying capacity $K_{\text{new}}$ is not reached. Then the growth rate of the mutant in these patches, $a_{\text{new}}$, is not affected by the different dispersal schemes.}

\chg{Of course there may be parameter configurations, typically high emigration rates $m$ and a bias of the wild type towards the new habitat ($\pi_w < 0$), where our assumption of density-independent reproduction is violated. Then our approximation and the numerical solution of eq.~\eqref{eq:ext_prob} in the main text strongly deviate from the simulation results (e.g. the New-New dispersal scheme in Fig.~\ref{fig:vary_m_est}(d)).}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Establishment probability %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Approximation of the establishment probability}
We compute the survival probability \chg{of the lineage} of a single mutant starting either in an old- or in a new-habitat patch. We call this probability the establishment probability because it implies the successful establishment of a mutant population within the meta-population. It is denoted by $\varphi_k$ where the index $k$ indicates the initial habitat type of the mutant (old or new).

Our method is the same as used in \citet{tomasini_2018} with the exception that our growth rate in old-habitat patches, $a_{\text{old}}$, depends on the demography of the population. The method in general is based on the theory of multi-type branching processes, cf.~Chapter~5.5 in \citet{haccou_book}. For a detailed application of the theory we refer to the Supplementary Information of \citet{tomasini_2018}.

We start by recalling the mean reproduction matrix of a mutant that gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides. It is given as (see also eq.~\eqref{eq:mean_repro} in the main text)

\begin{equation}\label{Seq:mean_repro}
	\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
		\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
		\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
		}\ ,
\end{equation}
where the rows denote the parent locations, and the columns the patch type of the offspring.

Our goal is to apply Theorem 5.6 from \citet{haccou_book} which states that for a slightly super-critical branching process, \chg{i.e. where the survival probability is slightly above zero}, the establishment probability can be expressed in terms of the largest eigenvalue $\rho$ and the corresponding left- and right-eigenvectors of the mean reproduction matrix \chg{$\mathcal{M}$}, denoted by $u$ and $v$, respectively. The eigenvectors should be normalized in the following way: $u_1+u_2 = 1$ and $\sum_{i=1}^2 u_i v_i = 1$. The establishment probabilities are then given by 
\begin{equation}\label{Seq:theory}
	\varphi_i = \frac{2(\rho-1)}{B} v_i + O(\varepsilon), 
\end{equation}
with
\begin{equation} 
	B = \sum_{i=1}^2 u_i \sum_{j=1}^2 v_j \mathcal{M}_{ij} + \rho(1-\rho) \sum_{j=1}^2 u_j v_j^2\, . 
\end{equation}

\subsection*{Computing the largest eigenvalue}
We first approximate the largest eigenvalue of $\mathcal{M}$ denoted by $\rho$. It is given by (see \textit{Mathematica} notebook)
\begin{equation}
\begin{aligned}
    \rho &= \frac{1}{2}\left(2 + a_{\text{old}} + a_{\text{new}} - m -  m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}} \right.\\
    &\quad \left. + \sqrt{4 (m -1) (1+a_{\text{old}})(1+a_{\text{new}}) + (2+a_{\text{old}} + a_{\text{new}} - m - m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}}})^2 \right)
\end{aligned}
\end{equation}
%
In order to make analytical progress and to identify under which conditions the process is slightly super-critical, i.e. $\rho>1$, we rescale the parameters by a small parameter $\varepsilon$. We set $a_{\text{old}} = \varepsilon, a_{\text{new}} = \varepsilon \xi$ and $m = \varepsilon\mu$. Assuming that $\varepsilon$ is small enough, i.e. effectively a weak selection assumption in old-habitat patches, we can neglect higher orders of $\varepsilon$ and find
\begin{equation}\label{eq:eigenvalue}
\begin{aligned}
    \rho &\approx 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{(\xi -1 + \mu_m^{\text{new}})^2 + 2(1-\xi+\mu_m^{\text{new}})\mu_m^{\text{old}} + (\mu_m^{\text{old}})^2  }\right)\\
    &= 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{\frac{\gamma}{1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}}  }\right),
\end{aligned}
\end{equation}
where $\gamma$ is the rescaled version of the constant $C$ in the main text (eq.~\eqref{eq:normalization}), i.e.
\begin{equation}
    \gamma = (1-f_{\text{old}})(\xi-1+\mu)^2 + \widehat{\pi}_m f_{\text{old}} (\xi-1-\mu)^2\, .
\end{equation}
%
For $\varepsilon\to 0$ we find that $\rho\to 1$ \chg{(eq.~\eqref{eq:eigenvalue}) which means that the branching is slightly super-critical if $\rho>1$ and real. A sufficient condition for this to be true is}
\begin{equation}
    1+\xi-\mu > 0 \quad \Leftrightarrow \quad a_{\text{old}} + a_{\text{new}} - m > 0
\end{equation}
%
\chg{In case that the branching process is not super-critical the establishment probability in eq.~\eqref{Seq:theory} becomes negative and as such is not a probability anymore. Hence, we can simply reject negative solutions of the establishment probability and by that implicitly justify that our approximation is valid.}%\\

%Inserting $\rho$ into equation~\eqref{eq:theory} we see the the enumerator is of order $\varepsilon$.

\subsection*{Computing the establishment probability}
For the solution of eq.~\eqref{Seq:theory} it remains to compute the normalized eigenvectors. Their precise form is of not much insight. We therefore omit stating them explicitly but refer to the \textit{Mathematica} notebook. Solving eq.~\eqref{Seq:theory} to the first order of $\varepsilon$ we find

\begin{equation}
    \begin{aligned}
        \varphi_{\text{old}} &= \varepsilon + \frac{\varepsilon\left(1-\xi\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}  + \frac{\varepsilon\left(\mu_m^{\text{old}} -\mu_m^{\text{new}} + 2 \mu_m^{\text{new}}\xi  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}\, ,\\
        \varphi_{\text{new}} &= \varepsilon\xi + \frac{\varepsilon\xi \left(\xi-1\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}} + \frac{\varepsilon\left(\xi \mu_m^{\text{new}} -\xi \mu_m^{\text{old}} + 2 \mu_m^{\text{old}}  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}\, .
    \end{aligned}
\end{equation}
%
Transforming back to the original variables and replacing $\gamma$ by the constant $C$ from the main text (eq.~\eqref{eq:normalization})
\begin{equation}
    C = (1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + \widehat{\pi}_m f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, ,
\end{equation}
we obtain

\begin{equation}
    \begin{aligned}
        \varphi_{\text{old}} &= a_{\text{old}} + \frac{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})a_{\text{old}} \left(a_{\text{old}}-a_{\text{new}}\right)}{\sqrt{C}}  + \\
        &\qquad \qquad \qquad m \frac{\left(\widehat{\pi}_m f_{\text{old}} a_{\text{old}} - (1-f_{\text{old}}) a_{\text{old}} + 2 (1-f_{\text{old}}) a_{\text{new}}  \right)}{\sqrt{C}}\, ,\\
        \varphi_{\text{new}} &= a_{\text{new}} + \frac{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})a_{\text{new}} \left(a_{\text{new}} - a_{\text{old}}\right)}{\sqrt{C}} \\
        &\qquad \qquad \qquad + m \frac{\left(a_{\text{new}} (1-f_{\text{old}}) - a_{\text{new}} \widehat{\pi}_m f_{\text{old}} + 2 a_{\text{old}} \widehat{\pi}_m f_{\text{old}}  \right)}{\sqrt{C}}\, .
    \end{aligned}
\end{equation}
%
Slightly re-ordering the terms, this gives the establishment probability of a single mutant individual, eq.~\eqref{eq:estab_approx} in the main text:
\begin{equation}\label{Seq:estab_approx}
	\begin{aligned}
		\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}} \qquad \quad \;  \; +  \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
		& \qquad \qquad + \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}\widehat{\pi}_m f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
		\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} +  \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
	& \qquad \quad \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) \widehat{\pi}_m f_{\text{old}} \right)\, .}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
	\end{aligned}
\end{equation}

\chg{For $m=0$ we see that $\varphi_{\text{old}}=0$, i.e. terms (1) and (2) cancel out. For the establishment probability in the new habitat we recover Haldane's result for the establishment probability of a slightly advantageous mutant: $\varphi_{\text{new}}=2a_{\text{new}}$ \citep{haldane_1927}.}

\subsection{Disentangling the contributions to the establishment probability}
\chg{We now proceed to explain} the three regions of the establishment probability from Fig.~\ref{fig:vary_m_est}(a) in the main text. These were defined by: (i) an initial increase of the establishment probability at low dispersal rates $m$; (ii) a local maximum with a subsequent decrease of the establishment probability; (iii) an increase of the establishment probability for high dispersal rates.

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig2a.pdf}
  		\caption{$\omega_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig2c.pdf}
  	\caption{$\omega_m=1.45$ (small fecundity difference)}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{figS1c.pdf}
  		\caption{$\omega_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{figS1d.pdf}
  	\caption{$\omega_m=1.45$ (small fecundity difference)}
	\end{subfigure}
	\caption{\textbf{Contribution of the different terms in eq.~\eqref{Seq:estab_approx} to the establishment probability $\varphi_{\text{old}}$.} Subfigures (a,b) are the same as Figs.~\ref{fig:vary_m_est}(a,c) in the main text. They show the establishment probability for a single mutant individual arising in an old-habitat patch for varying emigration probabilities $m$. In subfigures (c,d) we plot the terms from eq.~\eqref{Seq:estab_approx} separately ($\pi_w=0.5,\pi_m=-0.5$). Term 1, the mutant growth rate in old habitats (solid), increases with increasing dispersal rates as a consequence of relaxed competition. Term 2, the environmental effect (dashed), captures the differences between the growth rates in the habitats. The larger the difference, the larger its contribution to the overall establishment probability. Term 3, the effect of dispersal (dotted), (largely) increases with increasing dispersal rates $m$. The sum of the three terms is plotted as a dash-dotted line.}
	\label{Sfig:contribution}
\end{figure}

For clarity, we re-plot Figs.~\ref{fig:vary_m_est}(a,c) in Fig.~\ref{Sfig:contribution}(a,b), respectively. We try to explain the ongoing processes for each of the regions through the approximations of the establishment probabilities in eq.~\eqref{Seq:estab_approx}, see also Fig.~\ref{Sfig:contribution}(c,d). Note that these explanations are only valid for the establishment probability of a mutant initially in an old-habitat patch, $\varphi_{\text{old}}$. For the intuition behind the shape of $\varphi_{\text{new}}$ we refer to the corresponding section in the main text.

Region (i) is explained by the positive effect of dispersal. Mutants disperse from old- to new-habitat patches where they \chg{have a higher growth rate}. This effect is mediated through the third term of the establishment probability in eq.~\eqref{Seq:estab_approx}. \chg{Note that the first two terms of the approximation cancel out for small emigration probabilities $m$.}
While the third term increases with increasing emigration rate $m$, the second term in eq.~\eqref{Seq:estab_approx} decreases, cf. Fig.~\ref{Sfig:contribution}. 
In the formula this is mediated through \chg{the increase of the local growth rate in the old-habitat, $a_{\text{old}}$, i.e.\ it becomes less negative. Then both factors of the second term, $a_{\text{old}}$ and the difference $(a_{\text{old}}-a_{\text{new}})$, increase (which in turn decreases term two). The }\chg{intuitive} reason is that due to larger emigration probabilities $m$, more individuals leave old habitats before the reproductive event. This relaxes competition in these patches and therefore increases the local growth rate of mutants in old-habitat patches.   
%
%The intuition is that habitat differences become less important under stronger population mixing -- recall that the dispersal rate $m$ appears in the constant $C$ in the denominator. This alone would explain the decreasing curve of the second term. Additionally though, the local growth rate in old-habitat patches, $a_{\text{old}}$, is linked to the population demography which changes for varying $m$. With larger dispersal rates, more individuals leave the densely populated old habitats. This results in alleviated competition pressure for the remaining individuals, thus increasing the local growth rate $a_{\text{old}}$. This decreases the difference of growth rates $a_{\text{old}}-a_{\text{new}}$. Therefore, the local maximum can be explained by the decreasing environmental influence and the increasing dispersal effect, the second and the third term in eq.~\eqref{eq:estab_approx}, respectively. Hence, region (ii), beginning with the local maximum, is dominated by the decreasing effect of the environment on the establishment probability. 
Finally, in region (iii) dispersal is so large that the population homogenizes. This results in even less competitive pressure in old-habitat patches. Eventually, this yields a positive growth rate $a_{\text{old}}$ (first term in eq.~\eqref{Seq:estab_approx}). Therefore, this region is driven by the local growth rate in old habitats.

Note that region (iii) can be shifted to the left by increasing the absolute number of offspring of mutants in old-habitat patches, $\omega_m$, and therefore decreasing the local disadvantage of the mutant. If shifted sufficiently to the left, like in Fig.~\ref{Sfig:contribution}(d), region (ii) might vanish due to this effect.

\chg{Contrarily, if we set the fecundity parameter of the mutant in the old habitat to $\omega_m=1.1$, we see that region (iii) disappears for most dispersal schemes, cf. Fig.~\ref{Sfig:weak_fecundity}. Due to the low fecundity of the mutant, even under relaxed competition in old habitats, establishment of a mutant population is very unlikely. The final increase of our approximation in the dispersal schemes New-New and New-Old is due to our density-independence assumption in new habitats. For these large values of $m$ and the bias of the wild type towards new-habitat patches, this assumption is violated in the simulations explaining the deviation from the simulation results and our prediction. }

\begin{figure}
	\centering
	 	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{figS2a.pdf}
  		\caption{$\varphi_{\text{old}}$}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{figS2b.pdf}
  	\caption{$\varphi_{\text{new}}$}
	\end{subfigure}
	\caption{\textbf{Disappearance of region (iii) for large fecundity differences in the old habitat.} If the mutant fecundity in old-habitat patches is too low, here $\omega_m=1.1$, the effect of relaxed competition is not strong enough to have an impact on the establishment probability for high dispersal rates. The establishment probability remains at zero.}
	\label{Sfig:weak_fecundity}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Native habitat type %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Habitat of origin of the adaptive mutation}
\chg{Here, we provide further insight into the origin of the adaptive mutation. Therefore we plot the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ when varying the frequency of old habitats, $f_{\text{old}}$. In Fig.~\ref{fig:origin} in the main text we have seen that most successful adaptive mutations arise in old-habitat patches. Here, we show that this is explained mostly due to the large mutational input that is provided by the much larger wild-type population sizes in old habitats, see Fig.~\ref{Sfig:vary_f_origin}(b). The establishment probability though is always larger for mutants that arise in new habitats than in old habitats (Fig.~\ref{Sfig:vary_f_origin}(a)). 
}

\begin{figure}[h!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{figS3a.pdf}
  		\caption{Establishment probabilities}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{figS3b.pdf}
  	\caption{Stationary wild-type population size}
	\end{subfigure}
	\caption{\textbf{Establishment probability and stationary wild-type population size when varying the old-habitat frequency.} In the simulations we have used the standard set of parameters as given in Table~\ref{tab:parameters} and the unbiased dispersal scheme ($\pi_w=\pi_m=0$). In (a) we additionally chose the large fecundity difference scenario ($\omega_m=1.35$).}
	\label{Sfig:vary_f_origin}
\end{figure}

%Additionally, we investigate the habitat type of the origin of the adaptive mutations for low fecundity values of the mutant in the old habitat. While for large fecundity values of the mutant in the old habitat most successful lineages arise in the old habitat (Fig.~\ref{fig:origin} in the main text), we see that for lower fecundity values $\omega_m$ the contribution of new-habitat mutants approaches and occasionally exceeds the number of successful mutant lineages that emerge from old-habitat patches. In the subsequent figure, besides the default parameters as stated in Table 1 in the main text, we chose $\omega_m=1.05$ and $\theta = 10/(MK)$. The plot corresponding to Fig.~\ref{fig:origin} in the main text then looks as follows:

%\begin{figure}[h!]
%	\centering
%	\includegraphics[width=0.6\textwidth]{figS4.pdf}
%  	\caption{\textbf{Habitat type of the origin for strong fecundity differences in the old habitat ($\omega_m=1.05$).}}
%	\label{Sfig:origin}
%\end{figure}
\pete{There was a figure where the contribution from new habitats was larger than from old habitats. With the changed fecundities, I only get this behavior if the fecundity of the mutant is larger in the new patches than in the old patches, i.e. $\omega_m < a_{\text{new}}$. Also, for some reason the approximations do not work that well in these scenarios (I did not investigate why the estimated lines are bad -- soft sweeps still work fine but the other approximations don't.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Large emigration rate %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Probability of establishment for large frequencies of old-habitat patches}
\chg{We plot the probability of establishment for a large frequency of old-habitat patches, $f_{\text{old}}$. As visible in Fig.~\ref{Sfig:vary_f} below, for high frequencies of old-habitat patches the probability of adaptation becomes very small, if not zero, for large emigration probabilities $m$. These high frequency of old-habitat patches are the patch configurations which mutants that were present before the first patch degrades, i.e. standing genetic variation mutants, experience. Therefore, it is very unlikely that the mutant, which will eventually rescue the population, was already present before the first degradation event.} This supports the explanation that homogenizing the population (large $m$) reduces the impact of standing genetic variation on the probability of evolutionary rescue, see Fig.~\ref{fig:sgv}(b) in the main text where $\pi_w=\pi_m=0$ was plotted.

\begin{figure}[h!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{figS5a.pdf}
  		\caption{$\varphi_{\text{old}}$}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{figS5b.pdf}
  	\caption{$\varphi_{\text{new}}$}
	\end{subfigure}
	\caption{\textbf{Probability of adaptation for a large frequency of old-habitat patches ($f_{\text{old}}=0.9$).} The fecundity of the mutant in old-habitat patches is set to $\omega_m=1.45$. Note also the difference between the scales of the y-axes in the two panels.}
	\label{Sfig:vary_f}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Comparison to pop gen %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Establishment probability in a model without demography}
Here, we consider a variation of our original model in order to investigate the impact of demography on the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$. The dispersal process and the dynamics in old-habitat patches remain as studied before. In new-habitat patches we now assume that the population remains at carrying capacity, i.e. there is no longer a declining wild-type population. This means that the local growth rate $a_{\text{old}}$ in eq.~\eqref{Seq:s_old} takes the form for $\widehat{N}_w^{\text{new}}=K_{\text{new}}$. For simplicity we will also assume that $K_{\text{new}}=K_{\text{old}}=K$. In order to maintain the divergent selection assumption, we assume that the fecundity of the wild-type in the new habitat is below the fecundity of the mutant. %For simplicity, we assume that the fecundity values are given as
%\begin{equation}
%    \omega_w^{\text{old}} = \omega_m^{\text{new}} \quad \text{and} \quad \omega_w^{\text{new}} = \omega_m^{\text{old}}.
%\end{equation}
Therefore, we need to adjust the local growth rate of a single mutant, $a_{\text{new}}$. The local growth rate in old habitats, $a_{\text{old}}$, remains as outlined in eq.~\eqref{Seq:s_old}. We have
\begin{equation}
    1+a_{\text{new}} = K \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}} \widetilde{N}_w^{\text{new}}+ \omega_m^{\text{new}}}\approx K \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}} \widetilde{N}_w^{\text{new}}},
\end{equation}
which with the help of eq.~\eqref{Seq:wt_eq1} yields
\begin{equation}
    a_{\text{new}} \approx \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}}\left(1+\frac{m f_{\text{old}} (1-\widehat{\pi}_w)}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\right)}.
\end{equation}

\chg{Note, that we again used that during the establishment phase the wild type is much more abundant than the mutant which explains the approximation in the two equations.}
Plugging this in the approximation of the establishment probability from eq.~\eqref{Seq:estab_approx} we find the curves in Fig.~\ref{Sfig:pop_gen}.

We see that, as briefly mentioned in the main text, region (iii) of the establishment probability disappears in these type of models except for the Old-New and the New-Old dispersal schemes. The reason for the disappearance of the region is that relaxed competition only plays a subordinate role for the symmetric dispersal schemes (Old-Old, New-New and 0-0). In other words, these dispersal schemes maintain the local frequencies of the mutant at the same level as before the dispersal step and by that do not change the population dynamics. In contrast, the Old-New dispersal scheme strongly increases the frequency of mutants in new-habitat patches and by that increases the establishment probability. It is worth mentioning though, that this is not an effect of relaxed competition but rather a biased dispersal of the mutant into the habitat where it is favored. Solely for the New-Old dispersal scheme where individuals prefer the habitat they are relatively less fit, we see a relaxed competition scenario in old-habitat patches. For high dispersal rates $m$ the majority of the wild-type individuals emigrate to new habitats and by that leave space for the rare mutants in old-habitat patches to reproduce. This explains why this dispersal scheme still shows the characteristic pattern of a three-stage establishment probability.   \newpage

\begin{figure}[h!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{figS6a.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{figS6b.pdf}
  	\caption{$\varphi_{\text{new}}$ with $\omega_m=1.35$}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{figS6c.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega_m=1.45$ (small fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{figS6d.pdf}
  	\caption{$\varphi_{\text{new}}$ with $\omega_m=1.45$}
	\end{subfigure}
  	\caption{\textbf{Establishment probability when populations in both habitats are at carrying capacity.} We plot the establishment probability for a single mutant either initially in an old-habitat patch (a,c) or in a new-habitat patch (b,d). The numerical solution (solid lines) still approximates the simulated data reasonably well. The analytical approximation (dashed lines) however deviates strongly from the data due to large growth rates ($a_{\text{new}}\approx 0.2$) so that the conditions for the approximation to hold are violated. In this case, in eq.~\eqref{Seq:estab_approx} higher order corrections would need to be taken into account. \chg{The fecundity values in the new habitat are given by $\omega_w^{\text{new}} = \omega_m^{\text{old}}$ and $\omega_m^{\text{new}} = \omega_w^{\text{old}}$ and the carrying capacity is $K_{\text{new}}=K_{\text{old}}=K=500$.} Missing data points (mostly for the negative density-dependent dispersal scheme -- green triangles) are explained by too large computation times. All data points are averages from $10^4$ independent runs. Note the varying y-axes scales.}
	\label{Sfig:pop_gen}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Origin dispersal schemes %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Habitat of origin dependent on the dispersal scheme}
The habitat type of the origin of the rescue mutation is largely independent of the considered dispersal scheme. For $\omega_m=1.35$ we have plotted the relative contribution of each natal habitat type to the probability of evolutionary rescue, Fig.~\ref{Sfig:natal_habitat}. We do not see large differences between the three symmetric dispersal schemes (0-0, Old-Old, and New-New) and the Old-New dispersal scheme. \chg{Only for the New-Old dispersal schemes we see an increase in the contribution of mutants emerging from old-habitat patches for very high emigration probabilities $m$. A possible explanation is the large probability of establishment for a mutant emerging in old-habitat patches for this parameter set (cf. Fig.~\ref{Sfig:contribution}(a)).}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{figS7.pdf}
  	\caption{\textbf{Habitat type of the origin of the rescue mutant dependent on the dispersal scheme.} Varying the emigration probability $m$ we plot the relative contributions of each habitat type to the probability of evolutionary rescue. The color-coding is as in the main text: black for 0-0, blue for the Old-Old, green for the New-New, orange for the Old-New, and purple for the New-Old dispersal scheme.}
	\label{Sfig:natal_habitat}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\bibliographystyle{plainnat}
\bibliography{dispersal.bib}

\end{document}

### 2020-03-25 ms/main.tex
first commit

\documentclass[a4paper,11pt]{article}

\usepackage[utf8x]{inputenc}	
\usepackage{fullpage}		
\usepackage[T1]{fontenc}				
\usepackage[english]{babel}			
%\usepackage{ulem}
\usepackage{xcolor}						
\usepackage{amsmath,amstext,amssymb,amsfonts,amsthm,mathrsfs,chemarrow,stackrel,dsfont}   % Math packages
\usepackage{graphicx} 					
\usepackage{subcaption}					
%\usepackage{cite}	
\usepackage[authoryear,sectionbib,sort]{natbib} % Bio citation format
\linespread{1.7}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

\usepackage{microtype}
\usepackage{array}
\usepackage{booktabs}
\usepackage{makecell}

%%%%%% XX To be removed before submission!
\usepackage{lineno}
\usepackage{showlabels}

\usepackage[pdftex, breaklinks, colorlinks, citecolor=black, urlcolor=black, linkcolor=black]{hyperref} 
\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
\usepackage{FiraSans}
\usepackage{soul}

% Packages to comment out before submitting!
%\usepackage[textsize = scriptsize, backgroundcolor = white, linecolor = black]{todonotes}
%\usepackage{showlabels}

%\numberwithin{equation}{section}
%\setlength{\parindent}{0pt}
\allowdisplaybreaks[2]

\newcolumntype{P}[1]{>{\centering\arraybackslash}m{#1}}     % Table formatting
\renewcommand{\arraystretch}{1.1}                           % Table formatting

\definecolor{darkgreen}{rgb}{0.0, 0.42, 0.24}
\definecolor{darkblue}{rgb}{0.0, 0.0, 0.55}
\definecolor{change}{rgb}{0.5,0.,0.6}

%%%%%% our commenting commands
\newcommand{\florence}[1]{\textcolor{purple}{(#1)}} % Changed red to purple
\newcommand{\francois}[1]{\textcolor{blue}{(#1)}}
\newcommand{\hildegard}[1]{\textcolor{darkgreen}{(#1)}}
\newcommand{\pete}[1]{\textcolor{orange}{(#1)}}
\newcommand{\chg}[1]{\textcolor{change}{#1}}

%opening
\title{The effect of habitat choice on evolutionary rescue in subdivided populations}
\author{Peter Czuppon$^{1,2,\ast}$, Fran\c{c}ois Blanquart$^{1,3}$, Hildegard Uecker$^{4,\dag}$,\\ Florence D\'{e}barre$^{2,\dag}$}
\date{}

\begin{document}

\maketitle

\vspace{-20pt}
\noindent $^1$ Center for Interdisciplinary Research in Biology, CNRS, Coll\`ege de France, PSL Research University, Paris, France.

\noindent $^2$ Institute of Ecology and Environmental Sciences of Paris, Sorbonne Universit\'e, UPEC, CNRS, IRD, INRA, Paris, France.

\noindent $^3$ IAME, UMR 1137, INSERM, Universit\'{e} Paris Diderot, Paris, France.

\noindent $^4$ Research Group ``Stochastic evolutionary dynamics'', Department of Evolutionary Theory, Max-Planck Institute for Evolutionary Biology, Pl\"{o}n, Germany.

\noindent $\ast$ Corresponding author; e-mail: peter.czuppon@college-de-france.fr.

\noindent $^\dag$ Equal contribution.

\noindent The authors wish to be identified to the reviewers.

\bigskip

\textit{Manuscript elements}: Figures~1-6, Table~1, Supplementary Information including figures S1-S6. All figures are to print in color.

\bigskip

\textit{Keywords}: evolutionary rescue, local adaptation, source-sink dynamics, dispersal, gene flow, habitat choice, density-dependent dispersal.

\bigskip

\textit{Manuscript type}: Article. %Or e-article, note, e-note, natural history miscellany, e-natural history miscellany, comment, reply, invited symposium, or countdown to 150.

\newpage

\linenumbers{}
\modulolinenumbers[2]

\begin{abstract}
    \noindent \textbf{Abstract} Evolutionary rescue is the process by which a declining population successfully adapts genetically\footnote{\francois{remove genetically}\florence{ok to remove}} to avoid extinction. In a structured environment that deteriorates patch by patch, dispersal can substantially alter the chances of evolutionary rescue of a \linelabel{R2-1}\chg{population whose wild type is} not viable in deteriorated patches. Here, we investigate the effect of different dispersal schemes and intensities on the probability of successful establishment of a mutant population adapted to the deteriorated environment. \linelabel{R1-2}\chg{We assume that local fitness is determined by a single haploid locus\florence{ and that d}\st{. D}ispersal is genotype-dependent\st{ and linked to the adaptive trait, i.e. dispersal does not evolve by itself}\footnote{\florence{[otherwise it sounds like $m$ is changing...] [need to add that it's the bias that is changing, not the probability to emigrate]}}. In this scenario, we }find that the probability of evolutionary rescue can undergo up to three phases when increasing the rate of dispersal\footnote{\florence{since you specify ``at low disp'', ``at intermediate disp'' etc., you can remove ``when increasing the rate of dispersal'' to save words}}: (i) at low dispersal rates, the probability of establishment of a mutant population increases; (ii) at intermediate dispersal rates, the establishment probability decreases; (iii) at large dispersal rates, the population homogenizes, \florence{, which either promotes or suppresses}\footnote{\florence{old version: either promoting or suppressing}} the process of evolutionary rescue, depending on the fitness difference between the mutant and the wild type. Our results show that habitat choice\footnote{\florence{up to now, the reader does not know this is about habitat choice (``schemes'' is not precise enough}}, when compared to \chg{unbiased} dispersal, impedes successful adaptation when the mutant has the same habitat preference as the wild type, but promotes adaptation when the mutant mainly immigrates into patches where it has a growth advantage over the wild type.
\end{abstract}

%\textit{Keywords:} evolutionary rescue, local adaptation, source-sink dynamics, dispersal, gene flow, habitat choice, density-dependent dispersal

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% INTRODUCTION %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section*{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and empirical evidence %%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Facing current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, the populations of many species are declining and might eventually go extinct \citep{bellard_2012, diniz_2019}. Pests and pathogens experience similarly strong selective pressures as a result of increased consumption of antibiotics and use of pesticides \citep{ramsayer_2013, kreiner_2018}. 
The process of genetic adaptation that saves populations from extinction is termed evolutionary rescue. This \linelabel{R1-3}\chg{process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, typically }resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
In recent years, empirical examples of evolutionary rescue have accumulated \citep[as reviewed by ][]{alexander_2014,carlson_2014,bell_2017}. Evolutionary rescue \linelabel{R1-4}\chg{has been observed directly in laboratory experiments \citep[e.g.][]{bell_2009,agashe_2011,lachapelle_2012,lindsey_2013,stelkens_2014}. In wild populations demographic and genotypic data is rarely monitored at the same time impeding direct observation of evolutionary rescue. Still, it has been suggested as a mechanism that has saved a few wild populations from extinction \citep[e.g.][]{vanderwal_2012,digiallonardo_2015,gignoux_2018}.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and dispersal %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality. In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and horizontal gene transfer \citep{tazzyman_2014}, reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\chg{\citep{yamamichi_2015}}, bottlenecks \citep{martin_2013}, different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}. \chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al. 1991. Adaptive and  demographic responses of plankton populations to environmental change. Limnology and Oceanography 36:1301--1312., B{\"u}rger and Lynch 1995. Evolution and extinction in a changing environment: a quantitative genetic analysis. Evolution 49:151--163 and Lande and Shannon 1996. The role of  genetic  variation  in adaptation and population persistence in a changing environment. Evolution 50:434--437? But please check; it's a long time that I read those...}}. 
In fragmented environments, habitat deterioration is not necessarily synchronized across patches: there can be a transient spatially heterogeneous environment consisting of old- and new-habitat patches until eventually the whole environment has deteriorated. If individuals that populate different patches are able to move between those, the effect of dispersal on evolutionary rescue needs to be taken into account \citep{uecker_2014, tomasini_2019}. Experiments that study the effect of dispersal on evolutionary rescue are rare but indicate that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation \citep{bell_2011}.

\linelabel{R1-1}\chg{Evolutionary rescue in a fragmented environment can be interpreted as the intermediate situation between abrupt change and gradual change: if dispersal is very low, patches are approximately isolate and thus each patch undergoes an abrupt change independent of the other patches; if the dispersal rate and the number of patches are very high, the whole environment is gradually changing since spatial structure and the contribution of a single patch are negligible.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Adaptation and dispersal %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Evolutionary rescue in a situation where one patch after \linelabel{R2-2}\chg{another} deteriorates is tightly linked to the study of adaptation to a heterogeneous environment with source-sink dynamics. These describe a spatially heterogeneous environment that is constant in time where a population in unfavorable habitats can be maintained by constant immigration of the wild type. Experimental and theoretical studies have found that \linelabel{R2-3}\chg{increasing} dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g. \citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000,ronce_2001,gallet_2018}} for both effects).
%Reason for different results: (i) demography or not (Kawecki); (ii) selection strength (Gallet), (iii) selection strength (Ronce)
\florence{suggestion: focus on discrete traits - thousands of possible papers to cite; avoid RK2001}

In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be \chg{unbiased}, i.e. dispersing individuals are distributed uniformly among patches. Only few investigations in the context of local adaptation in source-sink systems have taken into account non-\chg{uniform} dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}. This analytical focus on \chg{unbiased} dispersal is in stark contrast to actually observed dispersal schemes in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}. Here, we explore and compare the effects of several biologically motivated dispersal schemes on adaptation and evolutionary rescue.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Empirical examples for non-random dispersal %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the best documented modes of non-\chg{uniform} dispersal is density-dependent dispersal. Density dependence can be positive or negative: either individuals prefer to settle or stay in large groups (positive density-dependence), or they choose to remain in or move to less populated regions (negative density-dependence). Density-dependent dispersal, of either form, is ubiquitously found in nature and has been reported in many species across the tree of life, including insects \citep{endriss_2019}, spiders \citep{meester2010}, amphibians \citep{gautier_2006}, birds \citep{wilson_2017}, fishes \citep{turgeon_2012}, and mammals~\citep{stoen_2006}.

Another well-established dispersal scheme is habitat choice, whereby individuals tend to immigrate into habitats they are best adapted to. This mechanism has for example been reported in lizards \citep{bestion_2015}, birds \citep{dreiss_2011,benkman_2017}, fishes \citep{bolnick_2009}, worms \citep{mathieu_2010}, and ciliates \citep{jacob_2017,jacob_2018}.

\linelabel{R1-8}\chg{Although density-dependence or habitat choice can also affect the emigration behavior, i.e. the likelihood that an individual leaves its current location, or even the vagrant stage when the individual is in the process of dispersal \citep{bowler_2005,ronce_2007}, we ignore this here. Instead, we concentrate on \chg{non-uniform} dispersal effects of the immigration process.} \\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What do we do? %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the following, we explicitly account for these \chg{biased} dispersal schemes in the context of genetic adaptation to a deteriorating and spatially structured environment. We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}\chg{types, wild type and mutant, have a positive growth rate}, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct. In this framework, we study four biologically motivated dispersal patterns and compare them to the \chg{unbiased} dispersal scheme. We start by investigating their consequences for the establishment dynamics of an adapted mutant in a time-constant heterogeneous environment. 
From there, we derive an approximation for adaptation in a source-sink setting.
Using these results, we study the probability of evolutionary rescue, i.e. of adaptation in the scenario where patches, one after \chg{another}, deteriorate over time until all locations contain the new habitat. We find that dispersal bias has a direct effect on the local growth rates which in turn affect the probability of establishment and evolutionary rescue.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% MODEL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% General description %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a spatially structured environment consisting of $M$ patches all connected to each other. The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively. The transition from the old to the new habitat is assumed to be irreversible. Patches are populated by two types of asexually reproducing, haploid individuals, either wild type or mutant. Generations are measured in discrete time steps and are non-overlapping. \linelabel{R1-9-2}\linelabel{R2-6}\chg{Wild-type and mutant individuals both have a positive growth rate} in old-habitat patches; the mutant has a lower growth rate there. In new-habitat patches, the wild type population declines locally while the mutant has a positive growth rate -- we will therefore call it a `rescue mutant'. \chg{Both habitat types, old and new, have a (potentially different) carrying capacity. At the end of a life cycle, the local population size is reduced to the carrying capacity if necessary.}

We consider a time-varying environment, such that one after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates (from old to new state). Initially ($t<0$), all patches are of the old-habitat type. At time $t=0$, the first patch deteriorates. After $(M-1)\tau$ generations, all patches are of the new-habitat type. We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$. It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event until it eventually hits~$0$, when all patches have undergone the environmental change. This environmental setting corresponds to the one analyzed by \citet{uecker_2014} and more recently in the special case of just two patches by \citet{tomasini_2019}.

The individuals are assumed to go through the following life-cycle: (i) Dispersal: individuals may move between patches; (ii) Reproduction: individuals reproduce within patches. \linelabel{R2-7}\chg{The number of offspring per parent is modeled by a Poisson number}; (iii) Mutation: wild-type individuals mutate to the rescue mutant type with probability $\theta$ -- back mutations from the mutant to the wild type are neglected; (iv) Regulation: the population size is, \chg{if necessary, down-regulated to the carrying capacity $K_k$ where $k$ refers to the habitat type (old or new - the two habitat types can have different carrying capacities)}. If the population size is below the carrying capacity, the regulation step is ignored.

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Old habitat dynamics %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Old-habitat dynamics}
\linelabel{R1-10}We assume that in old-habitat patches, \linelabel{R1-9-3}\chg{both types have a positive growth rate such that the population in these patches is typically at carrying capacity $K_{\text{old}}$.} We denote the fecundities of the \textbf{w}ild-type and the \textbf{m}utant by $\omega_w$ and $\omega_m$, respectively. 
We impose an adaptation trade-off for the rescue mutant, i.e. the mutant is less fit than the wild type in old-habitat patches ($\omega_m<\omega_w$).
%
\linelabel{R2-9}\chg{We assume that mutation from the wild type to the mutant is tied to reproduction. The probability that an offspring of a wild-type individual is a mutant is given by $\theta$. We neglect back-mutation from mutant to wild type.} 
\linelabel{R2-8}\chg{After reproduction, if the population size exceeds the carrying capacity $K_{\text{old}}$ we randomly remove individuals from the population until the carrying capacity is reached\linelabel{AE-3}, i.e. we assume that the wild type and the mutant are competitively equivalent. The mean number of offspring of\footnote{\francois{produced by}} one mutant after reproduction and population regulation is then given by   
\begin{equation}\label{eq:sold}
     K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}^{\text{old}}_w + \omega_m}\approx K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}^{\text{old}}_w} =1 + a_{\text{old}} \, ,
\end{equation}
where the random variable $\widetilde{N}_{i}^{k}$ denotes the population sizes of type $i$ (mutant or wild-type) in patches with habitat $k$ (old or new) after the dispersal step. Additionally, we have defined the local growth rate of a single mutant, denoted by $1+a_{\text{old}}$. \linelabel{R1-11}The approximation in eq.~\eqref{eq:sold} holds if the number of mutants is much lower than the number of wild types in old-habitat patches -- a reasonable assumption since the establishment of the mutant is decided when it is still rare.}
%Then, while still rare, mutant individuals will reproduce independently of each other and their dynamics will be described by the local growth rate given in eq.~\eqref{eq:sold}. 
\chg{Compared to the wild type we assume a reduced fecundity of the mutant in the old habitat, i.e. $\omega_m<\omega_w$. Still, the local growth rate can be positive, that is, larger than one ($a_{\text{old}}>0$).} This happens when the \linelabel{R2-12}\chg{wild-type population in a single old-habitat patch after dispersal,} \linelabel{R1-13}$\widetilde{N}^{\text{old}}_w$, becomes very small. This strongly depends on the dispersal scheme and rate. \linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}

%%%%%%%%%%%%%%%%%%%%%%%%%%
% New habitat dynamics %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{New habitat dynamics}
In the new habitat and in the absence of immigration, the wild-type population will go extinct. To account for this, we assume that the number of wild-type \linelabel{R1-15}offspring \chg{per parent} in new-habitat patches is a Poisson distributed value with parameter $(1-r)$, where $r$ measures how strongly the wild type is affected by the environmental deterioration. For $r=1$ the wild-type population becomes locally extinct in new-habitat patches after one generation.

The mutant type, in contrast, has a local growth rate larger than one and will (in expectation) grow in numbers. The number of offspring is again Poisson distributed with per-capita rate $(1+a_{\text{new}})$, the local growth rate of the mutant in new-habitat patches.

\linelabel{R1-17}\linelabel{R2-13}\chg{In the following analysis we will assume that the mutant dynamics in new-habitat patches are density-independent. This is a valid assumption as long as the carrying capacity in these patches, $K_{\text{new}}$, is not reached. Of course, in the simulations the carrying capacity might sometimes be reached, especially for large frequencies of old-habitat patches. This assumption on density-independence also restricts our choice for the carrying capacity in the new habitat. If the carrying capacity in new-habitat patches is much smaller than in old-habitat patches, the new habitat can reach its carrying capacity merely due to immigration of wild-type individuals (even after reproduction where the wild-type population is reduced).\footnote{\francois{Rephrase, as this is not clear. What carrying capacity do you choose and whether your approximation holds or not. Maybe: `Here we examine a carrying capacity in new habitats which is half that in old habitats. If the carrying capacity were much smaller, our analytical approximation would break down.}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dispersal mechanisms %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Dispersal mechanisms}

We assume that dispersal is cost-free, i.e. all emigrating individuals will settle in a patch, leaving the global population size before and after dispersal unchanged. \linelabel{R1-18}\linelabel{R2-14}\chg{Note however, that our methods are in principle also applicable to a scenario where dispersal is associated with a (potentially type-dependent)\footnote{\francois{remove bracket}} cost.} We split the dispersal step into emigration and immigration. We focus on habitat choice during immigration. Individuals have a bias to immigrate into patches of a certain habitat type. The bias \chg{during} immigration \chg{for} new-habitat patches is set to \chg{zero}, without loss of generality.\footnote{\francois{I did not understand why this is needed}} The bias \chg{for immigration to} an old-habitat patch is denoted by $\pi_{i}$, with the index $i$ indicating the wild-type ($w$) or mutant ($m$) bias. For \chg{$\pi_i<0$}, individuals of type $i$ are more likely to settle in new-habitat patches, for \chg{$\pi_i>0$} the reverse is true. For $\pi_i=0$, individuals do not have a preference and dispersal is \chg{unbiased}.

We consider equal and constant emigration rates for both types and habitats throughout the manuscript. Note again that the method would allow for type- and habitat-dependent emigration rates. We denote the probability \chg{that an individual leaves} its natal patch by $m$. 
Then, the probability \chg{that an individual of type $i$} \chg{emigrates (from any habitat type) and immigrates into }the new habitat is given by 
\begin{equation}\label{eq:dispersal}
	\begin{aligned}
		m_i^{\text{new}} &= m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}}+\chg{e^{\pi_i}} f_{\text{old}}} = 1-m_i^{\text{old}},\\
	\end{aligned}
\end{equation}
where $f_{\text{old}}$ is the frequency of old-habitat patches. \linelabel{R1-19}\chg{
%Since we are working with a time-discrete model the dispersal step needs to be a probability (and not a rate as is common in continuous-time models). 
We transformed the bias $\pi_i$ which can take any real value to a positive number, $e^{\pi_i}$. This ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one. For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.}

From a biological perspective, a number of dispersal schemes are of particular interest. Precisely, we distinguish between the following five scenarios (see also Fig.~\ref{fig:disp_schemes} for an overview):

\begin{itemize}
	\item \textbf{Old-Old:} Individuals prefer to immigrate into the habitat where they have the largest number of offspring before regulation. \linelabel{R1-22}\linelabel{R2-16}\chg{Since the average number of offspring per parent is higher in the old habitat for both types ($\omega_w>1-r$ and $\omega_m>1+a_{\text{new}}$), this translates} to both $\pi_w$ and $\pi_m$ being larger than~$0$, i.e. a bias towards old-habitat patches. In practice, individuals are thought to use biotic or abiotic cues to preferentially immigrate into habitats where their fecundity is highest when compared to other habitats. This type of dispersal, i.e. matching habitat choice, has for example been observed with common lizards \textit{Zootoca vivipara} \citep{bestion_2015}, three-spine sticklebacks \textit{Gasterosteus aculeatus} \citep{bolnick_2009}, and barn owls \textit{Tyto alba} \citep{dreiss_2011}.

	The same range of parameters ($\pi_w,\pi_m>0$) is obtained when implementing positive density-dependent dispersal \chg{on immigration. I}ndividuals are more likely to immigrate into patches with higher population densities, so both, the wild type and the mutant, will prefer the old habitat.\footnote{\francois{I would turn this around, otherwise this sounds like a fact. `Indeed, if both the wt and the mutant prefer the old habitat, individuals are more likely to immigrate into patches with...}} Highly populated locations can be an indication for a safe shelter, relevant for prey species, and potentially increase the mating success of individuals. This type of positive density-dependent dispersal on immigration (also called conspecific attraction) is for example found in several amphibians, e.g. the salamander species \textit{Mertensiella luschani} \citep{gautier_2006} and \textit{Ambystoma maculatum} \citep{greene_2016} or the frogs \textit{Oophaga pumilio} \citep{folt_2018}.

	\item \textbf{Old-New:} Under this dispersal scheme, individuals tend to immigrate into habitats where they are fitter than the other type. This translates into the wild type preferring old patches, i.e. $\pi_w>0$, while mutant has a bias towards the new habitat, i.e. $\pi_m<0$. 
	Empirical evidence of this mechanism is scarce. It resembles the recently observed dispersal pattern of the ciliates \textit{Tetrahymena thermophila} with a specialist and generalist type \citep{jacob_2018}. The specialist disperses to its preferred habitat while the generalist prefers to immigrate to a suboptimal habitat where it outcompetes the specialist.

	\item \textbf{New-New:} \chg{This dispersal scheme can be\footnote{\francois{is}} motivated by negative density-dependent dispersal on immigration,} whereby individuals are more likely to move to less populated patches. In these locations, resources might be more abundant, intra-specific competition alleviated and the chance of infection transmission decreased, which may compensate for the potentially reduced habitat quality. The corresponding parameter choice in our model is $\pi_w,\pi_m<0$, i.e. both types have a higher likelihood to immigrate to new-habitat patches. \linelabel{R1-23}\chg{We emphasize that population density in the patches changes over time but the biases are constant. We use the comparison with density-dependent dispersal only to motivate our chosen biases during the relevant phase of rescue, i.e. when the population density in new-habitat patches is low.}
	The mutant being initially rare, it is unlikely that the carrying capacity is reached in new-habitat patches -- as assumed throughout the analysis.\footnote{\francois{Unclear to me. Suggestion: `Because the carrying capacity is not typically reached in new patches in the initial phase of rescue, this bias amounts to negative density-dependent dispersal in that phase.}} 
    Various empirical examples of negative density-dependent dispersal exist. Density-dependent immigration effects as described here, are for example found in the damselfish species \textit{Stegastes adustus} \citep{turgeon_2012} and the migratory birds \textit{Setophaga ruticilla} \citep{wilson_2017}. %In both species, immigrants fill the empty space left by dead or emigrated individuals. Notably, in the fish species, negative density-dependent immigration is further enhanced by residents defending their territory against potential immigrants.

	\item \textbf{New-Old:} For completeness, we also consider the scenario in which both types disperse preferentially into the habitat they are relatively less fit in, i.e. we have $\pi_w<0$ and $\pi_m>0$. This dispersal scheme is \chg{vaguely} related to the concept of an `ecological trap': Individuals tend to immigrate into patches that cannot sustain a population, in its most extreme form resulting in the extinction of the species \citep{battin_2004}. \chg{Note however,} that this does not exactly apply to our framework since here the mutant is able to maintain a population in old-habitat patches.\footnote{\francois{Consider removing the last part about the ecological trap.}}

	\item \textbf{Unbiased dispersal (0-0):} Individuals do not have a bias towards any of the two habitats, $\pi_w=\pi_m=0$. Most theoretical results examining the interplay of dispersal and establishment have \chg{used} this dispersal scheme. Therefore, we use it as a benchmark to which we compare the above defined dispersal patterns.
	%
\end{itemize}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{fig1.pdf}
    \caption{\textbf{Parameter sets for the different dispersal schemes.} \small The colors and markers are the same across all figures. The $x$ and $y$ axes show the dispersal bias of the wild type and mutant, respectively, to immigrate into old-habitat patches ($\pi_w$ and $\pi_m$). The markers are located at the parameter values used for the simulations: $\pi_m=\pi_w=0$ for unbiased dispersal (0-0; {\Large $\bullet$}), $\pi_m=\pi_w=0.5$ for a type-independent bias towards old-habitat patches (Old-Old; {\color{blue} $\blacksquare$}), $\pi_m=-0.5,\pi_w=0.5$ for a wild-type preference towards the old habitat, and a mutant-bias to the new habitat (Old-New; {\color{orange}$\blacktriangle$}), $\pi_m=\pi_w=-0.5$ for a type-independent bias towards the new habitat (New-New; {\color{darkgreen}$\blacktriangledown$}), and $\pi_m=0.5,\pi_w=-0.5$ for the mutant-bias towards the old habitat and a preference of the wild type for the new habitat (New-Old; {\color{purple}$\blacklozenge$}). All the results derived below depend continuously on these parameters, so that varying any of the parameters will not result in a sudden change or discontinuity of the corresponding curves. {\color{orange} Do we need this last sentence?\color{blue}No.}}
    \label{fig:disp_schemes}
\end{figure}

\linespread{1}
\begin{table}[t!]
	\begin{center}
		\begin{tabular}{c | P{6cm} | c | c}
			\textbf{Parameter} & \textbf{Interpretation} & \textbf{Range} & \textbf{Default value} \\
			\midrule
			$K_k$ & Carrying capacity in a patch of type $k$ & -- & \makecell{$K_{\text{old}}=1000$,\\ $K_{\text{new}}=500$} \\
			$\omega_j$ & \makecell{Expected per-capita number of\\ type $j$ offspring in the old habitat\\ before regulation} & $0\leq \omega_m < \omega_w$ & \makecell{$\omega_w = 1.5$,\\ $\omega_m=1.45$ {\footnotesize (weak sel.)},\\ $\omega_m=1.35$ {\footnotesize (strong sel.)}}\\
			$1+a_{\text{old}}$ & \makecell{Growth rate of the mutant\\ in the old habitat} & $-1\leq a_{\text{old}}$ & -- \\
			$1-r$ & \makecell{Mean number of wild-type offspring\\ in the new habitat} & $0<r\leq 1$ & \makecell{$0.75$\\ $(r=0.25)$}\\
			$1+a_{\text{new}}$ & \makecell{Growth rate of the mutant\\ in the new habitat} & $0<a_{\text{new}}$ & \makecell{$1.02$ \\ ($a_{\text{new}}=0.02$)} \\
			$m$ & Emigration probability & $0<m\leq 1$ & $0.06$\\
			$\pi_i$ & Type $i$ bias towards the old habitat & $\pi_i\in\mathbb{R}$ & see Fig.\ref{fig:disp_schemes}\\
			$\widehat{\pi}_i$ & Transformed dispersal bias of type $i$ & $0<\widehat{\pi}_i$ & see Fig.\ref{fig:disp_schemes}\\
			$M$ & Number of patches & $2\leq M$ & $10$\\
			$f_{\text{old}}$ & Frequency of old-habitat patches & $0\leq f_{\text{old}}\leq 1$ & $0.5$ \\
			$\theta$ & Mutation probability & $0 < \theta$ & $\frac{1}{25 M K_{\text{new}}}$\\
			$\tau$ & Time interval between two consecutive deterioration events& $0<\tau$ & $100$\\
			\midrule
			$\widehat{N}_{i}^{k}$ & \multicolumn{3}{p{11cm}}{Number of type $i$ individuals in type $k$ habitat patches at stationarity}  \\
			$\widetilde{N}_{i}^{k}$ & \multicolumn{3}{p{11cm}}{Number of type $i$ individuals in type $k$ habitat patches after dispersal}
		\end{tabular}
		\caption{\textbf{Model parameters.}}	
		\label{tab:parameters}
	\end{center}
\end{table}
\linespread{1.7}

All the model parameters are summarized in Table~\ref{tab:parameters} along with the default parameter values and ranges. If not stated otherwise, the default parameter values are used for the stochastic simulations.

\subsection*{Simulations}
The algorithm implements the life cycle described above. First, a random number of dispersing individuals \chg{from each patch} is drawn from a binomial distribution with success probability $m$ and sample size equal to the \chg{current patch} population size. The dispersing individuals \chg{are pooled and} distributed according to their type and the dispersal pattern. \linelabel{R2-18}\chg{More precisely, we draw a binomial number of individuals of each type to immigrate into an old-habitat patch with probability $\frac{m_i^{\text{old}}}{m}$ (eq.~\eqref{eq:dispersal}) and then distribute these individuals uniformly at random over the old-habitat patches. The same is done with the remaining dispersing individuals who are distributed uniformly into the new-habitat patches.} \linelabel{R2-19}\chg{In each patch, reproduction is simulated by drawing a Poisson distributed number for each type. The mean of this Poisson number is given by the number of individuals of type $i$ in that patch times the mean number of offspring of a single individual of type $i$ which depends on the patch type (old or new).} 
Mutation of the wild-type \linelabel{R2-20}\chg{offspring} \chg{happens} with \chg{probability} $\theta$, implemented using a binomial distribution. Lastly, if necessary, \linelabel{R1-25}\chg{patches are down-regulated back to carrying capacity by sampling individuals uniformly at random without replacement from the population in a patch until the carrying capacity is reached (hypergeometric sampling).}

In Figs.~\ref{fig:vary_m_est}-\ref{fig:origin}, we simulate a heterogeneous environment that is constant in time, i.e. no patches deteriorate. Initial population sizes are $K_{\text{old}}$ wild-type individuals in old-habitat patches, and the \footnote{\francois{add `analytical' or `theoretical'}}stationary wild-type population size $\widehat{N}_w^{\text{new}}$ in new-habitat patches (see Section S1 in the SI for details). \chg{Simulations for} Fig.~\ref{fig:vary_m_est} are started with initially one mutant either in a old- or a new-habitat patch and the corresponding wild-type population size is reduced by one, \chg{and} the mutation probability is set to zero. In \linelabel{R2-21}\chg{Figs.~\ref{fig:source_sink} and \ref{fig:origin}}, mutants solely arise due to mutations \chg{over a finite time period}. 
We \chg{call the mutant established if} the total number of mutants in all new-habitat patches exceeds $0.\chg{6} \times K_{\text{new}} \times M(1- f_{\text{old}})$\footnote{\francois{These thresholds are mysterious at this stage.}} (\chg{or in old-habitat patches $0.6\times K_{\text{old}}\times M f_{\text{old}}$ which is relevant for dispersal schemes where the mutant has a bias towards old-habitat patches, Old-Old and New-Old}).
In the rescue scenario, where one patch after \chg{another} deteriorates (Figs.~\ref{fig:rescue} and \ref{fig:sgv}), we initialize all patches with $K_{\text{old}}$ wild-type individuals. Simulations are run until either the population has gone extinct or the (global) mutant population size exceeds $0.6\times K_{\text{new}}\times M$ after the last deterioration event has happened.

Unless stated otherwise, the simulation results are averages of $10^5$ independent runs. All simulations are written in the C++ programming language and use the \textit{Gnu Scientific Library}. The codes and data to generate the figures are deposited on Gitlab\footnote{https://gitlab.com/pczuppon/evolutionary\_rescue\_and\_dispersal}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% RESULTS %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Results}

We investigate the effect of the different dispersal schemes on the probability of successful adaptation and evolutionary rescue. We first compute the \textit{establishment probability} of a single mutant individual arising either in an old- or in a new-habitat patch. We then link this probability to the dynamics of a source-sink system, i.e. a fixed environment with a certain number of old habitats (sources) and new-habitat patches (sinks). We derive an expression for the probability of a mutation to emerge and establish in a given time interval. We call this second quantity the \textit{probability of adaptation}. Lastly, we study the time-varying scenario where patches, one after \chg{another}, deteriorate. We consider a third quantity, the \textit{probability of evolutionary rescue}, which corresponds to the probability that a mutant appears by mutation and establishes in the new environment \chg{after} all patches have deteriorated. The theoretical results are complemented by stochastic simulations that support our predictions and help visualize the differences between the different dispersal schemes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Establishment Probability %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Establishment probability in a heterogeneous environment}\label{subsec:establishment}

We derive the probability of establishment of a mutant population starting with a single individual initially located either in an old- or a new-habitat patch. In this analysis, we ignore further mutations and are only concerned with the fate of this single mutant lineage. 
The dynamics of the mutant population can be described by a two-type branching process, i.e. \linelabel{R1-27}\chg{all mutants, the descendants of the initial mutant, reproduce, disperse and die independently of each other.} This is a reasonable assumption as long as the overall number of mutants is a lot smaller than the population size of the wild type. The two ``types'' of the two-type branching process correspond to the two habitat types. The process tracks the \linelabel{R2-25}\chg{total number of mutants in old- and new-habitat patches at the end of the life cycle.} The number of offspring of a single mutant can be approximated by Poisson distributed numbers (see eq.~\eqref{eq:sold}). The mean number of offspring of a single mutant, either in an old- or a new-habitat patch, is then given by the following mean reproduction matrix:
%
\begin{equation}\label{eq:mean_repro}
	\bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
		\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
		\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
		}\ ,
\end{equation}
%
where the rows denote the parent locations, and the columns the patch type of the offspring. 
For example, the top-left entry reads as the probability for the parent to stay in an old-habitat patch ($1-m_m^{\text{new}}$) times the average number of offspring in these patches, given by the mean of the corresponding Poisson distribution with rate $(1+a_{\text{old}})$, cf. eq.~\eqref{eq:sold} \linelabel{R2-26}\chg{(see Section S1 in SI for the derivation of $a_{\text{old}}$ at the wild-type equilibrium)}. The other entries are obtained analogously.

\linelabel{R1-24}The survival probability of this multi-type branching process, $\varphi_{k}$ with $k$ indicating the initial habitat type of the mutant, is then given by the unique positive solution of the following system of equations \citep[see][Chapters 5.3 and 5.6]{haccou_book}
\begin{equation}\label{eq:ext_prob}
	\begin{aligned}
		1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \\
		&\qquad \qquad \qquad \left.+ \mathbf{P}(j \text{ offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\\
		&=\exp\left[-\left(1-m^{\text{old}\to \text{new}}_m\right)(1+a_{\text{old}})\varphi_{\text{old}} - m^{\text{old}\to \text{new}}_m (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, ,   \\
		1-\varphi_{\text{new}} &= \exp\left[-m^{\text{new}\to \text{old}}_m(1+a_{\text{old}})\varphi_{\text{old}} - \left(1-m^{\text{new}\to \text{old}}_m\right) (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .
	\end{aligned}
\end{equation}
Intuitively the left hand side, the extinction probability, equals the sum over all possible scenarios of the trajectories towards extinction, e.g. the initial individual having $j$ offspring in a patch of type $k$, old or new, and all of these $j$ offspring becoming extinct ($(1-\varphi_{\text{k}})^j$). Unfortunately, a general analytical solution to these equations is not accessible, but they can be solved numerically. For weak selection and (potentially) weak dispersal -- i.e. $a_{\text{old}},a_{\text{new}},m\ll 1$ needs to hold for at least two of the three parameters -- an approximation is available: see for example~\citet[Theorem~5.6]{haccou_book} for the general theory and \citet{tomasini_2018} for an application in a similar setting. The detailed derivation is presented in the SI, Section~S2. We find
\begin{equation}\label{eq:estab_approx}
	\begin{aligned}
		\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}}&+ & \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
		& & &  \; \; \; + \; \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}\widehat{\pi}_m f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
		\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} &+ & \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
	& & &  \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) \widehat{\pi}_m f_{\text{old}} \right)\, ,}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
	\end{aligned}
\end{equation}
where $C$ is a scaling constant that depends on $m,f_{\text{old}},\widehat{\pi}_m,a_{\text{old}}$ and $a_{\text{new}}$ through
\begin{equation}\label{eq:normalization}
	C = (1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + \widehat{\pi}_m f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, . 
\end{equation}

The first term in the approximation of the establishment probabilities ($\varphi_{\text{old}}$ and $\varphi_{\text{new}}$) in eq.~\eqref{eq:estab_approx} describes the local growth depending on the habitat type under study. 
The second term captures the growth rate differences between the two habitat types.
The factor $(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})$ accounts for the biased dispersal patterns (when $\pi_m\neq 0$ so that $\widehat{\pi}_m\neq 1$). 
The third term in the equations corresponds to the direct effect of dispersal on the establishment probability. 
The first two summands in the bracket are the same for both establishment probabilities. They represent the general effect of dispersal due to the dynamics from new-habitat patches (first summand) and old-habitat patches (second summand). The dispersal bias induced by $\pi_m$ changes the relative impact of old- vs. new-habitat patches. 
Finally, the last summand in the bracket measures the growth rate loss (or gain) due to dispersal to the other patch type. It therefore differs between the two approximations.

Note that in these equations, the competition of the mutant with the wild type in old-habitat patches appears in the local growth rate $a_{\text{old}}$, defined in eq.~\eqref{eq:sold}. This quantity is not constant, but instead depends on the wild type population size in old-habitat patches, which itself depends on the wild type's dispersal and growth rate in both habitats \linelabel{R2-28}\chg{(see SI, Section S1)}.

If the emigration probability is zero ($m=0$), the subpopulations in each habitat evolve in isolation from each other and we recover Haldane's classical result for the establishment probability of a slightly advantageous mutant in new habitats: $\varphi_{\text{new}}=2a_{\text{new}}$ \citep{haldane_1927}. \linelabel{R1-69}\chg{In old-habitat patches, where the mutation is deleterious, we see that term (1) and (2) cancel each other so that the establishment probability is simply $\varphi_{\text{old}}=0$.} Furthermore,\francois{when the emigration probability is strictly positive (m>0)} in the case of \chg{unbiased} dispersal ($\pi_w=\pi_m=0$) and for equal number of old- and new-habitat patches ($f_{\text{old}}=1/2$), we obtain the approximation found in \cite{tomasini_2018} (compare system~\eqref{eq:estab_approx} to their eqs.~(4) and~(5)). Note that the approximation is independent of the actual number of patches, but only depends on the environmental configuration determined by the frequency of old-habitat patches $f_{\text{old}}$.

\subsubsection*{Comparison to simulations and qualitative behavior}
We compare in Fig.~\ref{fig:vary_m_est} our predictions from eqs.~\eqref{eq:ext_prob} and~\eqref{eq:estab_approx} to simulation results for different values of the emigration rate $m$. We find good agreement with the numerical solution of eq.~\eqref{eq:ext_prob} (solid lines). The approximation given in eq.~\eqref{eq:estab_approx} (dashed lines) deviates slightly from the simulation results in regions where $m,a_{\text{new}}$ and $a_{\text{old}}$ are not small, i.e. when the assumptions made in the analytical derivation do not hold anymore.

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig2a.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega_m = 1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig2b.pdf}
  		\caption{$\varphi_{\text{new}}$ with $\omega_m = 1.35$}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig2c.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega_m = 1.45$ (small fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig2d.pdf}
  		\caption{$\varphi_{\text{new}}$ with $\omega_m = 1.45$}
	\end{subfigure}
	\caption{\textbf{Establishment probability when varying the emigration rate.} \small We plot the theoretical results for $\varphi_{\text{old}}$ in (a) and (c), and for $\varphi_{\text{new}}$ in (b) and (d) for $\omega_m=1.35$ in (a), (b) and $\omega_m=1.45$ in (c), (d). Comparison with the results from stochastic simulations (symbols) show very good agreement with our approximation found in eq.~\eqref{eq:estab_approx} (dashed lines). The solid lines are the numerical solution of eq.~\eqref{eq:ext_prob}. The lines in (a) show a clear separation of the three regions of the establishment probability as discussed in the main text: {\color{change}(i) an initial increase due to higher probabilities of dispersal from old- to new-habitat patches; (ii) a local maximum due to increasing number of mutants emigrating from new-habitat patches; (iii) an increase due to relaxed competition in old-habitat patches, a result of wild-type emigration.}}
	\label{fig:vary_m_est}
\end{figure}

The qualitative dependence of the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ on the dispersal probability $m$ is similar for all dispersal schemes. The shape of the curve however strongly depends on the fecundity $\omega_m$ of mutants in the old habitat. Before discussing the differences between the dispersal schemes, we first provide a qualitative understanding of this general behavior.

For the probability of establishment of a single mutant in an old-habitat patch ($\varphi_{\text{old}}$), we observe up to three different regions, cf. Fig.~\ref{fig:vary_m_est}(a). This is in line with previous observations in the context of local adaptation \citep[e.g.][]{kawecki_1995,tomasini_2018} and evolutionary rescue \citep{uecker_2014}. We define the regions as follows: (i) an initial increase of the establishment probability at low dispersal rates $m$; (ii) a local maximum with a subsequent decrease of the establishment probability; (iii) an increase of the establishment probability for high dispersal rates.

A detailed assessment and explanation of the regions is provided in the SI, Section S2.1. Briefly, in region (i), the prevalent effect is the dispersal of mutants from old- to new-habitat patches where they are fitter than the wild type, thus increasing the establishment probability. This effect is mediated through the third term of the establishment probability in eq.~\eqref{eq:estab_approx}. 
Region (ii), beginning with the local maximum, is a result of two counteracting processes: %increased emigration of individuals from the old habitat and \linelabel{R2-30-1}\chg{an increasing probability of an individual that is born in a new habitat emigrates to an old-habitat patch.}
\linelabel{R2-30-1}\chg{If we increase the emigration probability $m$, a mutant is more likely to emigrate out of old-habitat patches and thus has a higher chance to establish. However, higher dispersal rates also have a negative effect on mutant establishment through \linelabel{R2-32}increasing the probability that a mutant migrates from a new- to an old-habitat patch. More precisely, the number of offspring in the new habitat on average is given by the product of the terms $(1-m_m^{\text{old}})$ and $(1+a_{\text{new}})$. This product can, for large emigration probabilities $m$, be smaller than $1$, i.e. a mutant in a new-habitat patch has on average less than one offspring. Finally, in region (iii), dispersal is so large that the population is close to being well-mixed. Most importantly, many wild-type individuals leave old-habitat patches which results in less competitive pressure in old-habitat patches. In this case the local growth rate of the mutant in these patches, $a_{\text{old}}$ (first term in eq.~\eqref{eq:estab_approx}), becomes positive. This effect called `relaxed competition'~\citep{uecker_2014}. The onset of this effect, in terms of the emigration probability $m$, is strongly dependent on the fecundity values of the mutant in the old habitat. The larger it is, the `earlier' (i.e. for smaller emigration rates $m$) relaxed competition becomes relevant.}

%lowering competitive pressure in old-habitat patches \linelabel{R1-28-1}(see $\widetilde{N}_{\text{old}}^w$ from eqs.~(S7) and~(S8) in the SI). As a result, local growth rates of the mutant in old-habitat patches, \linelabel{R1-30} $a_{\text{old}}$, become larger. This effect was termed `relaxed competition' by \citet{uecker_2014}.

%Depending on the onset of this effect, the establishment probability either has a local maximum or not, compare \ref{fig:vary_m_est}(a) and (c).

\chg{In line with this reasoning we find} that the width of region (ii) strongly depends on the fecundity of the mutant in the old habitat, $\omega_m$. If $\omega_m$ is high, the local growth rate of the mutant $a_{\text{old}}$ starts to increase at a lower dispersal rate 
%(see Fig.~S1). 
\chg{and }region (ii) disappears, as visible in Fig.~\ref{fig:vary_m_est}(c). In contrast, for low mutant fecundity values $\omega_m$, region (iii) might vanish (see Fig. S2 in SI). There, the mutant is strongly disadvantageous in old-habitat patches when compared to the wild type, and the wild-type population will always outcompete the mutant due to the much higher offspring numbers.

We analyze the qualitative behavior of the establishment probability of a mutant emerging in the new habitat, $\varphi_{\text{new}}$ in a similar way (Figs.~\ref{fig:vary_m_est}(b,d)). The establishment probability $\varphi_{\text{new}}$ decreases at low dispersal rates: Since we let the mutant start in a new-habitat patch, where it fares better than the wild type, there is no initial benefit due to dispersal (the third term in eq.~\eqref{eq:estab_approx} is negative for $\varphi_{\text{new}}$). The interpretation of region (ii), describing the trajectory for intermediate emigration rates $m$, is the same as for $\varphi_{\text{old}}$ above. For large emigration rates $m$, region (iii), the resulting establishment probability is a combination of the local growth rate and the dispersal effect, the first and third terms in eq.~\eqref{eq:estab_approx}. This is because the mutant can migrate to old-habitat patches, where it will enjoy relaxed competition. %%

\chg{Lastly, for large emigration probabilities $m$ in Fig.~\ref{fig:vary_m_est}(c,d) we observe a decrease in the establishment probability of the simulation results of the New-New dispersal scheme (and to a lesser extent in the unbiased scheme). This is explained by `gene swamping' whereby a to large number of immigrants of a less-well adapted type (wild type) inhibits the establishment of a locally better adapted type (mutant) \citep{nagylaki_1978,lenormand_2002}. Here, gene swamping occurs because of the different carrying capacities of the two patch types, $K_{\text{new}}<K_{\text{old}}$. A proper analysis of the gene swamping regime, where density-dependence in new-habitat patches is important, is beyond the scope of this manuscript, but see for instance \citet{tomasini_2018} and Section S5 in the SI where such a situation is studied.}\footnote{\pete{If I do not mention gene swamping here, I think people will wonder what this effect is and misinterpret it as a late relaxed competition effect.}}

\subsubsection*{Comparison of dispersal schemes}

We now compare the establishment probabilities for the different dispersal schemes. 
We consistently observe that \chg{a bias towards the new habitat (New-New} in Fig.~\ref{fig:vary_m_est}) enhances establishment when compared to the \chg{unbiased} dispersal scheme (0-0). This can be attributed to two reasons: (1) the mutant is more likely to disperse to the new habitat where it outcompetes the wild type (stronger weighting of new-habitat patches in the third term in eq.~\eqref{eq:estab_approx}); and (2) since the wild type prefers to settle in new-habitat patches ($\pi_w<0$), individuals in old-habitat patches experience relaxed competition for lower dispersal rates $m$, thus increasing the local growth rate $a_{\text{old}}$, i.e. region (iii) is shifted to the left. An analogous (but reversed) argument explains why \chg{a bias towards the old habitat (Old-Old}) is always lower than the \chg{unbiased} dispersal scheme.

The effect of \chg{the asymmetric dispersal scheme Old-New}, where each type has a bias to move to the habitat where it is relatively fitter than the other type, is more involved. We disentangle the effects separately for each region. In region (i) at low emigration probabilities, the curve is almost identical to that of the \chg{New-New} pattern. The mutant preference \chg{(new habitats)} of these two dispersal modes is the same. The movement of the rare mutant is the prevailing process in this parameter regime, which explains the alignment of the \chg{corresponding} curves. 
For intermediate dispersal rates, region (ii), the effect of the heterogeneous environment (second term in eq.~\eqref{eq:estab_approx}) becomes stronger, i.e. the local growth rate $a_{\text{old}}$ increases because of relaxed competition in old-habitat patches. 
For the \chg{Old-New scheme}, the\linelabel{R2-30-3} wild type is more likely to immigrate into old-habitat patches and to increase the population size in these locations \linelabel{R1-31}\chg{when compared to the New-New dispersal scheme}. This reduces the effect of relaxed competition. Therefore, the \chg{Old-New line} drops below the \chg{New-New} curve and even starts to decrease (Fig.~\ref{fig:vary_m_est}(a)). For high emigration probabilities $m$, region (iii), the competitive pressure in old-habitat patches relaxes. Again, this region is dominated by the movement of wild-type individuals. In the \chg{unbiased} dispersal scheme, the wild type has a lower likelihood to be in old-habitat patches than with the \chg{Old-New} scheme, explaining the \chg{order of these} curves. 
For very large dispersal rates, also\francois{moreover?} the \chg{Old-Old} scheme can lead to higher values of $\varphi_{\text{old}}$ than \chg{the type-dependent preference Old-New}, cf. Fig.~\ref{fig:vary_m_est}(c). For these parameters, the mutant has a larger growth rate in old- than in the new-habitat patches ($a_{\text{old}}>a_{\text{new}}$). It is therefore beneficial for the mutant to \linelabel{R2-30-4}\chg{be in} old-habitat patches.

Finally, for the \chg{asymmetric dispersal scheme New-Old, where the types have a bias to the habitat they are less well adapted to, }we see that for low dispersal rates $m$ the results match those obtained under the \chg{Old-Old} scheme. The mutant preferences being the same, this again shows that this region is dominated by the mutant's dispersal behavior. For intermediate to large dispersal rates though, the \chg{New-Old} scheme results in the largest establishment probabilities. This is explained by the effect of strongly relaxed competition (wild-type individuals preferentially move into the new habitat) and the bias of mutants to disperse into old-habitat patches where their number of offspring individuals is now very large. %Overall, this results in the highest establishment probabilities of all dispersal schemes. %We note though that the term `maladaptive dispersal' might be misleading for large dispersal values, since here the growth rate of the mutant in old-habitat patches, $a_{\text{old}}$, is actually larger than the growth rate in new-habitat patches, $a_{\text{new}}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Emergence probability %%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Probability of adaptation in a heterogeneous environment} \label{subsec:source_sink}

We now study the probability of adaptation when mutations occur recurrently. As in the previous section, we consider a heterogeneous environment with a fixed number of old- and new-habitat patches. This is effectively a source-sink system \citep{holt_1985,pulliam_1988}, where old- and new-habitat patches correspond to sources and sinks for the wild type, respectively. In the previous section, we initialized the system with one mutant in either an old- or a new-habitat patch and computed the establishment probability. Now, we let mutants appear randomly within a certain time frame. The last time point at which a mutation can occur is denoted by $t_{\text{fin}}$.
Later, in the analysis of the probability for evolutionary rescue, this time will be replaced by the time between two consecutive patch deterioration events, $\tau$.\footnote{\pete{I think we can get rid of this last sentence. Thoughts?}\francois{Yes.}}

The probability of adaptation in this setting, $P_{\mathrm{adapt}}$, is given by
\begin{equation}\label{eq:source_sink}
	P_{\mathrm{adapt}} = 1 - \exp\left(-\theta t_{\text{fin}} M \left(\underbrace{\varphi_{\text{old}} f_{\text{old}} K_{\text{old}}}_{\text{old habitat contribution}} + \underbrace{\varphi_{\text{new}} (1-f_{\text{old}}) \widehat{N}_w^{\text{new}}}_{\text{new habitat contribution}}\right)\right)\, .	
\end{equation}
In words, the right-hand side is one minus the probability of no mutant establishing within the $[0, t_{\text{fin}}]$ time interval.
More precisely, the exponential is the probability of zero successes of a Poisson distribution. The rate of this Poisson distribution is given by the expected number of successfully emerging mutant lineages until time $t_{\text{fin}}$. \linelabel{R2-34}\chg{Assuming a Poisson distribution for the number of successful mutant lineages is an approximation of a Binomial distribution with large sample size (the wild-type populations $t_{\text{fin}} M f_{\text{old}} K_{\text{old}}$ in the old and approximately $t_{\text{fin}} M (1-f_{\text{old}}) \widehat{N}_w^{\text{new}}$ in the new habitats, see SI eq.~(S5)) with small success probability (the establishment probabilities $\varphi_{k}$). The sample size denotes the number of mutations in the system until time $t_{\text{fin}}$ and each of these mutants has, independently of each other (another approximation), has a survival probability of $\varphi_{k}$, dependent on the habitat it is born in.}
%Correctly, at least in the old habitat, one would have a Binomial distribution with success probability $\theta \varphi_{\text{old}}$ and sample size $t_{\text{fin}}M f_{\text{old}} K_{\text{old}}$ which is well approximated by a Poisson distribution for large sample sizes and small success probabilities. Due to the varying population size in the new habitat, the exact probability for successful lineages emerging from that habitat type is more difficult to compute. Yet again, on average this number is well approximated by the a Poisson distribution with a rate as stated above.}\footnote{\pete{This is the long explanation. I would shorten but if you think it is worth keeping a long explanation (as suggested by the reviewer) we can leave it as it is.}}
Note also, that for $t_{\text{fin}}$ tending to infinity, there will almost surely be a successful mutant so that $P_{\text{adapt}}=1$. \\
%For finite values of $t_{\text{fin}}$, the average number of mutations that appear until that time is given by the product of the mutation probability $\theta$, the length of the time window of interest $t_{\text{fin}}$, the number of patches $M$ and the stationary number of wild-type individuals present in the different habitats, that is $K$ and $\widehat{N}_w^{\text{new}}$ for old- and new-habitat patches, respectively (cf. SI eq.~(S5)). Once such a mutant appears, it then has the previously computed establishment probability $\varphi_{k}$ to establish, depending on the patch type $k$ (old or new) it arises in. This explains the factor $\varphi_k$ in the exponential.  \\

In Fig.~\ref{fig:source_sink} we compare our predictions to simulation results. The shape of the probability of adaptation as function of the emigration probability $m$ depends on the fecundity of the mutant in old-habitat patches, $\omega_m$ (Figs.~\ref{fig:source_sink}(a,c)). This is similar to the behavior of the establishment probability $\varphi_{\text{old}}$ in Fig.~\ref{fig:vary_m_est}. Likewise, the qualitative effects of the different dispersal schemes are comparable to the ones observed for the establishment probability. \linelabel{R1-33}\chg{This holds because the probability of adaptation, $P_{\text{adapt}}$, is given by $1-\exp(-C_1 \varphi_{\text{old}} -C_2 \varphi_{\text{new}})$ where two constants $C_1$ and $C_2$ depend on the dispersal scheme (see eq.~\eqref{eq:source_sink}). \linelabel{R2-35}The small differences in the ranking of the dispersal schemes between subfigures Figs.~\ref{fig:vary_m_est}(c) and Figs.~\ref{fig:source_sink}(c) can be explained by the mutational input, i.e. larger stationary populations sizes of the wild type in new habitats, $\widehat{N}_w^{\text{new}}$ result in more mutations during the observed time frame. This population size is altered by the different dispersal schemes, most importantly it is largest for the wild type having a bias towards old habitats, i.e. the Old-Old and the Old-New dispersal schemes.}

\begin{figure}[t!]
	\centering
		\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig3a.pdf}
  		\caption{$\omega_m=1.35$ (large fecundity difference)}
	\end{subfigure}%
	    \begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig3b.pdf}
  		\caption{$\omega_m=1.35$}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig3c.pdf}
  		\caption{$\omega_m=1.45$  (small fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig3d.pdf}
  		\caption{$\omega_m=1.45$}
	\end{subfigure}
	\caption{\textbf{Probability of adaptation in a heterogeneous environment.} \small In (a) and (c), we vary the emigration rate $m$ and observe a similar qualitative behavior as for the establishment probability $\varphi_k$ in Fig.~\ref{fig:vary_m_est}. In (b) and (d), we vary the frequency of old-habitat patches. The maximum is the result of two counteracting processes. The higher the number of old-habitat patches (the greater $f_{\text{old}}$), the larger the wild-type population. As a consequence, more mutants appear in the studied time-frame. In contrast, the less old-patch habitats there are in the environment (the lower $f_{\text{old}}$), the higher the probability of successful establishment of a mutant population. The curves labeled `approx.' are given by eq.~\eqref{eq:source_sink}, the exact solution refers to solving the establishment probabilities $\varphi_k$ from eq.~\eqref{eq:ext_prob} numerically and plugging these solutions into eq.~\eqref{eq:source_sink}. In all panels, the mutation probability is $u=1/(M K)$ and the final time for a mutant to appear is $t_{\text{fin}}=100$. }
	\label{fig:source_sink}
\end{figure}

In subfigures (b) and (d), we plot the probability of adaptation as a function of the frequency of old-habitat patches $f_{\text{old}}$. We observe a maximum which is the result of two effects: (1) the likelihood for a mutation to appear increases with the number of wild-type individuals present in the system, which is highest for high frequencies of old-habitat patches $f_{\text{old}}$, and (2) the probability of establishment of a mutant decreases with the number of old-habitat patches.

The different dispersal schemes alter both effects. \chg{A general bias towards the new habitat (New-New)}, when compared to \chg{unbiased} dispersal (0-0), always shifts the maximum to higher frequencies of old habitats while also increasing its quantitative value. \chg{Since for the New-New dispersal scheme, both types prefer settling in new-habitat patches ($\pi_w,\pi_m<0$), the local population before reproduction in these patches is increased. Thus, the overall population size is higher compared to the other dispersal schemes and therefore a higher number of mutants is generated under this scheme.} Additionally, the probability of establishment is also increased for \chg{the New-New} dispersal scheme, further increasing the probability of adaptation (see also the discussion around Fig.~\ref{fig:vary_m_est}). Again, a reversed argument explains why \chg{the general preference for old-habitat patches (Old-Old)} always yields lower probabilities of adaptation than \chg{unbiased} dispersal.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Origin of rescue mutation %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Habitat of origin of the adaptive mutation}
We now identify the habitat type where the successful mutation arises. The established mutant population often arises from a single mutant that is born in an old- or in a new-habitat patch. However, the rescued population can sometimes be traced back to two (or more) mutant individuals, \chg{at least} one from either patch type. \linelabel{R1-35}\chg{This is a special type of a soft selective sweep \citep[see][for a review]{hermisson_2017}. The approximate probability to observe a soft sweep from mutants from both habitat types is 
\begin{equation}\label{eq:origin}
    \begin{aligned}
	&\mathbf{P}\big(\text{successful adaptation from old habitat}\big)\ \mathbf{P}\big(\text{successful adaptation from new habitat}\big) \\
	& \qquad \qquad \approx \left(1 - \exp\left(-\theta t_{\text{fin}} M \varphi_{\text{old}} f_{\text{old}} K\right)\right)\left(1- \exp\left(-\theta t_{\text{fin}} M \varphi_{\text{new}} (1-f_{\text{old}}) \widehat{N}_w^{\text{new}}\right)\right)\, .
	\end{aligned}
\end{equation} 
}
\linelabel{R1-36}\chg{The approximation uses our basic assumption that different mutant individuals and consequently also the offspring of different mutants do not affect each others dynamics (branching process). In the simulations, we label a run as a soft sweep from mutants arising in different locations if these lineages are still alive after $1000$ generations. This ensures that we do not count any false-positives where a mutant in one of the habitats has just arisen right before the mutant population exceeds the establishment threshold.}

%The probabilities for the other two scenarios can be computed analogously. 
In Fig.~\ref{fig:origin} we compare simulation results with our predictions for the origin of a successful mutant when varying the old habitat frequency $f_{\text{old}}$. \chg{\linelabel{R2-36}We see that for both mutant fecundity values most successful mutations arise in old-habitat patches. The difference in the input from the two different patches is the factor $\varphi_{k} \widehat{N}_w^k f_k$. In Fig.~SX we plot the establishment probability $\varphi_k$ and the stationary population sizes $\widehat{N}_w^k$ when varying the frequency of old-habitat patches $f_{\text{old}}$. We see that the establishment probability for mutants arising in new-habitat patches, $\varphi_{\text{new}}$, is always larger than the corresponding probability for mutants from old-habitat patches, $\varphi_{\text{old}}$. Conversely, the stationary population size of the wild type in the old habitat measured over the whole environment is always larger then that of the wild type in new habitats. Combining these two observations we find that for this choice of parameters the mutational input, i.e. the absolute number of rescue mutants, has a larger influence on the origin of the rescue mutant than the corresponding establishment probability.}

%Even for a relatively strong fecundity disadvantage of the mutant in the old habitat, i.e. $\omega_m$ is $10\%$ smaller than $\omega_w$, we find that most successful mutations arise in old-habitat patches (Fig.~\ref{fig:origin}(a)). Decreasing the fecundity disadvantage of the mutant by increasing its fecundity $\omega_m$ further increases the number of successful mutant lineages from old-habitat patches (Fig.~\ref{fig:origin}(b)). 
%If we instead increase the fecundity difference in old habitats by decreasing $\omega_m$, we observe the opposite, i.e. \chg{the two lines approach each other} (see Fig.~S4 in SI).

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig4a.pdf}
  		\caption{$w_m = 1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig4b.pdf}
  		\caption{$w_m = 1.45$ (small fecundity difference)}
	\end{subfigure}
	\caption{\textbf{Origin of the adaptive mutant.} \small The origin of the adaptive mutant is strongly affected by the fecundity differences in the old habitat. If the difference is large as illustrated in panel (a), mutants appear more often in old-habitat patches than in new-habitat patches. Still, mutants arising in new-habitat patches contribute to the overall probability of adaptation. If fecundity differences are small like in panel (b), the successful mutant largely arises in old-habitat patches. In this case, the contribution from new-habitat patches is negligible. Circles correspond to simulations identified as soft selective sweeps with (at least) one lineage \francois{successful mutation} arising in an old- and one in a new-habitat patch. The curves are given by eq.~\eqref{eq:origin} (or the adjusted versions of it) under the unbiased dispersal scheme ($\pi_w=\pi_m=0$). Note the different scaling on the y-axes.}
	\label{fig:origin}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evolutionary Rescue %%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evolutionary rescue}
Finally, we consider a time-inhomogeneous environment where patches deteriorate one after \chg{another} at regular time intervals $\tau$, until all patches have switched to the new habitat. If the wild-type population fails to generate a successful mutant, the population will \chg{inevitably} go extinct. The probability of evolutionary rescue is therefore tightly linked to the probabilities of adaptation and establishment that we have computed in eqs.~\eqref{eq:estab_approx} and~\eqref{eq:source_sink}. Typically, in formulas expressing the probability of evolutionary rescue, one splits the contributions into mutations arising \textit{de novo} and evolutionary rescue due to standing genetic variation, i.e. mutations that are present in the population before the environmental change \citep{alexander_2014}. We will discuss the effect of standing genetic variation in our system in the following section. For now, we focus on evolutionary rescue due to de novo mutations. We approximate the probability of evolutionary rescue, denoted by $P_{\text{rescue}}$, as
\begin{equation}\label{eq:evol_rescue}
    \begin{aligned}
		P_{\text{rescue}} &\approx 1-\exp\left(- \theta \sum_{i=0}^{M-2} \left(\underbrace{\varphi_{\text{old}}\left(f_{\text{old}}(i)\right) \sum_{j=i\tau}^{(i+1)\tau -1} N_w^{\text{old}}(j)}_{\text{old habitat contribution}} + \underbrace{\varphi_{\text{new}}(f_{\text{old}}(i)) \sum_{j=i\tau}^{(i+1)\tau -1} N_w^{\text{new}}(j)}_{\text{new habitat contribution}}\right)\right.\\
		&\qquad\qquad \qquad\qquad\qquad\qquad\qquad \qquad \qquad \qquad\quad  \left.\underbrace{-\theta \varphi_{\text{new}}(0) \sum_{j = \tau (M-1) }^\infty  N_w^{\text{new}}(j)}_{\text{contribution after the last patch deteriorated}}\right),
	\end{aligned}
\end{equation}
where \linelabel{R1-40}$f_{\text{old}}(i) = (M-i-1)/M$ is the frequency of old-habitat patches after the $(i+1)-$th deterioration event\linelabel{R1-39}\chg{, the establishment probability is given as a function of the patch frequency, $\varphi_k(f_{\text{old}}(i))$,} and $N_w^{k}(j)$ denotes the overall number of wild-type individuals living in habitat $k$ (old or new) in generation~$j$ \chg{(see SI, Section S1 for an approximation)}. The interpretation of this equation is the same as for eq.~\eqref{eq:source_sink}. The only difference is that we now need to account for a changing environment, which alters the population sizes, $N_w^k$, and the establishment probabilities $\varphi_k$ \chg{over time}. In the formula, these changes are accounted for by the sums that iterate through the (discrete) time steps and by the time dependence of the corresponding quantities. We further note that we follow the \chg{expected value of} the wild-type population size deterministically over time, instead of assuming it to be already in its steady state as in eq.~\eqref{eq:source_sink} \chg{(see also Section S1 in the SI)}.

As visible in Fig.~\ref{fig:rescue}, the approximation explains the \linelabel{R1-42}\chg{order of dispersal schemes}. Yet, it does not accurately predict the simulated data. This discrepancy can be explained: in the formula we \linelabel{R2-42}\chg{assume that for a mutant born in a certain patch configuration, say with $j$ old-habitat patches, the environment does not change anymore. That is, a mutant born in patch-type $k$ (old or new) in this environment contributes $\varphi_{k}(j)$ to the probability of evolutionary rescue despite further patches deteriorating. Thus, this mutants probability of establishment is underestimated. This is especially true for mutants that emerge just before a deterioration event. These mutants are more likely to establish during the subsequent environmental configuration giving them a higher probability of survival than approximated by our formula. Additionally, $\varphi_k(j)$ assumes stationary wild-type population sizes and therefore does not reflect the decreasing wild-type population size right after the deterioration of a patch.} 
%However, mutants that arise very shortly before a deterioration event just need to survive until this event happens and are thus carried over to the following environmental configuration with a higher probability than predicted by our formula. 
This explains why the simulation results are higher than our approximation. 
%which ignores these ``left-over'' mutants present prior to the deterioration event. 
\chg{A time-dependent establishment probability could account for these effects but unfortunately is not amenable to approximations in our framework.}
%As a result, instead of assuming a constant establishment probability between two deterioration events, we should rather use a time-dependent establishment probability -- which can however not be approximated in our general framework. 
\linelabel{R1-43}\chg{In \citet{uecker_2014}, scenarios with an accessible time-dependent solution were studied, more precisely situations with either full mixing of the global population ($m=1$) or a non-viable mutant in old-habitat patches ($\omega_m=0$).}

\begin{figure}[t!]
	\centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig5a.pdf}
  		\caption{$\omega_m=1.45$ (small fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig5b.pdf}
  		\caption{$\omega_m=1.35$ (large fecundity difference)}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig5c.pdf}
  		\caption{$\omega_m=1.1$ (very large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{fig5d.pdf}
  		\caption{$\omega_m=0$ (lethal mutant)}
	\end{subfigure}
	\caption{\textbf{The probability of evolutionary rescue compared to simulation results.} \small Our predictions, computed with eq.~\eqref{eq:evol_rescue}, match the qualitative behavior of the simulated data for the probability of evolutionary rescue. All rankings of the dispersal schemes align well. Quantitatively though, we find that our predictions tend to underestimate the simulated data. In (a,b) the mutation probability is set to $u=1/(25MK_{\text{new}})$ while in (c,d) it is $u=1/(MK_{\text{new}})$. The label `exact' refers to the exact solution of eq.~\eqref{eq:ext_prob} which is then plugged into the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue}.}
	\label{fig:rescue}
\end{figure}

The different dispersal schemes exhibit substantial differences. The dispersal schemes affect the dispersal pattern of both types and as such alter their respective population dynamics. Both small and large fecundity differences in the old habitat, $\omega_m=1.45$ and $\omega_m=1.35$, conserve the ranking of dispersal schemes obtained for the probabilities of establishment and adaptation\footnote{\francois{difficult to follow}\pete{We could also get rid of the explict description of the figure and remove the following sentence}}: \linelabel{R2-43}\chg{New-New (green) is higher than unbiased dispersal (black) which is higher than Old-Old (blue), with Old-New decreasing and New-Old increasing their positions with increasing emigration probability $m$}, cf. Fig.~\ref{fig:rescue}(a,b). Based on our discussion of the same behavior in Fig.~\ref{fig:vary_m_est}(a,c), \linelabel{R2-44}\chg{i.e. the change in hierarchy of the dispersal schemes Old-New and New-Old,} The most influential factor in these parameter sets is the growth rate of the mutant in old-habitat patches, $a_{\text{old}}$, \chg{as discussed around Fig.~\ref{fig:vary_m_est}.} \chg{The ranking of the symmetric dispersal schemes is mostly determined by the mutant dispersal bias.}

For very large fecundity differences in the old habitat ($\omega_m = 1.1$ and $\omega_m=0$), the ranking of dispersal schemes is the same for all emigration probabilities; from highest to lowest, we find \chg{New-New, Old-New, unbiased dispersal, New-Old, and Old-Old} (cf. Fig.~\ref{fig:rescue}(c,d)). In this parameter setting, the probability of evolutionary rescue is dominated by the dispersal behavior of the mutant. Since the mutant is barely viable in old-habitat patches, it is always preferential for it to disperse towards new-habitat patches. Two schemes follow this rule, \chg{the New-New and Old-New dispersal scheme}, but differ in the preference of the wild type. 
Since wild-type individuals also preferentially disperse to new-habitat patches under \chg{the New-New scheme}, more individuals are present in those patches. Therefore, the total amount of mutations over the deterioration time is increased which generates more mutants on average. This explains the New-New dispersal scheme being consistently higher than the Old-New dispersal scheme.
%This effect is even stronger for large dispersal rates that increase the wild-type population size in the whole environment. Together with the effect of relaxed competition, this explains the strong increase of the \chg{New-Old}, and the \chg{unbiased} dispersal scheme for large emigration probabilities $m$ in Fig.~\ref{fig:rescue}(c).

Lastly, the probability of evolutionary rescue often reaches a local (or global) maximum for intermediate emigration probabilities (Figs.~\ref{fig:rescue}(c,d)). This extends previous results \citep{uecker_2014,tomasini_2019} to arbitrary dispersal schemes affecting the immigration process. The maximum can be attributed to the interaction of the three regions we identified when analyzing the establishment probability $\varphi_k$ (see also Fig.~\ref{fig:vary_m_est}). As such it is a result of the largely positive effect of dispersal (initial increase of the probability of evolutionary rescue) and the negative effect of population mixing \chg{(dispersal to old patches)}. \footnote{\francois{This discussion is still difficult.}}

\subsection*{Habitat of origin of the rescue mutant and standing genetic variation}
Similar to what \chg{we} found for the probability of adaptation, rescue mutants mainly originate from old-habitat patches. Mutations are more likely to appear in the more populated patches (old-habitat). However, a low mutant fecundity $\omega_m$ decreases the chance of establishment of these mutants that appear in old-habitat patches (compare black and yellow symbols in Fig.~\ref{fig:sgv}(a)). Lower fecundity values $\omega_m$ therefore comparatively increase the probability for the rescue mutation to have appeared in the already deteriorated habitat, i.e. in new-habitat patches. \linelabel{R1-35-2}\chg{The probability of a soft selective sweep for evolutionary rescue from mutants of the different habitat types is very low (circles in Fig.~\ref{fig:sgv}(a)). Our choice of a small mutation rate implies a hard selective sweep regime ($\theta K_{\text{old}} M = 0.08<1$) \citep{wilson_2018,hermisson_2017}. }

\begin{figure}[t]
	\centering
		\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig6a.pdf}
  		\caption{}
	\end{subfigure}%
    \begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{fig6b.pdf}
  		\caption{}
	\end{subfigure}
	\caption{\textbf{Habitat of origin of the rescue mutation and the impact of standing genetic variation.} \small (a) We compare the source of successful mutations under small (black) and large (yellow) fecundity differences in the old habitat. Decreasing the fecundity of the mutant results in more successful mutations emerging in new-habitat patches ($+$) when compared to the contribution from old-habitat patches ($\times$). We have chosen $\pi_m=\pi_w=0$. (b) For slower environmental degradation, i.e. $\tau=200$, the influence of standing genetic variation (sgv) on the probability of evolutionary rescue decreases. The simulations are done by letting the system evolve for $1,000$ generations before the first deterioration event happens. Parameters: $\pi_m=\pi_w=0$ in all scenarios and $\omega_m=1.45$. The relative contribution is then determined by $(P_{\text{rescue with sgv}}-P_{\text{rescue only de novo}})/P_{\text{rescue with sgv}}$.}
	\label{fig:sgv}
\end{figure}

So far, we have considered settings where evolutionary rescue is exclusively due to de novo mutations. To explore the role of standing genetic variation\footnote{\pete{or variance? is there a difference? why use one over the other?}}, we ran simulations \linelabel{R1-45}where we let the system evolve for $1,000$ generations before the first degradation event happened. Fig.~\ref{fig:sgv}(b) shows the relative contribution of de novo mutations and of standing genetic variation, i.e. mutations that appeared before the first degradation event happened. \linelabel{R2-39}\chg{For a successful rescue event due to standing genetic variation, mutants that were initially present (at time $t=0$) need to survive at least until sufficiently many patches have deteriorated that the probability of adaptation, $P_{\text{adapt}}$, becomes positive, compare to Fig.~\ref{fig:source_sink}(b,d) and Fig.~SX in the SI.} \linelabel{R1-46}\chg{Decreasing the value of $\tau$ reduces the time between two consecutive degradation events and the overall time-span of the entire environmental change. Therefore, the impact of standing genetic variation on the probability of evolutionary rescue increases for smaller values of $\tau$.}

Additionally, the relative contribution of standing genetic variation declines as the emigration rate $m$ increases. 
This holds because \linelabel{R1-49}\chg{the mutant is almost exclusively found in old-habitat patches if the frequency of these patches, $f_{\text{old}}$, and emigration rates $m$ are high} (see Fig.~SX in SI). Thus, mutants that existed prior to the first deterioration event are very unlikely to survive even for a rapidly changing environment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% DISCUSSION %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Discussion}
%
% Brief summary
We have studied the probabilities of establishment, adaptation and evolutionary rescue under four non-\chg{uniform} dispersal schemes and compared them to \chg{unbiased} dispersal. Our analysis builds on the probability of establishment of a single mutant lineage in a heterogeneous environment with a fixed patch configuration. In line with previous results, we find that the probabilities of establishment, adaptation and evolutionary rescue can display up to three different phases when varying the dispersal rate $m$.
The different dispersal schemes, by changing the population dynamics, \chg{alter the parameter range of these regions.}\footnote{\francois{Unclear. Suggestion (not great but a bit better maybe): "The dispersal scheme determines the population dynamics and consequently the parameter regions corresponding to the three phases."}}
%broaden, tighten or shift these regions.

\subsection*{Dispersal and adaptation}
% General comparison to other models and our resulting formula
Theoretical studies that investigated the effects of spatial subdivision on the adaptation of a population in a heterogeneous environment can be classified into two types. One type of models, classically analyzed in a population genetic context\francois{framework}, assumes constant population sizes in all patches, independent of their local habitat type and of dispersal strength. Results obtained in this framework show that larger dispersal rates tend to decrease the probability of successful establishment of a rare mutant favored in some part of the environment \citep[e.g.][]{garcia_1997}. This \st{inhibiting effect of dispersal on adaptation, also} \francois{effect} termed ``gene swamping'', is a result of an increase in absolute numbers of non-adapted individuals in the habitat where the rare mutant is beneficial \linelabel{R1-51}\chg{before reproduction and density regulation. This results in a lower mutant frequency \citep{lenormand_2002,tomasini_2018} and thus in lower reproductive success.} Additionally, for very high dispersal rates, the population homogenizes and individuals encounter an averaged environment. Therefore, the type with the largest overall growth rate, averaged over the environment, is favored.  \francois{In our model, gene swamping was rarely observed; the dominant effects were explained by the impact of dispersal on demography.} \footnote{\francois{Would consider shortening this paragraph which is not directly relevant. e.g. remove the sentence "this results in a lower mutant frequency...", but keep the references}}.

The second type of models explicitly takes into account demographic effects due to dispersal, often in the context of source-sink systems \citep{holt_1985,pulliam_1988}. 
Here, the effect of dispersal on adaptation depends on the growth rate differences of the mutant and the wild type in the two habitats \citep{kawecki_2000}. 
In accordance with this result \francois{these results}, we find that dispersal \francois{monotonically} increases the probability of adaptation if the mutant is just slightly less fit than the wild type in the old habitat %, $\omega_m / \omega_w= 0.99$
(Fig.~\ref{fig:vary_m_est}(c)). 
When the mutant's fecundity in old habitats, $\omega_m$, is smaller \francois{than what?}, establishment probabilities are non-monotonic with a local maximum at intermediate dispersal rates \francois{and increases at large dispersal rates thanks to relaxed competition} (Fig.~\ref{fig:vary_m_est}(a)). 
\st{However, if the mutant's disadvantage in old-habitat patches is even greater, i.e. $\omega_m$} \francois{When the fecundity of the mutant is even smaller, the local maximum remains but relaxed competition no longer occurs} \st{adaptation is also hindered for large dispersal rates. The establishment probability becomes hump-shaped } (cf. Fig.~S2 in SI). \footnote{\francois{I was confused by this paragraph so rephrased to make it more uniform and better describe the figures and the key differences}}.

\st{Hence, models with and without demographic change differ most at higher emigration rates. This corresponds to region (iii) in our analysis; we illustrate this point in} \francois{Compared to models with implicit demography, explicitly modelling the population dynamics causes relaxed competition that increases the probability of establishment at high dispersal rates }Fig.~S5 in SI, the non-demographic version of Fig.~\ref{fig:vary_m_est}.

\francois{In figure S5, there does seem to have relaxed competition for curves orange and purple. So it's a bit confusing. Maybe remove those? Also I think a more balanced discussion of the three phases: initial rise thanks to exportation of mut in new habitats; decline due to back-migration; final increase due to relaxed competition, and which effect appears or not in the pop gen model, would be nice. Right now we focus a lot on relaxed competition.}

\subsection*{\linelabel{R1-52}\chg{Standing genetic variation} and evolutionary rescue}

\st{Besides the general structure of the probability of evolutionary rescue when varying the emigration probability $m$, }we also studied the effect \francois{contribution} of standing genetic variation on this quantity \francois{to evolutionary rescue}. 
We find that the contribution of standing genetic variation on the overall probability of evolutionary rescue increases with the speed of the environmental change, determined by $\tau$ (Fig.~\ref{fig:sgv}(b)). 
This observation has also been made in a quantitative genetics setting where the adaptive trait is continuous (and not discrete as in our model) \citep{matuszewski_2015}. Experimental results with \textit{Caenorhabditis elegans} indicate that under slow environmental change the impact of standing genetic variation is small \citep{guzella_2018}. This is because the evolution of the trait is driven by \textit{de novo} mutations with small effects. For a sudden or fast environmental change (small $\tau$), standing genetic variation becomes increasingly important for the probability of evolutionary rescue. These observations are in line with our findings in Fig.~\ref{fig:sgv}(b). %and other theoretical results obtained in this context \citep{matuszewski_2015}.

\subsection*{The effect of \chg{biased} dispersal patterns on adaptation and evolutionary rescue}
The importance of considering dispersal schemes \chg{other than} \chg{unbiased} dispersal has been highlighted in several papers \citep{edelaar_2008,clobert_2009,edelaar_2012}. This has led to a number of simulation studies exploring the effect of various dispersal schemes onto \chg{(local)} adaptation %and niche width
\citep[e.g.][]{vuilleumier_2010,holt_2015,mortier_2018,pellerin_2018}.

Two of these simulation studies examined the effects of matching habitat choice on adaptation \chg{in} a heterogeneous environment \citep{vuilleumier_2010,holt_2015}. Both investigations indicate that matching habitat choice increases the probability of adaptation when compared to \chg{unbiased} dispersal. 
This is in line with our findings: we predict that type-dependent habitat choice \chg{where each type favors patches they are relatively more fit in (Old-New),} generates higher probabilities of establishment \chg{(}and evolutionary rescue\chg{)} than \chg{unbiased} dispersal \chg{(0-0)} (Figs.~\ref{fig:vary_m_est}, \ref{fig:source_sink}, \ref{fig:rescue}).

The dispersal schemes also slightly affect the origin of the successful mutant lineage, cf. SI, Fig.~S6. Population densities, especially in new-habitat patches, are altered when compared to the \chg{unbiased} dispersal scheme.\footnote{\pete{This paragraph could be dropped in line with a comment of a reviewer.} \francois{agree to drop it}}

To conclude, the effects of dispersal schemes are two-fold. By changing population densities in both habitat types, the dispersal schemes \chg{alter} the growth rate of the mutant in old-habitat patches. 
This is the primary reason for the ranking of the dispersal schemes. In addition, they also affect the number of mutations arising in either habitat type. This has a minor effect on the probability of evolutionary rescue for the explored parameter range but is relevant when studying the origin of the successful mutant\florence{mutation} lineage \chg{(see also SI, Fig.~S6)}. \francois{As the genetic background may vary across patches, the origin of a successful mutation will also affect which neutral and deleterious mutations will hitchhike with it}. \linelabel{R1-54}\chg{The origin of a successful mutant is of special interest once multiple loci are considered\francois{. Rescue by a single mutant implies} \st{which might promote} neutral (or even deleterious) hitchhiking effects. Similarly, in the case of polygenic rescue or under recombination \citep[e.g.][]{uecker_2015}, the origin of a mutant is likely to affect its success.}\florence{[mention genetic background]}

% \subsection*{Evolutionary consequences of habitat choice}
% The interplay of local adaptation and habitat choice has interested evolutionary ecologists for several decades, see for example \citet{rosenzweig_1981} for one of the first references. 
% Our analysis shows that absolute habitat choice lowers the probability of establishment when compared to random dispersal; see blue lines in Fig.~\ref{fig:vary_m_est}. 
% This indicates that habitat choice protects a population against locally deleterious mutations, that are potentially favorable elsewhere, if the mutant has the same dispersal pattern as the wild type, i.e. $\pi_m=\pi_w>0$. A similar observation, protection of the already established type, was made in \citet{ravigne_2009} where a co-evolutionary model of a habitat choice trait and a local adaptation trait is implemented. 
% %This is achieved by maintaining a large local density in old-habitat patches which increases the local selection strength. Therefore, fecundity differences between the mutant and the wild type have a strong impact on the composition of the next generation.

% Moreover, \citet{ravigne_2009} find that habitat choice can actually evolve and lead to evolutionary branching, i.e. the coexistence of two specialist types with their own habitat preferences (our relative habitat choice dispersal scheme). 
% %In a scenario where habitat choice and adaptation to the habitat are two independent traits, the adaptive mutant will share the habitat preference of the wild type at its emergence. For strong habitat preferences (large $\pi_w$) this can result in new patches remaining empty. This observation was also made in \citet{ravigne_2009}\footnote{\florence{Shorten; just say habitat choice could evolve (Ravigne 2009)}} -- even though in a different model not exhibiting source-sink dynamics and under a different regulation scheme. 
% %In that study, the authors analyzed the co-evolution of a habitat choice trait and a local adaptation trait. 
% %Translating their results (from their `model~1' that is closest to our life cycle) into our framework they find that relative habitat choice (orange) enhances evolutionary branching, i.e. the coexistence of two specialist types with their own habitat preferences. In contrast, absolute habitat choice protects the already established type (or at best creates a generalist type for the whole environment). 
% In line with this, our results show that in parameter regions where relative habitat choice yields higher probabilities for establishment than random dispersal, if a mutant is able to quickly evolve its own habitat preference as it does in \citet{ravigne_2009}, it has an increased chance of establishing in the meta-population. This is in agreement with recent experimental findings in the ciliate \textit{T. thermophila} \citep{jacob_2017}. Even though we do not explicitly consider speciation and the evolution of genetic polymorphisms, our results may suggest that type-dependent matching habitat choice enhances phenotypic divergence, if we use the probability of adaptation as a proxy for invasion probability. The probability of adaptation for relative habitat choice (orange) is higher than for absolute habitat choice (blue) or random dispersal (black), Fig.~\ref{fig:source_sink}. This idea has already been promoted in other theoretical and experimental studies \citep[e.g.][]{rosenzweig_1987,rice_1990,ravigne_2009,berner_2015,jacob_2018}.

\subsection*{Generality of our theoretical analysis and future directions}
Our mathematical results apply in the case where the mutant offspring numbers can be written in the form ``migration times reproduction (and potential regulation)'' with rates that are constant in time (\linelabel{R1-58-1}\chg{or as in case of evolutionary rescue, constant between two deterioration events}; see the mean reproduction matrix in eq.~\eqref{eq:mean_repro}). Biologically, this means that the resident population is stationary and the mutant is either at low numbers or unaffected by its own density. Furthermore, for our approximation in eq. \eqref{eq:estab_approx} to generate accurate predictions, it is essential that growth rate differences between the wild type and the mutant are weak and dispersal is low. Formally, just two of these parameters need to be small -- see also the corresponding discussion in \citet{tomasini_2018}.

Since we summarize the population dynamics in our parameters of the reproduction matrix, the approach taken here is quite general and can account for various dispersal schemes and local \chg{type-dependent} population dynamics, \linelabel{AE-3-1}\chg{i.e. different reproduction and competitive parameters.} However, it cannot account for \chg{type-dependent carrying capacities, explicit spatial structure %as for example the stepping stone model. It also cannot account for 
or \st{time-inhomogeneous} \francois{rapidly changing} environments.} \linelabel{R1-58-2} The \chg{latter} is the reason for our \chg{poor}\footnote{\pete{any better word here?}} approximation in the context of evolutionary rescue (Fig.~\ref{fig:rescue})\footnote{\francois{I would remove this sentence}}. Additionally, in order to obtain analytical solutions, it is important that the stationary population sizes of the wild type have an accessible solution. This is not the case if we consider non-linear emigration rates that depend on habitat choice like those incorporated in some simulation studies \citep[e.g][]{holt_2015,mortier_2018}.

In contrast, it \chg{is} possible to \chg{implement} a cost of dispersal \chg{or} a different life cycle. In particular, the variation of the life cycle could yield \chg{distinct} results regarding adaptation \citep{holt_2015} and, more generally, in the context of the evolution of dispersal \citep{massol_2015}. \\

In conclusion, we studied the effect of dispersal and different dispersal schemes on the probability of establishment, adaptation and evolutionary rescue of a mutant under divergent selection in a subdivided population. Our quantitative approach disentangles the interaction of dispersal and adaptation. We recover previous results on adaptation and provide a general framework for studying evolutionary dynamics of a subdivided population in heterogeneous environments \chg{in discrete time}. This unifying approach allows us to identify the forces responsible for the different predictions obtained in the population genetics literature and under source-sink dynamics, respectively. We find that including population demography significantly alters the results for high dispersal rates. For constant population sizes, high dispersal rates have a negative effect on establishment, while with explicit demography the effect is largely positive. The latter is a result of relaxed competition in old-habitat patches. Most importantly, we extend the existing literature by comparing different dispersal schemes and studying their effects on adaptation and evolutionary rescue. Our results indicate that habitat choice does not necessarily result in an increased adaptive potential and might even hinder successful establishment of a mutant population that would avoid population extinction. \linelabel{R1-60} %Negative density-dependent dispersal always increases the probability of adaptation and evolutionary rescue. 
These results show that non-\chg{uniform} dispersal patterns can have a strong influence on population survival and adaptation in a heterogeneous environment.

\subsection*{Acknowledgements}
PC and FD received funding from the Agence Nationale de la Recherche (ANR-14-ACHN-0003 to FD). HU appreciates funding from the Max Planck Society. We are grateful to the INRA MIGALE Bioinformatics Facility (MIGALE, INRA, 2018. Migale Bioinformatics Facility, doi: 10.15454/1.5572390655343293E12) for providing computational resources. Additionally, we thank J\'{e}r\^{o}me Mathieu for highlighting the connection of the `New-Old dispersal scheme' to the ecological trap literature, and Staffan Jacob and Pim Edelaar for fruitful discussion concerning the biological motivation of the dispersal schemes.

\bibliographystyle{plainnat}
\bibliography{dispersal.bib}

\end{document}

### 2020-03-25 ms/revision.tex
first commit

\documentclass[11pt]{article}

% Other package to deal with references
\usepackage{nameref}

% FONTS
\usepackage[charter]{mathdesign}
\renewcommand{\sfdefault}{phv}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel} % Language

\usepackage{amsmath} % Math
\usepackage{graphicx}   % Figures
\usepackage{subcaption}	
\usepackage[final]{pdfpages} % To include pdfs
\usepackage{float}
\usepackage{paralist}
\usepackage{natbib}
\usepackage{array}
\usepackage{rotating}
\usepackage{lineno}   % For line numbering
\usepackage{setspace} % For line spacing
\usepackage{xcolor}

%\usepackage{fourier}
\usepackage{microtype}
% Geometry - margins etc.
\usepackage[hscale=0.725, vscale=0.75, includehead=false, includefoot=false, marginparsep=-1.25cm]{geometry}

\usepackage[textsize=footnotesize, backgroundcolor=orange!50, bordercolor=white]{todonotes}
\usepackage{soul} % Highlighting

% SPECIFIC COMMANDS FOR THE DOCUMENT -------------------------------
\newcommand{\articletitle}[0]{} % Title of the manuscript
\newcommand{\authorlist}[0]{} % List of authors
\newcommand{\msid}[0]{\# 59535 (previous ID)} % Manuscript ID
\newcommand{\thejournal}[0]{The American Naturalist}
\newcommand{\thejournalshort}[0]{AmNat}
%-------------------------------------------------------------------

%%% HELPER CODE FOR DEALING WITH EXTERNAL REFERENCES
%---------------
\usepackage{xr-hyper}
\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}

%%% END HELPER CODE ------------------------------------
\myexternaldocument{main} %% add link to the maintext file of your revised manuscript in the file list (new file, from another project).

%% COLORS
\definecolor{darkblue}{rgb}{0,0,0.6}
\definecolor{darkred}{rgb}{0.6,0,0}
\definecolor{lgray}{gray}{0.95}
\definecolor{llgray}{gray}{0.925}
\definecolor{dgray}{gray}{0.15}
\definecolor{darkgreen}{rgb}{0.0, 0.42, 0.24} % FD I added this because it was not defined but used for Hildegard's comments
\definecolor{purple}{rgb}{0.75,0.,0.75}
\definecolor{change}{rgb}{0.5,0.,0.6}

% HEADERS & FOOTERS
\usepackage{lastpage}
 \usepackage{fancyhdr}
 \setlength{\headheight}{16pt} % Change head height otherwise warning
 \pagestyle{fancy}
 \fancyhead{}
 \fancyfoot{}
 \fancyhead[L]{\textcolor{dgray}{\textsc{\small Reply to reviewers -- Revision of manuscript \msid{}
 } }}
 \fancyhead[R]{}
 \fancyfoot[C] {\thepage{}}

% REFEREES TEXT
\definecolor{colreferee}{RGB}{0,30,50}
\definecolor{colchange}{RGB}{150, 0, 200}
\newenvironment{referee}{\vspace{0.cm} \sffamily \color{colreferee} \begin{quotation} }{\end{quotation} \vspace{0.cm}}

% NUMBER THE COMMENTS
% and add label to be able to refer to a specific comment
\newcommand{\lmarginpar}[1]{\reversemarginpar\marginpar{\textcolor{colreferee}{\textbf{\textsf{[#1]}}}}}
\newcommand{\addnb}[1]{\refstepcounter{numcom}\lmarginpar{\arabic{numcom}}\label{#1}}
\newcommand{\comref}[1]{[\ref{#1}]}

% SECTIONING
\usepackage{titlesec}
\titleformat{\section}
{\normalfont\Large\bfseries}
{\thesection\hskip 9pt\textpipe\hskip 9pt}
{0pt}
{}

\newcommand{\refsubsec}[1]{\vspace{0.25cm}{\noindent \textsf{\normalsize{\textcolor{colreferee}{#1} }}} \vspace{0.0cm}}

\newcommand{\refsec}[1]{{\textcolor{colreferee}{\section*{#1}}}}

\newlength{\opari}
\setlength{\opari}{\parindent}
% \setlength{\parindent}{0pt}

% SPECIFIC COMMANDS FOR THE COMMENTS
\newcommand{\rt}[1]{\textcolor{change}{#1}}

\newcommand{\florence}[1]{\textcolor{purple}{(#1)}} 
\newcommand{\francois}[1]{\textcolor{blue}{(#1)}}
\newcommand{\hildegard}[1]{\textcolor{darkgreen}{(#1)}}
\newcommand{\pete}[1]{\textcolor{orange}{(#1)}}
\newcommand{\chg}[1]{\textcolor{change}{#1}}

\newcounter{numcom}
\setcounter{numcom}{0}

%\renewcommand\thefigure{S.\arabic{figure}}

% TEMPORARY STUFF
\usepackage{showlabels}

% OTHER TYPESETTING COMMANDS
\newcommand{\ie}{\textit{i.e.}}

% Figure numbering
\renewcommand\thefigure{R.\arabic{figure}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% BEGIN DOCUMENT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\pagenumbering{roman}

\noindent{\Large \articletitle{The effect of habitat choice on evolutionary rescue in subdivided populations}} \vspace{5pt}\\
\authorlist{Peter Czuppon, Fran\c{c}ois Blanquart, Hildegard Uecker, Florence D\'{e}barre} \vspace{5pt}\\
(\thejournal)\\
%
%\listoftodos

\florence{General comments: 1) Use topic sentences at the beginning of each paragraph; 2) do not just write "we updated the lines" but briefly say what you did, e.g. "we removed the term xx". Why? to make the life of the AE as easy as possible: s/he should be able to understand what you did without having to go back and forth between the ms and the reply. Ideally, the reply should be self-sufficient} \florence{Pete: look for \texttt{st} commands before resubmitting, to make sure there is none left!}

We are grateful for the very detailed and constructive feedback on our manuscript. Please find below point-by-point answers to the comments that we received, indicating how we updated our manuscript.
%

\refsec{Editor's recommendation (Jennifer Lau)}

\begin{referee}
Dear Authors, \addnb{editor} \vspace{-10pt}\\

One of our associate editors (Dr. David Vasseur), two expert reviewers, and I have now read your manuscript. At this time, the Editorial Board remains unsure, but optimistic, of the manuscript's suitability for the American Naturalist. However, we will require a major revision of your manuscript "The effect of habitat choice on evolutionary rescue in subdivided populations" before we can make a decision. Consequently, I have assigned a decision of Decline Without Prejudice. This decision means that if you feel you can successfully revise your manuscript based on the extensive suggestions provided by Dr. Vasseur and the two reviewers, then we would be willing to consider a revised manuscript. I emphasize, however, that such revisions would require substantial effort (see the long list of suggestions and comments below). That said, there is no deadline for resubmission, if you do choose to resubmit.

The reviews and accompanying Associate Editor Dr. David Vasseur's comments might be the most thorough I have ever seen. Each raises good points, which I think will increase both the accessibility of your manuscript to a broad audience and also allow readers to better identify the novelty of this work. If you choose to resubmit your manuscript to The American Naturalist, the revision should carefully address the concerns raised by the reviewers and Dr. Vasseur. Because they are considered new submissions, revision response letters are not typically included with resubmitted mansucripts receiving a decision of Decline Without Prejudice. However, in this case, it may be useful to include text describing your responses to our concerns in your cover letter. The AE and reviewers’ comments are at the end of this letter. I hope you find the comments to be constructive and that they help improve your paper. 
\end{referee}
%
Thank you for your positive assessment of our work and for giving us the opportunity to revise and resubmit our manuscript.

\begin{referee}
Because \addnb{editor2}the reviewers and Dr. Vasseur were so thorough in their comments, I will focus on the strengths of your manuscript and provide only a single minor comment for you to consider as you revise. As an empiricist, I was immediately impressed by the clear introduction to the model. It is accessible to even a plant ecologist like me. I particularly liked the first paragraph in the Results section that helped prepare me for the key results that followed. I think that this manuscript could ultimately be one that both more theoretical and more empirical readers of the American Naturalist would appreciate. I suspect that this would be even more likely if Dr. Vasseur's concerns about the complexity of the results and resulting manuscript length could be addressed. 
\end{referee}
%
We are very glad that our \st{intention of making}\florence{efforts to make}  the model and its results as accessible as possible, especially to empiricists, seems to have \st{worked out}\florence{been successful (work out is colloquial)}.

\begin{referee}
One minor concern for me was the figure 2 legend. I would have appreciated an additional sentence on the biological interpretation of the figures, as is so nicely done in the other figure legends. 
\end{referee}
%
In the caption of Fig.~\ref{fig:vary_m_est}, we have added a \florence{\st{very}} brief description about the emergence of the three possible regions of the establishment probability.
\florence{ALT: We added in the caption of Fig.~\ref{fig:vary_m_est} a brief interpretation of the results, describing the three possible regions of the establishment probability.}

\begin{referee}
As \addnb{editor3}you revise your manuscript, please be careful in your revision to add as little as possible to the length of the text. Our journal is under extreme competition for space among many excellent articles, so we are forced to consider the value of a paper relative to the number of pages it requires. In addition, as Dr. Vasseur mentions, the manuscript is quite lengthy so any ability to condense a bit here and there would be appreciated.

I want to be completely straightforward about this manuscript's prospects. The manuscript seems to have the potential to meet the American Naturalist's goals to publish papers that are of broad interest to the readership, to pose a new and significant problem or introduce a novel subject to the readership, to develop conceptual unification, and to change the way people think about the topic of the manuscript. Therefore, I want to encourage submission of a new manuscript that addresses the concerns outlined below. But, the review process has identified important weaknesses in the present submission. We would like to see these weaknesses fixed, at which point we will be in a better position to judge the paper’s suitability. We must emphasize that, even in revised form, the manuscript ultimately may not be accepted.

Thank you for submitting your work to The American Naturalist. I look forward to seeing your work in print whether you decide to embark on the necessary revisions for it to be considered further at The American Naturalist or whether you decide to resubmit to an alternate journal instead. 
\end{referee}
%
\st{Concerning the length of the manuscript, we have shortened it by being more concise where possible. Additionally, we deleted a section in the discussion (`Evolutionary consequences of habitat choice'). In response to the criticism of the referees we have added some more details (see our point by point answers below), still in total we have maintained (if not even shortened a bit) the length of the manuscript. }
\florence{ALT: The revised manuscript is slightly shorter than the previous version (WORD COUNT). Some details were added in response to the reviewers' comments (see our point by point answers below), but we compensated for these additions by being more concise elsewhere and by deleting the `Evolutionary consequences of habitat choice' section of the Discussion. }

We have also addressed all the modeling concerns that were raised, especially about the large fecundity differences. We now also allow for different carrying capacities in the two habitat types (old and new) and have renamed the different dispersal schemes, cf. our responses to comments~\comref{AE2}, \comref{AE3} and \comref{AE4} and references therein.

\newpage
\refsec{Associate Editor's recommendation (David Vasseur)}

\begin{referee}
I \addnb{AE}have had the opportunity to read the manuscript in detail and to consider the comments provided by the reviewers. This manuscript is rich in detail and provides a very thorough analysis of the different dispersal scenarios and relevant parameters for fixation of a mutant and probability of rescue.   The combination of analytical solutions and numerical simulation show a tight correspondence in most cases, and the authors do a good job noting when and why deviations between the two occur. The fact that the relationships between dispersal scenarios, migration rate and the relative fitness of the mutant/wildtype interact to generate curves of differing shapes and with differing rankings generates a complex landscape of results. This has the unfortunate consequence of making the paper hard to follow and rather lengthy.
\end{referee}
%
We thank the associate editor for his constructive feedback and positive assessment of our analysis. We agree that the interaction between dispersal rate, dispersal scheme, population dynamics and evolutionary processes are non-trivial. In order to make this connection as accessible as possible, we spend many lines explaining and disentangling these dynamics.

Throughout the revision of our manuscript we have paid close attention to simplify the heuristic intuition of all the results as much as possible (without losing accuracy) and adding as little as possible to the manuscript length. Despite adding details in response to the referees below, the length has not changed.\hildegard{Above, we say that the manuscript got slightly shorter.}

\begin{referee}
Part \addnb{AE2}of the challenge of reading this paper, which is shared by both reviewers, begins with the section on dispersal regimes. The authors title these regimes according to the relative and absolute fitness of the mutant and wild-type, but in some cases, particularly for “negative density dependent” dispersal, this language makes the concept tricky. For instance, density dependent dispersal, in other papers, refers to a probability of emigration that changes with density in the host patch. Here the density dependence is not a driver but instead a realization of the outcome of habitat choice. The other regions suffer less from an issue with terminology. As stated by one of the reviewers, rescaling the axis of the choice parameter would make for a clearer interpretation here and better show the symmetry of the different assumptions around the “random” choice.    
\end{referee}
%
We have renamed all dispersal schemes in order to avoid any ambiguity with the existing terminology. They now indicate the bias of the wild type and the mutant directly, e.g. `Old-Old' if both types have a bias towards old-habitat patches. \st{In the section introducing the dispersal schemes, we still provide the possible explanation of them being linked to habitat choice and density-dependent dispersal (when affecting the immigration rate).} \florence{[I think we can remove the sentence]}

Additionally, we also changed the parameter ranges and definitions of the dispersal schemes as suggested by the referees (comments~\comref{R1_23} and \comref{R2_4}). The parameter space now reflects the symmetry of our choice around the unbiased choice (compare Fig.~\ref{fig:disp_schemes}).

\begin{referee}
The \addnb{AE3}model steps are generally described very clearly and I have no issues with updating scheme described by the authors.   The one major issue I have is with the implementation of competition among individuals.   Within a patch reproduction of both wild-type and mutants occur according to their fitnesses, and the population, if greater than K, is downregulated to K.   This assumes first that both the wild type and mutant have the same K, and second, that individuals are competitively equivalent (I assume) such that downregulation maintains the relative frequencies of the two types generated in the reproduction step. Second, it is assumed that the degraded patches also have the same K as the “old” patches.    This is a fairly nuanced set of assumptions about competition that fits well to systems which may have a limited number of territories or nesting sites, and the numbers of these are not altered by mutation or by environmental degradation.   These assumptions however do not
fit well for resource-extraction driven competition; here the wild-type and mutant would quite likely have different carrying capacities and environmental degradation would potentially alter both r and K for these.   I understand the author’s reasoning for making these assumptions – given that they maintain the analytical tractability of the model; however it is worthwhile to identify in the MS how competition occurs in this model. 
\end{referee}
%
We are glad that our general description of the model is perceived as very clear. In order to address the inaccuracies regarding the competition we have added more details, lines~\ref{AE-3}ff.

\florence{Like the editor writes, we consider that $K$ is linked to a number of territories or nesting sites. This is a common viewpoint in evolutionary models (admittedly less so in ecological models); this is in particular the case in the models we based our work on (T\&P, Hildegard). 
Since it corresponds to a number of territories or nesting sites, $K$ does not depend on the genotype. It can however depend on the type of patch, which we did not consider in the previous version of the manuscript. [Then add paragraph on $Kold$ $Knew$}.

\st{It is true, as the associate editor states, that both types have the same carrying capacity. This is necessary for the analysis to work. Also, w}\florence{W}e do not assume differences in the competitiveness between the two types \florence{for the sake of simplicity,} even though this could be accounted for in our framework. In order to do so we would have to change the definition of $a_{\text{old}}$ (formerly $s_{\text{old}}$) in eq.~\eqref{eq:sold} to 
\begin{equation}
    1 + a_{\text{old}} = K_{\text{old}} \frac{\omega_m c_m}{\omega_w c_w \widetilde{N}_{\text{old}}^w}, 
\end{equation}
where $c_w$ and $c_m$ denote the (inverse of the) strength of competition experienced by the wild type and the mutant, respectively. In order to not add any further complexity to the model we did not pursue this extension but comment on it in the discussion (see line~\ref{AE-3-1}). \florence{We think that the editor will agree that our model is already complicated enough and the article long enough [to be reformulated but that's the idea]}.

\florence{[move up]}On the other hand, we changed the model and now allow for two different carrying capacities in the two patch types $K_{\text{old}}$ and $K_{\text{new}}$. As long as $K_{\text{new}}$ is not too small (in order to avoid gene swamping for high emigration rates $m$), our analysis extends to this situation without any further adjustment. In the simulations we have now chosen $K_{\text{old}}=1000$ and $K_{\text{new}}=500$.

\pete{Should we also say something regarding the biological interpretation? We are assuming a weak selection scenario, so $r$ and $K$ are not likely to change (much) between the two types anyway.}\florence{I don't have a strong opinion about this! maybe the reply to this point is long enough already?}
\hildegard{But $r$ changes a lot between the types, no? I think I do not understand the comment.}

\begin{referee}
As \addnb{AE3-b}a follow up to this: Line 226 – Down regulation using a hypergeometric distribution.   Unclear what is actually happening here.   Do wild type and mutants have the same chance of survival under this scheme?
\end{referee}
%
We have added more details on what we mean by the hypergeometric sampling scheme (drawing uniformly at random among the types without replacement, see line~\ref{R1-25}). \florence{You are not replying precisely to his question here (who is more likely to be chosen). Maybe explain fitness affects fecundity (number of offspring that are produced) but not survival to the culling step (who among the offspring make it to the next generation)? only if this is correct of course :)}

\begin{referee} 
I \addnb{AE4}share one of the concerns identified by the reviewer regarding the choice of parameters which contrast the “new” and “old” habitats.   In the model, the expected number of wild-type and mutant offspring is 10 and 9.9-9.0 in the “old” patches but this drops to <1 and slightly >1 in the “new” patches.   This represents a very large disparity, but the effect of this choice is not clear. I would assume that the large values in the “old” patches ensure that the patch always maintains a population at K, which assists the calculation of an analytical assumption. In the “new” patches, the population would take a very long time to reach K mutants, which could strongly alter the potential for fixation and for evolutionary rescue.   Furthermore, in the evolutionary rescue scenarios, “new” patches where the mutant is already or partially fixed would have little potential to assist patches that have more recently degraded since the number of offspring produced by mutants in “new”
patches is vastly less than the number of both mutant and wild-type offspring produced in “old” patches. Some more justification of this set of assumptions is necessary.    
\end{referee}
%
We have changed the growth rates in the old habitat to $\omega_w=1.5$ and $\omega_m=1.35$ in the strong fecundity difference scenario and $\omega_m=1.45$ in the weak fecundity difference scenario, respectively. The figures remain similar and the approximation and its validity are as good as for the larger number of offspring. (Initially, we had chosen the large number of offspring in order to reduce the computational time.) In contrast, we do not increase the fecundity of the mutant in new patches since otherwise evolutionary rescue would be almost certain once the mutant arises.

Concerning the contribution of mutants from new patches to the probability of evolutionary rescue, this was (and still is) addressed by Fig.~\ref{fig:sgv}(a). There, we plot the relative contribution to the probability of evolutionary rescue dependent on the origin of the successful mutant. It is visible that as the emigration probability increases the effect on evolutionary rescue from a mutant born in a new-habitat patch increases.
\pete{I am not sure whether this addresses the editor's last point...}\florence{I think he does not fully understand the meaning of evolutionary rescue. Maybe you need to explain what this means (e.g. not about fixation)}

\begin{referee}
With \addnb{AE5}regard to the description of the regions in figure 2 – the mechanism described for region ii seems somewhat tenuous. The reduction in establishment probability is credited to the effect of back migration of mutant types to the “old” habitat patches. This would only be a possibility if essentially all of the offspring of a mutant were to disperse back to “old” patches – but it is unclear if this is really such a likely outcome.   It might be useful to show what the probability of successive loss of the population is via this mechanism after a successful migration of a mutant individual in the previous step. I suspect this is true only for very low rates of reproduction – but not for greater rates which ties back into the choice of parameters discussed above. 
\end{referee}
%
The associate editor is correct in his assessment that basically all of the offspring generated (if there are any) are leaving the new habitat. \hildegard{But that does not mean that all of them are migrating back to the old environment. E.g. for $f_old=0.5$, only half would back-migrate.} The probability of successive loss of the population via offspring emigrating from new habitats is given by the probability $\varphi_{\text{new}}$ in Fig.~\ref{fig:vary_m_est}(b), where the probability of establishment might even be $0$ for too high emigration probabilities $m$.

\hildegard{Shall we give somewhere in the response the intuition that the local growth rate becomes something like (1+s)(1-m) + (1+s)*m/D?}

As we state in the manuscript, the local maximum for intermediate dispersal probabilities does not appear for all choices of fecundity values. In the manuscript we vary the fecundity of the mutant in the old habitat and by that move the `boundary' of relaxed competition to the left. Equally, if we were to increase the fecundity of the mutant in new habitats\florence{,} we would move the `boundary' of offspring back-migration (i.e. the parameter configuration where less offspring are produced in \hildegard{word missin, no?} than leave the new habitat on average) to the right, which would cause the local maximum to disappear as well.

\st{Our choice for this particular mutant fecundity is due to the stochastic nature of our model, i.e. i}\florence{I}f we increase the fecundity of the mutant too much, \st{the problem of} evolutionary rescue \st{does not arise anymore} \florence{is not a question anymore: it nearly always happen } since nearly every mutant would produce more than one offspring with high probability. Also, both in Uecker et al. (2014) and Tomasini \& Peischl (2019+) the same fecundity (0.02) is chosen. Therefore our choice also simplifies the comparison \st{between the} \hildegard{to} results obtained in \st{the different manuscripts}\florence{previous work on the topic}.

\begin{referee}
Both \addnb{AE6}reviewers have provided extensive line-by-line comments related to the presentation of the paper and terminology used in the model. In many cases these comments aim to simplify and streamline the presentation and I urge the authors to consider these in close detail.   
\end{referee}
%
We have addressed all the comments of the two reviewers. Our answers and changes are detailed below.

\newpage
\refsec{Reviewer \#1:}
\begin{referee}
\noindent Summary:\addnb{R1intro} \vspace{-10pt}\\

Here the authors explore how the pattern of dispersal affects the probability a mutation establishes and leads to evolutionary rescue in a metapopulation experiencing a uniform rate of patch "deterioration". They do so using multi-type branching process techniques, which appropriately weight the fitness of a mutant in the two patch types. The key result from this (and for the paper) is equation 5 (and its approximation in eq 6), which is then applied to the question of adaptation and rescue from de novo mutations in equations 8 and 10. These numerical solutions and analytical approximations are then compared to extensive simulation work. The study appears to mainly be an extension of Uecker et al (2014 Am Nat) to non-random dispersal. While this extension is an important one, some of the key takeaways are relatively obvious (at least a priori): the mutant is most likely to establish and rescue the population when it migrates to patches where it has higher fitness and when
the wildtype migrates to patches where it has lower fitness (relaxing competition for the mutant). However the importance of various pathways (e.g., reduced competition with the wildtype vs more direct impacts of migration on mutant fitness; figure S1), the highly non-monotonic effect of dispersal on establishment and rescue (e.g, figure 2), and the relative contributions of the two patches to successful mutants (figure 4) are more complex, and carefully explored here.

I think the analysis is well-done and correct and I do not have any major concerns. Below is a (relatively long) list of smaller changes I would like to see made and discussions I think could be expanded. Of these, my biggest concerns are that \\

1) the demographic equations used to derive the probability of rescue are given (see L470), \\

2) the assumptions of each dispersal scheme are clearly stated when mentioned in the Discussion as I find their names and sometimes their explanations unintuitive (see L164, L188, L600), and \\

3) an approximation for rescue by standing genetic variance be given (see L529), or an argument for why this is too difficult be stated in the text.
\end{referee}
%
We thank the reviewer for the very detailed comments and his/her positive assessment of our work. In the following we provide answers, justifications and comments to all of the raised points. Within these we have very thoroughly addressed his/her three biggest concerns from above, see the responses to remarks \comref{R1_41}; \comref{R1_22}, \comref{R1_23}, \comref{R1_55}; and \comref{R1_50}, respectively.

\hildegard{Can we give a summary of the most important changes that address his concerns here? Since these are his biggest concerns, this could be good (and then refer to the more detailed answers in responses to the specific comments).\florence{+1}}

\begin{referee}
\noindent Comments:\addnb{R1_1} \vspace{-10pt}\\

General: Evolutionary rescue is typically modeled as an abrupt or gradual environmental change. Here there is a possible bridge between the two: with $M=1$ we have an abrupt change but with large m (such that fitness is an average across patches) and large M (such that $f_{\text{old}}$ changes slowly) we essentially recover the well-mixed gradual change scenario. Worth exploring/discussing?
\end{referee}

We thank the reviewer for this observation\st{ which we were not aware of}\florence{[rm]}. We have added a \st{short}\florence{[rm]} paragraph explaining how the fragmented environment relates to the abrupt change and gradual change scenarios (lines~\ref{R1-1}ff.). Since the manuscript is already quite long (comments~\comref{editor3} and~\comref{AE}), we do not investigate this connection in more detail.

\begin{referee}
Abstract:\addnb{R1_2} I think it is important to tell the readers a bit more about the model. In particular, given we are dealing with dispersal, it seems essential to mention this is a 1-locus model of haploid selection, implying that dispersal has no direct impact on genotypes, greatly limiting maladaptive gene flow. 
\end{referee}

We have added a sentence in the Abstract explaining the coupling of a fixed dispersal strategy to a locus that determines the local fitness, cf. line~\ref{R1-2}. \florence{I do not understand the reviewer's comment on genotypes. do you?} \hildegard{Does he mean that locally adaptive allele combinations can be broken up? E.g., if AB is needed for rescue but there is constant immigration of wild-type ab individuals, this hampers establishment of the AB type. Likewise, in a quantitative genetics setting where the optimal trait is different in the two habitats, immigration and mating could always pull the trait distribution back.}\florence{but this explanation sounds more like considering two alleles, which you could still do with a haploid model}

\begin{referee}
L23:\addnb{R1_3} I find this definition of rescue to be too vague (what, for example, are "rescue types"?). Rescue is persistence with evolution when extinction would occur otherwise (not just decline and recovery). Further the U-shaped curve is only for abrupt environmental change, but rescue can occur in many other scenarios.
\end{referee}

We have changed the description of evolutionary rescue such that it sounds less restrictive, see lines~\ref{R1-3}ff. We still write that the `U-shape' is typical since abrupt environmental change is, at least theoretically, the most studied scenario \hildegard{not sure about this}\florence{[I don't know and let the expert (H) decide what is best]} and because evolutionary rescue was introduced as such in the paper by Gomulkiewicz and Holt in 1995.

\begin{referee}
L27:\addnb{R1_4} I would argue that rescue in the wild can at most be suggested/inferred (rather than observed) as we don't know if populations would have went extinct and it is difficult to say evolution prevented extinction in any case. But perhaps this is too pessimistic...
\end{referee}

We agree that in wild populations the process of evolutionary rescue can at best be inferred and have changed the wording accordingly (lines~\ref{R1-4}ff.).

\begin{referee}
L35:\addnb{R1_5} Osmond et al 2017 is about gradual environmental change. A more appropriate (abrupt) model is Yamamichi \& Miner 2015 Evol. Appl.
\end{referee}

\st{We are grateful for this new reference that we did not know. Indeed it }\florence{Thanks for this reference, which }fits better to our purpose. We have replaced our previous reference Osmond et al. 2017 by Yamamichi \& Miner 2015 (line~\ref{R1-5}).

\begin{referee}
L43:\addnb{R1_6} Drop "even"? This seems to assume readers will expected lots of dispersal to help rescue/adaptation, but I imagine many will already be thinking about swamping, etc.
\end{referee}

We followed the referee's suggestion (line~\ref{R1-6}).

\begin{referee}
L52:\addnb{R1_7} Is Ronce \& Kirkpatrick 2001 Evol. another case of "both"?
\end{referee}

\florence{Well, some of us are very well aware of the paper!! Can you please explain why it shows dispersal can have a positive effect? I do not like citing it because there are many issues with this paper [but do not write this in the reply of course!!]. }Yes, this is another reference for the different effects of dispersal on local adaptation \st{(which we were not aware of)}. We have added it to our list of examples in line~\ref{R1-7}.

\begin{referee}
L72:\addnb{R1_8} I find "keep in mind" to be potentially misleading, as it might suggest that you deal with it/control for it. Perhaps rephrase to say "although ... emigration .., we ignore this here."
\end{referee}

We followed this suggestion and rephrased the corresponding paragraph, cf. lines~\ref{R1-8}ff.\florence{Explain the rephrasing, e.g., we removed the term "Wright-Fisher"}

\begin{referee}
L80:\addnb{R1_9} I find the word "viable" to be confusing in this context as it is often implied as an antonym of lethal, but here is used to mean supercritical. I also find the phrase "assume a type is able to maintain a population at carrying capacity" (e.g. L95, L115) a bit clunky. Perhaps it is most clear to just speak of growth rates being positive in both instances.
\end{referee}

We changed all these formulations to the suggested wording, lines~\ref{R1-9-1}, \ref{R1-9-2} and \ref{R1-9-3}.

\begin{referee}
L95:\addnb{R1_10} Discrete non-overlapping generations does not imply Wright-Fisher. This is also confusing given your discussion of population dynamics above (while WF assumes constant N).
\end{referee}

Due to rephrasing of the whole section this terminology problem does not arise anymore, see lines~\ref{R1-10}ff.

\begin{referee}
Eq2:\addnb{R1_11} State this further assumes the wildtype is sufficiently common, $\omega_w N_w^{old} \gg  \omega_m$ (to simplify the denominator)
\end{referee}

We now say that the approximation holds whenever the mutant is much less abundant than the wild type (line~\ref{R1-11}ff).

\begin{referee}
General:\addnb{R1_12} All references to equations in the main text have the wrong equation number (all 1 too big)
\end{referee}

We are very sorry for this blunder and resolved it. \florence{blunder: a stupid or careless mistake. Maybe too strong a word!! We solved the issue, thanks for noticing [do you know how this happened?]}

\begin{referee}
L134:\addnb{R1_13} Add tilde to $N_w$
\end{referee}

We \st{corrected this }\florence{added the missing tilde} (line~\ref{R1-13}).

\begin{referee}
L135:\addnb{R1_14} I would have liked a comment such as "see SI S1" here, for the dynamics of $N_w^{\text{old}}$ 
\end{referee}

We followed the suggestion and added a phrase referring to the Supplementary Information (SI), see line~\ref{R1-14} and at other places.

\begin{referee}
L138:\addnb{R1_15} Add "per parent" after "offspring"
\end{referee}

We specified the description as suggested (line~\ref{R1-15}).

\begin{referee}
L143:\addnb{R1_16} I'm not sure $s_{\text{new}}$ is a great choice of variable here for two reasons: 1) the actual selective advantage of the mutant in the new environment is $s_{\text{new}} + r$ and 2) throughout the manuscript this subtlety is lost and $s_{\text{new}}$ is referred to as the advantage over the wildtype (e.g. L324 L356). This is important because $s_{\text{new}}$ is what matters (absolute fitness), not $s_{\text{new}} + r$ (relative fitness), for rescue in this density-independent patch type and this fact is lost in the manuscript.
\end{referee}

We have changed the notation from $s_{\text{old}}$ and $s_{\text{new}}$ to $a_{\text{old}}$ and $a_{\text{new}}$, respectively. The reason being that, as the reviewer points out correctly, we deal with two different selective pressures. In the old habitat the mutant fitness is measured relative to the wild type ($a_{\text{old}}$ depends on the reproductive success of the wild type) whereas in new-habitat patches all that matters in our analysis is absolute fitness of the mutant, i.e. its reproductive success independent of the wild type. Our original choice was aligned with the corresponding parameters in the analysis of Tomasini \& Peischl (2018,2019+) to facilitate the comparison of the results. But we agree that this might be misleading in terms of the interpretation (which is not done by Tomasini \& Peischl)\florence{[phrasing issue: it reads like you are saying T\&P do not interpret their results]}, explaining our change to the neutral letter `$a$' (as the first letter of the alphabet).

\begin{referee}
L145:\addnb{R1_17} You also need the wildtype to be rare to ignore (any form of negative) density-dependence. Rephrase to say you assume population size in new patches to be less than carrying capacity, so with simple density dependence you can treat mutant dynamics as density independent?
\end{referee}

Given that we now also vary the carrying capacity between the two habitat types\florence{,} we added a brief discussion also covering this comment, see lines~\ref{R1-17}ff.

\begin{referee}
L148:\addnb{R1_18} It seems like adding a dispersal cost would greatly generalize the results and also be relatively straight-forward?? Can you include some constant cost (say survive dispersal with probability $1-c$), or say why this is difficult to do?
\end{referee}

Indeed we could add a cost on dispersal (even type-dependent dispersal costs are possible, as are different emigration rates $m$). Given the already large complexity of the model (and the length of the manuscript) we refrained from adding another parameter. Still, we mention this possible extension in line~\ref{R1-18}ff. \florence{[Perfect reply!]}

\begin{referee}
L153:\addnb{R1_19} I would find it more intuitive if bias=0 implied random dispersal. Perhaps you could take bias $= \log(\pi)$, which would mean that no bias is bias=0 and allows bias to take on any real value (as it is pi is restricted to positive values, else eq3 can become negative, which should be stated if this change to a log scale isn't made). The log scale would also make the chosen parameter values (fig1) more evenly spaced (which is what you want since it is the proportional rates of migration into the two patches you care about, not their difference).
\end{referee}

This is a great idea. We followed the \st{suggestion of the reviewer}\florence{reviewer's suggestion} and now allow for the bias to take any real value. When working with it we transform it, as suggested, by the exponential to ensure that the enumerator $1-f_{\text{old}} + \pi_i f_{\text{old}}$ remains positive (see also our answer to the subsequent comment \comref{R1_20}). Consequently, we changed Figure 1 and the description of the dispersal mechanism, cf. lines~\ref{R1-19}ff.

\begin{referee}
Eq3:\addnb{R1_20} Why not just $m(1-f)\pi$ [or $m(1-f)*\exp(\text{bias})$ if my suggestion above is followed]? 
\end{referee}

Since we work with a time-discrete model\florence{,} the dispersal of individuals needs to be written as a probability -- they need to be somewhere at the end of the dispersal step. \florence{[consequence of this?]}
Along the same lines, another argument is that of mass-conservation. Since we consider dispersal to be a two-step process, the amount of individuals emigrating needs to equal the number of immigrating individuals. If we do not normalize the emigration step there would be individuals missing or added during this step. \florence{[so write what you did -- don't make the AE guess :)]}

\begin{referee}
L161:\addnb{R1_21} Given that you only model emigration, per capita dispersal rates into any patch do not depend on the starting patch. You can therefore drop four parameters and consider only $m_i^{j}$ with $i=\{m,w\}$ the genotype index and $j=\{\text{old,new}\}$ the receiving patch.
\end{referee}

We followed the suggestion and simplified the notation, cf. eq.~\eqref{eq:dispersal}.

\begin{referee}
L164:\addnb{R1_22} ABS assumes $1+s_{\text{new}} < \omega_m$, which should be very clearly stated. This is especially so since it is further assumed that typically $s_{\text{old}} < s_{\text{new}}$ (e.g. in the discussion on L361) -- which should also be very clearly stated -- such that "absolute habitat choice" assumes the mutant migrates to the patch where it has lower fitness.
\end{referee}

We have changed the description of this dispersal scheme and now emphasize that the number of offspring before density regulation is compared (lines~\ref{R1-22}). Throughout the manuscript, we also tried to make a better distinction between the different interpretations of $a_{\text{old}}$ and $a_{\text{new}}$ (formerly $s_{\text{old}}$ and $s_{\text{new}}$).

\begin{referee}
L188:\addnb{R1_23} I don't think "density-dependent" dispersal is a good name given that population densities change but per capita dispersal is assumed to be constant. It should at least be made very clear that the bias is held constant in any case.
\end{referee}

We have renamed all the dispersal schemes to avoid confusion with the general use of density-dependent dispersal. Yet, we still use it to motivate the different schemes by clearly stating that we consider density dependence to act on the immigration rate and that it is held constant, see also lines~\ref{R1-23}ff.

\begin{referee}
Fig1:\addnb{R1_24} It would be helpful to spell out the acronyms in the caption. 
\end{referee}

We have adjusted the description of Figure~1 to the new dispersal scheme names.

\begin{referee}
I'm also confused about the last sentence; is there some proof that there cannot be bifurcations in the system?
\end{referee}
%
\florence{[I split the point in 2]}
\st{As to the question of bifurcations, these can not} \florence{Bifurcations cannot} arise since the fixed point equation in Eq.~\eqref{eq:ext_prob} has at most one positive solution. For a bifurcation to arise\florence{,} the equation would need to have multiple fixed points dependent on the parameter choices\florence{ --} but this is not possible due to branching process theory, see also lines~\ref{R1-24}ff.

\begin{referee}
L226:\addnb{R1_25} Spell out what the hypergeometric sampling implies (e.g. sample $K$ without replacement).
\end{referee}
%
We have followed the suggestion of the referee (line~\ref{R1-25}).

\begin{referee}
L243:\addnb{R1_26} Thanks for sharing the code!
\end{referee}
%
\florence{Thanks!} \florence{[NB I avoid line breaks between ref comments and reply to avoid the first paragraph indent in the reply]}

\begin{referee}
L261:\addnb{R1_27} Maybe replace "evolve" with "act" or "behave" as individuals don't evolve. Or make it clear you are talking about a lineage rather than an individual.
\end{referee}

We replaced `evolve' by `behave' (line~\ref{R1-27}). \florence{[I'd rather talk about lineages and keep writing 'evolve' -- 'behave' sounds like we are studying behavior.]}

\begin{referee}
L329:\addnb{R1_28} Here and elsewhere I would find it helpful to reference terms in equations. For example, you could add "($\tilde{N}_w^{\text{old}}$, eqs S7 and S8)" at the end of this sentence. Another example would be to add (S10) at the end of L337.
\end{referee}

We have added more detailed references to the SI, especially when talking about the population dynamics throughout the manuscript. The sentence to which the referee is referring \st{has vanished.} \florence{does not appear in the revised version of the manuscript anymore.}

\begin{referee}
Fig2:\addnb{R1_29} Here and in the other figures it seems like putting the y-axis on a log scale would really help showcase the regimes (especially the local maximum).
\end{referee}
%
\florence{Using a log-log scale unfortunately has disadvantages: }\st{We refrained from changing onto a log-log-scale since} for the \st{F}\florence{f}igures where the establishment probability hits zero, the estimated lines disappear and re-appear for intermediate dispersal rates\florence{, as can be seen in Figure~\ref{fig:loglog}}\st{, see the Figure below}. This is, at least to us, more disturbing than the benefit that is generated from changing the scales.

\begin{figure}[h!]
    \centering
	\begin{subfigure}{.5\textwidth}
  		\centering
  		\includegraphics[width=\linewidth]{loglog_a.pdf}
  		\caption{$\varphi_{\text{old}}$ with $\omega_m = 1.35$ (large fecundity difference)}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
 		 \centering
 		 \includegraphics[width=\linewidth]{loglog_b.pdf}
  		\caption{$\varphi_{\text{new}}$ with $\omega_m = 1.35$}
	\end{subfigure}
    \caption{\textbf{Establishment probability on a log-log-scale.} This is Figure~\ref{fig:vary_m_est}(a,b) from the main text with the y-axis changed to a log-scale.}
    \label{fig:loglog}
\end{figure}

\begin{referee}
L330:\addnb{R1_30} Missing comma after $s_{\text{old}}$
\end{referee}
%
\st{When shortening the manuscript, this sentence was deleted.}\florence{We deleted the sentence as we shortened the manuscript. }

\begin{referee}
L372:\addnb{R1_31} Increases the population size relative to what? Pre-dispersal? Random migration? 
\end{referee}

In that paragraph we are discussing the differences between the New-New and New-Old dispersal schemes. Therefore, the increase in population size is relative to the New-New dispersal scheme. We clarified this (line~\ref{R1-31}).

\begin{referee}
L403:\addnb{R1_32} You can stay general and just give the $t_{\text{fin}}$ value used in the figure caption.
\end{referee}

We removed the corresponding sentence.

\begin{referee}
L421:\addnb{R1_33} You could mention this is because $P_{\text{adapt}}$ is essentially just $1-\exp(-C *\text{establishment})$, eq 8, with C a constant. 
\end{referee}

We followed the suggestion and added a sentence explaining this relationship (line~\ref{R1-33}).

\begin{referee}
Fig3:\addnb{R1_34} Please give equation numbers for predictions (goes for all figures).
\end{referee}

We have \st{now }added references to the equations which we are plotting in the corresponding labels of the figures. \florence{[we have added / there are now]}

\begin{referee}
L441:\addnb{R1_35} Mention "soft sweeps". It might also be interesting to plot the probability of a soft sweep.
\end{referee}

We have added a brief reference to soft selective sweeps when discussing the origin of the rescue mutation, see lines~\ref{R1-35} and~\ref{R1-35-2}. The probability for a soft selective sweep in our context would be given by 
\begin{equation}
    1-\mathbf{P}(\text{no rescue}) - \mathbf{P}(\text{rescue from exactly one lineage}),
\end{equation}
which can be computed by the help of eq.~\eqref{eq:evol_rescue} and the Poisson distribution. The first two terms are simply the probability for evolutionary rescue and the last one is given by $\lambda \mathbf{P}(-\lambda)$, where $\lambda$ is the sum within the exponential in eq.~\eqref{eq:evol_rescue}.

However, \florence{we do not provide this formula in the manuscript, because the manuscript is already long.}\st{ since we do not want to go into more details of soft selective sweeps (also due to the length of the manuscript), we do not provide this formula in the manuscript.}

\begin{referee}
Eq9:\addnb{R1_36} You might want to note this is an approximation that assumes any mutation that has already established (or is on its way to establishing) does not interfere with the establishment of future mutations. Essentially it assumes all mutations establish at the same time (while they are all rare). Interesting to look at where the first establishing mutant is likely to arise (ratio of the rates)?
\end{referee}

The reviewer is correct in emphasizing that eq.~\eqref{eq:origin} is an approximation. The independence and non-interference of different mutant individuals and lineages is the basic assumption of \st{our whole methodology which relies on }branching process theory\florence{, the framework used in our study}. To emphasize this point yet again we have added a sentence around the description of eq.~\eqref{eq:origin} (line~\ref{R1-36}).

Further, we agree that it is interesting to look at where the first establishing mutant is likely to arise which is given, as the reviewer points out, by the ratio of the dashed and the solid line in Figure~\ref{fig:origin}. However, since both values can take the value zero\florence{,} this ratio can not be computed over the entire x-axis\florence{,} which is why we did not plot the ratio. In fact, we had a similar discussion on this point \florence{among all authors} before our initial submission and \florence{had }decided to simply plot eq.~\eqref{eq:origin}. \florence{[not sure the In fact, ... sentence is necessary]}

\begin{referee}
L446:\addnb{R1_37} The surprise only arises when one tries to reason this out from the omega's rather than the s's, as it is relative fitness that matters most in the old environment and absolute fitness in the new environment. Framing this discussion around $s_{\text{old}}$ vs $s_{\text{new}}$ will be more intuitive (i.e. less surprising).
\end{referee}

\pete{hm... I am a bit confused. I could rephrase it in terms of $a_{\text{old}}$ but the surprise remains that even for large fecundity differences ($a_{\text{old}}\approx -0.1$ vs. $a_{\text{new}}=0.02$) a lot of successful mutant lineages emerge from the old habitat. Does anybody know what the referee means? I have deleted `surprisingly' (which also shortens the manuscript ;))} \florence{Then you can 1) remove "surprisingly", 2) explain here in the reply that things are still complicated/not straightforward when considering omega; 3) keep the text as it is (modulo "surprising")}

\begin{referee}
Fig4:\addnb{R1_38} I'd find it more interesting to sub out panel b for FigS3, which shows an example where the new patch contributes more.
\end{referee}
%
\st{Due to our change of fecundity values in the old habitat we do not observe the mentioned flip anymore.} \florence{The flip [or other word, I am not sure to understand what you mean by flip] does not happen anymore in FigXX with the updated fecundity values (changed following comment RefToCommentOnFecundities}. 
Therefore, we did not change the corresponding figure.

\begin{referee}
Eq10:\addnb{R1_39} I was confused for a while here because it is not made clear that $\varphi$ is now a function of $f_{\text{old}}$
\end{referee}

We have added a comment that we now consider the establishment probability to be a function of the frequency of old-habitat patches, see line~\ref{R1-39}.

\begin{referee}
Eq10:\addnb{R1_40} I think the index inside $f_{\text{old}}$ should be $i+1$
\end{referee}

We adapted the definition of $f_{\text{old}}(i)$ such that now the correct old-habitat patch frequency is used (line~\ref{R1-40})\florence{[explain so that the AE does not need to check the text]}. The implementation of the formula in Figure~\ref{fig:rescue} was not affected by this mistake\florence{[typo or mistake?]}.

\begin{referee}
L470:\addnb{R1_41} Where is this deterministic population size calculated? Given this is an essential component of the probability of evolutionary rescue this should be given explicitly in the main text.
\end{referee}

\st{We are sorry that we did not provide the necessary equations in our initial submission. }We have now added a section in the SI, `Wild-type population sizes during the environmental change' (lines SXXff.) where we \st{state}\florence{provide} the equation for the deterministic population dynamics of the wild type in a changing environment\florence{ (sorry it was not included in the initial submission)}.

\begin{referee}
L473:\addnb{R1_42} What does "intersection" refer to? The y-axis, or each line with another?
\end{referee}

Intersection referred to the crossings of the different dispersal scheme lines. We now avoid using this term and rather speak of changes in the ranking of dispersal schemes, see also line~\ref{R1-42}.

\begin{referee}
L481:\addnb{R1_43} Can you state why you can't apply the Uecker result here? Can you apply the Uecker result to your RAND scheme? If so it might be nice to plot this in Fig5.
\end{referee}
%
Uecker et al. (2014) studied specific scenarios where a solution of the probability of evolutionary rescue was accessible, namely either $m=1$ or $\omega_m=0$. In these cases the branching process becomes one dimensional and the probability of establishment can be solved explicitly even in a time-varying environment. \florence{In our case, .... [explain why this is not the case for us]. }We now comment on this and give a bit more details, cf. line~\ref{R1-43}.

\begin{referee}
L497:\addnb{R1_44} Here viable is used in the more traditional sense, with non-viable meaning lethal (cf L80)
\end{referee}

We have changed the previous use of viable (see also our answer to comment~\comref{R1_9}) such that it \st{should be used coherently now} \florence{is now used consistently} throughout the manuscript.

\begin{referee}
L521:\addnb{R1_45} "were" ->  "where"
\end{referee}

We corrected the typo (line~\ref{R1-45}).

\begin{referee}
L525:\addnb{R1_46} Add a note about what changing tau does.
\end{referee}

We added a brief discussion about the effects of varying $\tau$, see lines~\ref{R1-46}ff.

\begin{referee}
Fig6a:\addnb{R1_47} As with Fig4, it would be more exciting to show a flip in the rank order of contributions (e.g. compare $\omega=9.9$ with $\omega=5$)
\end{referee}

Similar to Fig.~\ref{fig:origin}\florence{,} we do not longer observe a flip in the rank order of contributions \florence{because of}\st{due to }our new choice of fecundity values \florence{(a change suggested in comment REF [add ref to reviewer/AE comment])}.

\begin{referee}
Fig6b:\addnb{R1_48} If you plot this as a function of tau do you see a flip, i.e. at low tau SGV is more important than de novo while at high tau the opposite is true? 
\end{referee}

We have changed the time-intervals between consecutive deterioration events in Fig.~\ref{fig:sgv}(b). For $\tau=1$\florence{,} we see a flip of the contribution of de novo and standard genetic variation. Fort $\tau=5$ (results not shown)\florence{[show them here in the reply then - as a reviewer/AE I fight against "results not shown"...]} the effect of de novo mutants is again larger than that of standard genetic variation.

\begin{referee}
L528:\addnb{R1_49} I think it's worth mentioning why: mutant spends most of the time in old patches.
\end{referee}

We restructured the sentence and now include the suggested wording (line~\ref{R1-49}). \florence{[are you sure about this? how is it compatible with high migration rates?]}

\begin{referee}
L529:\addnb{R1_50} It seems like the number before deterioration is a key aspect here, and this is not manipulated in Fig6b. Perhaps if theta is larger there will be enough mutants in the SGV that some establish, increasing the importance of SGV? This makes me wonder if you have tried to calculate the probability of rescue from SGV. For instance, you might just assume mutation-selection balance prior to the first deterioration event (giving N mutants across all M old patches) and then use $1 - \exp(-\varphi_{\text{old}}(f_{\text{old}}(1))*N)$ as the probability of rescue, where $\varphi_{\text{old}}(f_{\text{old}}(1))$ is the probability of establishment when there is just 1 new patch.
\end{referee}

\st{We had the same idea as the referee. However, }\florence{Unfortunately, }this formula does not capture the probability for rescue due to standing genetic variation since the probability of adaptation for a mutant individual initially in an old-habitat patch is equal to zero (or at least very close to it) over the whole parameter range of the emigration probability $m$ (see the new Fig.~SXX in the SI). Additionally, in the scenario where standing genetic variation is most likely to affect the probability of evolutionary rescue, i.e. for a fast environmental change (small $\tau$), the suggested approximation does not take into account the environmental change which is crucial for an evolutionary rescue event to happen from standing genetic variation. \florence{[that's additional work but maybe you could plot the suggested formula, and put the figure here to illustrate that it does not work?]}

\begin{referee}
L540-547:\addnb{R1_51} I'm a bit confused here. You state that population size in each patch is constant regardless of dispersal (L541) but then say that more dispersal leads to more non-adapted individuals in the new environment (L546). 
\end{referee}

The population sizes after the whole life cycle remain constant at the respective carrying capacity.\florence{[specify when/under which conditions, otherwise they may this is all the time]} After dispersal the number of individuals may be above the carrying capacity but after reproduction and density regulation the population is again at the initial value. In our scenario the population sizes may be below the carrying capacity even after reproduction and density regulation\florence{[hmm need to reformulate the reply because this seems to contradict the first sentence]}. We added a brief comment in line~\ref{R1-51}.

\begin{referee}
L563:\addnb{R1_52} Replace "Dispersal" with "Standing genetic variance"
\end{referee}

We changed the title of that section following the suggestion (line~\ref{R1-52}).

\begin{referee}
L589-591:\addnb{R1_53} Needs elaboration or drop.
\end{referee}

\pete{I am happy to drop these three lines. The part in the SI (Section S6) can stay without further explanation I think. }\florence{[no strong opinion on this - maybe drop to shorten?]}

\begin{referee}
L596:\addnb{R1_54} This also seems important once more loci are considered, both for the case of neutral and deleterious hitchhikers (i.e. does the rescue mutation arise on an "old" or "new" haplotype) as well as in sexual populations where mutations at multiple loci are required for rescue (in which case depending on the "old" and "new" haplotypes recombination could help or hinder rescue). Is it worth adding a paragraph about sex/recombination/polygenic rescue here? Also makes sense after L640.
\end{referee}

We have added a brief reference to this area of research but do not want to go into the details of the rescue process in these situations (line~\ref{R1-54}). We think that more explanations would be needed and decided to not extend the manuscript length any further (see also the general comments of the two editors).
\pete{I think if we make more than a reference we need to go deeper into explanations of how rescue works in these situations which I would like to avoid. How do you feel about that? Maybe add another reference -- Hildegard, do you know something off the top of your head?}

\hildegard{Have a look at the abstracts. Not sure they are suitable to cite in the context.}
\begin{itemize}
    \item Schiffers, K., E. C. Bourne, S. Lavergne, W. Thuiller, and J. M. J. Travis, 2013 Limited evolutionary rescue of locally adapted
populations facing climate change. Philos. Trans. R. Soc. Lond.
B Biol. Sci. 368(1610): 20120083.
    \item Bourne, E. C., G. Bocedi, J. M. J. Travis, R. J. Pakeman, R. W.
Brooker et al., 2014 Between migration load and evolutionary
rescue: dispersal, adaptation and the response of spatially structured populations to environmental change. Proc. Biol. Sci. 281
(1778): 20132795.
    \item Pease, C. M., R. Lande, and J. J. Bull.1989. A model of populationgrowth, dispersal, and evolution in a changing environment. Ecology 70:1657--1664.
    \item Polechov{\'a}, J., N.H. Barton, and G. Marion. 2009. Species'
    range: adaptation in space and time. American Naturalist 174:E186--E204.
    \item Duputi{\'e}, A., F. Massol, I. Chuine, M. Kirkpatrick, and O.  Ronce.2012. How do genetic correlations affect species range shifts in achanging environment. Ecology Letters 15:251--259.
\end{itemize}

\begin{referee}
L600-604:\addnb{R1_55} As stated in my comment about L164, the assumptions of ABS should be laid out very clearly: the mutant migrates to habitat patches where it has lower fitness (except if m is really large). This is lost here, where the authors state "habitat choice protects a population against locally deleterious mutations" -- which is only true when the mutations migrate to patches where they have lower fitness. This clear explanation should be stated. \\

L611:\addnb{R1_56} If the trait is not which type of patch to enter (here acting like ABS) but to enter the patch one is best at (here REL) this trait does not have to evolve, but perhaps this "all-seeing" trait is unlikely to exist...\\

L613:\addnb{R1_57} Can you say a few more words about the ciliates? e.g., what were the habitats, how fast did preference evolve?
\end{referee}

In our attempt to shorten the manuscript we deleted the section ``Evolutionary consequences of habitat choice'', which the\florence{se} \st{previous }comments refer to.

\begin{referee}
L622 and L632:\addnb{R1_58} I'm a bit confused here as $s_{\text{old}}$ is not constant in time. Perhaps you mean constant between deterioration events? But even then you say population dynamics are changing (L470). 
\end{referee}

Indeed, as the referee mentions, the population size changes\florence{[need to discuss this as I am not sure to understand: doesn't it change all the time anyway?]} when we consider the evolutionary rescue scenario which in turn alters the parameter $a_{\text{old}}$\florence{(previously written  $s_{\text{old}}$)}. However, \florence{we approximate xxx as constant between two deterioration events. }in the analysis we assume $a_{\text{old}}$ to be constant between two deterioration events (which is the major reason for the \st{disagreement}\florence{difference/discrepancy/other term?} of our approximation and the simulation results, see line~\ref{R1-58-2}). We now specify what we mean by constant parameters by highlighting that this does not necessarily hold in our rescue situation (line~\ref{R1-58-1}).

\begin{referee}
L646:\addnb{R1_59} Is this a general framework or an application of multi-type branching processes?
\end{referee}

\pete{I would argue that our framework is very general when it comes to modeling populations in discrete time since we can vary basically all parameters and add/remove steps in the life-cycle arbitrarily. I agree that the analysis is an application of multi-type branching processes but the model itself is very general.}\florence{[fine with me]}

\begin{referee}
L656:\addnb{R1_60} Better to give example of dispersal scheme that hinders rescue first (to follow up on your previous sentence). I think it is also important, throughout the Discussion, to spell out what the schemes are, i.e. to say that "negative density-dependent" dispersal means the mutants migrate to patches where they have higher fitness and the wildtypes do the opposite.
\end{referee}

We deleted the sentence since we believe that throughout the manuscript we have made our point clear that non-uniform dispersal can alter the evolutionary outcome.

\begin{referee}
EqS7\addnb{R1_61} (and many equations that follow): $f$ should be $f_{\text{old}}$\\

LS35:\addnb{R1_62} $\pi$ should be $\pi_w$
\end{referee}

\st{We are very sorry for this shortcoming.}\florence{Thanks for noticing.} We corrected both typos in all the formulas (and also changed $\pi_w$ to the newly defined $\widehat{\pi}_w$).

\begin{referee}
LS38:\addnb{R1_63} should be equation 2, not 3 (all equation references to the main text are wrong in the SI too)
\end{referee}

Unfortunately, this was due to the same mistake\florence{[compilation error on the website?]} that led to the wrong numbering in the main text, see also comment~\comref{R1_12}. This problem is resolved now.

\begin{referee}
LS44:\addnb{R1_64} It is not that you "do not assume density regulation" but that you assume no density-dependence
\end{referee}

We rephrased the sentence (line~SXX).

\begin{referee}
LS53:\addnb{R1_65} Add comma after $s_{\text{old}}$
\end{referee}

Corrected.\florence{[this review would be so much more pleasant to read if they had written some "please" instead of given orders like this!]}

\begin{referee}
eqS17:\addnb{R1_66} Where do you prove that the branching process is only "slightly" supercritical? Is this implied by zeta and mu being small in this equation?
\end{referee}

The process is slightly super-critical because for $\varepsilon\to 0$ the eigenvalue converges to $1$ (cf. eq.~(SXX)). This translates to the rates $a_{\text{old}},a_{\text{new}}$ and $m$ being very small which is modeled by their scaling with $\varepsilon$. We added a bit more details in the corresponding section to highlight this (lines~SXXff.).

\begin{referee}
LS91:\addnb{R1_67} "We now proceed to explain"
\end{referee}

We rephrased the sentence following the suggestion (line~SXX).

\begin{referee}
LS103:\addnb{R1_68} Again, relative fitness in the new patch is irrelevant, it is absolute fitness that matters
\end{referee}

\st{This is entirely right. W}\florence{We agree and w}e adjusted the phrasing accordingly, see line~SXX.\florence{[do we? I cannot check because I do not have the SI with line numbers...]}

\begin{referee}
FigS1 and EqS20 and L301: \addnb{R1_69} Worth mentioning that $\varphi_{\text{old}}$ is zero when $m=0$ (in contrast to $\varphi_{\text{new}}=2s)$? When I first saw FigS1c I thought term 1 and 2 must be driving the patterns because they are so much larger in absolute value than term 3, but then I saw from EqS20 that they must exactly cancel at small enough m.    
\end{referee}

We have added a brief sentence in the main text (line~\ref{R1-69}) and also in the SI (lines~SXX) to highlight that terms (1) and (2) cancel for $m=0$.

\begin{referee}
FigS1cd:\addnb{R1_70} Would it be clearer to plot REL rather than RAND if you are interested in regime ii?
\end{referee}

We have followed the suggestion. Unfortunately, still the local maximum is not well visible in these figures due to the scaling of the y-axis which is necessary to capture terms (1) and (2).\florence{[so does it make sense to follow the suggestion?]}

\begin{referee}
LS153:\addnb{R1_71} So you are now assuming hypergeometric sampling rather than independent Poisson processes, right?
\end{referee}

Since we have changed the dynamics in the old habitat to follow the same rules as in the new habitat (fecundity modeled by a Poisson distribution and then down-regulation by hypergeometric sampling) this also applies to the scenario without demography.\florence{[I do not understand the sentence - think about the AE, put the reply back into context so that the AE does not have to check the paper to understand what this is about. It is longer to write but pays off because a happy AE is more likely to accept your paper!]} (For our chosen population sizes $K_{\text{old}}=K_{\text{new}}=500$ and fecundities $\omega_k^i\approx 1.5$ the carrying capacity is always reached with high probability.)

\begin{referee}
EqS21:\addnb{R1_72} These could just be given in the figure caption, no? Or does this impact eq S22 and S23?
\end{referee}

As suggested, we have moved the fecundity values to the caption of Figure~SXX.

\begin{referee}
L163:\addnb{R1_73} This isn't really highlighted in the main text, instead L561 just refers readers back to this SI. Perhaps more detail should be moved to the Discussion?
\end{referee}

We changed the word ``highlighted'' in the SI (line~SXX). We refrained from adding more details about the reason for the different outcomes in region (iii) between models with and without demography in the main text because \st{we do not visualize them there} \florence{this is not where they are seen (they are in figXXX)}. Instead we leave the detailed discussion, as it was before, in the SI.

\newpage
\refsec{Reviewer \#2:}

\begin{referee}
This\addnb{Ref2Intro} is an interesting paper on effects of different kinds of dispersal on evolutionary rescue. The habitat consists of a number of patches which start out with a favorable environment but each in turn deteriorates. The organism has two types (wild type favored in the original environment, mutant type favored in the degraded environment) that reproduce clonally. Each type can have unbiased dispersal or dispersal biased toward one habitat type (the one they have greater absolute or greater relative fitness). Branching processes are used for approximate analytical results and complemented by simulations. Dispersal magnitude and type affects the probability that a mutant arrives in the degraded habitat (where a lineage must persist for rescue) and the distribution of the wild type, which affects the fitness of the mutant in patches that have not yet degraded. These both contribute to rescue, of course.

Previous evolutionary rescue papers often assume either a single abrupt change in the environment or a continual linear change (sometime with random variation around one of these). This paper uses a step-wise change that has features of each (spread out over time like the continual change, but limited in total magnitude of change like the abrupt change). In deriving the evolutionary rescue results, the paper also uses analyses of source-sink systems, since between the starting and ending states there are both good patches (sources for the wild type) and bad patches (sinks). The authors reference species in natural systems that match many of their assumptions.\\

\vspace{-10pt} Specific comments (\_ indicates that what follows is a subscript, and \^{} a superscript). \\

line 4.\addnb{R2_1} Change 'wild-type population' to 'population whose wild type is'.
\end{referee}

We changed the wording as suggested (line~\ref{R2-1}).

\begin{referee}
line 45\addnb{R2_2} (and other places). Change 'the other' to 'another' ('the other' suggests that there are only 2).
\end{referee}

We followed the suggestion and changed all these occurrences of `the other'.

\begin{referee}
line 49.\addnb{R2_3} Add 'increasing' before 'dispersal' if this is what you mean, since the current wording suggests that the comparison is between systems with and without dispersal.
\end{referee}

We clarified the sentence by adding `increasing' (line~\ref{R2-3}).

\begin{referee}
line 54.\addnb{R2_4} Using 'random' to mean 'with a uniform distribution' is common but I would use random to mean the opposite of deterministic, for example constant, dispersal. Later, the paper mentions biased dispersal, so I think it would be better to use 'unbiased' instead of 'random'.
\end{referee}

When changing the range of the dispersal bias parameters $\pi_w$ and $\pi_m$, we also changed the names of the dispersal schemes (compare also our response to \comref{AE2}). In \st{alignment}\florence{line} with this, we also replaced all the occurrences of `random' by either `uniform' or `unbiased'. \florence{Thanks for the suggestion.}

\begin{referee}
line 74.\addnb{R2_5} I would change 'for an individual to leave' to 'that an individual leaves'. I think this same construction occurs elsewhere (sometimes with probability instead of likelihood), and I would change those also.
\end{referee}

\st{We are very grateful for the plenty stylistic comments made by the reviewer. If not mentioned differently, w}\florence{W}e followed the \florence{reviewer's}suggestions \st{of the him/her in all cases}\florence{and thank her/him for them. [NB: `the rule, at least in French, is to list options alphabetically, so her/him]} (comments \comref{R2_17}, \comref{R2_23}, \comref{R2_24}, \comref{R2_36}, \comref{R2_38} and \comref{R2_41}).

\begin{referee}
lines 95-96.\addnb{R2_6} I would write that in old-habitat patches both types have a mean reproduction >  1 with the wild type being higher, and there is ceiling density regulation (only K individuals reproduce in each patch each generation). 
\end{referee}

We have changed the initial description of the model in the spirit of the suggestion, lines~\ref{R2-6}ff.

\begin{referee}
lines 108-109.\addnb{R2_7} Change 'amount of offspring that is' to 'number of offspring that are'.\\

line 121.\addnb{R2_8} 
Add 'so the number of mutants is determined using' before 'a binomial'.\\

lines 123-124.\addnb{R2_9} Change '(wild-type individuals turning into mutants)' to '(probability that an offspring of a wild-type individual is a mutant)'.
\end{referee}

\florence{[maybe use one number for the three lines then (i.e. just R2-7, and jump to R2-10 later), otherwise it first looks like you have ignored the comments]}
Since we have changed the fecundity values in old-habitat patches to be a lot smaller than before (compare \comref{AE4} and \comref{R2_22}) we now model the dynamics in these patches differently so that the phrasing has changed and these comments do not apply anymore, see lines~\ref{R2-7}, \ref{R2-8} and \ref{R2-9}. Briefly, we now model the population in old habitats in the same way as in new habitats, i.e. we first assume reproduction by drawing a Poisson number of offspring for each individual and then we regulate the population size (if necessary) by sampling hypergeometrically, that is, without replacement.\florence{[Maybe make this really clear above as well!]}

\begin{referee}
equation (2).\addnb{R2_10} Using equation (1), the denominator would have '$+ \omega_m$' at the end. If this is assumed to be negligible, that should be stated. This also applies to equation (S22).
\end{referee}

We have added an explanatory sentence at both approximations (lines~\ref{R1-11} and SXX\florence{; yes it is assumed to be negligible [if true]}).

\begin{referee}
line 131.\addnb{R2_11} 'eq. (3)' here should be 'equation (2)'. From this point on, very many (perhaps all) of the equation numbers are 1 too high. So these all need to be corrected.
\end{referee}

We are very sorry for this \st{blunder}\florence{compilation issue} and resolved it.

\begin{referee}
line 134.\addnb{R2_12} It is not clear what $N_w^{\text{old}}$ refers to here. Is it the total over all old habitats, and at what point in the life cycle?
\end{referee}

There was a tilde missing above the variable (see also \comref{R1_13}). We now specified the point in the life cycle and that it is the average wild-type population size in one single patch (line~\ref{R2-12}).

\begin{referee}
lines 145-146.\addnb{R2_13} Some of the equations (such as [S5]) can have new patches at K.
\end{referee}

This is true. However, to simplify the analysis we assume density-independent reproduction in new-habitat patches\florence{,} which is reasonable as long as the carrying capacity is not reached. In the simulations\florence{,} the carrying capacity might be reached\florence{,} in which case our approximation fails (this partly explains the worse fit for large frequencies of old-habitat patches, $f_{\text{old}}$). We now provide more details explaining this point, cf. lines~\ref{R2-13}ff.

\begin{referee}
lines 151ff.\addnb{R2_14} Delete 'The bias of immigration into new-habitat patches is set equal to one, without loss of generality.' I think $\pi_i$ is the ratio of the probability of entering a specific old patch to that of entering a specific new patch for type i. If so, it should be defined in that way, since 'bias in immigration' is less clear.
\end{referee}

We have changed the meaning of the parameter $\pi_i$ (see also comments \comref{AE2} and \comref{R1_19}). Now, $\pi_i$ is indeed a bias, and the value in the formulas is given by $\widehat{\pi}_i=e^{\pi_i}$. As such, we continue describing $\pi_i$ as a bias. Still, we rephrased parts of the corresponding paragraph to be more clear about the meaning of $\pi_i$ and $\widehat{\pi}_i$ (lines~\ref{R2-14}ff.).

\begin{referee}
line 161.\addnb{R2_15} The equation for this is (S2), and it is not that clear how this is analogous to (3), so I think the equation should be given (or refer to the supplement if it is not important for the main text).
\end{referee}

We have simplified the dispersal notation since emigration is independent of the type of the individual and also the habitat-type (cf. comment~\comref{R1_21}). Now both dispersal rates, into new and into old habitats are given in the main text, see eq.~\eqref{eq:dispersal}.

\begin{referee}
lines 164-166.\addnb{R2_16} Obviously, it is being assumed that $\omega_m >  1 + s_{\text{new}}$, but that should be stated explicitly. 
\end{referee}

We followed the suggestion and now explicitly state the assumption underlying this dispersal scheme (line~\ref{R2-16}).

\begin{referee}
line 211.\addnb{R2_17} Change 'been using' to 'used'.\\
\end{referee}
\florence{done [otherwise it looks like you have omitted the comments; they won't remember that you listed the number above]}

\begin{referee}
lines 216-219. \addnb{R2_18} This needs to be explained in more detail. There are 10 patches and 2 types, so I presume 20 random numbers of dispersers were picked, one for each type in each patch, and the population size of each type in each patch decremented by that number. For each type, the dispersers from all patches can be summed, and then the number that go to each patch type can be determined using a binomial with p value determined by the $\pi_i$ value for that type and $f_{\text{old}}$ (the fractions in eqq. [S1] and [S2]). Then, these dispersers can each be added to a patch of the appropriate type (I assume randomly with equal probabilities). I would also cut 'Then reproduction happens locally.', because in old-habitat patches, reproduction, mutation and density regulation are done together. 
\end{referee}

We have added more details when describing the dispersal step in the simulations, see lines~\ref{R2-18}ff. We also followed the suggestion of cutting `Then reproduction happens locally' since the next sentence now begins with `In each patch...'.

\begin{referee}
lines 221-224.\addnb{R2_19} The Poisson distributed number for each type is the product of its mean offspring number and the number of adults of that type. I would also start with 'In each new-habitat patch' and change 'both types' to 'each type' [so there are $2(1-f_{\text{old}})$ Poissons if each new-habitat patch has both types]. 
\end{referee}

We have rephrased the corresponding sentence to clarify the mean of the Poisson numbers that we draw randomly (lines~\ref{R2-19}ff.). It depends on the type of the individual giving birth, the number of those individuals in the focal patch and the patch type.

\begin{referee}
line 224.\addnb{R2_20} Change 'individuals' to 'offspring' and 'rate' to 'probability'.
\end{referee}

We followed the suggestion (line~\ref{R2-20}).

\begin{referee}
lines 232.\addnb{R2_21} Add 'over a finite time period' after 'mutations'. Also, since this paragraph is stated to be about figures 2-4, it would be better to write 'figures 3-4' instead of 'all the other figures'.
\end{referee}

We followed both suggestions and rephrased the sentence accordingly, see line~\ref{R2-21}.

\begin{referee}
Table 1.\addnb{R2_22} It appears that $s_{\text{new}}$ was chosen so that the approximation in (6) is valid, and the values of $\omega_m$ so the mutants in the source to be determined by a binomial using (1). It is a little odd that the mutant goes from a birth rate of at least 9 in the old habitat to 1.02 in the new habitat. The results obviously would be vastly different if the latter were a little lower. With a birth rate of 1.02, each mutant lineage in the new habitat has a high probability of going extinct, and the result could be significantly different if the new habitat birth rate were not so close to 1.
\end{referee}

We now reduced the average number of offspring in old-habitat patches to $\omega_w=1.5$ and $\omega_m=1.35$ or $\omega_m=1.45$ dependent on the studied fecundity difference to the wild type (see also comment~\comref{AE4}). For these values our approximation is still applicable (preliminary simulations suggest that $\omega_m=1.1$ still gives reasonable results). \st{Concerning the choice of the local growth rate of the mutant in new-habitat patches w}\florence{W}e followed the previous choices made in Uecker et al. (2014) and more recently in Tomasini \& Peischl (2019+)\florence{ for the [value of the?] local growth rate of the mutant in new-habitat patches [try to avoid ``Concerning..., ...'']}. The particular choice of $1.02$ is interesting since for this choice of parameters we observe a local maximum in the establishment probability when varying the emigration probability and the mutant fecundity in the old habitat (Figure~2). \hildegard{The last sentence sounds as if we tuned the parameters and it is really important to have 1.02 to obtain the results. I think it is mainly important that it is small enough such that rescue is not guaranteed. Which pattern we observe also depends on the mutant fitness in the old habitat. If this is low, the fitness in the new patches can be larger, and we still see a maximum.}\florence{+1, the phrasing makes it sound like this is an exceptional event, only happening at this value} If we increase the local growth rate in the new habitat\florence{,} the local maximum disappears since we increase the positive effect due to dispersal (term (3)) and the local growth rate in new habitats (term (1) in $\varphi_{\text{new}}$) (see also our response to comment~\comref{AE5}  from the associate editor).

\begin{referee}
line 253.\addnb{R2_23} I would change 'where' to 'after'.%\\
\end{referee}
done
\begin{referee}
line 261.\addnb{R2_24} Change 'evolve' to 'reproduce, disperse and die' or something like that - individuals do not evolve.%\\
\end{referee}
done
\begin{referee}
line 264.\addnb{R2_25} Add 'total' before 'numbers' if that is the meaning. Also, specify the point in the life cycle at which $N_m^k$ is defined.
\end{referee}

We have rephrased that sentence in order to clarify the two-type branching process, see line~\ref{R2-25}.

\begin{referee}
line 271.\addnb{R2_26} I would add '(see SM1 for the derivation of $s_{\text{old}}$ at the wild-type equilibrium)' after the equation number (which should be 2).
\end{referee}

We followed the suggestion and added the reference to the SI (line~\ref{R2-26}).

\begin{referee}
top of page 15.\addnb{R2_27} See comments on lines 102ff in the supplement on the interpretation of the terms in these equations.
\end{referee}

Our explanation in the main text is intentionally very heuristic and not focused too strongly on the formula. In the SI, we \st{now }added \florence{/now provide} more details on the effects of the dispersal rate on the single terms of the establishment probability, see also our response to comment~\comref{R2_50}.

\begin{referee}
line 298.\addnb{R2_28} Add '(see SM1)' at the end.
\end{referee}

We added the reference as suggested (line~\ref{R2-28}).

\begin{referee}
line 309.\addnb{R2_29} Add 'in the old habitat' after 'reproduction'.
\end{referee}

The approximation of the establishment probability deviates for both estimates, $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$, for large values of the emigration probability $m$. Therefore, we did not change the sentence.

\begin{referee}
lines 327-328.\addnb{R2_30} I would change 'larger amounts of mutant individuals that, once emigrated to a new habitat, reimmigrate into an' to 'a larger probability that an individual born in a new habitat migrates to an'. As I understand the model, individuals can only migrate once, so individuals don't migrate and then re-immigrate. Also, extinction is most likely when there is only one mutant, so I think it is better to frame this statement in terms of probability rather than numbers. ('Re-immigrate' is used elsewhere also, and it would be better to re-word these passages to avoid this term, or make explicit that it is a lineage, or descendant, that re-immigrates.)
\end{referee}

We agree with the reviewer in all points and have changed the wording accordingly.\florence{[re-immigration makes sense where talking about a lineage, doesn't it? I think it is still an interesting perspective to have and suggest talking about lineages - i.e. differentiate between what happens to individuals (cannot reimmigrate) and lineages, and mention both. What do you think?]} We now avoid speaking of re-immigration since indeed an individual only migrates at most once before its death (lines \ref{R2-30-3} and \ref{R2-30-4}). We also followed the suggestion to speak of migration probabilities rather than proportions of individuals (line~\ref{R2-30-1}) since this is indeed, as the reviewer points out, the crucial point.

\begin{referee}
Figure 2\addnb{R2_31} legend, third line. Add 'symbols' after 'simulations' (assuming that is correct). 
\end{referee}

As suggested, we added `(symbols)' to clarify that simulation results are depicted as the different symbols.

\begin{referee}
lines 332-333.\addnb{R2_32} I would change 'number of mutants in the new-habitat patches through back-migration' to 'probability that a mutant migrates from new to old habitat'. It would be good to point out that the expected number of offspring of a mutant in new habitat that remain in new habitat is the product of $1 + s_{\text{new}}$ and $1 - m^{(\text{new}\to\text{old})}$, the latter of which increases with m. For high enough m, this can be less than 1, in which case persistence requires descendants of mutants that leave migrate back into new habitat at a rate sufficient to make up for the deficit, which is unlikely (assuming that the reproduction number is less than 1 in old habitat also). This probably explains the 0 values for establishment rate in some of the figures, even when the mutant starts in new habitat. 
\end{referee}

We agree with the reviewer and therefore changed the phrasing to the suggested form. We also added a more details on the mechanism as described by the reviewer, see lines~\ref{R2-32}ff.\florence{[please see comment above \comref{R2_30}]}

\begin{referee}
lines 378-379.\addnb{R2_33} Remind the reader that with absolute habitat choice, both types prefer the old habitat, so the behavior of the wild type is the same and the mutant type the opposite than with relative habitat choice. This generally leads to lower establishment than the other so far discussed, because the mutant behavior is maladaptive (which also explains the similarity to the maladaptive curves noted in the next paragraph). The exception is the case you note, in which the mutant does better in old habitat at high m, in which absolute habitat choice can lead to higher establishment than relative habitat choice.   
\end{referee}

Since we have changed the names of the dispersal schemes to the corresponding preference, i.e. `Old-Old' for both types having a bias towards the old habitat, we think that we do not need any further reminder concerning the dispersal preferences.

\begin{referee}
lines 407-408.\addnb{R2_34} If only mutants arising in old habitat are considered, I think the probability of adaptation should be the probability that a binomial with parameters $M*f_{\text{old}}*K*t_{\text{fin}}$ (number of offspring that survive density regulation) and $\theta*\varphi_{\text{old}}$ (probability that a birth produces a mutant whose lineage survives) is 0. A Poisson should give a very good approximation, but assuming it is an approximation, that should be stated. (The situation is more complicated for mutants arising in new habitats, because the number of wild-type individuals is a random variable with I think a complicated distribution.)
\end{referee}

We added a \st{very brief }comment\st{, not going into too many details,}\florence{} that using the Poisson distribution is indeed, as the reviewer correctly pointed out, an approximation (line~\ref{R2-34}).

\begin{referee}
lines 421-422.\addnb{R2_35} It could be pointed out that there are some differences in the order in panel (c) that likely results from effects of dispersal type on the equilibrium $N_w^{\text{new}}$. 
\end{referee}

We have added a \st{brief }\florence{}comment where we point out that the small differences between Fig.~\ref{fig:vary_m_est}(c) and Fig.~\ref{fig:source_sink}(c) in terms of the ranking of dispersal schemes goes back to the equilibrium sizes of the wild type in new-habitat patches, as pointed out by the reviewer (cf. lines~\ref{R2-35}ff.).

\begin{referee}
lines 441-3.\addnb{R2_36} Add 'at least' before 'one' and change 'a mutant from an old-habitat patches' to 'mutants from old-habitat patches only'.\\

lines 446-448.\addnb{R2_37} It is helpful in explaining this to note that the rate at which successful lineages are generated in each habitat type is proportional to $\varphi_i*f_i*N(hat)_w^i$. If figure 4(a), adaptation requires $f_{\text{old}} < 0.5$, and is higher for old habitat even for $f_{\text{old}} = 0.1$. The reason must be much higher values in old habitat of either or both of the equilibrium wild-type population (larger in old habitat) and the probability of establishment of a mutant in each habitat. Plots of those as a function of $f_{\text{old}}$ would be helpful here. Unfortunately, although $\varphi_i$ is plotted as a function of $f_{\text{old}}$ in figure S4, that figure uses a much higher m. $N(hat)_w^{\text{new}}$, however, seems to be about 2-7\% of K for the points in (a), if my calculations are correct.
\end{referee}

In order to disentangle the effects of population size and establishment probability on the origin of the successful mutant we now plot both values when varying the frequency of old habitats, $f_{\text{old}}$ (see Fig.~SXX in the SI). Indeed, the mutational input which is scaled by the population sizes in the corresponding patches seems to be responsible for the dominance of the old habitats concerning the origin of the rescue mutant.

We also added a brief explanation in the main text (lines~\ref{R2-36}ff.) and refer to the figure in the~SI.

\begin{referee}
line 456.\addnb{R2_38} Change 'unevitably' to 'inevitably'.%\\
\end{referee}
\florence{done}
\begin{referee}
equation (10).\addnb{R2_39} I think i should go from 0 to M - 1 and the last summation should start at M*tau. For M - 2, the limit of i in the equation, $f_{\text{old}} = 2/M$, but the last term has $f_{\text{old}}$ (the argument of $\varphi_{\text{new}}) = 0.$ So $f_{\text{old}} = 1/M$ is missing. The various terms in the exponential give the expected number of mutants arising at each time whose lineage survives. This should peak at intermediate values, because for high $f_{\text{old}}$ the rate of mutant generation is high but the lineage persistence is low, and the opposite is true for low $f_{\text{old}}$. In the middle, both are intermediate and so that product likely reaches a maximum for intermediate $f_{\text{old}}$, such as seen in figure 3b,d. This is relevant to the results for standing variation, because the lineages of initially present mutants must survive hostile conditions until $f_{\text{old}}$ drops at least to 0.8 (as implied by fig. 3d for conditions in fig. 6b).
\end{referee}

The frequency of old habitats, $f_{\text{old}}(i)$, was \st{wrongly}\florence{inaccurately} defined (see also \comref{R1_40}). We corrected \st{that mistake}\florence{it}\st{ so that the formula should be correct now} (eq.~\ref{eq:evol_rescue}). We also added the explanation that the reviewer is giving for the low impact of standing genetic variation around Figure~\ref{fig:sgv} (line~\ref{R2-39}).

\begin{referee}
Figure 5\addnb{R2_40} is based on equation (10), which is approximate. However, as in most other figures, there is a curve labeled 'exact'. I think the meaning is a numerical solution of equation (5) is used for $\varphi_i$ rather than the approximate values in equation (6). It should be made clear that exact refers only to $\varphi_i$, not the whole curve.
\end{referee}

We now clarify in the caption of Figure~\ref{fig:rescue} that the label `exact' refers to the exact solution of eq.~\eqref{eq:ext_prob}.

\begin{referee}
line 470.\addnb{R2_41} Add 'expected value of' after 'follow'.%\\
\end{referee}
\florence{done}
\begin{referee}
lines 474-479.\addnb{R2_42} I think it would be better to write that the approximation assumes that a mutant has the establishment probability assuming the environment is fixed for the environment ($f_{\text{old}}$) in which it is born (and I think also at the equilibrium population sizes for that environment). In the initial environment, the rate of creation of mutants is high but the establishment rate is low (0), and as deterioration progresses, the former decreases and the latter increases. The approximation captures the decrease in mutant creation but not the increase in establishment probability of a mutant whose lineage experiences a change in the environment (which as the paper states is likely for a mutant arising just before a change). 
\end{referee}

In line with the reviewers comment we now explain more explicitly which assumptions of the establishment probability $\varphi_k$ are violated in the evolutionary rescue situation, cf. lines~\ref{R2-42}ff.

\begin{referee}
lines 486-489.\addnb{R2_43} The wording here is not clear. It might be better to state something like negative density-dependent is higher than random which is higher than absolute habitat choice, with relative habitat choice decreasing and maladaptive dispersal increasing their positions with increasing m. 
\end{referee}

We followed the suggestion of the reviewer and rephrased the sentence accordingly (lines~\ref{R2-43}ff.).

\begin{referee}
lines 489-491.\addnb{R2_44} I assume this means that $s_{\text{old}}$ is the most influential factor in the change in position of relative and maladaptive dispersal, but if so this should be made clear. The relative position of the other three strategies I think depends on mutant dispersal probability into new habitat. 
\end{referee}

The reviewer is correct in \st{his/her}\florence{[alphabetical] her/his} assessment. We changed the wording, now differentiating between the ranking of the symmetric dispersal schemes and the change of the order for the asymmetric schemes, lines~\ref{R2-44}ff.

\begin{referee}
lines 527-529.\addnb{R2_45} Before the first degradation event, all patches are old, and I think dispersal is irrelevant. There should be an expected equilibrium number of mutants in each patch given by the solution of $X_{\text{mut}} = K p_m$, where $p_m$ as given in equation (1) with the number of mutants equal to $X_{\text{mut}}$ and the number of the wild type $= K - X_{\text{mut}}$. So the standing genetic variance should not depend on m. One reason (perhaps the major one) for the drop in the relative contribution on standing variation with increasing m is the fact that the rescue probability increases with increasing m, by a factor of about 10. If the number of populations rescued by mutants from the standing variation didn't change with m, this would cause the relative contribution to decrease by a factor of 10. Of course, once the first patch deteriorates, the loss of lineages from standing variation mutants does depend on m, but I would guess higher m is better, since it gets more mutants into the new patch and might relax competition in the old patches a little.
\end{referee}

\florence{[maybe you can subdivide the point? what about the first sentence btw?]}
The referee correctly points out that the probability of establishment after the first degradation changes with $m$. We have added a Figure in the SI to highlight this, see Fig.~SXX. There we plot the establishment probabilities $\varphi_{k}$ for varying emigration rates $m$ in the setting where one patch has deteriorated ($f_{\text{old}}=0.9$ in our standard parameter set). This is the environment which the mutants experience at time $t=0$. It can be seen that increasing $m$ indeed does not change the probability of establishment for mutants in old-habitat patches. It is in fact very close to $0$. Yet, for the mutants in the patch that deteriorates first, the establishment probability decreases with increasing $m$ for most dispersal schemes. Furthermore, for very large values of $m$ we enter the gene swamping scenario which also inhibits the establishment of mutant lineages in the new habitat.

Therefore, we are still convinced that our argument, for increasing dispersal rates the effect of standing genetic variation becomes less important for the overall rescue probability, remains valid. %To further validate this, we now also plot the effect of standing genetic variance for smaller values of $\tau$, the time between two consecutive deterioration events (see Fig.~\ref{fig:sgv}(b)). Here, the probability of evolutionary rescue (from de-novo mutants) changes less than the impact of standing genetic variation does \todo{figure!}.

\begin{referee}
line 545.\addnb{R2_46} 'Gene swamping' is often used for populations with sexual reproduction. In this case, dispersal can inhibit adaptation in a source-sink system because source individuals dispersing to the sink mate with each other but also with individuals in the sink. Therefore, if an adapted type arises in the sink but is still rare, it is likely to mate with an immigrant, which is maladapted in the sink. Their offspring will often be much less fit in the sink than the adapted parent, inhibiting adaptation. This factor is not present with clonal reproduction.   
\end{referee}

We \florence{respectfully }disagree with the reviewer. Gene swamping is also used in the context of clonal reproduction, see for instance Nagylaki, Genetics (1978) and Lenormand, TREE (2002). We therefore did not change the wording.

\begin{referee}
Supplement\addnb{R2_47} %\\
\end{referee}
\florence{done}
\begin{referee}
lines 4-5. Add 'when alone' after 'type' and 'when rare' after 'mutant'.
\end{referee}

We have specified the wording in the suggested way (line~SXX).

\begin{referee}
Equation (S10).\addnb{R2_48} There is an $f$ that should be $f_{\text{old}}$.
\end{referee}

We corrected the entire SI for typos concerning the frequency of old habitats, $f_{\text{old}}$, and the dispersal bias, $\pi_i$.

\begin{referee}
line 45.\addnb{R2_49} Add 'the lineage of' after 'of'.
\end{referee}

We changed the phrase according to the suggestion (line~SXX).

\begin{referee}
lines 102ff.\addnb{R2_50} Starting from low dispersal, increasing dispersal definitely increases the third term in the equation for $\varphi_{\text{old}}$. But this might not be the only reason for the increase in region (i). As the authors point out, the second term decreases with increasing m, and this is mostly a consequence of the increase in $s_{\text{old}}$. However, it is not mainly due to a corresponding decrease in $s_{\text{old}} - s_{\text{new}}$, but to the increase in the leading $s_{\text{old}}$ factor in the second term (line 107 says the difference $s_{\text{old}} - s_{\text{new}}$ decreases, but it actually increases, becoming less negative, when $s_{\text{old}}$ increases). C is called a scaling constant on line 284 of the main text, but it is not constant with respect to changes in m (the equation for C should also be given in the supplemental material). For $\pi_m = 1$ and $f_{\text{old}} = 0.5$ (the situation plotted in fig. S1), $C = \sqrt{(s_{\text{new}} - s_{\text{old}})^2 + m^2},$ which for $m = 0$ is ABS($s_{\text{new}} - s_{\text{old}}$), and so the second term is $-s_{\text{old}}$ (assuming $s_{\text{new}} >  s_{\text{old}}$), which is positive since $s_{\text{old}}$ is negative. So the first two terms cancel out. As m increases, $s_{\text{old}}$ increases (becomes less negative at first), so the first term increases and the second term decreases faster, because $s_{\text{new}} - s_{\text{old}}$ decreases with increasing m while $\sqrt{C}$ increases or decreases more slowly. So the sum of the first two terms for low m decreases with increasing m, which can cause $\varphi_{\text{old}}$ to decrease as in region (ii). However, when $s_{\text{old}}$ becomes $>  s_{\text{new}}$, the second term reverses and increases. Then, all three terms can increase in region (iii). 
\end{referee}

\pete{I do not really know what to do with this comment. Everything that the reviewer writes is correct. I have added the equation for $C$ in the SI and also added a few more comments concerning the points that the referee mentions. But the general punchline of the referee's comment is not clear to me...what does (s)he want us to do?} \florence{they are maybe trying to explain what happens to themselves? You can check that what you write is in line with the comment?}

\begin{referee}
Figure S1 legend.\addnb{R2_51} Change 'local' to 'mutant old-patch' on 6th line.
\end{referee}

We changed the caption of the Figure following the suggestion.

\end{document}

### 2020-06-08 ms/SI.tex
revision en cours

@@
 \begin{equation}
-    1 + a_{\text{old}} = K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}_w^{\text{old}}}\, .    
+    1 + a_{\text{old}} = K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}_w^{\text{old}}}\, .    
 \end{equation}
@@
     a_{\text{old}} = \left\{ \begin{array}{ll}
-        \frac{\omega_m K_{\text{old}}}{\omega_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
-        \frac{\omega_m}{\omega_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}. 
+        \frac{\omega^\text{old}_m K_{\text{old}}}{\omega^\text{old}_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
+        \frac{\omega^\text{old}_m}{\omega^\text{old}_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}.
     \end{array}
@@
   		\includegraphics[width=\linewidth]{fig2a.pdf}
-  		\caption{$\omega_m=1.35$ (large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{fig2c.pdf}
-  	\caption{$\omega_m=1.45$ (small fecundity difference)}
+  	\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
 	\end{subfigure}
@@
   		\includegraphics[width=\linewidth]{figS1c.pdf}
-  		\caption{$\omega_m=1.35$ (large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{figS1d.pdf}
-  	\caption{$\omega_m=1.45$ (small fecundity difference)}
+  	\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
 	\end{subfigure}
@@
 
-Note that region (iii) can be shifted to the left by increasing the absolute number of offspring of mutants in old-habitat patches, $\omega_m$, and therefore decreasing the local disadvantage of the mutant.
+Note that region (iii) can be shifted to the left by increasing the absolute number of offspring of mutants in old-habitat patches, $\omega^\text{old}_m$, and therefore decreasing the local disadvantage of the mutant.
 If shifted sufficiently to the left, like in Fig.~\ref{Sfig:contribution}(d), region (ii) might vanish due to this effect.
 
-\chg{Contrarily, if we set the fecundity parameter of the mutant in the old habitat to $\omega_m=1.1$, we see that region (iii) disappears for most dispersal schemes, cf.
+\chg{Contrarily, if we set the fecundity parameter of the mutant in the old habitat to $\omega^\text{old}_m=1.1$, we see that region (iii) disappears for most dispersal schemes, cf.
 Fig.~\ref{Sfig:weak_fecundity}.
@@
 	\end{subfigure}
-	\caption{\textbf{Disappearance of region (iii) for large fecundity differences in the old habitat.} If the mutant fecundity in old-habitat patches is too low, here $\omega_m=1.1$, the effect of relaxed competition is not strong enough to have an impact on the establishment probability for high dispersal rates.
+	\caption{\textbf{Disappearance of region (iii) for large fecundity differences in the old habitat.} If the mutant fecundity in old-habitat patches is too low, here $\omega^\text{old}_m=1.1$, the effect of relaxed competition is not strong enough to have an impact on the establishment probability for high dispersal rates.
 The establishment probability remains at zero.}
@@
 	\caption{\textbf{Establishment probability and stationary wild-type population size when varying the old-habitat frequency.} In the simulations we have used the standard set of parameters as given in Table~\ref{tab:parameters} and the unbiased dispersal scheme ($\pi_w=\pi_m=0$).
-In (a) we additionally chose the large fecundity difference scenario ($\omega_m=1.35$).}
+In (a) we additionally chose the large fecundity difference scenario ($\omega^\text{old}_m=1.35$).}
 	\label{Sfig:vary_f_origin}
@@
 %Additionally, we investigate the habitat type of the origin of the adaptive mutations for low fecundity values of the mutant in the old habitat.
-While for large fecundity values of the mutant in the old habitat most successful lineages arise in the old habitat (Fig.~\ref{fig:origin} in the main text), we see that for lower fecundity values $\omega_m$ the contribution of new-habitat mutants approaches and occasionally exceeds the number of successful mutant lineages that emerge from old-habitat patches.
-In the subsequent figure, besides the default parameters as stated in Table 1 in the main text, we chose $\omega_m=1.05$ and $\theta = 10/(MK)$.
+While for large fecundity values of the mutant in the old habitat most successful lineages arise in the old habitat (Fig.~\ref{fig:origin} in the main text), we see that for lower fecundity values $\omega^\text{old}_m$ the contribution of new-habitat mutants approaches and occasionally exceeds the number of successful mutant lineages that emerge from old-habitat patches.
+In the subsequent figure, besides the default parameters as stated in Table 1 in the main text, we chose $\omega^\text{old}_m=1.05$ and $\theta = 10/(MK)$.
 The plot corresponding to Fig.~\ref{fig:origin} in the main text then looks as follows:
@@
 %	\includegraphics[width=0.6\textwidth]{figS4.pdf}
-%  	\caption{\textbf{Habitat type of the origin for strong fecundity differences in the old habitat ($\omega_m=1.05$).}}
+%  	\caption{\textbf{Habitat type of the origin for strong fecundity differences in the old habitat ($\omega^\text{old}_m=1.05$).}}
 %	\label{Sfig:origin}
@@
 With the changed fecundities, I only get this behavior if the fecundity of the mutant is larger in the new patches than in the old patches, i.e.
-$\omega_m < a_{\text{new}}$.
+$\omega^\text{old}_m < a_{\text{new}}$.
 Also, for some reason the approximations do not work that well in these scenarios (I did not investigate why the estimated lines are bad -- soft sweeps still work fine but the other approximations don't.)}
@@
 	\end{subfigure}
-	\caption{\textbf{Probability of adaptation for a large frequency of old-habitat patches ($f_{\text{old}}=0.9$).} The fecundity of the mutant in old-habitat patches is set to $\omega_m=1.45$.
+	\caption{\textbf{Probability of adaptation for a large frequency of old-habitat patches ($f_{\text{old}}=0.9$).} The fecundity of the mutant in old-habitat patches is set to $\omega^\text{old}_m=1.45$.
 Note also the difference between the scales of the y-axes in the two panels.}
@@
 %\begin{equation}
-%    \omega_w^{\text{old}} = \omega_m^{\text{new}} \quad \text{and} \quad \omega_w^{\text{new}} = \omega_m^{\text{old}}.
+%    \omega^\text{old}_w^{\text{old}} = \omega^\text{old}_m^{\text{new}} \quad \text{and} \quad \omega^\text{old}_w^{\text{new}} = \omega^\text{old}_m^{\text{old}}.
 %\end{equation}
@@
 \begin{equation}
-    1+a_{\text{new}} = K \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}} \widetilde{N}_w^{\text{new}}+ \omega_m^{\text{new}}}\approx K \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}} \widetilde{N}_w^{\text{new}}},
+    1+a_{\text{new}} = K \frac{\omega^\text{old}_m^{\text{new}}}{\omega^\text{old}_w^{\text{new}} \widetilde{N}_w^{\text{new}}+ \omega^\text{old}_m^{\text{new}}}\approx K \frac{\omega^\text{old}_m^{\text{new}}}{\omega^\text{old}_w^{\text{new}} \widetilde{N}_w^{\text{new}}},
 \end{equation}
@@
 \begin{equation}
-    a_{\text{new}} \approx \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}}\left(1+\frac{m f_{\text{old}} (1-\widehat{\pi}_w)}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\right)}.
+    a_{\text{new}} \approx \frac{\omega^\text{old}_m^{\text{new}}}{\omega^\text{old}_w^{\text{new}}\left(1+\frac{m f_{\text{old}} (1-\widehat{\pi}_w)}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\right)}.
 \end{equation}
@@
   		\includegraphics[width=\linewidth]{figS6a.pdf}
-  		\caption{$\varphi_{\text{old}}$ with $\omega_m=1.35$ (large fecundity difference)}
+  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m=1.35$ (large fecundity difference)}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{figS6b.pdf}
-  	\caption{$\varphi_{\text{new}}$ with $\omega_m=1.35$}
+  	\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m=1.35$}
 	\end{subfigure}
@@
   		\includegraphics[width=\linewidth]{figS6c.pdf}
-  		\caption{$\varphi_{\text{old}}$ with $\omega_m=1.45$ (small fecundity difference)}
+  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m=1.45$ (small fecundity difference)}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{figS6d.pdf}
-  	\caption{$\varphi_{\text{new}}$ with $\omega_m=1.45$}
+  	\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m=1.45$}
 	\end{subfigure}
@@
 In this case, in eq.~\eqref{Seq:estab_approx} higher order corrections would need to be taken into account.
-\chg{The fecundity values in the new habitat are given by $\omega_w^{\text{new}} = \omega_m^{\text{old}}$ and $\omega_m^{\text{new}} = \omega_w^{\text{old}}$ and the carrying capacity is $K_{\text{new}}=K_{\text{old}}=K=500$.} Missing data points (mostly for the negative density-dependent dispersal scheme -- green triangles) are explained by too large computation times.
+\chg{The fecundity values in the new habitat are given by $\omega^\text{old}_w^{\text{new}} = \omega^\text{old}_m^{\text{old}}$ and $\omega^\text{old}_m^{\text{new}} = \omega^\text{old}_w^{\text{old}}$ and the carrying capacity is $K_{\text{new}}=K_{\text{old}}=K=500$.} Missing data points (mostly for the negative density-dependent dispersal scheme -- green triangles) are explained by too large computation times.
 All data points are averages from $10^4$ independent runs.
@@
 The habitat type of the origin of the rescue mutation is largely independent of the considered dispersal scheme.
-For $\omega_m=1.35$ we have plotted the relative contribution of each natal habitat type to the probability of evolutionary rescue, Fig.~\ref{Sfig:natal_habitat}.
+For $\omega^\text{old}_m=1.35$ we have plotted the relative contribution of each natal habitat type to the probability of evolutionary rescue, Fig.~\ref{Sfig:natal_habitat}.
 We do not see large differences between the three symmetric dispersal schemes (0-0, Old-Old, and New-New) and the Old-New dispersal scheme.

### 2020-06-08 ms/main.tex
revision en cours

@@
 \usepackage{fullpage}		
-\usepackage[T1]{fontenc}				
 \usepackage[english]{babel}			
@@
 
+\usepackage{enumitem}
+
 \usepackage{microtype}
@@
 \usepackage{FiraSans}
+\usepackage[T1]{fontenc}
+
 \usepackage{soul}
@@
 % Packages to comment out before submitting!
-%\usepackage[textsize = scriptsize, backgroundcolor = white, linecolor = black]{todonotes}
+\usepackage[textsize = scriptsize, backgroundcolor = white, linecolor = black]{todonotes}
 %\usepackage{showlabels}
@@
 
-\noindent The authors wish to be identified to the reviewers.
+\noindent The authors wish to be identified to the reviewers (the manuscript is already on bioRxiv so the identities of all authors can be easily found).
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Facing current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, the populations of many species are declining and might eventually go extinct \citep{bellard_2012, diniz_2019}.
+Current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, contribute to the decline of the populations of many species, that might eventually go extinct \citep{bellard_2012, diniz_2019}.
 Pests and pathogens experience similarly strong selective pressures as a result of increased consumption of antibiotics and use of pesticides \citep{ramsayer_2013, kreiner_2018}.
-The process of genetic adaptation that saves populations from extinction is termed evolutionary rescue.
-This \linelabel{R1-3}\chg{process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, typically }resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
+The process of genetic\footnote{\florence{rescue by plasticity as well}} adaptation that saves populations from extinction is termed evolutionary rescue.
+This \linelabel{R1-3}\chg{process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, } \florence{classically} resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
 In recent years, empirical examples of evolutionary rescue have accumulated \citep[as reviewed by ][]{alexander_2014,carlson_2014,bell_2017}.
-Evolutionary rescue \linelabel{R1-4}\chg{has been observed directly in laboratory experiments \citep[e.g.][]{bell_2009,agashe_2011,lachapelle_2012,lindsey_2013,stelkens_2014}.
-In wild populations demographic and genotypic data is rarely monitored at the same time impeding direct observation of evolutionary rescue.
-Still, it has been suggested as a mechanism that has saved a few wild populations from extinction \citep[e.g.][]{vanderwal_2012,digiallonardo_2015,gignoux_2018}.}
+Laboratory experiments have provided direct evidence of evolutionary rescue \linelabel{R1-4} \citep[e.g.][]{bell_2009, agashe_2011, lachapelle_2012, lindsey_2013, stelkens_2014}.
+In the wild, however, demographic and genotypic data are rarely monitored together at the same time, which impedes direct observation of evolutionary rescue.
+Still, evolutionary rescue has been suggested as a mechanism that has saved a few wild populations from extinction \citep[e.g.][]{vanderwal_2012, digiallonardo_2015, gignoux_2018}.
+
+%% Genetic setting
+
+\florence{[Need a paragraph on the genetic makeup.
+We are not explaining this at all at the moment and this is lacking] Rescue can occur through plasticity (refs Chevin?
+\footnote{https://royalsocietypublishing.org/doi/full/10.1098/rspb.2016.1690 , https://royalsocietypublishing.org/doi/abs/10.1098/rstb.2012.0089 and/or https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2009.00875.x}) or genetic adaptation (refs).
+The traits involves can be continuous (ref) or discrete (ref).
+In this work, we consider genetic adaptation mediated by discrete traits.
+Specifically, we consider that a single mutation can prevent extinction, and...}
 
@@
 But please check; it's a long time that I read those...}}.
-In fragmented environments, habitat deterioration is not necessarily synchronized across patches: there can be a transient spatially heterogeneous environment consisting of old- and new-habitat patches until eventually the whole environment has deteriorated.
+Such gradual changes can in particular occur in fragmented environments.
+
+In fragmented environments, habitat deterioration is not necessarily synchronized across patches: there can be a transient spatially heterogeneous environment consisting of a mosaic of old and of degraded habitat patches, until eventually the whole environment has deteriorated.
 If individuals that populate different patches are able to move between those, the effect of dispersal on evolutionary rescue needs to be taken into account \citep{uecker_2014, tomasini_2019}.
-Experiments that study the effect of dispersal on evolutionary rescue are rare but indicate that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation \citep{bell_2011}.
-
-\linelabel{R1-1}\chg{Evolutionary rescue in a fragmented environment can be interpreted as the intermediate situation between abrupt change and gradual change: if dispersal is very low, patches are approximately isolate and thus each patch undergoes an abrupt change independent of the other patches; if the dispersal rate and the number of patches are very high, the whole environment is gradually changing since spatial structure and the contribution of a single patch are negligible.}
+\linelabel{R1-1}\chg{The intensity of dispersal among patches tunes how abruptly environmental change is experienced.
+With very low dispersal, patches are essentially isolated from each other, and each patch undergoes an abrupt change independently of the other patches.
+With higher dispersal, asynchronous deterioration among patches is experienced as a more gradual change overall.}
+Experiments that study the effect of dispersal on evolutionary rescue are rare, but, for instance, \citet{bell_2011} found that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Evolutionary rescue in a situation where one patch after \linelabel{R2-2}\chg{another} deteriorates is tightly linked to the study of adaptation to a heterogeneous environment with source-sink dynamics.
-These describe a spatially heterogeneous environment that is constant in time where a population in unfavorable habitats can be maintained by constant immigration of the wild type.
-Experimental and theoretical studies have found that \linelabel{R2-3}\chg{increasing} dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g.
-\citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000,ronce_2001,gallet_2018}} for both effects).
+The transient mosaic of degraded and non-degraded patches that results from asynchronous degradation in a fragmented habitat is similar to the setting of models of source-sink dynamics.
+These models represent a spatially heterogeneous environment, constant in time, in which wild-type populations in unfavorable habitats can only be maintained thanks to recurrent immigration from favorable habitats.
+Experimental and theoretical studies have found that \linelabel{R2-3}\chg{increasing} dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g., for studies on discrete traits, \florence{[refocus the citations with papers on discrete traits; otherwise there are too many possible citations -- it's fine to do so given that we now explicitly say the paper is focused on discrete traits]}\citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000, gallet_2018}} for both effects).
 %Reason for different results: (i) demography or not (Kawecki); (ii) selection strength (Gallet), (iii) selection strength (Ronce)
-\florence{suggestion: focus on discrete traits - thousands of possible papers to cite; avoid RK2001}
-
+
+%%% DISPERSAL, biased or not
 In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be \chg{unbiased}, i.e.
@@
 Only few investigations in the context of local adaptation in source-sink systems have taken into account non-\chg{uniform} dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}.
-This analytical focus on \chg{unbiased} dispersal is in stark contrast to actually observed dispersal schemes in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}.
-Here, we explore and compare the effects of several biologically motivated dispersal schemes on adaptation and evolutionary rescue.
+This analytical focus on \chg{unbiased} dispersal is in stark contrast to dispersal schemes observed in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}.
 
@@
 
-Another well-established dispersal scheme is habitat choice, whereby individuals tend to immigrate into habitats they are best adapted to.
+Another well-established dispersal scheme is a type of habitat choice, whereby individuals tend to immigrate into habitats they are best adapted to.
 This mechanism has for example been reported in lizards \citep{bestion_2015}, birds \citep{dreiss_2011,benkman_2017}, fishes \citep{bolnick_2009}, worms \citep{mathieu_2010}, and ciliates \citep{jacob_2017,jacob_2018}.
 
-\linelabel{R1-8}\chg{Although density-dependence or habitat choice can also affect the emigration behavior, i.e.
-the likelihood that an individual leaves its current location, or even the vagrant stage when the individual is in the process of dispersal \citep{bowler_2005,ronce_2007}, we ignore this here.
-Instead, we concentrate on \chg{non-uniform} dispersal effects of the immigration process.} \\
+\linelabel{R1-8} Dispersal biases can affect the different steps of dispersal (the probability to emigrate, the vagrant stage, and immigration \citep{bowler_2005,ronce_2007}).
+In this work, we focus on effects on the immigration step.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In the following, we explicitly account for these \chg{biased} dispersal schemes in the context of genetic adaptation to a deteriorating and spatially structured environment.
 We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}\chg{types, wild type and mutant, have a positive growth rate}, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct.
-In this framework, we study four biologically motivated dispersal patterns and compare them to the \chg{unbiased} dispersal scheme.
-We start by investigating their consequences for the establishment dynamics of an adapted mutant in a time-constant heterogeneous environment.
-From there, we derive an approximation for adaptation in a source-sink setting.
-Using these results, we study the probability of evolutionary rescue, i.e.
-of adaptation in the scenario where patches, one after \chg{another}, deteriorate over time until all locations contain the new habitat.
-We find that dispersal bias has a direct effect on the local growth rates which in turn affect the probability of establishment and evolutionary rescue.
+We study four biologically motivated dispersal schemes, which correspond to the four combinations of biases towards old vs.\ new patches for wild type and mutants, and we compare these dispersal schemes to \chg{unbiased} dispersal.
+Our analysis is carried out step-wise.
+We first consider a temporally constant but spatially heterogeneous environment with two (`old' and `new') patch types.
+In this setting, we first study the probability of establishment of a single mutant, assuming there are no further mutations between types.
+We then relax the assumption of no further mutations, and compute a probability of adaptation, i.e.
+of establishment of the mutant lineage\footnote{\florence{not sure about this defn, please check}}.
+Finally, we let habitat degradation proceed, assuming that patches, one after \chg{another}, deteriorate over time until all locations contain the new habitat.
+Using the previous results, we approximate a probability of evolutionary rescue, i.e.\ that a mutation appears, establishes, thereby allowing the population to persist in spite of environmental degradation.
+We find that dispersal biases affect the probabilities of establishment and of evolutionary rescue, because it directly affects the local growth rates.
 
@@
 The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively.
-The transition from the old to the new habitat is assumed to be irreversible.
-Patches are populated by two types of asexually reproducing, haploid individuals, either wild type or mutant.
-Generations are measured in discrete time steps and are non-overlapping.
-\linelabel{R1-9-2}\linelabel{R2-6}\chg{Wild-type and mutant individuals both have a positive growth rate} in old-habitat patches; the mutant has a lower growth rate there.
-In new-habitat patches, the wild type population declines locally while the mutant has a positive growth rate -- we will therefore call it a `rescue mutant'.
-\chg{Both habitat types, old and new, have a (potentially different) carrying capacity.
-At the end of a life cycle, the local population size is reduced to the carrying capacity if necessary.}
-
-We consider a time-varying environment, such that one after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates (from old to new state).
+One after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible.
 Initially ($t<0$), all patches are of the old-habitat type.
@@
 We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$.
-It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event until it eventually hits~$0$, when all patches have undergone the environmental change.
-This environmental setting corresponds to the one analyzed by \citet{uecker_2014} and more recently in the special case of just two patches by \citet{tomasini_2019}.
-
-The individuals are assumed to go through the following life-cycle: (i) Dispersal: individuals may move between patches; (ii) Reproduction: individuals reproduce within patches.
-\linelabel{R2-7}\chg{The number of offspring per parent is modeled by a Poisson number}; (iii) Mutation: wild-type individuals mutate to the rescue mutant type with probability $\theta$ -- back mutations from the mutant to the wild type are neglected; (iv) Regulation: the population size is, \chg{if necessary, down-regulated to the carrying capacity $K_k$ where $k$ refers to the habitat type (old or new - the two habitat types can have different carrying capacities)}.
-If the population size is below the carrying capacity, the regulation step is ignored.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Old habitat dynamics %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Old-habitat dynamics}
-\linelabel{R1-10}We assume that in old-habitat patches, \linelabel{R1-9-3}\chg{both types have a positive growth rate such that the population in these patches is typically at carrying capacity $K_{\text{old}}$.} We denote the fecundities of the \textbf{w}ild-type and the \textbf{m}utant by $\omega_w$ and $\omega_m$, respectively.
-We impose an adaptation trade-off for the rescue mutant, i.e. the mutant is less fit than the wild type in old-habitat patches ($\omega_m<\omega_w$).
-%
-\linelabel{R2-9}\chg{We assume that mutation from the wild type to the mutant is tied to reproduction.
-The probability that an offspring of a wild-type individual is a mutant is given by $\theta$.
-We neglect back-mutation from mutant to wild type.}
-\linelabel{R2-8}\chg{After reproduction, if the population size exceeds the carrying capacity $K_{\text{old}}$ we randomly remove individuals from the population until the carrying capacity is reached\linelabel{AE-3}, i.e.
-we assume that the wild type and the mutant are competitively equivalent.
-The mean number of offspring of\footnote{\francois{produced by}} one mutant after reproduction and population regulation is then given by
-\begin{equation}\label{eq:sold}
-     K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}^{\text{old}}_w + \omega_m}\approx K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}^{\text{old}}_w} =1 + a_{\text{old}} \, ,
-\end{equation}
-where the random variable $\widetilde{N}_{i}^{k}$ denotes the population sizes of type $i$ (mutant or wild-type) in patches with habitat $k$ (old or new) after the dispersal step.
-Additionally, we have defined the local growth rate of a single mutant, denoted by $1+a_{\text{old}}$.
-\linelabel{R1-11}The approximation in eq.~\eqref{eq:sold} holds if the number of mutants is much lower than the number of wild types in old-habitat patches -- a reasonable assumption since the establishment of the mutant is decided when it is still rare.}
-%Then, while still rare, mutant individuals will reproduce independently of each other and their dynamics will be described by the local growth rate given in eq.~\eqref{eq:sold}. 
-\chg{Compared to the wild type we assume a reduced fecundity of the mutant in the old habitat, i.e.
-$\omega_m<\omega_w$.
-Still, the local growth rate can be positive, that is, larger than one ($a_{\text{old}}>0$).} This happens when the \linelabel{R2-12}\chg{wild-type population in a single old-habitat patch after dispersal,} \linelabel{R1-13}$\widetilde{N}^{\text{old}}_w$, becomes very small.
-This strongly depends on the dispersal scheme and rate.
-\linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-% New habitat dynamics %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{New habitat dynamics}
-In the new habitat and in the absence of immigration, the wild-type population will go extinct.
-To account for this, we assume that the number of wild-type \linelabel{R1-15}offspring \chg{per parent} in new-habitat patches is a Poisson distributed value with parameter $(1-r)$, where $r$ measures how strongly the wild type is affected by the environmental deterioration.
-For $r=1$ the wild-type population becomes locally extinct in new-habitat patches after one generation.
-
-The mutant type, in contrast, has a local growth rate larger than one and will (in expectation) grow in numbers.
-The number of offspring is again Poisson distributed with per-capita rate $(1+a_{\text{new}})$, the local growth rate of the mutant in new-habitat patches.
-
-\linelabel{R1-17}\linelabel{R2-13}\chg{In the following analysis we will assume that the mutant dynamics in new-habitat patches are density-independent.
-This is a valid assumption as long as the carrying capacity in these patches, $K_{\text{new}}$, is not reached.
-Of course, in the simulations the carrying capacity might sometimes be reached, especially for large frequencies of old-habitat patches.
-This assumption on density-independence also restricts our choice for the carrying capacity in the new habitat.
-If the carrying capacity in new-habitat patches is much smaller than in old-habitat patches, the new habitat can reach its carrying capacity merely due to immigration of wild-type individuals (even after reproduction where the wild-type population is reduced).\footnote{\francois{Rephrase, as this is not clear.
-What carrying capacity do you choose and whether your approximation holds or not.
-Maybe: `Here we examine a carrying capacity in new habitats which is half that in old habitats.
-If the carrying capacity were much smaller, our analytical approximation would break down.}}}
+It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event, until it eventually hits~$0$, when all patches have undergone the environmental change.
+This setting corresponds to the one analyzed by \citet{uecker_2014}, and more recently by \citet{tomasini_2019} in the special case of just two patches\footnote{\florence{what about carrying capacities in these papers?}}.
+The maximum numbers of individuals that can live in a patch of a given habitat type, i.e.\ the carrying capacities, are denoted $K_{\text{old}}$ and $K_{\text{new}}$ for old- and new-habitat patches respectively; $K_{\text{old}}$ and $K_{\text{new}}$ may differ.
+
+The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping.
+There are two possible types of individuals, wild types and mutants.
+The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
+	\item Dispersal: individuals may move between patches. Further details about this step are given below.
+	\item Reproduction: individuals reproduce within patches.
+\linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$.
+Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average.
+On the contrary, the local population is viable if the expected fecundity is greater than $1$.
+We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}.
+In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$.
+All parents die.
+	\item Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
+	\item Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$.
+One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent.
+If the number of offspring is below the carrying capacity, the regulation step is ignored.
+	At the end of this step, all offspring become adults, and a new cycle then begins.
+\end{enumerate}
 
@@
 	\item \textbf{Old-Old:} Individuals prefer to immigrate into the habitat where they have the largest number of offspring before regulation.
-\linelabel{R1-22}\linelabel{R2-16}\chg{Since the average number of offspring per parent is higher in the old habitat for both types ($\omega_w>1-r$ and $\omega_m>1+a_{\text{new}}$), this translates} to both $\pi_w$ and $\pi_m$ being larger than~$0$, i.e.
+\linelabel{R1-22}\linelabel{R2-16}\chg{Since the average number of offspring per parent is higher in the old habitat for both types ($\omega^\text{old}_w>1-r$ and $\omega^\text{old}_m>1+a_{\text{new}}$), this translates} to both $\pi_w$ and $\pi_m$ being larger than~$0$, i.e.
 a bias towards old-habitat patches.
@@
 			$K_k$ & Carrying capacity in a patch of type $k$ & -- & \makecell{$K_{\text{old}}=1000$,\\ $K_{\text{new}}=500$} \\
-			$\omega_j$ & \makecell{Expected per-capita number of\\ type $j$ offspring in the old habitat\\ before regulation} & $0\leq \omega_m < \omega_w$ & \makecell{$\omega_w = 1.5$,\\ $\omega_m=1.45$ {\footnotesize (weak sel.)},\\ $\omega_m=1.35$ {\footnotesize (strong sel.)}}\\
+			$\omega_j$ & \makecell{Expected per-capita number of\\ type $j$ offspring in the old habitat\\ before regulation} & $0\leq \omega^\text{old}_m < \omega^\text{old}_w$ & \makecell{$\omega^\text{old}_w = 1.5$,\\ $\omega^\text{old}_m=1.45$ {\footnotesize (weak sel.)},\\ $\omega^\text{old}_m=1.35$ {\footnotesize (strong sel.)}}\\
 			$1+a_{\text{old}}$ & \makecell{Growth rate of the mutant\\ in the old habitat} & $-1\leq a_{\text{old}}$ & -- \\
@@
 If not stated otherwise, the default parameter values are used for the stochastic simulations.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Old habitat dynamics %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Old-habitat dynamics}
+\florence{It is not clear whether population size is at Kold or not}
+\linelabel{R1-10}\chg{We assume that the populations in old-habitat patches \linelabel{R1-9-3}are at carrying capacity $K_{\text{old}}$, since both types have a positive growth rate in these patches.} We denote the fecundities of the \textbf{w}ild-type and the \textbf{m}utant by $\omega^\text{old}_w$ and $\omega^\text{old}_m$, respectively.
+We assume that adaptation to the new habitat comes at a cost, such that the mutant is less fit than the wild type in old-habitat patches ($\omega^\text{old}_m<\omega^\text{old}_w$).
+%
+\linelabel{R2-9}Mutations occur upon reproduction\footnote{\florence{please check my phrasing}}.
+The probability that the offspring of a wild-type individual is mutant is given by $\theta$.
+We neglect back-mutation from mutant to wild type.
+\linelabel{R2-8}\chg{After reproduction, if the population size exceeds the carrying capacity $K_{\text{old}}$ we randomly remove individuals from the population until the carrying capacity is reached\linelabel{AE-3}, i.e.
+we assume that the wild type and the mutant are competitively equivalent.
+The mean number of offspring of\footnote{\francois{produced by}} one mutant after reproduction and population regulation is then given by
+\begin{equation}\label{eq:sold}
+     K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m}\approx K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} =1 + a_{\text{old}} \, ,
+\end{equation}
+where the random variable $\widetilde{N}_{i}^{k}$ denotes the population sizes of type $i$ (mutant or wild-type) in patches with habitat $k$ (old or new) after the dispersal step.
+Additionally, we have defined the local growth rate of a single mutant, denoted by $1+a_{\text{old}}$.
+\linelabel{R1-11}The approximation in eq.~\eqref{eq:sold} holds if the number of mutants is much lower than the number of wild types in old-habitat patches -- a reasonable assumption since the establishment of the mutant is decided when it is still rare.}
+%Then, while still rare, mutant individuals will reproduce independently of each other and their dynamics will be described by the local growth rate given in eq.~\eqref{eq:sold}. 
+\chg{Compared to the wild type we assume a reduced fecundity of the mutant in the old habitat, i.e.
+$\omega^\text{old}_m<\omega^\text{old}_w$.
+Still, the local growth rate can be positive, that is, larger than one ($a_{\text{old}}>0$).} This happens when the \linelabel{R2-12}\chg{wild-type population in a single old-habitat patch after dispersal,} \linelabel{R1-13}$\widetilde{N}^{\text{old}}_w$, becomes very small.
+This strongly depends on the dispersal scheme and rate.
+\linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+% New habitat dynamics %%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{New habitat dynamics}
+In the new habitat and in the absence of immigration, the wild-type population will go extinct.
+To account for this, we assume that the number of wild-type \linelabel{R1-15}offspring \chg{per parent} in new-habitat patches is a Poisson distributed value with parameter $(1-r)$, where $r$ measures how strongly the wild type is affected by the environmental deterioration.
+For $r=1$ the wild-type population becomes locally extinct in new-habitat patches after one generation.
+
+The mutant type, in contrast, has a local growth rate larger than one and will (in expectation) grow in numbers.
+The number of offspring is again Poisson distributed with per-capita rate $(1+a_{\text{new}})$, the local growth rate of the mutant in new-habitat patches.
+
+\linelabel{R1-17}\linelabel{R2-13}\chg{In the following analysis we will assume that the mutant dynamics in new-habitat patches are density-independent.
+This is a valid assumption as long as the carrying capacity in these patches, $K_{\text{new}}$, is not reached.
+Of course, in the simulations the carrying capacity might sometimes be reached, especially for large frequencies of old-habitat patches.
+This assumption on density-independence also restricts our choice for the carrying capacity in the new habitat.
+If the carrying capacity in new-habitat patches is much smaller than in old-habitat patches, the new habitat can reach its carrying capacity merely due to immigration of wild-type individuals (even after reproduction where the wild-type population is reduced).\footnote{\francois{Rephrase, as this is not clear.
+What carrying capacity do you choose and whether your approximation holds or not.
+Maybe: `Here we examine a carrying capacity in new habitats which is half that in old habitats.
+If the carrying capacity were much smaller, our analytical approximation would break down.}}}
 
@@
   		\includegraphics[width=\linewidth]{fig2a.pdf}
-  		\caption{$\varphi_{\text{old}}$ with $\omega_m = 1.35$ (large fecundity difference)}
+  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.35$ (large fecundity difference)}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{fig2b.pdf}
-  		\caption{$\varphi_{\text{new}}$ with $\omega_m = 1.35$}
+  		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.35$}
 	\end{subfigure}
@@
   		\includegraphics[width=\linewidth]{fig2c.pdf}
-  		\caption{$\varphi_{\text{old}}$ with $\omega_m = 1.45$ (small fecundity difference)}
+  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.45$ (small fecundity difference)}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{fig2d.pdf}
-  		\caption{$\varphi_{\text{new}}$ with $\omega_m = 1.45$}
+  		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.45$}
 	\end{subfigure}
-	\caption{\textbf{Establishment probability when varying the emigration rate.} \small We plot the theoretical results for $\varphi_{\text{old}}$ in (a) and (c), and for $\varphi_{\text{new}}$ in (b) and (d) for $\omega_m=1.35$ in (a), (b) and $\omega_m=1.45$ in (c), (d).
+	\caption{\textbf{Establishment probability when varying the emigration rate.} \small We plot the theoretical results for $\varphi_{\text{old}}$ in (a) and (c), and for $\varphi_{\text{new}}$ in (b) and (d) for $\omega^\text{old}_m=1.35$ in (a), (b) and $\omega^\text{old}_m=1.45$ in (c), (d).
 Comparison with the results from stochastic simulations (symbols) show very good agreement with our approximation found in eq.~\eqref{eq:estab_approx} (dashed lines).
@@
 The qualitative dependence of the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ on the dispersal probability $m$ is similar for all dispersal schemes.
-The shape of the curve however strongly depends on the fecundity $\omega_m$ of mutants in the old habitat.
+The shape of the curve however strongly depends on the fecundity $\omega^\text{old}_m$ of mutants in the old habitat.
 Before discussing the differences between the dispersal schemes, we first provide a qualitative understanding of this general behavior.
@@
 
-\chg{In line with this reasoning we find} that the width of region (ii) strongly depends on the fecundity of the mutant in the old habitat, $\omega_m$.
-If $\omega_m$ is high, the local growth rate of the mutant $a_{\text{old}}$ starts to increase at a lower dispersal rate
+\chg{In line with this reasoning we find} that the width of region (ii) strongly depends on the fecundity of the mutant in the old habitat, $\omega^\text{old}_m$.
+If $\omega^\text{old}_m$ is high, the local growth rate of the mutant $a_{\text{old}}$ starts to increase at a lower dispersal rate
 %(see Fig.~S1). 
 \chg{and }region (ii) disappears, as visible in Fig.~\ref{fig:vary_m_est}(c).
-In contrast, for low mutant fecundity values $\omega_m$, region (iii) might vanish (see Fig.
+In contrast, for low mutant fecundity values $\omega^\text{old}_m$, region (iii) might vanish (see Fig.
 S2 in SI).
@@
 In Fig.~\ref{fig:source_sink} we compare our predictions to simulation results.
-The shape of the probability of adaptation as function of the emigration probability $m$ depends on the fecundity of the mutant in old-habitat patches, $\omega_m$ (Figs.~\ref{fig:source_sink}(a,c)).
+The shape of the probability of adaptation as function of the emigration probability $m$ depends on the fecundity of the mutant in old-habitat patches, $\omega^\text{old}_m$ (Figs.~\ref{fig:source_sink}(a,c)).
 This is similar to the behavior of the establishment probability $\varphi_{\text{old}}$ in Fig.~\ref{fig:vary_m_est}.
@@
  		 \includegraphics[width=\linewidth]{fig3a.pdf}
-  		\caption{$\omega_m=1.35$ (large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{fig3b.pdf}
-  		\caption{$\omega_m=1.35$}
+  		\caption{$\omega^\text{old}_m=1.35$}
 	\end{subfigure}
@@
  		 \includegraphics[width=\linewidth]{fig3c.pdf}
-  		\caption{$\omega_m=1.45$  (small fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.45$  (small fecundity difference)}
 	\end{subfigure}%
@@
   		\includegraphics[width=\linewidth]{fig3d.pdf}
-  		\caption{$\omega_m=1.45$}
+  		\caption{$\omega^\text{old}_m=1.45$}
 	\end{subfigure}
@@
 %Even for a relatively strong fecundity disadvantage of the mutant in the old habitat, i.e.
-$\omega_m$ is $10\%$ smaller than $\omega_w$, we find that most successful mutations arise in old-habitat patches (Fig.~\ref{fig:origin}(a)).
-Decreasing the fecundity disadvantage of the mutant by increasing its fecundity $\omega_m$ further increases the number of successful mutant lineages from old-habitat patches (Fig.~\ref{fig:origin}(b)).
-%If we instead increase the fecundity difference in old habitats by decreasing $\omega_m$, we observe the opposite, i.e. \chg{the two lines approach each other} (see Fig.~S4 in SI).
+$\omega^\text{old}_m$ is $10\%$ smaller than $\omega^\text{old}_w$, we find that most successful mutations arise in old-habitat patches (Fig.~\ref{fig:origin}(a)).
+Decreasing the fecundity disadvantage of the mutant by increasing its fecundity $\omega^\text{old}_m$ further increases the number of successful mutant lineages from old-habitat patches (Fig.~\ref{fig:origin}(b)).
+%If we instead increase the fecundity difference in old habitats by decreasing $\omega^\text{old}_m$, we observe the opposite, i.e. \chg{the two lines approach each other} (see Fig.~S4 in SI).
 
@@
 %As a result, instead of assuming a constant establishment probability between two deterioration events, we should rather use a time-dependent establishment probability -- which can however not be approximated in our general framework.
-\linelabel{R1-43}\chg{In \citet{uecker_2014}, scenarios with an accessible time-dependent solution were studied, more precisely situations with either full mixing of the global population ($m=1$) or a non-viable mutant in old-habitat patches ($\omega_m=0$).}
+\linelabel{R1-43}\chg{In \citet{uecker_2014}, scenarios with an accessible time-dependent solution were studied, more precisely situations with either full mixing of the global population ($m=1$) or a non-viable mutant in old-habitat patches ($\omega^\text{old}_m=0$).}
 
@@
   		\includegraphics[width=\linewidth]{fig5a.pdf}
-  		\caption{$\omega_m=1.45$ (small fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
 	\end{subfigure}%
@@
   		\includegraphics[width=\linewidth]{fig5b.pdf}
-  		\caption{$\omega_m=1.35$ (large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
 	\end{subfigure}
@@
   		\includegraphics[width=\linewidth]{fig5c.pdf}
-  		\caption{$\omega_m=1.1$ (very large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.1$ (very large fecundity difference)}
 	\end{subfigure}%
@@
   		\includegraphics[width=\linewidth]{fig5d.pdf}
-  		\caption{$\omega_m=0$ (lethal mutant)}
+  		\caption{$\omega^\text{old}_m=0$ (lethal mutant)}
 	\end{subfigure}
@@
 The dispersal schemes affect the dispersal pattern of both types and as such alter their respective population dynamics.
-Both small and large fecundity differences in the old habitat, $\omega_m=1.45$ and $\omega_m=1.35$, conserve the ranking of dispersal schemes obtained for the probabilities of establishment and adaptation\footnote{\francois{difficult to follow}\pete{We could also get rid of the explict description of the figure and remove the following sentence}}: \linelabel{R2-43}\chg{New-New (green) is higher than unbiased dispersal (black) which is higher than Old-Old (blue), with Old-New decreasing and New-Old increasing their positions with increasing emigration probability $m$}, cf.
+Both small and large fecundity differences in the old habitat, $\omega^\text{old}_m=1.45$ and $\omega^\text{old}_m=1.35$, conserve the ranking of dispersal schemes obtained for the probabilities of establishment and adaptation\footnote{\francois{difficult to follow}\pete{We could also get rid of the explict description of the figure and remove the following sentence}}: \linelabel{R2-43}\chg{New-New (green) is higher than unbiased dispersal (black) which is higher than Old-Old (blue), with Old-New decreasing and New-Old increasing their positions with increasing emigration probability $m$}, cf.
 Fig.~\ref{fig:rescue}(a,b).
@@
 
-For very large fecundity differences in the old habitat ($\omega_m = 1.1$ and $\omega_m=0$), the ranking of dispersal schemes is the same for all emigration probabilities; from highest to lowest, we find \chg{New-New, Old-New, unbiased dispersal, New-Old, and Old-Old} (cf.
+For very large fecundity differences in the old habitat ($\omega^\text{old}_m = 1.1$ and $\omega^\text{old}_m=0$), the ranking of dispersal schemes is the same for all emigration probabilities; from highest to lowest, we find \chg{New-New, Old-New, unbiased dispersal, New-Old, and Old-Old} (cf.
 Fig.~\ref{fig:rescue}(c,d)).
@@
 Mutations are more likely to appear in the more populated patches (old-habitat).
-However, a low mutant fecundity $\omega_m$ decreases the chance of establishment of these mutants that appear in old-habitat patches (compare black and yellow symbols in Fig.~\ref{fig:sgv}(a)).
-Lower fecundity values $\omega_m$ therefore comparatively increase the probability for the rescue mutation to have appeared in the already deteriorated habitat, i.e.
+However, a low mutant fecundity $\omega^\text{old}_m$ decreases the chance of establishment of these mutants that appear in old-habitat patches (compare black and yellow symbols in Fig.~\ref{fig:sgv}(a)).
+Lower fecundity values $\omega^\text{old}_m$ therefore comparatively increase the probability for the rescue mutation to have appeared in the already deteriorated habitat, i.e.
 in new-habitat patches.
@@
 The simulations are done by letting the system evolve for $1,000$ generations before the first deterioration event happens.
-Parameters: $\pi_m=\pi_w=0$ in all scenarios and $\omega_m=1.45$.
+Parameters: $\pi_m=\pi_w=0$ in all scenarios and $\omega^\text{old}_m=1.45$.
 The relative contribution is then determined by $(P_{\text{rescue with sgv}}-P_{\text{rescue only de novo}})/P_{\text{rescue with sgv}}$.}
@@
 Here, the effect of dispersal on adaptation depends on the growth rate differences of the mutant and the wild type in the two habitats \citep{kawecki_2000}. 
-In accordance with this result \francois{these results}, we find that dispersal \francois{monotonically} increases the probability of adaptation if the mutant is just slightly less fit than the wild type in the old habitat %, $\omega_m / \omega_w= 0.99$
+In accordance with this result \francois{these results}, we find that dispersal \francois{monotonically} increases the probability of adaptation if the mutant is just slightly less fit than the wild type in the old habitat %, $\omega^\text{old}_m / \omega^\text{old}_w= 0.99$
 (Fig.~\ref{fig:vary_m_est}(c)). 
-When the mutant's fecundity in old habitats, $\omega_m$, is smaller \francois{than what?}, establishment probabilities are non-monotonic with a local maximum at intermediate dispersal rates \francois{and increases at large dispersal rates thanks to relaxed competition} (Fig.~\ref{fig:vary_m_est}(a)).
+When the mutant's fecundity in old habitats, $\omega^\text{old}_m$, is smaller \francois{than what?}, establishment probabilities are non-monotonic with a local maximum at intermediate dispersal rates \francois{and increases at large dispersal rates thanks to relaxed competition} (Fig.~\ref{fig:vary_m_est}(a)).
 \st{However, if the mutant's disadvantage in old-habitat patches is even greater, i.e.
-$\omega_m$} \francois{When the fecundity of the mutant is even smaller, the local maximum remains but relaxed competition no longer occurs} \st{adaptation is also hindered for large dispersal rates.
+$\omega^\text{old}_m$} \francois{When the fecundity of the mutant is even smaller, the local maximum remains but relaxed competition no longer occurs} \st{adaptation is also hindered for large dispersal rates.
 The establishment probability becomes hump-shaped } (cf.

### 2020-06-12 ms/main.tex
model almost finished

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Model}
+\section*{Model and methods}
+
+\subsection*{Main assumptions and life-cycle}
 %%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$.
-All parents die.
-	\item Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
-	\item Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$.
+	All parents die at the end of this step.
+	\item \linelabel{R2-9} Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
+	\item \linelabel{R2-8}Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$.
 One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent.
@@
 
-We assume that dispersal is cost-free, i.e.
-all emigrating individuals will settle in a patch, leaving the global population size before and after dispersal unchanged.
-\linelabel{R1-18}\linelabel{R2-14}\chg{Note however, that our methods are in principle also applicable to a scenario where dispersal is associated with a (potentially type-dependent)\footnote{\francois{remove bracket}} cost.} We split the dispersal step into emigration and immigration.
-We focus on habitat choice during immigration.
-Individuals have a bias to immigrate into patches of a certain habitat type.
-The bias \chg{during} immigration \chg{for} new-habitat patches is set to \chg{zero}, without loss of generality.\footnote{\francois{I did not understand why this is needed}} The bias \chg{for immigration to} an old-habitat patch is denoted by $\pi_{i}$, with the index $i$ indicating the wild-type ($w$) or mutant ($m$) bias.
-For \chg{$\pi_i<0$}, individuals of type $i$ are more likely to settle in new-habitat patches, for \chg{$\pi_i>0$} the reverse is true.
-For $\pi_i=0$, individuals do not have a preference and dispersal is \chg{unbiased}.
-
-We consider equal and constant emigration rates for both types and habitats throughout the manuscript.
-Note again that the method would allow for type- and habitat-dependent emigration rates.
-We denote the probability \chg{that an individual leaves} its natal patch by $m$.
-Then, the probability \chg{that an individual of type $i$} \chg{emigrates (from any habitat type) and immigrates into }the new habitat is given by 
+We split the dispersal step into emigration and immigration.
+Emigration is unbiased: all individuals have the same probability $m$ of leaving the patch they were born in.
+We assumed that dispersal biases affect the immigration step.
+We denote by $\pi_{i}$ the bias for immigration to an old-habitat patch, where the index $i$ refers to the type of the dispersing individual ($w$ for wild type, $m$ for mutant).
+When $\pi_i<0$, individuals of type $i$ are relatively more likely to settle in new-habitat patches than in old-habitat patches; conversely, their bias is towards old-habitat patches when $\pi_i>0$.
+The case $\pi_i=0$ corresponds to unbiased dispersal.
+For simplicity, we assume that dispersal is cost-free.
+While local population sizes may be affected by dispersal, the global size of the metapopulation remains the same before and after dispersal.
+\linelabel{R1-18}\linelabel{R2-14} Note that our methods can readily be applied to costly dispersal (including to costs that differ among wild-type and mutant individuals), and also to type- and habitat-dependent emigration probabilities.
+
+The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is
+%
 \begin{equation}\label{eq:dispersal}
@@
 \end{equation}
-where $f_{\text{old}}$ is the frequency of old-habitat patches. \linelabel{R1-19}\chg{
-	%Since we are working with a time-discrete model the dispersal step needs to be a probability (and not a rate as is common in continuous-time models). 
-	We transformed the bias $\pi_i$ which can take any real value to a positive number, $e^{\pi_i}$.
-This ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one.
-For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.}
-
-From a biological perspective, a number of dispersal schemes are of particular interest.
-Precisely, we distinguish between the following five scenarios (see also Fig.~\ref{fig:disp_schemes} for an overview):
-
-\begin{itemize}
-	\item \textbf{Old-Old:} Individuals prefer to immigrate into the habitat where they have the largest number of offspring before regulation.
-\linelabel{R1-22}\linelabel{R2-16}\chg{Since the average number of offspring per parent is higher in the old habitat for both types ($\omega^\text{old}_w>1-r$ and $\omega^\text{old}_m>1+a_{\text{new}}$), this translates} to both $\pi_w$ and $\pi_m$ being larger than~$0$, i.e.
-a bias towards old-habitat patches.
-In practice, individuals are thought to use biotic or abiotic cues to preferentially immigrate into habitats where their fecundity is highest when compared to other habitats.
-This type of dispersal, i.e.
-matching habitat choice, has for example been observed with common lizards \textit{Zootoca vivipara} \citep{bestion_2015}, three-spine sticklebacks \textit{Gasterosteus aculeatus} \citep{bolnick_2009}, and barn owls \textit{Tyto alba} \citep{dreiss_2011}.
-
-	The same range of parameters ($\pi_w,\pi_m>0$) is obtained when implementing positive density-dependent dispersal \chg{on immigration.
-I}ndividuals are more likely to immigrate into patches with higher population densities, so both, the wild type and the mutant, will prefer the old habitat.\footnote{\francois{I would turn this around, otherwise this sounds like a fact.
-`Indeed, if both the wt and the mutant prefer the old habitat, individuals are more likely to immigrate into patches with...}} Highly populated locations can be an indication for a safe shelter, relevant for prey species, and potentially increase the mating success of individuals.
-This type of positive density-dependent dispersal on immigration (also called conspecific attraction) is for example found in several amphibians, e.g.
+where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches.
+The use of an exponential $e^{\pi_i}$ \linelabel{R1-19} ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one.
+For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.\footnote{\florence{TODO remove hat notation}}
+
+Qualitatively, there are four possible combinations of dispersal biases.
+We name them according to the preferences of wild type and then of the mutant (e.g., ``Old-New'', wild-type individuals have a bias toward old-habitat patches, and mutant individuals toward new-habitat patches).
+We add to these four dispersal schemes the case of unbiased dispersal.
+Fig.~\ref{fig:disp_schemes} provides an overview of the different schemes, together with the parameter values used in the numerical simulations \footnote{(note that all our results depend continuously on the dispersal parameters, so that varying any of the parameters will not result in a sudden change or discontinuity of the corresponding curves; in other words.).\florence{I am not sure whether this should be said somewhere, and where -- maybe discussion?}}
+
+Each of these dispersal schemes can be related to a biological illustration:
+
+\begin{description}
+	\item[Old-Old ($\pi_w>0$, $\pi_m>0$)] Both types of individuals have a bias towards old-habitat patches.
+\linelabel{R1-22}\linelabel{R2-16}If we assume that mutant individuals have a higher fecundity in old-habitat patches than in new-habitat patches (i.e., $\omega_m^{\text{old}} > \omega_m^{\text{new}}$, which is the case in our numerical examples), then this dispersal scheme correspond to biases towards the habitat where individuals have the highest fecundity.
+This type of dispersal, which can be described as matching habitat choice, has for example been observed with common lizards \textit{Zootoca vivipara} \citep{bestion_2015}, three-spine sticklebacks \textit{Gasterosteus aculeatus} \citep{bolnick_2009}, and barn owls \textit{Tyto alba} \citep{dreiss_2011}.
+
+	Population densities being high in the old-habitat patches, this dispersal scheme can also be interpreted as positive density-dependent immigration.
+For prey species, highly populated locations can be an indication for a safe shelter, or of a place with numerous mating opportunities.
+This type of positive density-dependent immigration (also called conspecific attraction) is for example found in several amphibians, e.g.
 the salamander species \textit{Mertensiella luschani} \citep{gautier_2006} and \textit{Ambystoma maculatum} \citep{greene_2016} or the frogs \textit{Oophaga pumilio} \citep{folt_2018}.
 
-	\item \textbf{Old-New:} Under this dispersal scheme, individuals tend to immigrate into habitats where they are fitter than the other type.
-This translates into the wild type preferring old patches, i.e.
-$\pi_w>0$, while mutant has a bias towards the new habitat, i.e.
-$\pi_m<0$.
-	Empirical evidence of this mechanism is scarce.
-It resembles the recently observed dispersal pattern of the ciliates \textit{Tetrahymena thermophila} with a specialist and generalist type \citep{jacob_2018}.
-The specialist disperses to its preferred habitat while the generalist prefers to immigrate to a suboptimal habitat where it outcompetes the specialist.
-
-	\item \textbf{New-New:} \chg{This dispersal scheme can be\footnote{\francois{is}} motivated by negative density-dependent dispersal on immigration,} whereby individuals are more likely to move to less populated patches.
-In these locations, resources might be more abundant, intra-specific competition alleviated and the chance of infection transmission decreased, which may compensate for the potentially reduced habitat quality.
-The corresponding parameter choice in our model is $\pi_w,\pi_m<0$, i.e.
-both types have a higher likelihood to immigrate to new-habitat patches.
-\linelabel{R1-23}\chg{We emphasize that population density in the patches changes over time but the biases are constant.
-We use the comparison with density-dependent dispersal only to motivate our chosen biases during the relevant phase of rescue, i.e.
-when the population density in new-habitat patches is low.}
-	The mutant being initially rare, it is unlikely that the carrying capacity is reached in new-habitat patches -- as assumed throughout the analysis.\footnote{\francois{Unclear to me.
-Suggestion: `Because the carrying capacity is not typically reached in new patches in the initial phase of rescue, this bias amounts to negative density-dependent dispersal in that phase.}}
-	Various empirical examples of negative density-dependent dispersal exist.
+	\item[Old-New ($\pi_w>0$, $\pi_m<0$)] Wild-type individuals preferentially immigrate into old-habitat patches, while mutants prefer new-habitat patches.
+This corresponds to immigration to patches where the focal type is fitter than the other type (since $\omega_w^{\text{old}} >\omega_m^{\text{old}}$ and $\omega_m^{\text{new}} >\omega_w^{\text{new}}$).
+	A similar dispersal scheme was recently observed for the ciliates \textit{Tetrahymena thermophila} with a specialist and generalist type \citep{jacob_2018}, and where the specialist disperses to its preferred habitat while the generalist prefers to immigrate to a suboptimal habitat where it outcompetes the specialist.
+
+	\item[New-New ($\pi_w<0$, $\pi_m<0$)] \linelabel{R1-23} Both types of individual preferentially immigrate into new-habitat patches.
+Population densities being on average lower in the new-habitat patches, and in particular, because the carrying capacity is not typically reached in new-habitat patches during the initial phase of evolutionary rescue, this dispersal scheme can be interpreted as negative density-dependent immigration, whereby individuals are more likely to move to less populated patches.
+In such locations indeed, resources might be more abundant, intra-specific competition alleviated and the chance of infection transmission decreased, which may compensate for the potentially reduced habitat quality.
 	Density-dependent immigration effects as described here, are for example found in the damselfish species \textit{Stegastes adustus} \citep{turgeon_2012} and the migratory birds \textit{Setophaga ruticilla} \citep{wilson_2017}.
@@
 
-	\item \textbf{New-Old:} For completeness, we also consider the scenario in which both types disperse preferentially into the habitat they are relatively less fit in, i.e.
-we have $\pi_w<0$ and $\pi_m>0$.
-This dispersal scheme is \chg{vaguely} related to the concept of an `ecological trap': Individuals tend to immigrate into patches that cannot sustain a population, in its most extreme form resulting in the extinction of the species \citep{battin_2004}.
-\chg{Note however,} that this does not exactly apply to our framework since here the mutant is able to maintain a population in old-habitat patches.\footnote{\francois{Consider removing the last part about the ecological trap.}}
-
-	\item \textbf{Unbiased dispersal (0-0):} Individuals do not have a bias towards any of the two habitats, $\pi_w=\pi_m=0$.
-Most theoretical results examining the interplay of dispersal and establishment have \chg{used} this dispersal scheme.
-Therefore, we use it as a benchmark to which we compare the above defined dispersal patterns.
+	\item[New-Old ($\pi_w<0$, $\pi_m>0$)] Wild-type individual preferentially immigrate into new-habitat patches, while mutants prefer old-habitat patches.
+This dispersal scheme is considered mostly for completeness, because it is biologically quite unlikely, although it can be related to the concept of an `ecological trap', wherein individuals tend to immigrate into patches that cannot sustain a population, in its most extreme form resulting in the extinction of the species \citep{battin_2004}.
+
+	\item[Unbiased dispersal (0-0) ($\pi_w=0$, $\pi_m=0$)] Neither typed has a dispersal bias.
+Most theoretical results examining the interplay of dispersal and establishment have used this dispersal scheme.
+We therefore use it as a benchmark to which we compare the biased dispersal schemes.
 	%
-\end{itemize}
+\end{description}
 
@@
 	\includegraphics[width=0.5\textwidth]{fig1.pdf}
-	\caption{\textbf{Parameter sets for the different dispersal schemes.} \small The colors and markers are the same across all figures.
-The $x$ and $y$ axes show the dispersal bias of the wild type and mutant, respectively, to immigrate into old-habitat patches ($\pi_w$ and $\pi_m$).
-The markers are located at the parameter values used for the simulations: $\pi_m=\pi_w=0$ for unbiased dispersal (0-0; {\Large $\bullet$}), $\pi_m=\pi_w=0.5$ for a type-independent bias towards old-habitat patches (Old-Old; {\color{blue} $\blacksquare$}), $\pi_m=-0.5,\pi_w=0.5$ for a wild-type preference towards the old habitat, and a mutant-bias to the new habitat (Old-New; {\color{orange}$\blacktriangle$}), $\pi_m=\pi_w=-0.5$ for a type-independent bias towards the new habitat (New-New; {\color{darkgreen}$\blacktriangledown$}), and $\pi_m=0.5,\pi_w=-0.5$ for the mutant-bias towards the old habitat and a preference of the wild type for the new habitat (New-Old; {\color{purple}$\blacklozenge$}).
-All the results derived below depend continuously on these parameters, so that varying any of the parameters will not result in a sudden change or discontinuity of the corresponding curves.
-{\color{orange} Do we need this last sentence?\color{blue}No.}}
+	\caption{\textbf{Parameter sets and legends for the different dispersal schemes.} \small The colors and markers are the same across all figures.
+The horizontal axis is the dispersal bias of the wild type, $\pi_w$ (positive values corresponding to preferential immigration into old-habitat patches), and the vertical axis that of the mutant, $\pi_m$.
+The markers are located at the parameter values used in the simulations.
+		%: $\pi_m=\pi_w=0$ for unbiased dispersal (0-0; {\Large $\bullet$}), $\pi_m=\pi_w=0.5$ for a type-independent bias towards old-habitat patches (Old-Old; {\color{blue} $\blacksquare$}), $\pi_m=-0.5,\pi_w=0.5$ for a wild-type preference towards the old habitat, and a mutant-bias to the new habitat (Old-New; {\color{orange}$\blacktriangle$}), $\pi_m=\pi_w=-0.5$ for a type-independent bias towards the new habitat (New-New; {\color{darkgreen}$\blacktriangledown$}), and $\pi_m=0.5,\pi_w=-0.5$ for the mutant-bias towards the old habitat and a preference of the wild type for the new habitat (New-Old; {\color{purple}$\blacklozenge$}).
+		}
 	\label{fig:disp_schemes}
@@
 
-All the model parameters are summarized in Table~\ref{tab:parameters} along with the default parameter values and ranges.
+\todo{to be moved} All the model parameters are summarized in Table~\ref{tab:parameters} along with the default parameter values and ranges.
 If not stated otherwise, the default parameter values are used for the stochastic simulations.
+
+\subsection*{Analysis steps}
+We decompose our analysis into several steps of increasing complexity. 
+\begin{enumerate}
+	\item We first consider an environment that is constant over time, and heterogeneous over space, with a fraction $f_{\text{old}}$ of old-habitat patches and $1-f_{\text{old}}$ of new-habitat patches.
+The population is initiated with wild-type individuals at carrying capacity in old-habitat patches, and at the migration-selection equilibrium $\widehat{N}_w^{\text{new}}$ in new-habitat patches (see Section S1 in the SI for details),
+	and with a single mutant individual, either in an old- or in a new-habitat patch. There are no further mutations ($\theta = 0$), and we compute the probability of establishment of a mutant lineage. 
+	\item We then consider the same environmental setting, but initialize the population with only wild-type individuals.
+Mutants can appear by mutation during the simulation ($\theta >0$).
+We compute the probability of adaptation, i.e.
+that, during a fixed time interval, a mutant appears by mutation and then establishes.
+	\item Finally, we consider the full scenario where each patch degrades one after the other, as described above.
+The environment is spatially and temporally variable.
+The population is initialized with only old-habitat patches, all at carrying capacity, with wild-type individuals only.
+We compute the probability of rescue, i.e.
+that a mutant appears by mutation and establishes before the population goes extinct.
+\end{enumerate}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Old-habitat dynamics}
-\florence{It is not clear whether population size is at Kold or not}
-\linelabel{R1-10}\chg{We assume that the populations in old-habitat patches \linelabel{R1-9-3}are at carrying capacity $K_{\text{old}}$, since both types have a positive growth rate in these patches.} We denote the fecundities of the \textbf{w}ild-type and the \textbf{m}utant by $\omega^\text{old}_w$ and $\omega^\text{old}_m$, respectively.
-We assume that adaptation to the new habitat comes at a cost, such that the mutant is less fit than the wild type in old-habitat patches ($\omega^\text{old}_m<\omega^\text{old}_w$).
+\subsection*{Additional assumptions for the analytical part}
+We make a few additional assumptions in the analytical part of our work; these assumptions are relaxed in the stochastic simulations.
+
+Our calculations assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents.
+We also consider that mutants are rare compared to wild-type individuals in old-habitat patches.
+Temporarily ignoring mutations, and denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of mutants in this old-habitat patch at the beginning of the next generation is
 %
-\linelabel{R2-9}Mutations occur upon reproduction\footnote{\florence{please check my phrasing}}.
-The probability that the offspring of a wild-type individual is mutant is given by $\theta$.
-We neglect back-mutation from mutant to wild type.
-\linelabel{R2-8}\chg{After reproduction, if the population size exceeds the carrying capacity $K_{\text{old}}$ we randomly remove individuals from the population until the carrying capacity is reached\linelabel{AE-3}, i.e.
-we assume that the wild type and the mutant are competitively equivalent.
-The mean number of offspring of\footnote{\francois{produced by}} one mutant after reproduction and population regulation is then given by
 \begin{equation}\label{eq:sold}
-     K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m}\approx K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} =1 + a_{\text{old}} \, ,
+K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m \widetilde{N}^{\text{old}}_m}
+%
+\approx K_{\text{old}}\, \frac{\omega^\text{old}_m \widetilde{N}^{\text{old}}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} 
+%
+= \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
 \end{equation}
-where the random variable $\widetilde{N}_{i}^{k}$ denotes the population sizes of type $i$ (mutant or wild-type) in patches with habitat $k$ (old or new) after the dispersal step.
-Additionally, we have defined the local growth rate of a single mutant, denoted by $1+a_{\text{old}}$.
-\linelabel{R1-11}The approximation in eq.~\eqref{eq:sold} holds if the number of mutants is much lower than the number of wild types in old-habitat patches -- a reasonable assumption since the establishment of the mutant is decided when it is still rare.}
-%Then, while still rare, mutant individuals will reproduce independently of each other and their dynamics will be described by the local growth rate given in eq.~\eqref{eq:sold}. 
-\chg{Compared to the wild type we assume a reduced fecundity of the mutant in the old habitat, i.e.
-$\omega^\text{old}_m<\omega^\text{old}_w$.
-Still, the local growth rate can be positive, that is, larger than one ($a_{\text{old}}>0$).} This happens when the \linelabel{R2-12}\chg{wild-type population in a single old-habitat patch after dispersal,} \linelabel{R1-13}$\widetilde{N}^{\text{old}}_w$, becomes very small.
-This strongly depends on the dispersal scheme and rate.
+%
+The approximation results from the assumption that mutants are rare compared to wild-type individuals in old-habitat patches.
+Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$.
+\hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal.
+Mutants are less fecund that wild-type individuals in old-habitat patches ($\omega^\text{old}_m < \omega^\text{old}_w$), yet the number of mutants can increase in old-habitat patches if $\widetilde{N}^{\text{old}}_w$ is low enough.
+In this case indeed, the population is far from carrying capacity, and the offspring of all types of individuals can make it to the next generation.
+Competition is fiercer as carrying capacity is nearer.
+}\footnote{\florence{I think this is more Results than Model}}.
 \linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-% New habitat dynamics %%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{New habitat dynamics}
-In the new habitat and in the absence of immigration, the wild-type population will go extinct.
-To account for this, we assume that the number of wild-type \linelabel{R1-15}offspring \chg{per parent} in new-habitat patches is a Poisson distributed value with parameter $(1-r)$, where $r$ measures how strongly the wild type is affected by the environmental deterioration.
-For $r=1$ the wild-type population becomes locally extinct in new-habitat patches after one generation.
-
-The mutant type, in contrast, has a local growth rate larger than one and will (in expectation) grow in numbers.
-The number of offspring is again Poisson distributed with per-capita rate $(1+a_{\text{new}})$, the local growth rate of the mutant in new-habitat patches.
-
-\linelabel{R1-17}\linelabel{R2-13}\chg{In the following analysis we will assume that the mutant dynamics in new-habitat patches are density-independent.
-This is a valid assumption as long as the carrying capacity in these patches, $K_{\text{new}}$, is not reached.
-Of course, in the simulations the carrying capacity might sometimes be reached, especially for large frequencies of old-habitat patches.
-This assumption on density-independence also restricts our choice for the carrying capacity in the new habitat.
-If the carrying capacity in new-habitat patches is much smaller than in old-habitat patches, the new habitat can reach its carrying capacity merely due to immigration of wild-type individuals (even after reproduction where the wild-type population is reduced).\footnote{\francois{Rephrase, as this is not clear.
-What carrying capacity do you choose and whether your approximation holds or not.
-Maybe: `Here we examine a carrying capacity in new habitats which is half that in old habitats.
-If the carrying capacity were much smaller, our analytical approximation would break down.}}}
+In new-habitat patches, on the contrary, we assume that population size remains much lower than carrying capacity, so that all offspring make it to the next generation (no regulation is necessary after reproduction).\footnote{\florence{I am getting rid of notation $r$}} The expected number of mutants in a new-habitat patch at the beginning of the next generation, given that there are $\widetilde{N}^{\text{new}}_m$ mutants in the patch right after dispersal, is
+%
+\begin{equation}\label{eq:defanew}
+\omega_{m}^{\text{new}} \widetilde{N}^{\text{new}}_m = \left(1+a_{\text{new}} \right) \widetilde{N}^{\text{new}}_m,
+\end{equation}
+which defines $a_{\text{new}}$, the per-capita growth rate of mutants in new-habitat patches.
+This assumption constrains the differences in carrying capacities among old- and new-habitat patches under which our analysis remains valid.
+In particular, $K_{\text{new}}$ cannot be too much lower than $K_{\text{old}}$, because in this case, the flow of migrants from old- to new-habitat patches would substantially increase local population sizes in new-habitat patches.
+
+\linelabel{R1-17}\linelabel{R2-13} These assumptions are relaxed in our stochastic simulations.
 
 \subsection*{Simulations}
-The algorithm implements the life cycle described above.
-First, a random number of dispersing individuals \chg{from each patch} is drawn from a binomial distribution with success probability $m$ and sample size equal to the \chg{current patch} population size.
-The dispersing individuals \chg{are pooled and} distributed according to their type and the dispersal pattern.
-\linelabel{R2-18}\chg{More precisely, we draw a binomial number of individuals of each type to immigrate into an old-habitat patch with probability $\frac{m_i^{\text{old}}}{m}$ (eq.~\eqref{eq:dispersal}) and then distribute these individuals uniformly at random over the old-habitat patches.
-The same is done with the remaining dispersing individuals who are distributed uniformly into the new-habitat patches.} \linelabel{R2-19}\chg{In each patch, reproduction is simulated by drawing a Poisson distributed number for each type.
-The mean of this Poisson number is given by the number of individuals of type $i$ in that patch times the mean number of offspring of a single individual of type $i$ which depends on the patch type (old or new).}
-Mutation of the wild-type \linelabel{R2-20}\chg{offspring} \chg{happens} with \chg{probability} $\theta$, implemented using a binomial distribution.
-Lastly, if necessary, \linelabel{R1-25}\chg{patches are down-regulated back to carrying capacity by sampling individuals uniformly at random without replacement from the population in a patch until the carrying capacity is reached (hypergeometric sampling).}
-
-In Figs.~\ref{fig:vary_m_est}-\ref{fig:origin}, we simulate a heterogeneous environment that is constant in time, i.e.
+The simulation algorithm implements the life cycle described above. We specify here the sampling distributions\footnote{\florence{please check the wording of this sentence}} that we use. 
+\begin{enumerate}[label = (\roman*)]
+	\item Dispersal: for each patch, a random number of dispersing individuals is drawn from a binomial distribution with success probability $m$.
+%and sample size equal to the patch's current population size.
+[removed because obvious; and for consistency would have to be specified for all binomial dists]
+	The dispersing individuals from all patches are pooled together and redistributed into patches according to their type and the dispersal pattern.
+\linelabel{R2-18} For each type of individuals (wild-type and mutant), immigration patches are assigned by first drawing the number of individuals who immigrate into old-habitat patches from a binomial distribution with success probability $\frac{m_i^{\text{old}}}{m}$ (eq.~\eqref{eq:dispersal}), and then distributing these individuals uniformly at random over the old-habitat patches.
+The remaining dispersing individuals are then distributed uniformly a random into the new-habitat patches.
+	\item Reproduction: \linelabel{R2-19}In each patch, reproduction is simulated by drawing a Poisson distributed number for each type.
+The mean of this Poisson number is the product of the number of individuals of type $i$ in that patch times $\omega_i^k$, the mean number of offspring of a single individual of type $i$ in a patch of habitat $k$ (old or new).
+	All adults are then removed. 
+	\item Mutation: the number of wild-type \linelabel{R2-20}offspring mutating into the mutant type is drawn from a binomial distribution, with success probability $\theta$, the mutation probability. 
+	\item Density regulation: if the number of offspring in a patch is higher than the local carrying capacity ($K_k$ for a patch of habitat-type $k$ (old or new)), \linelabel{R1-25}patches, we sample $K_k$ individuals uniformly at random without replacement from the offspring population of the patch (hypergeometric sampling).
+Otherwise, the local population is left unchanged.
+\end{enumerate}
+
+% Initial conditions
+%\paragraph{Initial conditions} 
+%In Figs.~\ref{fig:vary_m_est}-\ref{fig:origin}, we simulate a heterogeneous environment that is constant in time, i.e.
 no patches deteriorate.
-Initial population sizes are $K_{\text{old}}$ wild-type individuals in old-habitat patches, and the \footnote{\francois{add `analytical' or `theoretical'}}stationary wild-type population size $\widehat{N}_w^{\text{new}}$ in new-habitat patches (see Section S1 in the SI for details).
-\chg{Simulations for} Fig.~\ref{fig:vary_m_est} are started with initially one mutant either in a old- or a new-habitat patch and the corresponding wild-type population size is reduced by one, \chg{and} the mutation probability is set to zero.
-In \linelabel{R2-21}\chg{Figs.~\ref{fig:source_sink} and \ref{fig:origin}}, mutants solely arise due to mutations \chg{over a finite time period}.
-We \chg{call the mutant established if} the total number of mutants in all new-habitat patches exceeds $0.\chg{6} \times K_{\text{new}} \times M(1- f_{\text{old}})$\footnote{\francois{These thresholds are mysterious at this stage.}} (\chg{or in old-habitat patches $0.6\times K_{\text{old}}\times M f_{\text{old}}$ which is relevant for dispersal schemes where the mutant has a bias towards old-habitat patches, Old-Old and New-Old}).
-In the rescue scenario, where one patch after \chg{another} deteriorates (Figs.~\ref{fig:rescue} and \ref{fig:sgv}), we initialize all patches with $K_{\text{old}}$ wild-type individuals.
-Simulations are run until either the population has gone extinct or the (global) mutant population size exceeds $0.6\times K_{\text{new}}\times M$ after the last deterioration event has happened.
+Initial population sizes are $K_{\text{old}}$ wild-type individuals in old-habitat patches, and $\widehat{N}_w^{\text{new}}$ in new-habitat patches, which is the analytical stationary wild-type population size (see Section S1 in the SI for details).
+Simulations for Fig.~\ref{fig:vary_m_est} are started with initially one mutant either in a old- or a new-habitat patch and the corresponding wild-type population size is reduced by one; the mutation probability is set to zero.
+In \linelabel{R2-21} Figs.~\ref{fig:source_sink} and \ref{fig:origin}, mutants solely arise due to mutations.
+
+We consider that the mutant population has established if its total population size in patches of a given habitat type (old or new) is greater than $60\%$ the total carrying capacity of patches of that type ($(0.\chg{6} \times K_{\text{new}} \times M(1- f_{\text{old}}))$ for new-habitat patches, $(0.6\times K_{\text{old}}\times M f_{\text{old}})$  for old-habitat patches).
+\footnote{\florence{simulation duration in non-rescue scenarios?}}
 

### 2020-06-13 ms/main.tex
added SI and more text edits

@@
-\documentclass[a4paper,11pt]{article}
+\documentclass[11pt]{article}
 
@@
 \usepackage[english]{babel}			
+%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
+\usepackage{fourier}
+
 %\usepackage{ulem}
@@
 \linespread{1.7}
+%\usepackage{titlesec}
+%\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
+%\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
+%\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
+%\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}
+
 \usepackage{titlesec}
@@
 \usepackage[pdftex, breaklinks, colorlinks, citecolor=black, urlcolor=black, linkcolor=black]{hyperref} 
-\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
-\usepackage{FiraSans}
-\usepackage[T1]{fontenc}
+%\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
+%\usepackage{FiraSans}
+%\usepackage[T1]{fontenc}
 
@@
 
+%%%APPENDIX
+\usepackage[toc,page,header]{appendix}
+\usepackage{minitoc}
+
+\usepackage{chapterbib}
+
+%%%
 %opening
@@
 
+\doparttoc % Tell to minitoc to generate a toc for the parts
+\faketableofcontents % Run a fake tableofcontents command for the partocs
+
+% Make the "Part I" text invisible
+\renewcommand \thepart{}
+\renewcommand \partname{}
+
+%\part{} % Start the document part
+%\parttoc % Insert the document TOC
+
 \maketitle
@@
 \modulolinenumbers[2]
+\part{}
 
@@
 \newpage
+
 \section*{Introduction}
@@
 In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality.
-In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and horizontal gene transfer \citep{tazzyman_2014}, reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\chg{\citep{yamamichi_2015}}, bottlenecks \citep{martin_2013}, different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}.
+In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and
+horizontal gene transfer \citep{tazzyman_2014}, 
+reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\citep{yamamichi_2015}, 
+bottlenecks \citep{martin_2013}, 
+different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}.
 \chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al.
@@
 If the number of offspring is below the carrying capacity, the regulation step is ignored.
+We call ``successful offspring'' offspring that survive the density regulation step, and become adults at the next generation.
 	At the end of this step, all offspring become adults, and a new cycle then begins.
@@
 The population is initiated with wild-type individuals at carrying capacity in old-habitat patches, and at the migration-selection equilibrium $\widehat{N}_w^{\text{new}}$ in new-habitat patches (see Section S1 in the SI for details),
-	and with a single mutant individual, either in an old- or in a new-habitat patch. There are no further mutations ($\theta = 0$), and we compute the probability of establishment of a mutant lineage. 
+	and with a single mutant individual, either in an old- or in a new-habitat patch.
+There are no further mutations ($\theta = 0$), and we compute the \textit{probability of establishment} of a mutant lineage.
 	\item We then consider the same environmental setting, but initialize the population with only wild-type individuals.
 Mutants can appear by mutation during the simulation ($\theta >0$).
-We compute the probability of adaptation, i.e.
+We compute the \textit{probability of adaptation}, i.e.
 that, during a fixed time interval, a mutant appears by mutation and then establishes.
@@
 The population is initialized with only old-habitat patches, all at carrying capacity, with wild-type individuals only.
-We compute the probability of rescue, i.e.
+We compute the \textit{probability of evolutionary rescue}, i.e.
 that a mutant appears by mutation and establishes before the population goes extinct.
@@
 We also consider that mutants are rare compared to wild-type individuals in old-habitat patches.
-Temporarily ignoring mutations, and denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of mutants in this old-habitat patch at the beginning of the next generation is
+Denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of successful offspring of mutant individuals in this old-habitat patch (i.e., of offspring that survive density regulation and become adults in the next generation) is
 %
@@
 Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$.
+For the analysis, we also further assume that the number of successful offspring of mutant individuals is Poisson distributed, with mean $1+ a_{\text{old}}$.
+
 \hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal.
@@
 }\footnote{\florence{I think this is more Results than Model}}.
-\linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}
-
-In new-habitat patches, on the contrary, we assume that population size remains much lower than carrying capacity, so that all offspring make it to the next generation (no regulation is necessary after reproduction).\footnote{\florence{I am getting rid of notation $r$}} The expected number of mutants in a new-habitat patch at the beginning of the next generation, given that there are $\widetilde{N}^{\text{new}}_m$ mutants in the patch right after dispersal, is
+\linelabel{R1-14}The population dynamics of the wild type are analyzed explicitly in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI); formulas for $a_{\text{old}}$ are given in eq.~\ref{Seq:s_old}.
+
+In new-habitat patches, on the contrary, we assume that population size remains much lower than carrying capacity, so that all offspring make it to the next generation (no regulation is necessary after reproduction).\footnote{\florence{I am getting rid of notation $r$}} The expected number of successful offspring of mutant individuals in a new-habitat patch, given that there are $\widetilde{N}^{\text{new}}_m$ mutants in the patch right after dispersal, is
 %
@@
 
-\linelabel{R1-17}\linelabel{R2-13} These assumptions are relaxed in our stochastic simulations.
+To summarize, for the analysis part, we assume that the number of successful offspring of mutant individuals are Poisson-distributed, with means $(1+ a_{\text{old}})$ in old-habitat patches, and $1+ a_{\text{new}}$ in new-habitat patches, where both quantities are treated as constant over time\footnote{\florence{please check this statement}}.
+\linelabel{R1-17}\linelabel{R2-13} These assumptions, made for the sake of mathematical analysis, are relaxed in our stochastic simulations.
 
@@
 
-We investigate the effect of the different dispersal schemes on the probability of successful adaptation and evolutionary rescue.
+% Removed because the different steps are now described above
+%We investigate the effect of the different dispersal schemes on the probability of %successful adaptation and 
+%evolutionary rescue.
 We first compute the \textit{establishment probability} of a single mutant individual arising either in an old- or in a new-habitat patch.
-We then link this probability to the dynamics of a source-sink system, i.e.
+\hl{We then link this probability to the dynamics of a source-sink system}\footnote{\florence{I don't understand what this means -- link?}}, i.e.
 a fixed environment with a certain number of old habitats (sources) and new-habitat patches (sinks).
@@
 
+We proceed step-wise towards the computation of a probability of evolutionary rescue.
+For each step, we first present a mathematical analysis, for then compare our results to the output of simulations that relax the assumptions made for mathematical purposes.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We derive the probability of establishment of a mutant population starting with a single individual initially located either in an old- or a new-habitat patch.
-In this analysis, we ignore further mutations and are only concerned with the fate of this single mutant lineage.
-The dynamics of the mutant population can be described by a two-type branching process, i.e.
-\linelabel{R1-27}\chg{all mutants, the descendants of the initial mutant, reproduce, disperse and die independently of each other.} This is a reasonable assumption as long as the overall number of mutants is a lot smaller than the population size of the wild type.
-The two ``types'' of the two-type branching process correspond to the two habitat types.
-The process tracks the \linelabel{R2-25}\chg{total number of mutants in old- and new-habitat patches at the end of the life cycle.} The number of offspring of a single mutant can be approximated by Poisson distributed numbers (see eq.~\eqref{eq:sold}).
-The mean number of offspring of a single mutant, either in an old- or a new-habitat patch, is then given by the following mean reproduction matrix:
+In this first step, we consider that there is initially a single mutant individual in the population, located either in an old- or a new-habitat patch, and we compute the probability of establishment of the mutant population.
+In this step, we ignore further mutations and are only concerned with the fate of this single mutant lineage.
+
+\subsubsection*{Mathematical analysis}
+The dynamics of the mutant population can be described by a two-type branching process.
+A key assumption of this framework is that \linelabel{R1-27}all mutants reproduce, disperse and die independently of each other.
+This is a reasonable assumption as long as the overall number of mutants is a lot smaller than the population size of the wild type.
+The two ``types'' in the name of method, two-type branching process, correspond to the two habitat types.
+The process tracks the \linelabel{R2-25}total number of mutants in old- and new-habitat patches at the end of each generation.
+As mentioned above, the numbers of successful offspring of mutant individuals, i.e., of offspring that survive density regulation and become adults at the next generation, are approximated by Poisson-distributed numbers, with means $(1+ a_{\text{old}})$ in old-habitat patches (eq.~\eqref{eq:sold} and eq.~\eqref{Seq:s_old}), and $1+ a_{\text{new}}$ in new-habitat patches (eq.~\eqref{eq:defanew}).
+After one generation, the mean number of successful offspring of a single mutant, either in an old- or a new-habitat patch, is then given by the following mean reproduction matrix:
 %
@@
 %
-where the rows denote the parent locations, and the columns the patch type of the offspring. 
-For example, the top-left entry reads as the probability for the parent to stay in an old-habitat patch ($1-m_m^{\text{new}}$) times the average number of offspring in these patches, given by the mean of the corresponding Poisson distribution with rate $(1+a_{\text{old}})$, cf.
-eq.~\eqref{eq:sold} \linelabel{R2-26}\chg{(see Section S1 in SI for the derivation of $a_{\text{old}}$ at the wild-type equilibrium)}.
+where the rows denote the parent locations, and the columns the patch type of the successful offspring. 
+For example, the top-left entry is the probability for the parent to stay in an old-habitat patch ($1-m_m^{\text{new}}$, see eq.\eqref{eq:dispersal}) times the average number of successful offspring in these patches \linelabel{R2-26}(see Section~\ref{sec:app:modeldyn} in SI for the derivation of $a_{\text{old}}$ at the wild-type equilibrium).
 The other entries are obtained analogously.
@@
 In the formula, these changes are accounted for by the sums that iterate through the (discrete) time steps and by the time dependence of the corresponding quantities.
-We further note that we follow the \chg{expected value of} the wild-type population size deterministically over time, instead of assuming it to be already in its steady state as in eq.~\eqref{eq:source_sink} \chg{(see also Section S1 in the SI)}.
+We further note that we follow the \chg{expected value of} the wild-type population size deterministically over time, instead of assuming it to be already in its steady state as in eq.~\eqref{eq:source_sink} \chg{(see also Section \ref{sec:app:modeldyn} in the SI)}.
 
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{amnatnat.bst}
 \bibliography{dispersal.bib}
 
+\clearpage
+
+\pagenumbering{arabic}% resets `page` counter to 1
+\renewcommand*{\thepage}{S\arabic{page}}
+
+%\titleformat{\section}[hang]{\Large\bfseries\filcenter}{\thesection}{1em}{}
+%%%%%% Putting an S before the equation, section and figure numbering
+\renewcommand\theequation{S\arabic{equation}}
+\renewcommand\thesection{S\arabic{section}}
+\renewcommand\thefigure{S\arabic{figure}}
+
+\resetlinenumber
+\renewcommand\thelinenumber{S\arabic{linenumber}}
+\linenumbers{}
+\modulolinenumbers[2]
+
+\appendix
+
+\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
+\part{Appendix} % Start the appendix part
+\parttoc % Insert the appendix TOC
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Model dynamics %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+
+\renewcommand{\theequation}{A\arabic{equation}}
+\setcounter{equation}{0}  % reset counter 
+% redefine the command that creates the equation number.
+\renewcommand{\thetable}{A\arabic{table}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+
+\section{Deriving the model dynamics \label{sec:app:modeldyn}}
+\linelabel{test}In this section we provide the mathematical \chg{details} of the model that is verbally described in the main text.
+We start by deriving the population dynamics of the wild type \chg{when alone} in both patch types.
+This will allow us to compute the local growth rate of the mutant \chg{when rare} in old-habitat patches, $a_{\text{old}}$.
+
+Before we go into the details of the computation, we recall the form of the dispersal rates. For a wild-type individual to disperse to a new-habitat patch, this probability is given by
+
+\begin{equation}\label{Seq:dispersal_rates_old}
+m_w^{\text{new}} = m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, ,
+\end{equation}
+%
+where $m$ denotes the emigration probability, $f_{\text{old}}$ the frequency of old-habitat patches and $\widehat{\pi}_w$ the transformed wild-type bias towards old-habitat patches.
+Analogously, the probability for a wild-type individual \chg{to emigrate and} to move to an old-habitat patch reads
+%
+\begin{equation}\label{Seq:dispersal_rate_new}
+m_w^{\text{old}} = m\, \frac{\widehat{\pi}_w f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, .
+\end{equation}
+
+All the subsequent computations can be checked with a symbolic programming language (e.g.
+\textit{Mathematica}).
+A \textit{Mathematica} notebook is deposited on Gitlab\footnote{\url{https://gitlab.com/pczuppon/evolutionary\_rescue\_and\_dispersal}}.
+
+\subsection*{Stationary wild-type population sizes}
+We derive the (deterministic) stationary population sizes of the wild-type in both patch types.
+These values are denoted by $\widehat{N}_w^{k}$, where $k$ corresponds to the patch type (old or new).
+In old-patch habitats we assume that the population is always at carrying capacity.
+Therefore, we have $\widehat{N}_w^{\text{old}} = K_{\text{old}}$.
+
+In new-habitat patches, the stationary value is given by the solution of the following equation:
+
+\begin{equation}\label{Seq:wt_deme2}
+\begin{aligned}
+\widehat{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)(1-r)\widehat{N}_w^{\text{new}} \\
+&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} (1-r)  \widehat{N}_w^{\text{old}}\\
+&= \left(1-m+m_w^{\text{new}} \right)(1-r)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \left(\frac{f_{\text{old}}}{1-f_{\text{old}}}\right) (1-r)  K_{\text{old}},
+\end{aligned}
+\end{equation}
+%
+which simplifies to
+%
+\begin{equation}
+\widehat{N}_w^{\text{new}} = \frac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\ .
+\end{equation}
+%
+Since we assume density regulation, this value cannot be larger than $K_{\text{new}}$, the carrying capacity of new-habitat patches. Hence, we find
+\begin{equation}\label{Seq:Nhatnew}
+\widehat{N}_w^{\text{new}} =\left\{ \begin{array}{ll} 
+K_{\text{new}}, & \text{if }\dfrac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\geq K_{\text{new}}; \\
+%
+\dfrac{m (1-r) f_{\text{old}} K_{\text{old}}
+}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}, & \text{otherwise}.
+\end{array}\right.
+\end{equation}
+
+\subsection*{Wild-type population sizes after dispersal}
+In order to explicitly compute the growth rate of the mutant in old-habitat patches, we need an analytical expression for the number of wild-type individuals after the dispersal step.
+Later on, in the approximation of the probability of adaptation, we also use the number of wild-type individuals in new-habitat patches before reproduction.
+We denote these values by $\widetilde{N}_w^k$, where $k$ corresponds to the habitat type (old or new).
+These are given as the solutions\florence{solution sounds like there is something to solve, which does not seem to be the case.
+Am I missing something?} to the following equations
+
+\begin{equation}
+\begin{aligned}
+\widetilde{N}_w^{\text{old}} &= \left(1-m + m\, \frac{\widehat{\pi}_w f_{\text{old}}}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})}\,  \frac{f_{\text{old}}M}{f_{\text{old}}M} \right) \widehat{N}_w^{\text{old}} + m_w^{\text{old}} \frac{(1-f_{\text{old}})M}{f_{\text{old}}M}\widehat{N}_w^{\text{new}}\ , \\
+\widetilde{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{1-f_{\text{old}}+\pi f_{\text{old}}} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \widehat{N}_w^{\text{old}}\ .
+\end{aligned}
+\end{equation}
+Using $\widehat{N}_w^{\text{old}}=K_{\text{old}}$ and in the case of $\widehat{N}_w^{\text{new}} = K_{\text{new}}$ this reduces to
+\begin{equation}\label{Seq:wt_eq1}
+\begin{aligned}
+\widetilde{N}_w^{\text{old}} &= \frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \\ 
+\widetilde{N}_w^{\text{new}}&= \frac{(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w (1-m))K_{\text{new}} + m f_{\text{old}} K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\ .
+\end{aligned}
+\end{equation}
+%
+If on the other hand $\widehat{N}_w^{\text{new}}<K_{\text{new}}$ holds, using eq.~\eqref{Seq:Nhatnew}\footnote{\florence{please check the eqref I wrote}}, we obtain\florence{separate into 2 eqs for more precise reference}
+\begin{equation}\label{Seq:wt_eq2}
+\begin{aligned}
+\widetilde{N}_w^{\text{old}} &= \left(1-m+m_w^{\text{old}} \right)K_{\text{old}}\\
+&\qquad \qquad + \left(m_w^{\text{old}} \frac{m (1-r) f_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\left(\frac{1-f_{\text{old}}}{f_{\text{old}}}\right)\right) K_{\text{old}} \\
+%&= \left(1-m_1+\frac{m_1 m_2 (1-r)}{(1-(1-m_2)(1-r))}\right) K \\ 
+&= \frac{m \widehat{\pi}_w f_{\text{old}} + r (1-m) (1-f_{\text{old}} (1-\widehat{\pi}_w))}{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} K_{\text{old}}\ \\
+&= \left(1 - \frac{r m (1-f_{\text{old}}) }{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} \right) K_{\text{old}}\, ,\\
+\widetilde{N}_w^{\text{new}} &= \left(1-m+m_w^{\text{new}}\right) \frac{m (1-r) f_{\text{old}} }{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\, K_{\text{old}}\\
+&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}}}{(1-f_{\text{old}})} K_{\text{old}}\\
+%&= \left(\frac{(1-m_2) m_1 (1-r) f }{(1-f)(1-\left(1-m_2 \right)(1-r))} + \frac{m_1 f}{(1-f)}\right) K\\
+&= \frac{m f_{\text{old}}}{m \widehat{\pi}_w f_{\text{old}} + r(1- f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}  (1-m)) }\, K_{\text{old}}\, .
+\end{aligned}
+\end{equation}
+
+\subsection*{Wild-type population sizes during the environmental change}
+\chg{Lastly, compute the (deterministic) wild-type population size over time during the environmental change.
+This value is used in the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue} in the main text, more precisely to estimate the number of rescue mutants that appear during the deterioration of patches.}
+
+\chg{If a patch deteriorates its current population size is given by the carrying capacity of the old habitat, $K_{\text{old}}$.
+Subsequently, it is reduced by $(1-r)$, the fecundity of a single wild-type individual in a new-habitat patch.
+That is, neglecting dispersal for the moment, in generation $j$ after the degradation of a patch, we would have
+	\begin{equation}
+	N_w^{\text{new}}(j) = K_{\text{old}}(1-r)^j
+	\end{equation}
+	wild-type individuals in this patch.
+Including dispersal between the patches then results in the following number of wild-type individuals in a patch $i$ at time $j$, given that there are $k-1$ other new-habitat patches
+	\begin{equation}
+	N_w^{i,k}(j) = (1-r)\left((1-m)N_w^{i,k}(j-1) + \frac{m_w^{\text{new}}}{k} (M-k) K_{\text{old}} + \frac{m_w^{\text{new}}}{k} \sum_{l=1}^k N_w^{l,k}(j-1) \right)\ .
+	\end{equation}
+	The first term represents the remaining individuals after emigration, $(1-m)$, the second and third term are immigrants from old- and new-habitat patches (distributed equally among the $k$ new-habitat patches), respectively.
+}
+
+\subsection*{The local growth rate $a_{\text{old}}$}
+As stated in eq.~\eqref{eq:sold} in the main text, we define the growth rate of a single mutant in the old habitat by
+\begin{equation}
+1 + a_{\text{old}} = K_{\text{old}}\, \dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}_w^{\text{old}}}\, .    
+\end{equation}
+%
+Having computed the number of wild-type individuals after dispersal in an old-habitat patch, $\widetilde{N}_w^{\text{old}}$, we can write the growth rate as follows
+
+\begin{equation}\label{Seq:s_old}
+a_{\text{old}} = \left\{ \begin{array}{ll}
+\dfrac{\omega^\text{old}_m K_{\text{old}}}{\omega^\text{old}_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
+\dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}.
+\end{array}
+\right.
+\end{equation}
+
+\subsection*{The local growth rate $a_{\text{new}}$}
+\chg{In new-habitat patches we assume that during the relevant phase of establishment of a rare mutant the carrying capacity $K_{\text{new}}$ is not reached.
+Then the growth rate of the mutant in these patches, $a_{\text{new}}$, is not affected by the different dispersal schemes.}
+
+\chg{Of course there may be parameter configurations, typically high emigration rates $m$ and a bias of the wild type towards the new habitat ($\pi_w < 0$), where our assumption of density-independent reproduction is violated.
+Then our approximation and the numerical solution of eq.~\eqref{eq:ext_prob} in the main text strongly deviate from the simulation results (e.g.
+the New-New dispersal scheme in Fig.~\ref{fig:vary_m_est}(d)).}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Establishment probability %%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\renewcommand{\theequation}{B\arabic{equation}}
+\setcounter{equation}{0}  % reset counter
+
+\section{Approximation of the establishment probability}
+We compute the survival probability \chg{of the lineage} of a single mutant starting either in an old- or in a new-habitat patch.
+We call this probability the establishment probability because it implies the successful establishment of a mutant population within the meta-population.
+It is denoted by $\varphi_k$ where the index $k$ indicates the initial habitat type of the mutant (old or new).
+
+Our method is the same as used in \citet{tomasini_2018} with the exception that our growth rate in old-habitat patches, $a_{\text{old}}$, depends on the demography of the population.
+The method in general is based on the theory of multi-type branching processes, cf.~Chapter~5.5 in \citet{haccou_book}.
+For a detailed application of the theory we refer to the Supplementary Information of \citet{tomasini_2018}.
+
+We start by recalling the mean reproduction matrix of a mutant that gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides.
+It is given as (see also eq.~\eqref{eq:mean_repro} in the main text)
+
+\begin{equation}\label{Seq:mean_repro}
+\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
+	\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
+	\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
+}\ ,
+\end{equation}
+where the rows denote the parent locations, and the columns the patch type of the offspring.
+
+Our goal is to apply Theorem 5.6 from \citet{haccou_book} which states that for a slightly super-critical branching process, \chg{i.e.
+where the survival probability is slightly above zero}, the establishment probability can be expressed in terms of the largest eigenvalue $\rho$ and the corresponding left- and right-eigenvectors of the mean reproduction matrix \chg{$\mathcal{M}$}, denoted by $u$ and $v$, respectively.
+The eigenvectors should be normalized in the following way: $u_1+u_2 = 1$ and $\sum_{i=1}^2 u_i v_i = 1$.
+The establishment probabilities are then given by
+\begin{equation}\label{Seq:theory}
+\varphi_i = \frac{2(\rho-1)}{B} v_i + O(\varepsilon), 
+\end{equation}
+with
+\begin{equation} 
+B = \sum_{i=1}^2 u_i \sum_{j=1}^2 v_j \mathcal{M}_{ij} + \rho(1-\rho) \sum_{j=1}^2 u_j v_j^2\, . 
+\end{equation}
+
+\subsection*{Computing the largest eigenvalue}
+We first approximate the largest eigenvalue of $\mathcal{M}$ denoted by $\rho$. It is given by (see \textit{Mathematica} notebook)
+\begin{equation}
+\begin{aligned}
+\rho &= \frac{1}{2}\left(2 + a_{\text{old}} + a_{\text{new}} - m -  m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}} \right.\\
+&\quad \left. + \sqrt{4 (m -1) (1+a_{\text{old}})(1+a_{\text{new}}) + (2+a_{\text{old}} + a_{\text{new}} - m - m_m^{\text{new}} a_{\text{old}} - m_m^{\text{old}} a_{\text{new}}})^2 \right)
+\end{aligned}
+\end{equation}
+%
+In order to make analytical progress and to identify under which conditions the process is slightly super-critical, i.e.
+$\rho>1$, we rescale the parameters by a small parameter $\varepsilon$.
+We set $a_{\text{old}} = \varepsilon, a_{\text{new}} = \varepsilon \xi$ and $m = \varepsilon\mu$.
+Assuming that $\varepsilon$ is small enough, i.e.
+effectively a weak selection assumption in old-habitat patches, we can neglect higher orders of $\varepsilon$ and find
+\begin{equation}\label{eq:eigenvalue}
+\begin{aligned}
+\rho &\approx 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{(\xi -1 + \mu_m^{\text{new}})^2 + 2(1-\xi+\mu_m^{\text{new}})\mu_m^{\text{old}} + (\mu_m^{\text{old}})^2  }\right)\\
+&= 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{\frac{\gamma}{1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}}  }\right),
+\end{aligned}
+\end{equation}
+where $\gamma$ is the rescaled version of the constant $C$ in the main text (eq.~\eqref{eq:normalization}), i.e.
+\begin{equation}
+\gamma = (1-f_{\text{old}})(\xi-1+\mu)^2 + \widehat{\pi}_m f_{\text{old}} (\xi-1-\mu)^2\, .
+\end{equation}
+%
+For $\varepsilon\to 0$ we find that $\rho\to 1$ \chg{(eq.~\eqref{eq:eigenvalue}) which means that the branching is slightly super-critical if $\rho>1$ and real.
+A sufficient condition for this to be true is}
+\begin{equation}
+1+\xi-\mu > 0 \quad \Leftrightarrow \quad a_{\text{old}} + a_{\text{new}} - m > 0
+\end{equation}
+%
+\chg{In case that the branching process is not super-critical the establishment probability in eq.~\eqref{Seq:theory} becomes negative and as such is not a probability anymore.
+Hence, we can simply reject negative solutions of the establishment probability and by that implicitly justify that our approximation is valid.}%\\
+
+%Inserting $\rho$ into equation~\eqref{eq:theory} we see the the enumerator is of order $\varepsilon$.
+
+\subsection*{Computing the establishment probability}
+For the solution of eq.~\eqref{Seq:theory} it remains to compute the normalized eigenvectors.
+Their precise form is of not much insight.
+We therefore omit stating them explicitly but refer to the \textit{Mathematica} notebook.
+Solving eq.~\eqref{Seq:theory} to the first order of $\varepsilon$ we find
+
+\begin{equation}
+\begin{aligned}
+\varphi_{\text{old}} &= \varepsilon + \frac{\varepsilon\left(1-\xi\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}  + \frac{\varepsilon\left(\mu_m^{\text{old}} -\mu_m^{\text{new}} + 2 \mu_m^{\text{new}}\xi  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}\, ,\\
+\varphi_{\text{new}} &= \varepsilon\xi + \frac{\varepsilon\xi \left(\xi-1\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}} + \frac{\varepsilon\left(\xi \mu_m^{\text{new}} -\xi \mu_m^{\text{old}} + 2 \mu_m^{\text{old}}  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}\, .
+\end{aligned}
+\end{equation}
+%
+Transforming back to the original variables and replacing $\gamma$ by the constant $C$ from the main text (eq.~\eqref{eq:normalization})
+\begin{equation}
+C = (1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + \widehat{\pi}_m f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, ,
+\end{equation}
+we obtain
+
+\begin{equation}
+\begin{aligned}
+\varphi_{\text{old}} &= a_{\text{old}} + \frac{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})a_{\text{old}} \left(a_{\text{old}}-a_{\text{new}}\right)}{\sqrt{C}}  + \\
+&\qquad \qquad \qquad m \frac{\left(\widehat{\pi}_m f_{\text{old}} a_{\text{old}} - (1-f_{\text{old}}) a_{\text{old}} + 2 (1-f_{\text{old}}) a_{\text{new}}  \right)}{\sqrt{C}}\, ,\\
+\varphi_{\text{new}} &= a_{\text{new}} + \frac{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})a_{\text{new}} \left(a_{\text{new}} - a_{\text{old}}\right)}{\sqrt{C}} \\
+&\qquad \qquad \qquad + m \frac{\left(a_{\text{new}} (1-f_{\text{old}}) - a_{\text{new}} \widehat{\pi}_m f_{\text{old}} + 2 a_{\text{old}} \widehat{\pi}_m f_{\text{old}}  \right)}{\sqrt{C}}\, .
+\end{aligned}
+\end{equation}
+%
+Slightly re-ordering the terms, this gives the establishment probability of a single mutant individual, eq.~\eqref{eq:estab_approx} in the main text:
+\begin{equation}\label{Seq:estab_approx}
+\begin{aligned}
+\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}} \qquad \quad \;  \; +  \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
+& \qquad \qquad + \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}\widehat{\pi}_m f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
+\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} +  \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
+& \qquad \quad \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) \widehat{\pi}_m f_{\text{old}} \right)\, .}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
+\end{aligned}
+\end{equation}
+
+\chg{For $m=0$ we see that $\varphi_{\text{old}}=0$, i.e.
+terms (1) and (2) cancel out.
+For the establishment probability in the new habitat we recover Haldane's result for the establishment probability of a slightly advantageous mutant: $\varphi_{\text{new}}=2a_{\text{new}}$ \citep{haldane_1927}.}
+
+\subsection{Disentangling the contributions to the establishment probability}
+\chg{We now proceed to explain} the three regions of the establishment probability from Fig.~\ref{fig:vary_m_est}(a) in the main text.
+These were defined by: (i) an initial increase of the establishment probability at low dispersal rates $m$; (ii) a local maximum with a subsequent decrease of the establishment probability; (iii) an increase of the establishment probability for high dispersal rates.
+
+\begin{figure}[t!]
+	\centering
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{fig2a.pdf}
+		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
+	\end{subfigure}%
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{fig2c.pdf}
+		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
+	\end{subfigure}
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS1c.pdf}
+		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
+	\end{subfigure}%
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS1d.pdf}
+		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
+	\end{subfigure}
+	\caption{\textbf{Contribution of the different terms in eq.~\eqref{Seq:estab_approx} to the establishment probability $\varphi_{\text{old}}$.} Subfigures (a,b) are the same as Figs.~\ref{fig:vary_m_est}(a,c) in the main text.
+They show the establishment probability for a single mutant individual arising in an old-habitat patch for varying emigration probabilities $m$.
+In subfigures (c,d) we plot the terms from eq.~\eqref{Seq:estab_approx} separately ($\pi_w=0.5,\pi_m=-0.5$).
+Term 1, the mutant growth rate in old habitats (solid), increases with increasing dispersal rates as a consequence of relaxed competition.
+Term 2, the environmental effect (dashed), captures the differences between the growth rates in the habitats.
+The larger the difference, the larger its contribution to the overall establishment probability.
+Term 3, the effect of dispersal (dotted), (largely) increases with increasing dispersal rates $m$.
+The sum of the three terms is plotted as a dash-dotted line.}
+	\label{Sfig:contribution}
+\end{figure}
+
+For clarity, we re-plot Figs.~\ref{fig:vary_m_est}(a,c) in Fig.~\ref{Sfig:contribution}(a,b), respectively.
+We try to explain the ongoing processes for each of the regions through the approximations of the establishment probabilities in eq.~\eqref{Seq:estab_approx}, see also Fig.~\ref{Sfig:contribution}(c,d).
+Note that these explanations are only valid for the establishment probability of a mutant initially in an old-habitat patch, $\varphi_{\text{old}}$.
+For the intuition behind the shape of $\varphi_{\text{new}}$ we refer to the corresponding section in the main text.
+
+Region (i) is explained by the positive effect of dispersal.
+Mutants disperse from old- to new-habitat patches where they \chg{have a higher growth rate}.
+This effect is mediated through the third term of the establishment probability in eq.~\eqref{Seq:estab_approx}.
+\chg{Note that the first two terms of the approximation cancel out for small emigration probabilities $m$.}
+While the third term increases with increasing emigration rate $m$, the second term in eq.~\eqref{Seq:estab_approx} decreases, cf. Fig.~\ref{Sfig:contribution}. 
+In the formula this is mediated through \chg{the increase of the local growth rate in the old-habitat, $a_{\text{old}}$, i.e.\ it becomes less negative.
+Then both factors of the second term, $a_{\text{old}}$ and the difference $(a_{\text{old}}-a_{\text{new}})$, increase (which in turn decreases term two).
+The }\chg{intuitive} reason is that due to larger emigration probabilities $m$, more individuals leave old habitats before the reproductive event.
+This relaxes competition in these patches and therefore increases the local growth rate of mutants in old-habitat patches.
+%
+%The intuition is that habitat differences become less important under stronger population mixing -- recall that the dispersal rate $m$ appears in the constant $C$ in the denominator.
+This alone would explain the decreasing curve of the second term.
+Additionally though, the local growth rate in old-habitat patches, $a_{\text{old}}$, is linked to the population demography which changes for varying $m$.
+With larger dispersal rates, more individuals leave the densely populated old habitats.
+This results in alleviated competition pressure for the remaining individuals, thus increasing the local growth rate $a_{\text{old}}$.
+This decreases the difference of growth rates $a_{\text{old}}-a_{\text{new}}$.
+Therefore, the local maximum can be explained by the decreasing environmental influence and the increasing dispersal effect, the second and the third term in eq.~\eqref{eq:estab_approx}, respectively.
+Hence, region (ii), beginning with the local maximum, is dominated by the decreasing effect of the environment on the establishment probability.
+Finally, in region (iii) dispersal is so large that the population homogenizes.
+This results in even less competitive pressure in old-habitat patches.
+Eventually, this yields a positive growth rate $a_{\text{old}}$ (first term in eq.~\eqref{Seq:estab_approx}).
+Therefore, this region is driven by the local growth rate in old habitats.
+
+Note that region (iii) can be shifted to the left by increasing the absolute number of offspring of mutants in old-habitat patches, $\omega^\text{old}_m$, and therefore decreasing the local disadvantage of the mutant.
+If shifted sufficiently to the left, like in Fig.~\ref{Sfig:contribution}(d), region (ii) might vanish due to this effect.
+
+\chg{Contrarily, if we set the fecundity parameter of the mutant in the old habitat to $\omega^\text{old}_m=1.1$, we see that region (iii) disappears for most dispersal schemes, cf.
+Fig.~\ref{Sfig:weak_fecundity}.
+Due to the low fecundity of the mutant, even under relaxed competition in old habitats, establishment of a mutant population is very unlikely.
+The final increase of our approximation in the dispersal schemes New-New and New-Old is due to our density-independence assumption in new habitats.
+For these large values of $m$ and the bias of the wild type towards new-habitat patches, this assumption is violated in the simulations explaining the deviation from the simulation results and our prediction.
+}
+
+\begin{figure}
+	\centering
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS2a.pdf}
+		\caption{$\varphi_{\text{old}}$}
+	\end{subfigure}%
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS2b.pdf}
+		\caption{$\varphi_{\text{new}}$}
+	\end{subfigure}
+	\caption{\textbf{Disappearance of region (iii) for large fecundity differences in the old habitat.} If the mutant fecundity in old-habitat patches is too low, here $\omega^\text{old}_m=1.1$, the effect of relaxed competition is not strong enough to have an impact on the establishment probability for high dispersal rates.
+The establishment probability remains at zero.}
+	\label{Sfig:weak_fecundity}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Native habitat type %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\renewcommand{\theequation}{C\arabic{equation}}
+\setcounter{equation}{0}  % reset counter
+
+\section{Habitat of origin of the adaptive mutation}
+\chg{Here, we provide further insight into the origin of the adaptive mutation.
+Therefore we plot the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ when varying the frequency of old habitats, $f_{\text{old}}$.
+In Fig.~\ref{fig:origin} in the main text we have seen that most successful adaptive mutations arise in old-habitat patches.
+Here, we show that this is explained mostly due to the large mutational input that is provided by the much larger wild-type population sizes in old habitats, see Fig.~\ref{Sfig:vary_f_origin}(b).
+The establishment probability though is always larger for mutants that arise in new habitats than in old habitats (Fig.~\ref{Sfig:vary_f_origin}(a)).
+}
+
+\begin{figure}[h!]
+	\centering
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS3a.pdf}
+		\caption{Establishment probabilities}
+	\end{subfigure}%
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS3b.pdf}
+		\caption{Stationary wild-type population size}
+	\end{subfigure}
+	\caption{\textbf{Establishment probability and stationary wild-type population size when varying the old-habitat frequency.} In the simulations we have used the standard set of parameters as given in Table~\ref{tab:parameters} and the unbiased dispersal scheme ($\pi_w=\pi_m=0$).
+In (a) we additionally chose the large fecundity difference scenario ($\omega^\text{old}_m=1.35$).}
+	\label{Sfig:vary_f_origin}
+\end{figure}
+
+%Additionally, we investigate the habitat type of the origin of the adaptive mutations for low fecundity values of the mutant in the old habitat.
+While for large fecundity values of the mutant in the old habitat most successful lineages arise in the old habitat (Fig.~\ref{fig:origin} in the main text), we see that for lower fecundity values $\omega^\text{old}_m$ the contribution of new-habitat mutants approaches and occasionally exceeds the number of successful mutant lineages that emerge from old-habitat patches.
+In the subsequent figure, besides the default parameters as stated in Table 1 in the main text, we chose $\omega^\text{old}_m=1.05$ and $\theta = 10/(MK)$.
+The plot corresponding to Fig.~\ref{fig:origin} in the main text then looks as follows:
+
+%\begin{figure}[h!]
+%	\centering
+%	\includegraphics[width=0.6\textwidth]{figS4.pdf}
+%  	\caption{\textbf{Habitat type of the origin for strong fecundity differences in the old habitat ($\omega^\text{old}_m=1.05$).}}
+%	\label{Sfig:origin}
+%\end{figure}
+\pete{There was a figure where the contribution from new habitats was larger than from old habitats.
+With the changed fecundities, I only get this behavior if the fecundity of the mutant is larger in the new patches than in the old patches, i.e.
+$\omega^\text{old}_m < a_{\text{new}}$.
+Also, for some reason the approximations do not work that well in these scenarios (I did not investigate why the estimated lines are bad -- soft sweeps still work fine but the other approximations don't.)}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Large emigration rate %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\renewcommand{\theequation}{D\arabic{equation}}
+\setcounter{equation}{0}  % reset counter
+
+\section{Probability of establishment for large frequencies of old-habitat patches}
+\chg{We plot the probability of establishment for a large frequency of old-habitat patches, $f_{\text{old}}$.
+As visible in Fig.~\ref{Sfig:vary_f} below, for high frequencies of old-habitat patches the probability of adaptation becomes very small, if not zero, for large emigration probabilities $m$.
+These high frequency of old-habitat patches are the patch configurations which mutants that were present before the first patch degrades, i.e.
+standing genetic variation mutants, experience.
+Therefore, it is very unlikely that the mutant, which will eventually rescue the population, was already present before the first degradation event.} This supports the explanation that homogenizing the population (large $m$) reduces the impact of standing genetic variation on the probability of evolutionary rescue, see Fig.~\ref{fig:sgv}(b) in the main text where $\pi_w=\pi_m=0$ was plotted.
+
+\begin{figure}[h!]
+	\centering
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS5a.pdf}
+		\caption{$\varphi_{\text{old}}$}
+	\end{subfigure}%
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS5b.pdf}
+		\caption{$\varphi_{\text{new}}$}
+	\end{subfigure}
+	\caption{\textbf{Probability of adaptation for a large frequency of old-habitat patches ($f_{\text{old}}=0.9$).} The fecundity of the mutant in old-habitat patches is set to $\omega^\text{old}_m=1.45$.
+Note also the difference between the scales of the y-axes in the two panels.}
+	\label{Sfig:vary_f}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Comparison to pop gen %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\renewcommand{\theequation}{E\arabic{equation}}
+\setcounter{equation}{0}  % reset counter
+
+\section{Establishment probability in a model without demography}
+Here, we consider a variation of our original model in order to investigate the impact of demography on the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$.
+The dispersal process and the dynamics in old-habitat patches remain as studied before.
+In new-habitat patches we now assume that the population remains at carrying capacity, i.e.
+there is no longer a declining wild-type population.
+This means that the local growth rate $a_{\text{old}}$ in eq.~\eqref{Seq:s_old} takes the form for $\widehat{N}_w^{\text{new}}=K_{\text{new}}$.
+For simplicity we will also assume that $K_{\text{new}}=K_{\text{old}}=K$.
+In order to maintain the divergent selection assumption, we assume that the fecundity of the wild-type in the new habitat is below the fecundity of the mutant.
+%For simplicity, we assume that the fecundity values are given as
+%\begin{equation}
+%    \omega^\text{old}_w^{\text{old}} = \omega^\text{old}_m^{\text{new}} \quad \text{and} \quad \omega^\text{old}_w^{\text{new}} = \omega^\text{old}_m^{\text{old}}.
+%\end{equation}
+Therefore, we need to adjust the local growth rate of a single mutant, $a_{\text{new}}$. The local growth rate in old habitats, $a_{\text{old}}$, remains as outlined in eq.~\eqref{Seq:s_old}. We have
+\begin{equation}
+1+a_{\text{new}} = K \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}} \widetilde{N}_w^{\text{new}}+ \omega_m^{\text{new}}}\approx K \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}} \widetilde{N}_w^{\text{new}}},
+\end{equation}
+which with the help of eq.~\eqref{Seq:wt_eq1} yields
+\begin{equation}
+a_{\text{new}} \approx \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}}\left(1+\frac{m f_{\text{old}} (1-\widehat{\pi}_w)}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\right)}.
+\end{equation}
+
+\chg{Note, that we again used that during the establishment phase the wild type is much more abundant than the mutant which explains the approximation in the two equations.}
+Plugging this in the approximation of the establishment probability from eq.~\eqref{Seq:estab_approx} we find the curves in Fig.~\ref{Sfig:pop_gen}.
+
+We see that, as briefly mentioned in the main text, region (iii) of the establishment probability disappears in these type of models except for the Old-New and the New-Old dispersal schemes.
+The reason for the disappearance of the region is that relaxed competition only plays a subordinate role for the symmetric dispersal schemes (Old-Old, New-New and 0-0).
+In other words, these dispersal schemes maintain the local frequencies of the mutant at the same level as before the dispersal step and by that do not change the population dynamics.
+In contrast, the Old-New dispersal scheme strongly increases the frequency of mutants in new-habitat patches and by that increases the establishment probability.
+It is worth mentioning though, that this is not an effect of relaxed competition but rather a biased dispersal of the mutant into the habitat where it is favored.
+Solely for the New-Old dispersal scheme where individuals prefer the habitat they are relatively less fit, we see a relaxed competition scenario in old-habitat patches.
+For high dispersal rates $m$ the majority of the wild-type individuals emigrate to new habitats and by that leave space for the rare mutants in old-habitat patches to reproduce.
+This explains why this dispersal scheme still shows the characteristic pattern of a three-stage establishment probability.
+\newpage
+
+\begin{figure}[h!]
+	\centering
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS6a.pdf}
+		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m=1.35$ (large fecundity difference)}
+	\end{subfigure}%
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS6b.pdf}
+		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m=1.35$}
+	\end{subfigure}
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS6c.pdf}
+		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m=1.45$ (small fecundity difference)}
+	\end{subfigure}%
+	\begin{subfigure}{.5\textwidth}
+		\centering
+		\includegraphics[width=\linewidth]{figS6d.pdf}
+		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m=1.45$}
+	\end{subfigure}
+	\caption{\textbf{Establishment probability when populations in both habitats are at carrying capacity.} We plot the establishment probability for a single mutant either initially in an old-habitat patch (a,c) or in a new-habitat patch (b,d).
+The numerical solution (solid lines) still approximates the simulated data reasonably well.
+The analytical approximation (dashed lines) however deviates strongly from the data due to large growth rates ($a_{\text{new}}\approx 0.2$) so that the conditions for the approximation to hold are violated.
+In this case, in eq.~\eqref{Seq:estab_approx} higher order corrections would need to be taken into account.
+\chg{The fecundity values in the new habitat are given by $\omega_w^{\text{new}} = \omega_m^{\text{old}}$ and $\omega_m^{\text{new}} = \omega_w^{\text{old}}$ and the carrying capacity is $K_{\text{new}}=K_{\text{old}}=K=500$.} Missing data points (mostly for the negative density-dependent dispersal scheme -- green triangles) are explained by too large computation times.
+All data points are averages from $10^4$ independent runs.
+Note the varying y-axes scales.}
+	\label{Sfig:pop_gen}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Origin dispersal schemes %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+\renewcommand{\theequation}{F\arabic{equation}}
+\setcounter{equation}{0}  % reset counter 
+\section{Habitat of origin dependent on the dispersal scheme}
+The habitat type of the origin of the rescue mutation is largely independent of the considered dispersal scheme.
+For $\omega^\text{old}_m=1.35$ we have plotted the relative contribution of each natal habitat type to the probability of evolutionary rescue, Fig.~\ref{Sfig:natal_habitat}.
+We do not see large differences between the three symmetric dispersal schemes (0-0, Old-Old, and New-New) and the Old-New dispersal scheme.
+\chg{Only for the New-Old dispersal schemes we see an increase in the contribution of mutants emerging from old-habitat patches for very high emigration probabilities $m$.
+A possible explanation is the large probability of establishment for a mutant emerging in old-habitat patches for this parameter set (cf.
+Fig.~\ref{Sfig:contribution}(a)).}
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.6\textwidth]{figS7.pdf}
+	\caption{\textbf{Habitat type of the origin of the rescue mutant dependent on the dispersal scheme.} Varying the emigration probability $m$ we plot the relative contributions of each habitat type to the probability of evolutionary rescue.
+The color-coding is as in the main text: black for 0-0, blue for the Old-Old, green for the New-New, orange for the Old-New, and purple for the New-Old dispersal scheme.}
+	\label{Sfig:natal_habitat}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Bibliography %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\newpage
+%\bibliographystyle{plainnat}
+%\bibliography{dispersal.bib}
+
 \end{document}

### 2020-07-05 ms/main.tex
intermediate backup because the computer is heating up

@@
 \newcommand{\chg}[1]{\textcolor{change}{#1}}
+
+%%% Fix for lineno and align
+\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
+	\expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
+	\expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
+	\renewenvironment{#1}%
+	{\linenomath\csname old#1\endcsname}%
+	{\csname oldend#1\endcsname\endlinenomath}}% 
+\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
+	\patchAmsMathEnvironmentForLineno{#1}%
+	\patchAmsMathEnvironmentForLineno{#1*}}%
+\AtBeginDocument{%
+	\patchBothAmsMathEnvironmentsForLineno{equation}%
+	\patchBothAmsMathEnvironmentsForLineno{align}%
+	\patchBothAmsMathEnvironmentsForLineno{flalign}%
+	\patchBothAmsMathEnvironmentsForLineno{alignat}%
+	\patchBothAmsMathEnvironmentsForLineno{gather}%
+	\patchBothAmsMathEnvironmentsForLineno{multline}%
+}
+%%%
 
@@
 
-The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is
+The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is\florence{TODO change into simple bias (remove m)}
 %
@@
 \end{equation}
+\florence{there seems to be a problem with the second equation, should be $m -$ no? }
 where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches.
@@
 
-Our calculations assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents.
-We also consider that mutants are rare compared to wild-type individuals in old-habitat patches.
+A key assumption to our mathematical analysis is that the mutant individuals are rare enough that their dynamics do not affect the wild-type population, at least during the establishment phase of the mutant lineage.
+The wild-type population sets a demographic context that affects mutants dynamics.
+The mathematical analysis therefore focuses on the population dynamics of the mutant population, considering the wild-type population as constant over time.\footnote{\florence{please check this is correct}}
+
+We assume that the subpopulations in old-habitat patches are always at carrying capacity, i.e.\ that there are always enough offspring that are produced to at least replace all the parents.
 Denoting by $\widetilde{N}_{i}^{k}$ the number of type-$i$ individuals in a $k$-habitat patch right after dispersal, then the expected number of successful offspring of mutant individuals in this old-habitat patch (i.e., of offspring that survive density regulation and become adults in the next generation) is
@@
 %
-= \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
+\overset{\mathrm{def}}{=} \left(1 + a_{\text{old}}\right) \widetilde{N}^{\text{old}}_m \, .
 \end{equation}
@@
 Eq.~\eqref{eq:sold} defines the per-capita expected growth rate of mutants in old-habitat patches, $a_{\text{old}}$.
-For the analysis, we also further assume that the number of successful offspring of mutant individuals is Poisson distributed, with mean $1+ a_{\text{old}}$.
-
-\hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal.
+It depends on $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal, which is calculated in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI) (Eq.~\eqref{Seq:Ntildeold}).
+
+\st{For the analysis, we also further assume that the number of successful offspring of mutant individuals is Poisson distributed, with mean $1+ a_{\text{old}}$. }\florence{com eback to this}
+
+%\hl{The sign of this local growth rate depends on the value of $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal.
 Mutants are less fecund that wild-type individuals in old-habitat patches ($\omega^\text{old}_m < \omega^\text{old}_w$), yet the number of mutants can increase in old-habitat patches if $\widetilde{N}^{\text{old}}_w$ is low enough.
@@
 }\footnote{\florence{I think this is more Results than Model}}.
-\linelabel{R1-14}The population dynamics of the wild type are analyzed explicitly in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI); formulas for $a_{\text{old}}$ are given in eq.~\ref{Seq:s_old}.
+\linelabel{R1-14}The population dynamics of the wild type are analyzed explicitly in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI).
 
@@
 
-\linelabel{R1-24}The survival probability of this multi-type branching process, $\varphi_{k}$ with $k$ indicating the initial habitat type of the mutant, is then given by the unique positive solution of the following system of equations \citep[see][Chapters 5.3 and 5.6]{haccou_book}
+\linelabel{R1-24}We denote by $\varphi_{\text{old}}$ (resp.
+$\varphi_{\text{new}}$) the probability of establishment of this two-type branching process when the mutant is initially located in a old- (resp.
+new-) habitat patch.
+This probability is computed by considering all possible ways of going extinct: the initial individual having $j$ successful offspring in a patch of type $k$, but all lineages descending from these $j$ offspring eventually go extinct; then summing over $k$ and $j$.
+The establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ are then given by the unique positive solution of the following system of equations \citep[see][Chapters 5.3 and 5.6]{haccou_book}:
+%
 \begin{equation}\label{eq:ext_prob}
 	\begin{aligned}
-		1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \\
-		&\qquad \qquad \qquad \left.+ \mathbf{P}(j \text{ offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\\
+		1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ successful offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \\
+		&\qquad \qquad \qquad \left.+ \mathbf{P}(j \text{successful offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\\
 		&=\exp\left[-\left(1-m^{\text{old}\to \text{new}}_m\right)(1+a_{\text{old}})\varphi_{\text{old}} - m^{\text{old}\to \text{new}}_m (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, ,   \\
@@
 \end{equation}
-Intuitively the left hand side, the extinction probability, equals the sum over all possible scenarios of the trajectories towards extinction, e.g.
-the initial individual having $j$ offspring in a patch of type $k$, old or new, and all of these $j$ offspring becoming extinct ($(1-\varphi_{\text{k}})^j$).
-Unfortunately, a general analytical solution to these equations is not accessible, but they can be solved numerically.
-For weak selection and (potentially) weak dispersal -- i.e.
-$a_{\text{old}},a_{\text{new}},m\ll 1$ needs to hold for at least two of the three parameters -- an approximation is available: see for example~\citet[Theorem~5.6]{haccou_book} for the general theory and \citet{tomasini_2018} for an application in a similar setting.
-The detailed derivation is presented in the SI, Section~S2.
+This system of equations can be solved numerically.
+An analytical approximate solution is available in the case of weak selection and (potentially) weak dispersal (i.e.
+$a_{\text{old}},a_{\text{new}},m\ll 1$ needs to hold for at least two of the three parameters); see for example~\citet[Theorem~5.6]{haccou_book} for the general theory and \citet{tomasini_2018} for an application in a similar setting.
+The detailed derivation is presented in the SI, Section~\ref{sec:approxestabproba}.
 We find
@@
 \section{Deriving the model dynamics \label{sec:app:modeldyn}}
-\linelabel{test}In this section we provide the mathematical \chg{details} of the model that is verbally described in the main text.
-We start by deriving the population dynamics of the wild type \chg{when alone} in both patch types.
-This will allow us to compute the local growth rate of the mutant \chg{when rare} in old-habitat patches, $a_{\text{old}}$.
-
-Before we go into the details of the computation, we recall the form of the dispersal rates. For a wild-type individual to disperse to a new-habitat patch, this probability is given by
+\linelabel{test}In this section we provide the mathematical details of the model that is verbally described in the main text.
+We start by deriving the population dynamics when only the wild type is present.
+This will allow us to compute the local growth rate of a rare mutant.
+
+Before we go into the details of the computation, we recall the form of the dispersal rates. A dispersing wild-type individual immigrates to a new-habitat patch with probability
 
 \begin{equation}\label{Seq:dispersal_rates_old}
-m_w^{\text{new}} = m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, ,
-\end{equation}
-%
-where $m$ denotes the emigration probability, $f_{\text{old}}$ the frequency of old-habitat patches and $\widehat{\pi}_w$ the transformed wild-type bias towards old-habitat patches.
-Analogously, the probability for a wild-type individual \chg{to emigrate and} to move to an old-habitat patch reads
-%
-\begin{equation}\label{Seq:dispersal_rate_new}
-m_w^{\text{old}} = m\, \frac{\widehat{\pi}_w f_{\text{old}}}{1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}}\, .
-\end{equation}
+\mu_w^{\text{new}} = \frac{1-f_{\text{old}}}{1-f_{\text{old}} + e^{\pi_w} f_{\text{old}}} = 1 - \mu_w^{\text{old}} \, ,
+\end{equation}
+%
+where $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_w$ is the wild-type bias towards old-habitat patches.
+The complement, $\mu_w^{\text{old}}$, is the probability that the dispersing wild-type individual instead immigrates into an old-habitat patch.
 
@@
 \subsection*{Stationary wild-type population sizes}
-We derive the (deterministic) stationary population sizes of the wild-type in both patch types.
-These values are denoted by $\widehat{N}_w^{k}$, where $k$ corresponds to the patch type (old or new).
-In old-patch habitats we assume that the population is always at carrying capacity.
-Therefore, we have $\widehat{N}_w^{\text{old}} = K_{\text{old}}$.
-
-In new-habitat patches, the stationary value is given by the solution of the following equation:
-
+We denote by $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ the deterministic stationary population sizes of the wild type in old- and new-habitat patches.
+We assume that the population is always at carrying capacity in old-habitat patches, so that $\widehat{N}_w^{\text{old}} = K_{\text{old}}$.
+We now compute $\widehat{N}_w^{\text{new}}$ recursively: it is given by the solution of the following equation:
+%
 \begin{equation}\label{Seq:wt_deme2}
-\begin{aligned}
-\widehat{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)(1-r)\widehat{N}_w^{\text{new}} \\
-&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} (1-r)  \widehat{N}_w^{\text{old}}\\
-&= \left(1-m+m_w^{\text{new}} \right)(1-r)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \left(\frac{f_{\text{old}}}{1-f_{\text{old}}}\right) (1-r)  K_{\text{old}},
-\end{aligned}
-\end{equation}
-%
-which simplifies to
+\widehat{N}_w^{\text{new}} = \left(1-m + m \mu_w^{\text{new}}  \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)  \omega_w^{\text{new}}  \widehat{N}_w^{\text{new}}   + m \mu_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \omega_w^{\text{new}}  \widehat{N}_w^{\text{old}}, \nonumber
+\end{equation}
+where the first term on the right-hand side corresponds to individuals born in a new-habitat patch and staying in it or migrating and landing in a new-habitat patch, and the second term corresponds to individuals born in an old-habitat patch and migrating to a new-habitat patch.
+Simplifying, using eq.~\eqref{Seq:dispersal_rates_old} to replace $\mu_w^{\text{new}} $ and replacing $\widehat{N}_w^{\text{old}}$ by $K_{\text{old}}$ , we obtain
 %
 \begin{equation}
-\widehat{N}_w^{\text{new}} = \frac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\ .
-\end{equation}
-%
-Since we assume density regulation, this value cannot be larger than $K_{\text{new}}$, the carrying capacity of new-habitat patches. Hence, we find
+\widehat{N}_w^{\text{new}} = \frac{m \omega_w^{\text{new}} f_{\text{old}}  K_{\text{old}}}{
+1-f_{\text{old}} + e^{\pi_w} f_{\text{old}} - \omega_w^{\text{new}} (1-f_{\text{old}} + e^{\pi_w} f_{\text{old}} (1-m))
+}
+\ .
+\end{equation}
+%
+This value cannot be larger than $K_{\text{new}}$, the carrying capacity of new-habitat patches. So, 
 \begin{equation}\label{Seq:Nhatnew}
-\widehat{N}_w^{\text{new}} =\left\{ \begin{array}{ll} 
-K_{\text{new}}, & \text{if }\dfrac{m (1-r) f_{\text{old}}  K_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\geq K_{\text{new}}; \\
-%
-\dfrac{m (1-r) f_{\text{old}} K_{\text{old}}
-}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}, & \text{otherwise}.
-\end{array}\right.
+\widehat{N}_w^{\text{new}} = \min\left(K_{\text{new}},  \frac{m \omega_w^{\text{new}} f_{\text{old}}  K_{\text{old}}}{
+	1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} - \omega_w^{\text{new}} (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))
+} \right).
 \end{equation}
@@
 \subsection*{Wild-type population sizes after dispersal}
-In order to explicitly compute the growth rate of the mutant in old-habitat patches, we need an analytical expression for the number of wild-type individuals after the dispersal step.
-Later on, in the approximation of the probability of adaptation, we also use the number of wild-type individuals in new-habitat patches before reproduction.
-We denote these values by $\widetilde{N}_w^k$, where $k$ corresponds to the habitat type (old or new).
-These are given as the solutions\florence{solution sounds like there is something to solve, which does not seem to be the case.
-Am I missing something?} to the following equations
-
+
+We denote by $\widetilde{N}_w^{\text{old}}$ and $\widetilde{N}_w^{\text{new}}$ the numbers of wild-type individuals \textit{after the dispersal step}.
+These quantities are needed to explicitly compute the growth rate of the mutant in old-habitat patches, and to approximate the probability of adaptation.
+They are given by
+%
+\begin{subequations}\label{Seq:Ntilde}
+\begin{align}
+\widetilde{N}_w^{\text{old}} &= \left(1-m + m\, \mu_w^{\text{old}} \,  \frac{f_{\text{old}}M}{f_{\text{old}}M} \right) \widehat{N}_w^{\text{old}} + m \mu_w^{\text{old}} \frac{(1-f_{\text{old}})M}{f_{\text{old}}M}\widehat{N}_w^{\text{new}}\ , \label{Seq:Ntildeold}\\
+%
+%
+\widetilde{N}_w^{\text{new}} &= \left(1-m + m \mu_w^{\text{new}} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)\widehat{N}_w^{\text{new}} + m \mu_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \widehat{N}_w^{\text{old}}\ .\label{Seq:Ntildenew}
+\end{align}
+\end{subequations}
+%
+We then replace $\widehat{N}_w^{\text{old}}$ by $K_{\text{old}}$ (since old-habitat patches are assumed to be at carrying capacity after density regulation), and $\widehat{N}_w^{\text{new}}$ by the formula given in eq.~\eqref{Seq:Nhatnew}.
+
+\subsection*{Wild-type population sizes during the environmental change}
+Lastly, compute the (deterministic) wild-type population size over time during the environmental change.
+This value is used in the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue} in the main text, more precisely to estimate the number of rescue mutants that appear during the deterioration of patches.
+
+At the moment a patch deteriorate, its population size is still given by the carrying capacity of the old habitat, $K_{\text{old}}$, but only each adult has now on average $\omega_{m}^{\text{new}}<1$ offspring.
+As the size of the local population decreases, there is not need for density regulation anymore.
+Neglecting dispersal for the moment, in generation $\tau$ after the degradation of a patch\footnote{\florence{is this just said/shown for the sake of pedagogy, or do you actually need this equation?
+If not, I suggest removing it.}} , we would have
 	\begin{equation}
-\begin{aligned}
-\widetilde{N}_w^{\text{old}} &= \left(1-m + m\, \frac{\widehat{\pi}_w f_{\text{old}}}{(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}})}\,  \frac{f_{\text{old}}M}{f_{\text{old}}M} \right) \widehat{N}_w^{\text{old}} + m_w^{\text{old}} \frac{(1-f_{\text{old}})M}{f_{\text{old}}M}\widehat{N}_w^{\text{new}}\ , \\
-\widetilde{N}_w^{\text{new}} &= \left(1-m + m\frac{(1-f_{\text{old}})}{1-f_{\text{old}}+\pi f_{\text{old}}} \frac{(1-f_{\text{old}})M}{(1-f_{\text{old}})M}\right)\widehat{N}_w^{\text{new}} + m_w^{\text{new}} \frac{f_{\text{old}} M}{(1-f_{\text{old}})M} \widehat{N}_w^{\text{old}}\ .
-\end{aligned}
-\end{equation}
-Using $\widehat{N}_w^{\text{old}}=K_{\text{old}}$ and in the case of $\widehat{N}_w^{\text{new}} = K_{\text{new}}$ this reduces to
-\begin{equation}\label{Seq:wt_eq1}
-\begin{aligned}
-\widetilde{N}_w^{\text{old}} &= \frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \\ 
-\widetilde{N}_w^{\text{new}}&= \frac{(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w (1-m))K_{\text{new}} + m f_{\text{old}} K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\ .
-\end{aligned}
-\end{equation}
-%
-If on the other hand $\widehat{N}_w^{\text{new}}<K_{\text{new}}$ holds, using eq.~\eqref{Seq:Nhatnew}\footnote{\florence{please check the eqref I wrote}}, we obtain\florence{separate into 2 eqs for more precise reference}
-\begin{equation}\label{Seq:wt_eq2}
-\begin{aligned}
-\widetilde{N}_w^{\text{old}} &= \left(1-m+m_w^{\text{old}} \right)K_{\text{old}}\\
-&\qquad \qquad + \left(m_w^{\text{old}} \frac{m (1-r) f_{\text{old}}}{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\left(\frac{1-f_{\text{old}}}{f_{\text{old}}}\right)\right) K_{\text{old}} \\
-%&= \left(1-m_1+\frac{m_1 m_2 (1-r)}{(1-(1-m_2)(1-r))}\right) K \\ 
-&= \frac{m \widehat{\pi}_w f_{\text{old}} + r (1-m) (1-f_{\text{old}} (1-\widehat{\pi}_w))}{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} K_{\text{old}}\ \\
-&= \left(1 - \frac{r m (1-f_{\text{old}}) }{m \widehat{\pi}_w f_{\text{old}} + r (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))} \right) K_{\text{old}}\, ,\\
-\widetilde{N}_w^{\text{new}} &= \left(1-m+m_w^{\text{new}}\right) \frac{m (1-r) f_{\text{old}} }{m f_{\text{old}} \widehat{\pi}_w + r(1-f_{\text{old}}+f_{\text{old}}\widehat{\pi}_w -m \widehat{\pi}_w f_{\text{old}})}\, K_{\text{old}}\\
-&\qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad \qquad + m_w^{\text{new}} \frac{f_{\text{old}}}{(1-f_{\text{old}})} K_{\text{old}}\\
-%&= \left(\frac{(1-m_2) m_1 (1-r) f }{(1-f)(1-\left(1-m_2 \right)(1-r))} + \frac{m_1 f}{(1-f)}\right) K\\
-&= \frac{m f_{\text{old}}}{m \widehat{\pi}_w f_{\text{old}} + r(1- f_{\text{old}} + \widehat{\pi}_w f_{\text{old}}  (1-m)) }\, K_{\text{old}}\, .
-\end{aligned}
-\end{equation}
-
-\subsection*{Wild-type population sizes during the environmental change}
-\chg{Lastly, compute the (deterministic) wild-type population size over time during the environmental change.
-This value is used in the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue} in the main text, more precisely to estimate the number of rescue mutants that appear during the deterioration of patches.}
-
-\chg{If a patch deteriorates its current population size is given by the carrying capacity of the old habitat, $K_{\text{old}}$.
-Subsequently, it is reduced by $(1-r)$, the fecundity of a single wild-type individual in a new-habitat patch.
-That is, neglecting dispersal for the moment, in generation $j$ after the degradation of a patch, we would have
+	N_w^{\text{new}}(\tau) = K_{\text{old}}(\omega_w^{\text{new}})^{\tau}
+	\end{equation}
+	wild-type individuals in this patch.
+Including dispersal between the patches then results in the following number of wild-type individuals in a patch $i$ at time $\tau$ post-degradation, given that there are $k-1$ other new-habitat patches
 	\begin{equation}
-	N_w^{\text{new}}(j) = K_{\text{old}}(1-r)^j
-	\end{equation}
-	wild-type individuals in this patch.
-Including dispersal between the patches then results in the following number of wild-type individuals in a patch $i$ at time $j$, given that there are $k-1$ other new-habitat patches
+	N_w^{i,k}(\tau) = \omega_w^{\text{new}}\left((1-m)N_w^{i,k}(\tau-1) + \frac{m_w^{\text{new}}}{k} (M-k) K_{\text{old}} + \frac{m_w^{\text{new}}}{k} \sum_{l=1}^k N_w^{l,k}(\tau-1) \right)\ .
+	\end{equation}
+	The first term represents the remaining individuals after emigration, the second and third term are immigrants from old- and new-habitat patches (distributed equally among the $k$ new-habitat patches), respectively.
+
+\subsection*{The local per capita growth rate $a_{\text{old}}$}
+As stated in eq.~\eqref{eq:sold} in the main text, we define the per capita growth rate of rare mutant in the old habitat by
 \begin{equation}
-	N_w^{i,k}(j) = (1-r)\left((1-m)N_w^{i,k}(j-1) + \frac{m_w^{\text{new}}}{k} (M-k) K_{\text{old}} + \frac{m_w^{\text{new}}}{k} \sum_{l=1}^k N_w^{l,k}(j-1) \right)\ .
-	\end{equation}
-	The first term represents the remaining individuals after emigration, $(1-m)$, the second and third term are immigrants from old- and new-habitat patches (distributed equally among the $k$ new-habitat patches), respectively.
-}
-
-\subsection*{The local growth rate $a_{\text{old}}$}
-As stated in eq.~\eqref{eq:sold} in the main text, we define the growth rate of a single mutant in the old habitat by
+1 + a_{\text{old}} = K_{\text{old}}\, \dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}_w^{\text{old}}}\, ,    
+\end{equation}
+%
+where we replace $\widetilde{N}_w^{\text{old}}$ by the formula given in \eqref{Seq:Ntildeold}. 
+%Having computed the number of wild-type individuals after dispersal in an old-habitat patch, $\widetilde{N}_w^{\text{old}}$, we can write the growth rate as follows \florence{in which case would we have $\widehat{N}_w^{\text{new}} = K_{\text{new}}$ ?}
+
+%\begin{equation}\label{Seq:s_old}
+%a_{\text{old}} = \left\{ \begin{array}{ll}
+%\dfrac{\omega^\text{old}_m K_{\text{old}}}{\omega^\text{old}_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
+%\dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}.
+%\end{array}
+%\right.
+%\end{equation}
+
+\subsection*{The local per capita growth rate $a_{\text{new}}$}
+In new-habitat patches, we assume that the carrying capacity $K_{\text{new}}$ is not reached during the establishment phase of the mutant.
+In other words, we assume that there is no density regulation in new-habitat patches.
+As a result, the local per capita growth rate of the mutant only depends on its fecundity:
+%
 \begin{equation}
-1 + a_{\text{old}} = K_{\text{old}}\, \dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}_w^{\text{old}}}\, .    
-\end{equation}
-%
-Having computed the number of wild-type individuals after dispersal in an old-habitat patch, $\widetilde{N}_w^{\text{old}}$, we can write the growth rate as follows
-
-\begin{equation}\label{Seq:s_old}
-a_{\text{old}} = \left\{ \begin{array}{ll}
-\dfrac{\omega^\text{old}_m K_{\text{old}}}{\omega^\text{old}_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
-\dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}.
-\end{array}
-\right.
-\end{equation}
-
-\subsection*{The local growth rate $a_{\text{new}}$}
-\chg{In new-habitat patches we assume that during the relevant phase of establishment of a rare mutant the carrying capacity $K_{\text{new}}$ is not reached.
+1 + a_{\text{new}} = \omega_m^{\text{new}}.
+\end{equation}
+
+%\chg{In new-habitat patches we assume that during the relevant phase of establishment of a rare mutant the carrying capacity $K_{\text{new}}$ is not reached\footnote{\florence{This seems at odds with the commented out sentences just above}}.
 Then the growth rate of the mutant in these patches, $a_{\text{new}}$, is not affected by the different dispersal schemes.}
 
-\chg{Of course there may be parameter configurations, typically high emigration rates $m$ and a bias of the wild type towards the new habitat ($\pi_w < 0$), where our assumption of density-independent reproduction is violated.
+Of course, there may be parameter configurations, typically high emigration rates $m$ and a bias of the wild type towards the new habitat ($\pi_w < 0$), where our assumption of density-independent reproduction is violated.
 Then our approximation and the numerical solution of eq.~\eqref{eq:ext_prob} in the main text strongly deviate from the simulation results (e.g.
-the New-New dispersal scheme in Fig.~\ref{fig:vary_m_est}(d)).}
+the New-New dispersal scheme in Fig.~\ref{fig:vary_m_est}(d)).
 
@@
 
-\section{Approximation of the establishment probability}
-We compute the survival probability \chg{of the lineage} of a single mutant starting either in an old- or in a new-habitat patch.
-We call this probability the establishment probability because it implies the successful establishment of a mutant population within the meta-population.
-It is denoted by $\varphi_k$ where the index $k$ indicates the initial habitat type of the mutant (old or new).
-
-Our method is the same as used in \citet{tomasini_2018} with the exception that our growth rate in old-habitat patches, $a_{\text{old}}$, depends on the demography of the population.
+\section{Approximation of the establishment probability\label{sec:approxestabproba}}
+We compute the survival probability of the lineage of a single mutant starting either in an old- or in a new-habitat patch.
+We call this probability the establishment probability because it implies the successful establishment of a mutant population within the metapopulation.
+It is denoted by $\varphi_k$, $k$ indicating the initial habitat type of the mutant (old or new).
+
+Our method is the same as the one used in \citet{tomasini_2018}, with the exception that our per capita growth rate in old-habitat patches, $a_{\text{old}}$, depends on the demography of the population.
 The method in general is based on the theory of multi-type branching processes, cf.~Chapter~5.5 in \citet{haccou_book}.
-For a detailed application of the theory we refer to the Supplementary Information of \citet{tomasini_2018}.
-
-We start by recalling the mean reproduction matrix of a mutant that gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides.
-It is given as (see also eq.~\eqref{eq:mean_repro} in the main text)
+We refer the reader to the Supplementary Information of \citet{tomasini_2018} for a detailed application of the theory.
+
+The mean reproduction matrix $\mathcal{M}$ of a mutant gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides (see also eq.~\eqref{eq:mean_repro} in the main text):
 

### 2020-07-05 ms/main.tex
ca avance, bloque sur formule fixation

@@
 A key assumption to our mathematical analysis is that the mutant individuals are rare enough that their dynamics do not affect the wild-type population, at least during the establishment phase of the mutant lineage.
+Because of their rarity, we can also consider that \linelabel{R1-27}all mutants reproduce, disperse and die independently of each other.
 The wild-type population sets a demographic context that affects mutants dynamics.
@@
 It depends on $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal, which is calculated in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI) (Eq.~\eqref{Seq:Ntildeold}).
-
-\st{For the analysis, we also further assume that the number of successful offspring of mutant individuals is Poisson distributed, with mean $1+ a_{\text{old}}$. }\florence{com eback to this}
 
@@
 
-To summarize, for the analysis part, we assume that the number of successful offspring of mutant individuals are Poisson-distributed, with means $(1+ a_{\text{old}})$ in old-habitat patches, and $1+ a_{\text{new}}$ in new-habitat patches, where both quantities are treated as constant over time\footnote{\florence{please check this statement}}.
-\linelabel{R1-17}\linelabel{R2-13} These assumptions, made for the sake of mathematical analysis, are relaxed in our stochastic simulations.
+Combining the different steps of the life cycle, the expected per capita numbers of successful offspring (i.e., offspring that become adults at the next generation) can be gathered in a reproduction matrix
+%
+\begin{equation}\label{eq:repromatrix}
+\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
+	\text{old patch} & \left(1 - m \mu_m^{\text{new}} \right) (1+a_{\text{old}}) & m \mu_m^{\text{new}} (1+a_{\text{new}}) \cr
+	\text{new patch} & m \mu_m^{\text{old}} (1+a_{\text{old}}) & \left(1 - m \mu_m^{\text{old}}\right) (1+a_{\text{new}})
+}\ ,
+\end{equation}
+where the rows denote the parent locations, and the columns the patch type of the successful offspring.
+Our final assumption for the mathematical analysis is that the distributions of numbers of successful offspring are Poisson, with means given by the elements of matrix $\mathcal{M}$ (while in reality, only the production of offspring before density regulation is Poisson)\florence{please check that it is true that dispersal is also lumped in}; these means are treated as constant over time.
+This way, the dynamics of the mutant population can be described by a two-type branching process.
+(The two ``types'' in the name of method, two-type branching process, correspond to the two habitat types.)
+
+\linelabel{R1-17}\linelabel{R2-13} All of these assumptions, made for the sake of mathematical analysis, are relaxed in our stochastic simulations.
 
@@
 \subsubsection*{Mathematical analysis}
-The dynamics of the mutant population can be described by a two-type branching process.
-A key assumption of this framework is that \linelabel{R1-27}all mutants reproduce, disperse and die independently of each other.
-This is a reasonable assumption as long as the overall number of mutants is a lot smaller than the population size of the wild type.
-The two ``types'' in the name of method, two-type branching process, correspond to the two habitat types.
-The process tracks the \linelabel{R2-25}total number of mutants in old- and new-habitat patches at the end of each generation.
-As mentioned above, the numbers of successful offspring of mutant individuals, i.e., of offspring that survive density regulation and become adults at the next generation, are approximated by Poisson-distributed numbers, with means $(1+ a_{\text{old}})$ in old-habitat patches (eq.~\eqref{eq:sold} and eq.~\eqref{Seq:s_old}), and $1+ a_{\text{new}}$ in new-habitat patches (eq.~\eqref{eq:defanew}).
-After one generation, the mean number of successful offspring of a single mutant, either in an old- or a new-habitat patch, is then given by the following mean reproduction matrix:
-%
-\begin{equation}\label{eq:mean_repro}
-	\bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
-		\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
-		\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
-		}\ ,
-\end{equation}
-%
-where the rows denote the parent locations, and the columns the patch type of the successful offspring. 
-For example, the top-left entry is the probability for the parent to stay in an old-habitat patch ($1-m_m^{\text{new}}$, see eq.\eqref{eq:dispersal}) times the average number of successful offspring in these patches \linelabel{R2-26}(see Section~\ref{sec:app:modeldyn} in SI for the derivation of $a_{\text{old}}$ at the wild-type equilibrium).
-The other entries are obtained analogously.
 
@@
 This probability is computed by considering all possible ways of going extinct: the initial individual having $j$ successful offspring in a patch of type $k$, but all lineages descending from these $j$ offspring eventually go extinct; then summing over $k$ and $j$.
-The establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ are then given by the unique positive solution of the following system of equations \citep[see][Chapters 5.3 and 5.6]{haccou_book}:
-%
-\begin{equation}\label{eq:ext_prob}
-	\begin{aligned}
-		1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ successful offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \\
-		&\qquad \qquad \qquad \left.+ \mathbf{P}(j \text{successful offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\\
-		&=\exp\left[-\left(1-m^{\text{old}\to \text{new}}_m\right)(1+a_{\text{old}})\varphi_{\text{old}} - m^{\text{old}\to \text{new}}_m (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, ,   \\
-		1-\varphi_{\text{new}} &= \exp\left[-m^{\text{new}\to \text{old}}_m(1+a_{\text{old}})\varphi_{\text{old}} - \left(1-m^{\text{new}\to \text{old}}_m\right) (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .
-	\end{aligned}
-\end{equation}
+We obtain
+%
+\begin{subequations}\label{eq:ext_prob}
+	\begin{align}
+	1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ successful offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \nonumber \\
+	&\qquad \qquad \left.+ \mathbf{P}(j \text{successful offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\nonumber \\
+	%
+	\intertext{Recalling that the number of successful offspring over the life cycle is assumed to be Poisson with means given by the terms of matrix $\mathcal{M}$ (Eq.~\eqref{eq:repromatrix}), we obtain after symplifying}
+	%
+	1-\varphi_{\text{old}}  &=\exp\left[-\left(1 - m \mu_{m}^{\text{new}} \right)(1+a_{\text{old}})\varphi_{\text{old}} - m \mu_{m}^{\text{new}} (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .   \\
+	%
+	\intertext{Similarly in a new-habitat patch, }
+	1-\varphi_{\text{new}} &= \exp\left[- m \mu_{m}^{\text{old}} (1+a_{\text{old}})\varphi_{\text{old}} - \left(1- m \mu_{m}^{\text{old}} \right) (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .
+\end{align}
+\end{subequations}
+The establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ are then given by the unique positive solution of system~\eqref{eq:ext_prob} \citep[see][Chapters 5.3 and 5.6]{haccou_book}.
 This system of equations can be solved numerically.

### 2020-07-07 ms/main.tex
revision continue

@@
 It depends on $\widetilde{N}^{\text{old}}_w$, the size of the local wild-type population right after dispersal, which is calculated in Section~\ref{sec:app:modeldyn} of the Supplementary Information (SI) (Eq.~\eqref{Seq:Ntildeold}).
+\florence{I suggest adding plots of $a_{\text{old}}$ in the SI, with the same parameters and axes as in the other figures of the main text }
 
@@
 
-Combining the different steps of the life cycle, the expected per capita numbers of successful offspring (i.e., offspring that become adults at the next generation) can be gathered in a reproduction matrix
-%
-\begin{equation}\label{eq:repromatrix}
-\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
-	\text{old patch} & \left(1 - m \mu_m^{\text{new}} \right) (1+a_{\text{old}}) & m \mu_m^{\text{new}} (1+a_{\text{new}}) \cr
-	\text{new patch} & m \mu_m^{\text{old}} (1+a_{\text{old}}) & \left(1 - m \mu_m^{\text{old}}\right) (1+a_{\text{new}})
-}\ ,
-\end{equation}
-where the rows denote the parent locations, and the columns the patch type of the successful offspring.
-Our final assumption for the mathematical analysis is that the distributions of numbers of successful offspring are Poisson, with means given by the elements of matrix $\mathcal{M}$ (while in reality, only the production of offspring before density regulation is Poisson)\florence{please check that it is true that dispersal is also lumped in}; these means are treated as constant over time.
-This way, the dynamics of the mutant population can be described by a two-type branching process.
+We finally combine the different steps of the life cycle.
+The expected per capita numbers of successful offspring  (i.e., offspring that become adults at the next generation) in habitat $k'$ of an individual in a $k$-habitat patch at the beginning of the generation, $\lambda_{k, k'}$, are the following:
+%
+\begin{equation}\label{eq:lambda}
+\begin{array}{ll}
+\lambda_{\text{old}, \text{old}} = \left(1 - m \mu_m^{\text{new}} \right) (1+a_{\text{old}}) , & 
+\lambda_{\text{old}, \text{new}} = m \mu_m^{\text{new}} (1+a_{\text{new}}) ,\\
+%
+\lambda_{\text{new}, \text{old}} =  m \mu_m^{\text{old}} (1+a_{\text{old}}) , & 
+\lambda_{\text{new}, \text{new}} = \left(1 - m \mu_m^{\text{old}}\right) (1+a_{\text{new}}) .\\
+\end{array}
+\end{equation}
+Our final assumption for the mathematical analysis is that the distributions of numbers of successful offspring are Poisson, with means $\lambda_{k, k'}$ (counting the successful offspring in habitat $k'$ of a parent in a $k$-habitat patch at the beginning of the generation).
+In reality, only the production of offspring before density regulation is Poisson, here we lump in the effects of dispersal and of density regulation.
+These means are treated as constant over time.
+This way, the dynamics of the mutant population can be described by a two-type branching process, for which an established methodology exists \citep{haccou_book}.
 (The two ``types'' in the name of method, two-type branching process, correspond to the two habitat types.)
@@
 This probability is computed by considering all possible ways of going extinct: the initial individual having $j$ successful offspring in a patch of type $k$, but all lineages descending from these $j$ offspring eventually go extinct; then summing over $k$ and $j$.
-We obtain
+Denoting by $\mathbb{P}_{k, k'}(j)$ the probability that an individual in a $k$-habitat patch at the beginning of the generation has $j$ successful offspring in a $k'$-habitat patch after density regulation, the following system of equations holds:
+\begin{subequations}\label{eq:phidefs}
+\begin{align}
+1 - \varphi_{\text{old}} &= \sum_{j=0}^{\infty}\sum_{j'=0}^{\infty} \left( \mathbb{P}_{\text{old}, \text{old}}(j) (1-\varphi_{\text{old}})^j \times \mathbb{P}_{\text{old}, \text{new}}(j')  (1-\varphi_{\text{old}})^{j'} \right) \\
+%
+1 - \varphi_{\text{new}} &= \sum_{j=0}^{\infty}\sum_{j'=0}^{\infty} \left( \mathbb{P}_{\text{new}, \text{old}}(j) (1-\varphi_{\text{old}})^j \times \mathbb{P}_{\text{new}, \text{new}}(j')  (1-\varphi_{\text{old}})^{j'} \right)
+\end{align}
+\end{subequations}
+%
+As mentioned previously, we assume for our mathematical analysis that the numbers of successful offspring over the whole life cycle are Poisson distributed with means $\lambda_{k,k'}$ given in Eq.~\eqref{eq:lambda}:
+%
+\begin{equation}\label{eq:poissondist}
+\mathbb{P}_{k, k'}(j) = \frac{e^{- \lambda_{k,k'}} \lambda_{k,k'}^j}{j!}.
+\end{equation}
+%
+Inserting these expressions into system~\eqref{eq:phidefs} and simplifying, we obtain
 %
@@
 	\begin{align}
-	1-\varphi_{\text{old}} &= \sum_{j=0}^{\infty}\left( \mathbf{P}(j \text{ successful offspring in old habitat})(1-\varphi_{\text{old}})^j \right. \nonumber \\
-	&\qquad \qquad \left.+ \mathbf{P}(j \text{successful offspring in new habitat})(1-\varphi_{\text{new}})^j\right)\nonumber \\
-	%
-	\intertext{Recalling that the number of successful offspring over the life cycle is assumed to be Poisson with means given by the terms of matrix $\mathcal{M}$ (Eq.~\eqref{eq:repromatrix}), we obtain after symplifying}
-	%
 	1-\varphi_{\text{old}}  &=\exp\left[-\left(1 - m \mu_{m}^{\text{new}} \right)(1+a_{\text{old}})\varphi_{\text{old}} - m \mu_{m}^{\text{new}} (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .   \\
 	%
-	\intertext{Similarly in a new-habitat patch, }
 	1-\varphi_{\text{new}} &= \exp\left[- m \mu_{m}^{\text{old}} (1+a_{\text{old}})\varphi_{\text{old}} - \left(1- m \mu_{m}^{\text{old}} \right) (1+a_{\text{new}}) \varphi_{\text{new}}\right]\, .
@@
 We find
-\begin{equation}\label{eq:estab_approx}
-	\begin{aligned}
-		\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}}&+ & \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
-		& & &  \; \; \; + \; \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}\widehat{\pi}_m f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
-		\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} &+ & \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
-	& & &  \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) \widehat{\pi}_m f_{\text{old}} \right)\, ,}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
-	\end{aligned}
-\end{equation}
-where $C$ is a scaling constant that depends on $m,f_{\text{old}},\widehat{\pi}_m,a_{\text{old}}$ and $a_{\text{new}}$ through
+\begin{subequations}\label{eq:estab_approx}
+	\begin{align}
+\varphi_{\text{old}} \approx &%
+a_{\text{old}} %
++ a_{\text{old}} \frac{\left(1-f_{\text{old}} + e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \nonumber \\
+& + \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} e^{\pi_m} f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
+%
+%
+\varphi_{\text{new}} \approx & % 
+a_{\text{new}} %
++ a_{\text{new}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}}) \nonumber \\
+& + \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) e^{\pi_m} f_{\text{old}} \right)\, ,
+	\end{align}
+with
 \begin{equation}\label{eq:normalization}
-	C = (1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + \widehat{\pi}_m f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, . 
-\end{equation}
-
-The first term in the approximation of the establishment probabilities ($\varphi_{\text{old}}$ and $\varphi_{\text{new}}$) in eq.~\eqref{eq:estab_approx} describes the local growth depending on the habitat type under study.
-The second term captures the growth rate differences between the two habitat types.
-The factor $(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})$ accounts for the biased dispersal patterns (when $\pi_m\neq 0$ so that $\widehat{\pi}_m\neq 1$). 
-The third term in the equations corresponds to the direct effect of dispersal on the establishment probability. 
-The first two summands in the bracket are the same for both establishment probabilities.
-They represent the general effect of dispersal due to the dynamics from new-habitat patches (first summand) and old-habitat patches (second summand).
-The dispersal bias induced by $\pi_m$ changes the relative impact of old- vs.
-new-habitat patches.
-Finally, the last summand in the bracket measures the growth rate loss (or gain) due to dispersal to the other patch type. It therefore differs between the two approximations.
-
-Note that in these equations, the competition of the mutant with the wild type in old-habitat patches appears in the local growth rate $a_{\text{old}}$, defined in eq.~\eqref{eq:sold}.
-This quantity is not constant, but instead depends on the wild type population size in old-habitat patches, which itself depends on the wild type's dispersal and growth rate in both habitats \linelabel{R2-28}\chg{(see SI, Section S1)}.
-
-If the emigration probability is zero ($m=0$), the subpopulations in each habitat evolve in isolation from each other and we recover Haldane's classical result for the establishment probability of a slightly advantageous mutant in new habitats: $\varphi_{\text{new}}=2a_{\text{new}}$ \citep{haldane_1927}.
-\linelabel{R1-69}\chg{In old-habitat patches, where the mutation is deleterious, we see that term (1) and (2) cancel each other so that the establishment probability is simply $\varphi_{\text{old}}=0$.} Furthermore,\francois{when the emigration probability is strictly positive (m>0)} in the case of \chg{unbiased} dispersal ($\pi_w=\pi_m=0$) and for equal number of old- and new-habitat patches ($f_{\text{old}}=1/2$), we obtain the approximation found in \cite{tomasini_2018} (compare system~\eqref{eq:estab_approx} to their eqs.~(4) and~(5)).
-Note that the approximation is independent of the actual number of patches, but only depends on the environmental configuration determined by the frequency of old-habitat patches $f_{\text{old}}$.
+C = (1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + e^{\pi_m} f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, . 
+\end{equation}
+\end{subequations}
+%
+Recall that $a_{\text{old}}$, while considered constant in time, depends on the model's parameters, and in particular on the dispersal probability $m$.
+The establishment probabilities $\varphi_{\text{old}} $ and $\varphi_{\text{new}}$ in \eqref{eq:estab_approx} are therefore not affine functions of $m$ (although they look so in \eqref{eq:estab_approx}).
+
+When the emigration probability is zero ($m=0$), the subpopulations in each habitat evolve in isolation from each other. The establishment probabilities in \eqref{eq:estab_approx} become
+%
+\begin{subequations}\label{eq:estabnomig}
+	\begin{align}
+	\varphi_{\text{old}} (m=0) &= 0, \\
+	\varphi_{\text{new}} (m=0) &  = 2 a_{\text{new}}. \label{eq:estabnomigHaldane}
+	\end{align}
+\end{subequations}
+Eq.~\eqref{eq:estabnomigHaldane} corresponds to Haldane's classical result for the establishment probability of a slightly advantageous mutant in new habitats \citep{haldane_1927}.
+\linelabel{R1-69} The mutation being counter-selected in old-habitat patches, its probability of establishment is null.
+
+When the emigration probability is strictly positive ($m>0$), in the case of unbiased dispersal ($\pi_w=\pi_m=0$) and for equal number of old- and new-habitat patches ($f_{\text{old}}=1/2$), we recover the approximation found in \cite{tomasini_2018} (compare system~\eqref{eq:estab_approx} to their eqs.~(4) and~(5)).
+Note that the approximation is independent of the actual number of patches (two patches in total in \cite{tomasini_2018} \footnote{\florence{please check if this is correct}}), but only depends on the environmental configuration determined by the frequency of old-habitat patches $f_{\text{old}}$.
 
 \subsubsection*{Comparison to simulations and qualitative behavior}
-We compare in Fig.~\ref{fig:vary_m_est} our predictions from eqs.~\eqref{eq:ext_prob} and~\eqref{eq:estab_approx} to simulation results for different values of the emigration rate $m$.
+Our mathematical analysis provided two kinds of results for the establishment probability: an implicit solution in eq.\eqref{eq:ext_prob}, which we solve numerically, and an explicit but approximate solution in eq.\eqref{eq:estab_approx}.
+In Fig.~\ref{fig:vary_m_est}, we compare these solutions to simulation results for different values of the emigration rate $m$.
 We find good agreement with the numerical solution of eq.~\eqref{eq:ext_prob} (solid lines).
@@
   		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.35$ (large fecundity difference)}
+  		\label{fig:vary_m_est_old-large}
 	\end{subfigure}%
@@
   		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.35$}
+  		\label{fig:vary_m_est_new-large}
 	\end{subfigure}
@@
   		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.45$ (small fecundity difference)}
+  		\label{fig:vary_m_est_old-small}
 	\end{subfigure}%
@@
   		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.45$}
+  		\label{fig:vary_m_est_new-small}		
 	\end{subfigure}
@@
 
-The qualitative dependence of the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ on the dispersal probability $m$ is similar for all dispersal schemes.
-The shape of the curve however strongly depends on the fecundity $\omega^\text{old}_m$ of mutants in the old habitat.
+%The qualitative dependence of the establishment probabilities $\varphi_{\text{old}}$ and $\varphi_{\text{new}}$ on the dispersal probability $m$ is similar for all dispersal schemes.
+The shapes of the curves however strongly depend on the fecundity $\omega^\text{old}_m$ of mutants in the old habitat.
 Before discussing the differences between the dispersal schemes, we first provide a qualitative understanding of this general behavior.
 
-For the probability of establishment of a single mutant in an old-habitat patch ($\varphi_{\text{old}}$), we observe up to three different regions, cf.
-Fig.~\ref{fig:vary_m_est}(a).
+We identify up to three different regions that define how the probability of establishment of a single mutant initially in an old-habitat patch ($\varphi_{\text{old}}$) changes with the emigration probability $m$ (Fig.~\ref{fig:vary_m_est_old-large} ).
 This is in line with previous observations in the context of local adaptation \citep[e.g.][]{kawecki_1995,tomasini_2018} and evolutionary rescue \citep{uecker_2014}.
-We define the regions as follows: (i) an initial increase of the establishment probability at low dispersal rates $m$; (ii) a local maximum with a subsequent decrease of the establishment probability; (iii) an increase of the establishment probability for high dispersal rates.
-
-A detailed assessment and explanation of the regions is provided in the SI, Section S2.1.
-Briefly, in region (i), the prevalent effect is the dispersal of mutants from old- to new-habitat patches where they are fitter than the wild type, thus increasing the establishment probability.
-This effect is mediated through the third term of the establishment probability in eq.~\eqref{eq:estab_approx}.
-Region (ii), beginning with the local maximum, is a result of two counteracting processes: %increased emigration of individuals from the old habitat and \linelabel{R2-30-1}\chg{an increasing probability of an individual that is born in a new habitat emigrates to an old-habitat patch.}
-\linelabel{R2-30-1}\chg{If we increase the emigration probability $m$, a mutant is more likely to emigrate out of old-habitat patches and thus has a higher chance to establish.
-However, higher dispersal rates also have a negative effect on mutant establishment through \linelabel{R2-32}increasing the probability that a mutant migrates from a new- to an old-habitat patch.
-More precisely, the number of offspring in the new habitat on average is given by the product of the terms $(1-m_m^{\text{old}})$ and $(1+a_{\text{new}})$.
+We define the regions as follows: (i) at low dispersal rates $m$, an initial increase of the establishment probability with $m$ ; (ii) a local maximum with a subsequent decrease of the establishment probability; (iii) at high dispersal rates $m$, an increase of the establishment probability with $m$.
+
+A detailed assessment and explanation of the regions is provided in the Appendix, Section~\ref{sec:SI_contributions}.
+In region (i), the beneficial effect of dispersal on the establishment probability $\varphi_{\text{old}}$ is due to mutants dispersing from old- to new-habitat patches where they are fitter than the wild type.
+While this effect is still present in region (ii), the establishment probability $\varphi_{\text{old}}$ now decreases with dispersal \linelabel{R2-30-1} \linelabel{R2-32} because back-dispersal into the old-habitat patches is also possible.
+\florence{STOPPED HERE}
+More precisely, the expected per capita number of successful offspring in the new habitat of an adult in a new-habitat patch is $\lambda_{\text{new, new}} = (1- m \mu_m^{\text{old}}) (1+a_{\text{new}})$.
 This product can, for large emigration probabilities $m$, be smaller than $1$, i.e.
 a mutant in a new-habitat patch has on average less than one offspring.
+This is detrimental to the mutant because it means mutants do not efficiently reproduce in the habitat where they are fitter.
 Finally, in region (iii), dispersal is so large that the population is close to being well-mixed.
@@
 The larger it is, the `earlier' (i.e.
-for smaller emigration rates $m$) relaxed competition becomes relevant.}
+for smaller emigration rates $m$) relaxed competition becomes relevant.
 
@@
 The mean reproduction matrix $\mathcal{M}$ of a mutant gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides (see also eq.~\eqref{eq:mean_repro} in the main text):
-
+\begin{equation}\label{eq:repromatrix}
+\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
+	\text{old patch} & \left(1 - m \mu_m^{\text{new}} \right) (1+a_{\text{old}}) & m \mu_m^{\text{new}} (1+a_{\text{new}}) \cr
+	\text{new patch} & m \mu_m^{\text{old}} (1+a_{\text{old}}) & \left(1 - m \mu_m^{\text{old}}\right) (1+a_{\text{new}})
+}\ ,
+\end{equation}
 \begin{equation}\label{Seq:mean_repro}
@@
 
-\subsection*{Computing the largest eigenvalue}
+\subsection{Computing the largest eigenvalue}
 We first approximate the largest eigenvalue of $\mathcal{M}$ denoted by $\rho$. It is given by (see \textit{Mathematica} notebook)
@@
 
-\subsection*{Computing the establishment probability}
+\subsection{Computing the establishment probability}
 For the solution of eq.~\eqref{Seq:theory} it remains to compute the normalized eigenvectors.
@@
 
-\subsection{Disentangling the contributions to the establishment probability}
+\subsection{Disentangling the contributions to the establishment probability \label{sec:SI_contributions}}
 \chg{We now proceed to explain} the three regions of the establishment probability from Fig.~\ref{fig:vary_m_est}(a) in the main text.

### 2020-07-08 ms/main.tex
fin premiere grosse revision main text

@@
 \begin{abstract}
+	\florence{will do this in the very end}
     \noindent \textbf{Abstract} Evolutionary rescue is the process by which a declining population successfully adapts genetically\footnote{\francois{remove genetically}\florence{ok to remove}} to avoid extinction.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, contribute to the decline of the populations of many species, that might eventually go extinct \citep{bellard_2012, diniz_2019}.
+Current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, contribute to the decline of the populations of many species, that might eventually go extinct \citep{diniz_2019}.
 Pests and pathogens experience similarly strong selective pressures as a result of increased consumption of antibiotics and use of pesticides \citep{ramsayer_2013, kreiner_2018}.
-The process of genetic\footnote{\florence{rescue by plasticity as well}} adaptation that saves populations from extinction is termed evolutionary rescue.
-This \linelabel{R1-3}\chg{process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, } \florence{classically} resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
+The process of adaptation that saves populations from extinction is termed evolutionary rescue.
+This \linelabel{R1-3}process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, classically resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
 In recent years, empirical examples of evolutionary rescue have accumulated \citep[as reviewed by ][]{alexander_2014,carlson_2014,bell_2017}.
@@
 The traits involves can be continuous (ref) or discrete (ref).
-In this work, we consider genetic adaptation mediated by discrete traits.
-Specifically, we consider that a single mutation can prevent extinction, and...}
+In this work, we consider genetic adaptation mediated by a discrete trait.
+}
 
@@
 different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}.
-\chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al.
+In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al.
 1991.
@@
 If individuals that populate different patches are able to move between those, the effect of dispersal on evolutionary rescue needs to be taken into account \citep{uecker_2014, tomasini_2019}.
-\linelabel{R1-1}\chg{The intensity of dispersal among patches tunes how abruptly environmental change is experienced.
+\linelabel{R1-1}The intensity of dispersal among patches tunes how abruptly environmental change is experienced.
 With very low dispersal, patches are essentially isolated from each other, and each patch undergoes an abrupt change independently of the other patches.
-With higher dispersal, asynchronous deterioration among patches is experienced as a more gradual change overall.}
+With higher dispersal, asynchronous deterioration among patches is experienced as a more gradual change overall.
 Experiments that study the effect of dispersal on evolutionary rescue are rare, but, for instance, \citet{bell_2011} found that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation.
@@
 These models represent a spatially heterogeneous environment, constant in time, in which wild-type populations in unfavorable habitats can only be maintained thanks to recurrent immigration from favorable habitats.
-Experimental and theoretical studies have found that \linelabel{R2-3}\chg{increasing} dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g., for studies on discrete traits, \florence{[refocus the citations with papers on discrete traits; otherwise there are too many possible citations -- it's fine to do so given that we now explicitly say the paper is focused on discrete traits]}\citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000, gallet_2018}} for both effects).
+Experimental and theoretical studies have found that \linelabel{R2-3}increasing dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g., for studies on discrete traits, \florence{[refocus the citations with papers on discrete traits; otherwise there are too many possible citations -- it's fine to do so given that we now explicitly say the paper is focused on discrete traits]}\citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000, gallet_2018}} for both effects).
 %Reason for different results: (i) demography or not (Kawecki); (ii) selection strength (Gallet), (iii) selection strength (Ronce)
@@
 %%% DISPERSAL, biased or not
-In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be \chg{unbiased}, i.e.
+In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be unbiased, i.e.
 dispersing individuals are distributed uniformly among patches.
-Only few investigations in the context of local adaptation in source-sink systems have taken into account non-\chg{uniform} dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}.
-This analytical focus on \chg{unbiased} dispersal is in stark contrast to dispersal schemes observed in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}.
+Only few investigations in the context of local adaptation in source-sink systems have taken into account non-uniform dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}.
+This analytical focus on unbiased dispersal is in stark contrast to dispersal schemes observed in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-One of the best documented modes of non-\chg{uniform} dispersal is density-dependent dispersal.
+One of the best documented modes of non-uniform dispersal is density-dependent dispersal.
 Density dependence can be positive or negative: either individuals prefer to settle or stay in large groups (positive density-dependence), or they choose to remain in or move to less populated regions (negative density-dependence).
@@
 
-Another well-established dispersal scheme is a type of habitat choice, whereby individuals tend to immigrate into habitats they are best adapted to.
+Another well-established dispersal scheme is a type of habitat choice whereby individuals tend to immigrate into habitats they are best adapted to.
 This mechanism has for example been reported in lizards \citep{bestion_2015}, birds \citep{dreiss_2011,benkman_2017}, fishes \citep{bolnick_2009}, worms \citep{mathieu_2010}, and ciliates \citep{jacob_2017,jacob_2018}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}\chg{types, wild type and mutant, have a positive growth rate}, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct.
-We study four biologically motivated dispersal schemes, which correspond to the four combinations of biases towards old vs.\ new patches for wild type and mutants, and we compare these dispersal schemes to \chg{unbiased} dispersal.
+We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}types, wild type and mutant, have a positive growth rate, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct.
+We study four biologically motivated dispersal schemes, which correspond to the four combinations of biases towards old vs.\ new patches for wild type and mutants, and we compare these dispersal schemes to unbiased dispersal.
 Our analysis is carried out step-wise.
@@
 of establishment of the mutant lineage\footnote{\florence{not sure about this defn, please check}}.
-Finally, we let habitat degradation proceed, assuming that patches, one after \chg{another}, deteriorate over time until all locations contain the new habitat.
+Finally, we let habitat degradation proceed, assuming that patches, one after another, deteriorate over time until all locations contain the new habitat.
 Using the previous results, we approximate a probability of evolutionary rescue, i.e.\ that a mutation appears, establishes, thereby allowing the population to persist in spite of environmental degradation.
-We find that dispersal biases affect the probabilities of establishment and of evolutionary rescue, because it directly affects the local growth rates.
+We find that dispersal biases affect the probabilities of establishment and of evolutionary rescue.
 
@@
 The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively.
-One after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible.
+One after another every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible.
 Initially ($t<0$), all patches are of the old-habitat type.
@@
 On the contrary, the local population is viable if the expected fecundity is greater than $1$.
-We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}.
+We assume that wild-type and mutant populations are viable in old-habitat patches\footnote{\florence{this won't be the case in fig \ref{fig:rescue_lethal}}}, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}.
 In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$.
@@
 
-The probability that an individual of type $i$ migrates from a patch of any habitat type into a patch of the new-habitat type is\florence{TODO change into simple bias (remove m)}
+The probability that a dispersing individual of type $i$ settles in a patch of the new-habitat type is
 %
@@
 \begin{aligned}
-m_i^{\text{new}} &= m\, \frac{1-f_{\text{old}}}{1-f_{\text{old}}+\chg{e^{\pi_i}} f_{\text{old}}} = 1-m_i^{\text{old}},\\
+\mu_i^{\text{new}} &= \frac{1-f_{\text{old}}}{1-f_{\text{old}}+ e^{\pi_i} f_{\text{old}}} = 1 - \mu_i^{\text{old}},\\
 \end{aligned}
 \end{equation}
-\florence{there seems to be a problem with the second equation, should be $m -$ no? }
 where, as defined above, $f_{\text{old}}$ is the frequency of old-habitat patches and $\pi_i$ the dispersal bias into old-habitat patches.
 The use of an exponential $e^{\pi_i}$ \linelabel{R1-19} ensures that the fraction in eq.~\eqref{eq:dispersal} is positive and between zero and one.
-For the ease of notation we will write $\widehat{\pi}_i=e^{\pi_i}$.\footnote{\florence{TODO remove hat notation}}
 
@@
 		\begin{tabular}{c | P{6cm} | c | c}
-			\textbf{Parameter} & \textbf{Interpretation} & \textbf{Range} & \textbf{Default value} \\
-			\midrule
+			\textbf{Notation} & \textbf{Interpretation} & \textbf{Range} & \textbf{Default value} \\
+			\hline
 			$K_k$ & Carrying capacity in a patch of type $k$ & -- & \makecell{$K_{\text{old}}=1000$,\\ $K_{\text{new}}=500$} \\ 
-			$\omega_j$ & \makecell{Expected per-capita number of\\ type $j$ offspring in the old habitat\\ before regulation} & $0\leq \omega^\text{old}_m < \omega^\text{old}_w$ & \makecell{$\omega^\text{old}_w = 1.5$,\\ $\omega^\text{old}_m=1.45$ {\footnotesize (weak sel.)},\\ $\omega^\text{old}_m=1.35$ {\footnotesize (strong sel.)}}\\
-			$1+a_{\text{old}}$ & \makecell{Growth rate of the mutant\\ in the old habitat} & $-1\leq a_{\text{old}}$ & -- \\
-			$1-r$ & \makecell{Mean number of wild-type offspring\\ in the new habitat} & $0<r\leq 1$ & \makecell{$0.75$\\ $(r=0.25)$}\\
-			$1+a_{\text{new}}$ & \makecell{Growth rate of the mutant\\ in the new habitat} & $0<a_{\text{new}}$ & \makecell{$1.02$ \\ ($a_{\text{new}}=0.02$)} \\
+			$\omega_w^{\text{old}}$ & Fecundity of wild-type individuals in old-habitat patches & $1 < \omega^\text{old}_w$ & $1.5$\\
+			$\omega_m^{\text{old}}$ & Fecundity of mutant individuals in old-habitat patches & $0\leq \omega^\text{old}_m < \omega^\text{old}_w$ & $1.45$ or $1.35$\\
+			%			
+%			$\omega_j^{\text{old}}$ & Fecundity of type $j$ individuals in old-habitat patches & $0\leq \omega^\text{old}_m < \omega^\text{old}_w$ & \makecell{$\omega^\text{old}_w = 1.5$,\\ $\omega^\text{old}_m=1.45$ or $\omega^\text{old}_m=1.35$}\\
+			%
+			$\omega_w^{\text{new}}$ & Fecundity of wild-type individuals in new-habitat patches & $0\leq \omega^\text{new}_w < 1$ & $0.75$\\
+			%
+			$\omega_m^{\text{new}}$ & Fecundity of mutant individuals in new-habitat patches & $1 < \omega^\text{new}_m$ & $1.02$\\
+%		
+%			$\omega_j^{\text{new}}$ & Fecundity of type $j$ individuals in new-habitat patches & $0\leq \omega^\text{new}_w < 1 < \omega^\text{new}_m$ & \makecell{$\omega^\text{new}_w = 0.75$,\\ $\omega^\text{new}_m=1.02$}\\
 			$m$ & Emigration probability & $0<m\leq 1$ & $0.06$\\
 			$\pi_i$ & Type $i$ bias towards the old habitat & $\pi_i\in\mathbb{R}$ & see Fig.\ref{fig:disp_schemes}\\
-			$\widehat{\pi}_i$ & Transformed dispersal bias of type $i$ & $0<\widehat{\pi}_i$ & see Fig.\ref{fig:disp_schemes}\\
+			%$\widehat{\pi}_i$ & Transformed dispersal bias of type $i$ & $0<\widehat{\pi}_i$ & see Fig.\ref{fig:disp_schemes}\\
 			$M$ & Number of patches & $2\leq M$ & $10$\\
@@
 			$\tau$ & Time interval between two consecutive deterioration events& $0<\tau$ & $100$\\
-			\midrule
-			$\widehat{N}_{i}^{k}$ & \multicolumn{3}{p{11cm}}{Number of type $i$ individuals in type $k$ habitat patches at stationarity}  \\
-			$\widetilde{N}_{i}^{k}$ & \multicolumn{3}{p{11cm}}{Number of type $i$ individuals in type $k$ habitat patches after dispersal}
+			\hline
+			$a_{\text{old}}$ & \makecell{Growth rate of the mutant\\ in the old habitat} & $-1\leq a_{\text{old}}$ & eq.~\eqref{eq:sold} \\
+			%
+			$\widehat{N}_{i}^{k}$ & Number of type $i$ individuals in type $k$ habitat patches at stationarity & $0 \leq \widehat{N}_{i}^{k} \leq K_{i}$& eq.~\eqref{Seq:Nhat}  \\
+			%
+			$\widetilde{N}_{i}^{k}$ & Number of type $i$ individuals in type $k$ habitat patches after dispersal & $0\leq \widetilde{N}_{i}^{k}$  & eq.~\eqref{Seq:Ntilde}
 		\end{tabular}
-		\caption{\textbf{Model parameters.}}	
+		\caption{\textbf{Model parameters and variables.}}	
 		\label{tab:parameters}
@@
 
-\todo{to be moved} All the model parameters are summarized in Table~\ref{tab:parameters} along with the default parameter values and ranges.
+All the model parameters are summarized in Table~\ref{tab:parameters} along with the default parameter values and ranges.
 If not stated otherwise, the default parameter values are used for the stochastic simulations.
@@
 + a_{\text{new}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}}) \nonumber \\
-& + \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) e^{\pi_m} f_{\text{old}} \right)\, ,
+& + \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} e^{\pi_m} f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) e^{\pi_m} f_{\text{old}} \right)\, ,
 	\end{align}
@@
   		\includegraphics[width=\linewidth]{fig2a.pdf}
-  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.35$ (large fecundity difference)}
-  		\label{fig:vary_m_est_old-large}
+  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.35$}
+  		\label{fig:vary_m_est_old-35}
 	\end{subfigure}%
@@
   		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.35$}
-  		\label{fig:vary_m_est_new-large}
+  		\label{fig:vary_m_est_new-35}
 	\end{subfigure}
@@
   		\includegraphics[width=\linewidth]{fig2c.pdf}
-  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.45$ (small fecundity difference)}
-  		\label{fig:vary_m_est_old-small}
+  		\caption{$\varphi_{\text{old}}$ with $\omega^\text{old}_m = 1.45$}
+  		\label{fig:vary_m_est_old-45}
 	\end{subfigure}%
@@
   		\caption{$\varphi_{\text{new}}$ with $\omega^\text{old}_m = 1.45$}
-  		\label{fig:vary_m_est_new-small}		
+  		\label{fig:vary_m_est_new-45}		
 	\end{subfigure}
-	\caption{\textbf{Establishment probability when varying the emigration rate.} \small We plot the theoretical results for $\varphi_{\text{old}}$ in (a) and (c), and for $\varphi_{\text{new}}$ in (b) and (d) for $\omega^\text{old}_m=1.35$ in (a), (b) and $\omega^\text{old}_m=1.45$ in (c), (d).
-Comparison with the results from stochastic simulations (symbols) show very good agreement with our approximation found in eq.~\eqref{eq:estab_approx} (dashed lines).
-The solid lines are the numerical solution of eq.~\eqref{eq:ext_prob}.
-The lines in (a) show a clear separation of the three regions of the establishment probability as discussed in the main text: {\color{change}(i) an initial increase due to higher probabilities of dispersal from old- to new-habitat patches; (ii) a local maximum due to increasing number of mutants emigrating from new-habitat patches; (iii) an increase due to relaxed competition in old-habitat patches, a result of wild-type emigration.}}
+	\caption{\textbf{Establishment probability as a function of the emigration rate.} \small Panels (a) and (c) show the establishment probabilities when the first mutant arises in an old-habitat patch ($\varphi_{\text{old}}$), and panels (b) and (d) the establishment probabilities when the mutant arises in a new-habitat patch ($\varphi_{\text{new}}$), for mutant fecundity in old-habitat patches $\omega^\text{old}_m=1.35$ in (a), (b) and $\omega^\text{old}_m=1.45$ in (c), (d).
+Markers: simulations; full lines: numerical solution of eq.~\eqref{eq:ext_prob}; dashed lines, approximate solution shown eq.~\eqref{eq:estab_approx}.
+The colors and marker shapes correspond to the different dispersal schemes, with the same parameters as in Fig.\ref{fig:disp_schemes}.}
 	\label{fig:vary_m_est}
@@
 
-We identify up to three different regions that define how the probability of establishment of a single mutant initially in an old-habitat patch ($\varphi_{\text{old}}$) changes with the emigration probability $m$ (Fig.~\ref{fig:vary_m_est_old-large} ).
+We identify up to three different regions that define how the probability of establishment of a single mutant initially in an old-habitat patch ($\varphi_{\text{old}}$) changes with the emigration probability $m$ (Fig.~\ref{fig:vary_m_est_old-35} ).
 This is in line with previous observations in the context of local adaptation \citep[e.g.][]{kawecki_1995,tomasini_2018} and evolutionary rescue \citep{uecker_2014}.
@@
 
-A detailed assessment and explanation of the regions is provided in the Appendix, Section~\ref{sec:SI_contributions}.
+A detailed assessment and explanation of the regions is provided in the Appendix, Section~\ref{sec:SI_contributions}\footnote{\florence{reconsider whether to include}}.
 In region (i), the beneficial effect of dispersal on the establishment probability $\varphi_{\text{old}}$ is due to mutants dispersing from old- to new-habitat patches where they are fitter than the wild type.
-While this effect is still present in region (ii), the establishment probability $\varphi_{\text{old}}$ now decreases with dispersal \linelabel{R2-30-1} \linelabel{R2-32} because back-dispersal into the old-habitat patches is also possible.
-\florence{STOPPED HERE}
+While this effect is still present in region (ii), the establishment probability $\varphi_{\text{old}}$ now decreases with dispersal \linelabel{R2-30-1} \linelabel{R2-32} because back-dispersal into the old-habitat patches becomes more prevalent.
 More precisely, the expected per capita number of successful offspring in the new habitat of an adult in a new-habitat patch is $\lambda_{\text{new, new}} = (1- m \mu_m^{\text{old}}) (1+a_{\text{new}})$.
@@
 This is detrimental to the mutant because it means mutants do not efficiently reproduce in the habitat where they are fitter.
-Finally, in region (iii), dispersal is so large that the population is close to being well-mixed.
-Most importantly, many wild-type individuals leave old-habitat patches which results in less competitive pressure in old-habitat patches.
-In this case the local growth rate of the mutant in these patches, $a_{\text{old}}$ (first term in eq.~\eqref{eq:estab_approx}), becomes positive.
+Finally, in region (iii) at high dispersal, some many wild-type individuals leave old-habitat patches that competitive pressure in old-habitat patches is substantially decreases.
+The post-dispersal size of the wild-type population $\widetilde{N}^{\text{old}}_w$ is then low enough that the local growth rate of the mutant in these patches, $a_{\text{old}}$ (first term in eq.~\eqref{eq:estab_approx}), becomes positive\footnote{\florence{add ref to figures of aold as function of m}}.
 This effect called `relaxed competition'~\citep{uecker_2014}.
-The onset of this effect, in terms of the emigration probability $m$, is strongly dependent on the fecundity values of the mutant in the old habitat.
+The onset of this effect, in terms of the emigration probability $m$, is strongly dependent on the fecundity of the mutant in the old habitat, $\omega_{m}^{\text{old}}$.
 The larger it is, the `earlier' (i.e.
-for smaller emigration rates $m$) relaxed competition becomes relevant.
-
-%lowering competitive pressure in old-habitat patches \linelabel{R1-28-1}(see $\widetilde{N}_{\text{old}}^w$ from eqs.~(S7) and~(S8) in the SI).
-As a result, local growth rates of the mutant in old-habitat patches, \linelabel{R1-30} $a_{\text{old}}$, become larger.
-This effect was termed `relaxed competition' by \citet{uecker_2014}.
-
-%Depending on the onset of this effect, the establishment probability either has a local maximum or not, compare \ref{fig:vary_m_est}(a) and (c).
-
-\chg{In line with this reasoning we find} that the width of region (ii) strongly depends on the fecundity of the mutant in the old habitat, $\omega^\text{old}_m$.
-If $\omega^\text{old}_m$ is high, the local growth rate of the mutant $a_{\text{old}}$ starts to increase at a lower dispersal rate
-%(see Fig.~S1). 
-\chg{and }region (ii) disappears, as visible in Fig.~\ref{fig:vary_m_est}(c).
-In contrast, for low mutant fecundity values $\omega^\text{old}_m$, region (iii) might vanish (see Fig.
-S2 in SI).
-There, the mutant is strongly disadvantageous in old-habitat patches when compared to the wild type, and the wild-type population will always outcompete the mutant due to the much higher offspring numbers.
-
-We analyze the qualitative behavior of the establishment probability of a mutant emerging in the new habitat, $\varphi_{\text{new}}$ in a similar way (Figs.~\ref{fig:vary_m_est}(b,d)).
-The establishment probability $\varphi_{\text{new}}$ decreases at low dispersal rates: Since we let the mutant start in a new-habitat patch, where it fares better than the wild type, there is no initial benefit due to dispersal (the third term in eq.~\eqref{eq:estab_approx} is negative for $\varphi_{\text{new}}$).
-The interpretation of region (ii), describing the trajectory for intermediate emigration rates $m$, is the same as for $\varphi_{\text{old}}$ above.
-For large emigration rates $m$, region (iii), the resulting establishment probability is a combination of the local growth rate and the dispersal effect, the first and third terms in eq.~\eqref{eq:estab_approx}.
-This is because the mutant can migrate to old-habitat patches, where it will enjoy relaxed competition.
+for smaller emigration rates $m$) relaxed competition becomes relevant (compare panels \ref{fig:vary_m_est_old-35} to \ref{fig:vary_m_est_old-45}), to a point that region (ii) vanishes (panel \ref{fig:vary_m_est_old-45}) and there is no decrease of $\varphi_{\text{old}}$ with $m$ any longer.
+In contrast, for lower mutant fecundity values $\omega^\text{old}_m$, region (iii) might vanish (see Fig.~\ref{Sfig:weak_fecundity} in SI), because the mutant's fecundity in old-habitat patches is too low compared to the wild type's.
+
+The qualitative behavior of the establishment probability of a mutant emerging in the new habitat, $\varphi_{\text{new}}$, can be interpreted in a similar way (panels \ref{fig:vary_m_est_new-35}, \ref{fig:vary_m_est_new-45}).
+The establishment probability $\varphi_{\text{new}}$ decreases with the emigration probability $m$ at low $m$, because the mutant appeared in a new-habitat patch, where it fares better than the wild type, so there is no initial benefit due to dispersal.
+When the emigration probability is higher, however, mutants can back emigrate to new-habitat patches, while those that land in old-habitat patches can enjoy relaxed competition when $m$ is high.
+This is why $\varphi_{\text{new}}$ increases with $m$ at higher $m$.
 %%
 
-\chg{Lastly, for large emigration probabilities $m$ in Fig.~\ref{fig:vary_m_est}(c,d) we observe a decrease in the establishment probability of the simulation results of the New-New dispersal scheme (and to a lesser extent in the unbiased scheme).
-This is explained by `gene swamping' whereby a to large number of immigrants of a less-well adapted type (wild type) inhibits the establishment of a locally better adapted type (mutant) \citep{nagylaki_1978,lenormand_2002}.
-Here, gene swamping occurs because of the different carrying capacities of the two patch types, $K_{\text{new}}<K_{\text{old}}$.
-A proper analysis of the gene swamping regime, where density-dependence in new-habitat patches is important, is beyond the scope of this manuscript, but see for instance \citet{tomasini_2018} and Section S5 in the SI where such a situation is studied.}\footnote{\pete{If I do not mention gene swamping here, I think people will wonder what this effect is and misinterpret it as a late relaxed competition effect.}}
+An additional effect can take place at high dispersal and reduce mutant establishment probabilities, in particular when the wild type is biased toward new-habitat patches (see for instance the New-New scheme in Figs.~\ref{fig:vary_m_est_old-45}, \ref{fig:vary_m_est_new-45}).
+The high dispersal of wild types individuals into new-habitat patches can lead to `gene swamping`, inhibiting the establishment of a locally better adapted type (mutant) \citep{nagylaki_1978,lenormand_2002}.
+Note that the lower carrying capacity in new-habitat patches than in old-habitat patches ($K_{\text{new}}<K_{\text{old}}$) creates a favorable setting to gene swamping.
+A proper analysis of the gene swamping regime is beyond the scope of this manuscript, but see for instance \citet{tomasini_2018} {\color{gray}and Section S5\florence{I don't understand why S5 is about swamping} in the SI where such a situation is studied.\footnote{\pete{If I do not mention gene swamping here, I think people will wonder what this effect is and misinterpret it as a late relaxed competition effect.}}}
 
@@
 
-We now compare the establishment probabilities for the different dispersal schemes. 
-We consistently observe that \chg{a bias towards the new habitat (New-New} in Fig.~\ref{fig:vary_m_est}) enhances establishment when compared to the \chg{unbiased} dispersal scheme (0-0).
-This can be attributed to two reasons: (1) the mutant is more likely to disperse to the new habitat where it outcompetes the wild type (stronger weighting of new-habitat patches in the third term in eq.~\eqref{eq:estab_approx}); and (2) since the wild type prefers to settle in new-habitat patches ($\pi_w<0$), individuals in old-habitat patches experience relaxed competition for lower dispersal rates $m$, thus increasing the local growth rate $a_{\text{old}}$, i.e.
-region (iii) is shifted to the left.
-An analogous (but reversed) argument explains why \chg{a bias towards the old habitat (Old-Old}) is always lower than the \chg{unbiased} dispersal scheme.
-
-The effect of \chg{the asymmetric dispersal scheme Old-New}, where each type has a bias to move to the habitat where it is relatively fitter than the other type, is more involved.
-We disentangle the effects separately for each region.
-In region (i) at low emigration probabilities, the curve is almost identical to that of the \chg{New-New} pattern.
-The mutant preference \chg{(new habitats)} of these two dispersal modes is the same.
-The movement of the rare mutant is the prevailing process in this parameter regime, which explains the alignment of the \chg{corresponding} curves.
-For intermediate dispersal rates, region (ii), the effect of the heterogeneous environment (second term in eq.~\eqref{eq:estab_approx}) becomes stronger, i.e.
-the local growth rate $a_{\text{old}}$ increases because of relaxed competition in old-habitat patches.
-For the \chg{Old-New scheme}, the\linelabel{R2-30-3} wild type is more likely to immigrate into old-habitat patches and to increase the population size in these locations \linelabel{R1-31}\chg{when compared to the New-New dispersal scheme}.
-This reduces the effect of relaxed competition.
-Therefore, the \chg{Old-New line} drops below the \chg{New-New} curve and even starts to decrease (Fig.~\ref{fig:vary_m_est}(a)).
-For high emigration probabilities $m$, region (iii), the competitive pressure in old-habitat patches relaxes.
-Again, this region is dominated by the movement of wild-type individuals.
-In the \chg{unbiased} dispersal scheme, the wild type has a lower likelihood to be in old-habitat patches than with the \chg{Old-New} scheme, explaining the \chg{order of these} curves.
-For very large dispersal rates, also\francois{moreover?} the \chg{Old-Old} scheme can lead to higher values of $\varphi_{\text{old}}$ than \chg{the type-dependent preference Old-New}, cf.
-Fig.~\ref{fig:vary_m_est}(c).
-For these parameters, the mutant has a larger growth rate in old- than in the new-habitat patches ($a_{\text{old}}>a_{\text{new}}$).
-It is therefore beneficial for the mutant to \linelabel{R2-30-4}\chg{be in} old-habitat patches.
-
-Finally, for the \chg{asymmetric dispersal scheme New-Old, where the types have a bias to the habitat they are less well adapted to, }we see that for low dispersal rates $m$ the results match those obtained under the \chg{Old-Old} scheme.
-The mutant preferences being the same, this again shows that this region is dominated by the mutant's dispersal behavior.
-For intermediate to large dispersal rates though, the \chg{New-Old} scheme results in the largest establishment probabilities.
-This is explained by the effect of strongly relaxed competition (wild-type individuals preferentially move into the new habitat) and the bias of mutants to disperse into old-habitat patches where their number of offspring individuals is now very large.
-%Overall, this results in the highest establishment probabilities of all dispersal schemes.
-%We note though that the term `maladaptive dispersal' might be misleading for large dispersal values, since here the growth rate of the mutant in old-habitat patches, $a_{\text{old}}$, is actually larger than the growth rate in new-habitat patches, $a_{\text{new}}$.
+We now compare the establishment probabilities across the different dispersal schemes. 
+We consistently observe that a bias towards the new habitat (New-New in Fig.~\ref{fig:vary_m_est}) enhances mutant establishment compared to the unbiased dispersal scheme (0-0).
+This can be attributed to two reasons.
+First, the mutant is more likely to disperse to the new habitat where it outcompetes the wild type.
+Secondly, competition in old-habitat patches is relaxed starting at low emigration probabilities $m$ because the wild type preferentially disperses to new-habitat patches.
+Conversely, a bias towards the old habitat (Old-Old) lowers mutant establishment probabilities compared to the unbiased dispersal scheme.
+
+The rankings of the asymmetric dispersal schemes (Old-New and New-Old) compared the the unbiased scheme (0-0) depend on the amount of dispersal (compare the orange, purple and black curves in Fig.~\ref{fig:vary_m_est}).
+As explained above, at low dispersal probabilities $m$, the prevalent force is the dispersal of mutants to new-habitat patches.
+The establishment probability of the mutant is therefore higher for the Old-New scheme, where mutants preferentially dispersal to new-habitat patches.
+At high dispersal probabilities $m$ however, an important force is competitive release in old-habitat patches.
+The establishment probability of the mutant is therefore higher in the scheme where wild-type individuals preferentially disperse out of old-habitat patches, releasing competition there (New-Old scheme).
 
@@
 The last time point at which a mutation can occur is denoted by $t_{\text{fin}}$.
-Later, in the analysis of the probability for evolutionary rescue, this time will be replaced by the time between two consecutive patch deterioration events, $\tau$.\footnote{\pete{I think we can get rid of this last sentence.
-Thoughts?}\francois{Yes.}}
-
-The probability of adaptation in this setting, $P_{\mathrm{adapt}}$, is given by
+
+In this setting, the probability of adaptation $P_{\mathrm{adapt}}$ is approximated by
 \begin{equation}\label{eq:source_sink}
@@
 \end{equation}
-In words, the right-hand side is one minus the probability of no mutant establishing within the $[0, t_{\text{fin}}]$ time interval.
-More precisely, the exponential is the probability of zero successes of a Poisson distribution.
-The rate of this Poisson distribution is given by the expected number of successfully emerging mutant lineages until time $t_{\text{fin}}$.
-\linelabel{R2-34}\chg{Assuming a Poisson distribution for the number of successful mutant lineages is an approximation of a Binomial distribution with large sample size (the wild-type populations $t_{\text{fin}} M f_{\text{old}} K_{\text{old}}$ in the old and approximately $t_{\text{fin}} M (1-f_{\text{old}}) \widehat{N}_w^{\text{new}}$ in the new habitats, see SI eq.~(S5)) with small success probability (the establishment probabilities $\varphi_{k}$).
-The sample size denotes the number of mutations in the system until time $t_{\text{fin}}$ and each of these mutants has, independently of each other (another approximation), has a survival probability of $\varphi_{k}$, dependent on the habitat it is born in.}
-%Correctly, at least in the old habitat, one would have a Binomial distribution with success probability $\theta \varphi_{\text{old}}$ and sample size $t_{\text{fin}}M f_{\text{old}} K_{\text{old}}$ which is well approximated by a Poisson distribution for large sample sizes and small success probabilities.
-Due to the varying population size in the new habitat, the exact probability for successful lineages emerging from that habitat type is more difficult to compute.
-Yet again, on average this number is well approximated by the a Poisson distribution with a rate as stated above.}\footnote{\pete{This is the long explanation.
-I would shorten but if you think it is worth keeping a long explanation (as suggested by the reviewer) we can leave it as it is.}}
-Note also, that for $t_{\text{fin}}$ tending to infinity, there will almost surely be a successful mutant so that $P_{\text{adapt}}=1$. \\
+This is one minus the probability that zero mutant establishes within the $[0, t_{\text{fin}}]$ time interval, given by the probability of zero successes of a Poisson distribution.
+The rate of this Poisson distribution is the expected number of successfully emerging mutant lineages until time $t_{\text{fin}}$.
+Mutants arise with probability $\theta$; $M f_{\text{old}} K_{\text{old}}$ is the total number of wild-type individuals in old-habitat patches, $M (1 - f_{\text{old}}) \widehat{N}_w^{\text{new}}$ the total number of wild-type individuals in new-habitat patches.
+A mutant arising in a $k$-habitat patch has a probability $\varphi_{k}$ of establishing in the population; we assume that mutants establish independently of one another.
+ \linelabel{R2-34}Assuming a Poisson distribution for the number of successful mutant lineages is an approximation of a Binomial distribution with large sample size (the wild-type population size) and small success probability (the establishment probabilities $\varphi_{k}$).
+Note also that for $t_{\text{fin}}$ tending to infinity, there will almost surely be a successful mutant, so that $P_{\text{adapt}}=1$. 
 %For finite values of $t_{\text{fin}}$, the average number of mutations that appear until that time is given by the product of the mutation probability $\theta$, the length of the time window of interest $t_{\text{fin}}$, the number of patches $M$ and the stationary number of wild-type individuals present in the different habitats, that is $K$ and $\widehat{N}_w^{\text{new}}$ for old- and new-habitat patches, respectively (cf.
@@
 
-In Fig.~\ref{fig:source_sink} we compare our predictions to simulation results.
-The shape of the probability of adaptation as function of the emigration probability $m$ depends on the fecundity of the mutant in old-habitat patches, $\omega^\text{old}_m$ (Figs.~\ref{fig:source_sink}(a,c)).
-This is similar to the behavior of the establishment probability $\varphi_{\text{old}}$ in Fig.~\ref{fig:vary_m_est}.
-Likewise, the qualitative effects of the different dispersal schemes are comparable to the ones observed for the establishment probability.
-\linelabel{R1-33}\chg{This holds because the probability of adaptation, $P_{\text{adapt}}$, is given by $1-\exp(-C_1 \varphi_{\text{old}} -C_2 \varphi_{\text{new}})$ where two constants $C_1$ and $C_2$ depend on the dispersal scheme (see eq.~\eqref{eq:source_sink}).
-\linelabel{R2-35}The small differences in the ranking of the dispersal schemes between subfigures Figs.~\ref{fig:vary_m_est}(c) and Figs.~\ref{fig:source_sink}(c) can be explained by the mutational input, i.e.
-larger stationary populations sizes of the wild type in new habitats, $\widehat{N}_w^{\text{new}}$ result in more mutations during the observed time frame.
+The probability of adaptation $P_{\mathrm{adapt}}$ calculated with eq.~\ref{eq:source_sink} is compared to simulation results in Fig.~\ref{fig:source_sink}.
+In spite of our approximations, the fit to simulation results is still very good.
+\linelabel{R1-33}As $P_{\mathrm{adapt}}$ includes the probabilities of establishment $\varphi_{k}$, here again, the shapes of the curves as function of the emigration probability $m$ depend on the fecundity of the mutant in old-habitat patches, $\omega^\text{old}_m$ (Figs.~\ref{fig:source_sink_m-35}, \ref{fig:source_sink_m-45}).
+Likewise, the rankings of the different dispersal schemes are comparable to the ones observed for the establishment probability.
+\linelabel{R2-35}The small differences in the ranking of the dispersal schemes between panels~\ref{fig:vary_m_est_old-45} and \ref{fig:source_sink_m-45} can be explained by the mutational input.
+Larger stationary populations sizes of the wild type in new habitats, $\widehat{N}_w^{\text{new}}$ result in more mutations during the observed time frame.
 This population size is altered by the different dispersal schemes, most importantly it is largest for the wild type having a bias towards old habitats, i.e.
-the Old-Old and the Old-New dispersal schemes.}
+the Old-Old and the Old-New dispersal schemes \florence{It does not make intuitive sense why $\widehat{N}_w^{\text{new}}$ is higher with WT biased towards old...
+maybe then figures with $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ would help}.
 
@@
  		 \includegraphics[width=\linewidth]{fig3a.pdf}
-  		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.35$}
+  		\label{fig:source_sink_m-35}
 	\end{subfigure}%
@@
   		\caption{$\omega^\text{old}_m=1.35$}
+  		\label{fig:source_sink_f-35}
 	\end{subfigure}
@@
  		 \includegraphics[width=\linewidth]{fig3c.pdf}
-  		\caption{$\omega^\text{old}_m=1.45$  (small fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.45$ }
+  		\label{fig:source_sink_m-45}
 	\end{subfigure}%
@@
   		\caption{$\omega^\text{old}_m=1.45$}
+  		\label{fig:source_sink_f-45}
 	\end{subfigure}
@@
 
-In subfigures (b) and (d), we plot the probability of adaptation as a function of the frequency of old-habitat patches $f_{\text{old}}$.
-We observe a maximum which is the result of two effects: (1) the likelihood for a mutation to appear increases with the number of wild-type individuals present in the system, which is highest for high frequencies of old-habitat patches $f_{\text{old}}$, and (2) the probability of establishment of a mutant decreases with the number of old-habitat patches.
+Panels \ref{fig:source_sink_f-35} and \ref{fig:source_sink_f-45} show the probability of adaptation as a function of the frequency of old-habitat patches $f_{\text{old}}$.
+The maximum of $P_{\mathrm{adapt}}$ at intermediate  $f_{\text{old}}$ is the result of two antagonistic effects: (1) the likelihood for a mutation to appear increases with the number of wild-type individuals present in the system, which is highest for high frequencies of old-habitat patches $f_{\text{old}}$, and (2) the probability of establishment of a mutant decreases with the number of old-habitat patches.
 
 The different dispersal schemes alter both effects.
-\chg{A general bias towards the new habitat (New-New)}, when compared to \chg{unbiased} dispersal (0-0), always shifts the maximum to higher frequencies of old habitats while also increasing its quantitative value.
-\chg{Since for the New-New dispersal scheme, both types prefer settling in new-habitat patches ($\pi_w,\pi_m<0$), the local population before reproduction in these patches is increased.
-Thus, the overall population size is higher compared to the other dispersal schemes and therefore a higher number of mutants is generated under this scheme.} Additionally, the probability of establishment is also increased for \chg{the New-New} dispersal scheme, further increasing the probability of adaptation (see also the discussion around Fig.~\ref{fig:vary_m_est}).
-Again, a reversed argument explains why \chg{the general preference for old-habitat patches (Old-Old)} always yields lower probabilities of adaptation than \chg{unbiased} dispersal.
-
+The probability of adaptation is highest when there is a general bias towards the new habitat (New-New), due to a combination of high establishment probabilities (Fig.~\ref{fig:vary_m_est}) and high local population sizes thanks to the bias \florence{ref to figure of Nhatnew?}.
+Conversely, a general preference for old-habitat patches (Old-Old) yields lower probabilities of adaptation.
+
+\clearpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection*{Habitat of origin of the adaptive mutation}
-We now identify the habitat type where the successful mutation arises.
-The established mutant population often arises from a single mutant that is born in an old- or in a new-habitat patch.
-However, the rescued population can sometimes be traced back to two (or more) mutant individuals, \chg{at least} one from either patch type.
-\linelabel{R1-35}\chg{This is a special type of a soft selective sweep \citep[see][for a review]{hermisson_2017}.
-The approximate probability to observe a soft sweep from mutants from both habitat types is
+\subsubsection*{Habitat of origin of the adaptive mutation}
+We now ask in which habitat mutations leading to successful establishment appear.
+To do so, we distinguish in the simulations between mutants that appear in old-habitat patches and mutants that appear in new-habitat patches, and identify the habitat of origin of the mutation by considering the composition of the mutant population after establishment.
+We however do not distinguish between separate mutations that appear in the same type of habitat, meaning that we cannot rule out the presence of multiple lineages if they all appeared in the same type of patch: there may be \linelabel{R1-35}soft selective sweeps \citep[see][for a review]{hermisson_2017}.
+Only if the established mutant population contains mutants appeared in both old- and new-habitat patches can we be sure that multiple lineages contributed.
+
+Analytically, we approximate the probability to observe a mutant population that can be traced back to a mutant
+from an old-habitat patch by
 \begin{equation}\label{eq:origin}
     \begin{aligned}
-	&\mathbf{P}\big(\text{successful adaptation from old habitat}\big)\ \mathbf{P}\big(\text{successful adaptation from new habitat}\big) \\
-	& \qquad \qquad \approx \left(1 - \exp\left(-\theta t_{\text{fin}} M \varphi_{\text{old}} f_{\text{old}} K\right)\right)\left(1- \exp\left(-\theta t_{\text{fin}} M \varphi_{\text{new}} (1-f_{\text{old}}) \widehat{N}_w^{\text{new}}\right)\right)\, .
+	&\mathbb{P}\big(\text{successful adaptation from old habitat}\big)\  \left( 1 - \mathbb{P}\big(\text{successful adaptation from new habitat}\big) \right) \\
+	& \qquad \qquad \approx \left(1 - \exp\left(-\theta t_{\text{fin}} M \varphi_{\text{old}} f_{\text{old}} K\right)\right)\left(\exp\left(-\theta t_{\text{fin}} M \varphi_{\text{new}} (1-f_{\text{old}}) \widehat{N}_w^{\text{new}}\right)\right)\, .
 	\end{aligned}
 \end{equation}
-}
-\linelabel{R1-36}\chg{The approximation uses our basic assumption that different mutant individuals and consequently also the offspring of different mutants do not affect each others dynamics (branching process).
-In the simulations, we label a run as a soft sweep from mutants arising in different locations if these lineages are still alive after $1000$ generations.
-This ensures that we do not count any false-positives where a mutant in one of the habitats has just arisen right before the mutant population exceeds the establishment threshold.}
-
-%The probabilities for the other two scenarios can be computed analogously. 
-In Fig.~\ref{fig:origin} we compare simulation results with our predictions for the origin of a successful mutant when varying the old habitat frequency $f_{\text{old}}$.
-\chg{\linelabel{R2-36}We see that for both mutant fecundity values most successful mutations arise in old-habitat patches.
-The difference in the input from the two different patches is the factor $\varphi_{k} \widehat{N}_w^k f_k$.
-In Fig.~SX we plot the establishment probability $\varphi_k$ and the stationary population sizes $\widehat{N}_w^k$ when varying the frequency of old-habitat patches $f_{\text{old}}$.
-We see that the establishment probability for mutants arising in new-habitat patches, $\varphi_{\text{new}}$, is always larger than the corresponding probability for mutants from old-habitat patches, $\varphi_{\text{old}}$.
-Conversely, the stationary population size of the wild type in the old habitat measured over the whole environment is always larger then that of the wild type in new habitats.
-Combining these two observations we find that for this choice of parameters the mutational input, i.e.
-the absolute number of rescue mutants, has a larger influence on the origin of the rescue mutant than the corresponding establishment probability.}
+
+The corresponding probabilities for the other two scenarios can be computed analogously. 
+\linelabel{R1-36}The approximation uses our key assumption that different mutant individuals and their offspring do not affect each others dynamics (branching process).
+In the simulations, we label a run as having established lineages originated from different habitat types (``both'' in Fig.~\ref{fig:origin}) if these lineages are still alive after $1000$ generations.
+This ensures that we do not count any false-positives where a mutant in one of the habitats has just arisen right before the mutant population exceeds the establishment threshold.
+
+We compare our calculations to simulation results in Fig.~\ref{fig:origin}, varying the frequency of old-habitat patches $f_{\text{old}}$.
+\linelabel{R2-36}Most successful mutations arise in old-habitat patches.
+The contributions of old- vs.
+new-habitat patches depend on the product $\varphi_{k} \widehat{N}_w^k f_k$, which we decompose in Fig.~\ref{Sfig:vary_f_origin}.
+Even though the establishment probability from old-habitat patches is lower ($\varphi_{\text{old}} < \varphi_{\text{new}}$), the total population size of the wild type in the old-habitat patches is larger than that in new-habitat patches, so that more mutants appear in old-habitat patches than in new-habitat patches, which compensate their lower establishment probability.
 
@@
  		 \includegraphics[width=\linewidth]{fig4a.pdf}
-  		\caption{$w_m = 1.35$ (large fecundity difference)}
+  		\caption{$\omega_m^{\text{old}} = 1.35$ }
 	\end{subfigure}%
@@
   		\includegraphics[width=\linewidth]{fig4b.pdf}
-  		\caption{$w_m = 1.45$ (small fecundity difference)}
+  		\caption{$\omega_m^{\text{old}} = 1.45$}
 	\end{subfigure}
-	\caption{\textbf{Origin of the adaptive mutant.} \small The origin of the adaptive mutant is strongly affected by the fecundity differences in the old habitat.
-If the difference is large as illustrated in panel (a), mutants appear more often in old-habitat patches than in new-habitat patches.
-Still, mutants arising in new-habitat patches contribute to the overall probability of adaptation.
-If fecundity differences are small like in panel (b), the successful mutant largely arises in old-habitat patches.
-In this case, the contribution from new-habitat patches is negligible.
-Circles correspond to simulations identified as soft selective sweeps with (at least) one lineage \francois{successful mutation} arising in an old- and one in a new-habitat patch.
+	\caption{\textbf{Origin of the adaptive mutant,} \small depending on mutant fecundity in old-habitat patches $\omega_m^{\text{old}}$ (recall that $\omega_{w}^{\text{old}} = 1.5$).
+The points correspond to simulations, were mutants arising in old- vs.
+new-habitat patches are differently labeled, and were we consider the composition of the mutation population at the end of the simulation.
 The curves are given by eq.~\eqref{eq:origin} (or the adjusted versions of it) under the unbiased dispersal scheme ($\pi_w=\pi_m=0$).
@@
 \subsection*{Evolutionary rescue}
-Finally, we consider a time-inhomogeneous environment where patches deteriorate one after \chg{another} at regular time intervals $\tau$, until all patches have switched to the new habitat.
-If the wild-type population fails to generate a successful mutant, the population will \chg{inevitably} go extinct.
-The probability of evolutionary rescue is therefore tightly linked to the probabilities of adaptation and establishment that we have computed in eqs.~\eqref{eq:estab_approx} and~\eqref{eq:source_sink}.
+Finally, we assume that patches deteriorate one after another at regular time intervals $\tau$, until all patches have switched to the new habitat.
+If the wild-type population fails to generate a successful mutant, the population will inevitably go extinct, because the entire environment will consist of new-habitat patches, and because a wild-type population is assumed not to be viable there.
+%The probability of evolutionary rescue is tightly linked to the probabilities of adaptation and establishment that we have computed in eqs.~\eqref{eq:estab_approx} and~\eqref{eq:source_sink}.
 Typically, in formulas expressing the probability of evolutionary rescue, one splits the contributions into mutations arising \textit{de novo} and evolutionary rescue due to standing genetic variation, i.e.
@@
 We will discuss the effect of standing genetic variation in our system in the following section.
-For now, we focus on evolutionary rescue due to de novo mutations.
-We approximate the probability of evolutionary rescue, denoted by $P_{\text{rescue}}$, as
+We first focus on evolutionary rescue due to \textit{de novo} mutations. We approximate the probability of evolutionary rescue, denoted by $P_{\text{rescue}}$, as
 \begin{equation}\label{eq:evol_rescue}
@@
 \end{equation}
-where \linelabel{R1-40}$f_{\text{old}}(i) = (M-i-1)/M$ is the frequency of old-habitat patches after the $(i+1)-$th deterioration event\linelabel{R1-39}\chg{, the establishment probability is given as a function of the patch frequency, $\varphi_k(f_{\text{old}}(i))$,} and $N_w^{k}(j)$ denotes the overall number of wild-type individuals living in habitat $k$ (old or new) in generation~$j$ \chg{(see SI, Section S1 for an approximation)}.
-The interpretation of this equation is the same as for eq.~\eqref{eq:source_sink}.
-The only difference is that we now need to account for a changing environment, which alters the population sizes, $N_w^k$, and the establishment probabilities $\varphi_k$ \chg{over time}.
+where \linelabel{R1-40}$f_{\text{old}}(i) = (M-i-1)/M$ is the frequency of old-habitat patches after the $(i+1)^{\text{th}}$ deterioration event\linelabel{R1-39} ; the establishment probability is given as a function of the patch frequency, $\varphi_k(f_{\text{old}}(i))$, and $N_w^{k}(j)$ denotes the overall number of wild-type individuals living in habitat $k$ (old or new) in generation~$j$(see Appendix, Section~\ref{sec:app:WTpopsizeEnv} for an approximation).
+The interpretation of this equation is the same as for the probability of adaptation in eq.~\eqref{eq:source_sink}.
+The only difference is that we now need to account for a changing environment, which alters the population sizes, $N_w^k$, and the establishment probabilities $\varphi_k$ over time.
 In the formula, these changes are accounted for by the sums that iterate through the (discrete) time steps and by the time dependence of the corresponding quantities.
-We further note that we follow the \chg{expected value of} the wild-type population size deterministically over time, instead of assuming it to be already in its steady state as in eq.~\eqref{eq:source_sink} \chg{(see also Section \ref{sec:app:modeldyn} in the SI)}.
-
-As visible in Fig.~\ref{fig:rescue}, the approximation explains the \linelabel{R1-42}\chg{order of dispersal schemes}.
-Yet, it does not accurately predict the simulated data.
-This discrepancy can be explained: in the formula we \linelabel{R2-42}\chg{assume that for a mutant born in a certain patch configuration, say with $j$ old-habitat patches, the environment does not change anymore.
-That is, a mutant born in patch-type $k$ (old or new) in this environment contributes $\varphi_{k}(j)$ to the probability of evolutionary rescue despite further patches deteriorating.
-Thus, this mutants probability of establishment is underestimated.
+We further note that we follow the expected value of the wild-type population size deterministically over time, instead of assuming it to be already in its steady state as in eq.~\eqref{eq:source_sink} \chg{(see also Section \ref{sec:app:modeldyn} in the SI)}\florence{but this is still the stationary population size in phiold ?
+-- I'd remove this sentence as it is confusing (but would like to learn why not the stationary pop size)}.
+
+Comparison to simulated data indicates that the approximation in eq.~\eqref{eq:evol_rescue} correctly predicts the \linelabel{R1-42} ranking of dispersal schemes; the actual fit to data is however less good that for the previous steps of our analysis.
+This discrepancy can be explained: our analysis \linelabel{R2-42}assumes that for a mutant born in a certain patch configuration, say with $j$ old-habitat patches, the environment does not change anymore.
+That is, a mutant born in a $k$-habitat patch in this environment contributes $\varphi_{k}(j)$ to the probability of evolutionary rescue despite further patches deteriorating -- while more new-habitat patches increases the probability of establishment\footnote{\florence{this calls for plots of phinew and phiold as functions of fold}}.
+Thus, the probability of establishment is underestimated.
 This is especially true for mutants that emerge just before a deterioration event.
-These mutants are more likely to establish during the subsequent environmental configuration giving them a higher probability of survival than approximated by our formula.
-Additionally, $\varphi_k(j)$ assumes stationary wild-type population sizes and therefore does not reflect the decreasing wild-type population size right after the deterioration of a patch.}
+Additionally, $\varphi_k(j)$ assumes stationary wild-type population sizes and therefore does not reflect the decreasing wild-type population size right after the deterioration of a patch \florence{but the pop size depend on fold, don't they?
+it seems that this then only matters for about one generation in the just deteriorated patch?}.
 %However, mutants that arise very shortly before a deterioration event just need to survive until this event happens and are thus carried over to the following environmental configuration with a higher probability than predicted by our formula.
-This explains why the simulation results are higher than our approximation. 
+%This explains why the simulation results are higher than our approximation. 
 %which ignores these ``left-over'' mutants present prior to the deterioration event. 
-\chg{A time-dependent establishment probability could account for these effects but unfortunately is not amenable to approximations in our framework.}
+A time-dependent establishment probability could account for these effects but unfortunately is not amenable to approximations in our framework.
 %As a result, instead of assuming a constant establishment probability between two deterioration events, we should rather use a time-dependent establishment probability -- which can however not be approximated in our general framework.
-\linelabel{R1-43}\chg{In \citet{uecker_2014}, scenarios with an accessible time-dependent solution were studied, more precisely situations with either full mixing of the global population ($m=1$) or a non-viable mutant in old-habitat patches ($\omega^\text{old}_m=0$).}
+\linelabel{R1-43}\Citet{uecker_2014} considered scenarios with an accessible time-dependent solution: situations with either full mixing of the global population ($m=1$) or a non-viable mutant in old-habitat patches ($\omega^\text{old}_m=0$).
 
@@
   		\includegraphics[width=\linewidth]{fig5a.pdf}
-  		\caption{$\omega^\text{old}_m=1.45$ (small fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.45$}
+  		\label{fig:rescue_45}
 	\end{subfigure}%
@@
   		\includegraphics[width=\linewidth]{fig5b.pdf}
-  		\caption{$\omega^\text{old}_m=1.35$ (large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.35$}
+  		\label{fig:rescue_35}
 	\end{subfigure}
@@
   		\includegraphics[width=\linewidth]{fig5c.pdf}
-  		\caption{$\omega^\text{old}_m=1.1$ (very large fecundity difference)}
+  		\caption{$\omega^\text{old}_m=1.1$}
+  		\label{fig:rescue_1}
 	\end{subfigure}%
@@
   		\caption{$\omega^\text{old}_m=0$ (lethal mutant)}
+  		\label{fig:rescue_lethal}
 	\end{subfigure}
@@
 Quantitatively though, we find that our predictions tend to underestimate the simulated data.
-In (a,b) the mutation probability is set to $u=1/(25MK_{\text{new}})$ while in (c,d) it is $u=1/(MK_{\text{new}})$.
+In (a,b) the mutation probability is set to \florence{$\theta$ instead of $u$?} $u=1/(25MK_{\text{new}})$ while in (c,d) it is \florence{$\theta$ instead of $u$?} $u=1/(MK_{\text{new}})$.
 The label `exact' refers to the exact solution of eq.~\eqref{eq:ext_prob} which is then plugged into the approximation of the probability of evolutionary rescue in eq.~\eqref{eq:evol_rescue}.}
@@
 
-The different dispersal schemes exhibit substantial differences.
-The dispersal schemes affect the dispersal pattern of both types and as such alter their respective population dynamics.
-Both small and large fecundity differences in the old habitat, $\omega^\text{old}_m=1.45$ and $\omega^\text{old}_m=1.35$, conserve the ranking of dispersal schemes obtained for the probabilities of establishment and adaptation\footnote{\francois{difficult to follow}\pete{We could also get rid of the explict description of the figure and remove the following sentence}}: \linelabel{R2-43}\chg{New-New (green) is higher than unbiased dispersal (black) which is higher than Old-Old (blue), with Old-New decreasing and New-Old increasing their positions with increasing emigration probability $m$}, cf.
-Fig.~\ref{fig:rescue}(a,b).
-Based on our discussion of the same behavior in Fig.~\ref{fig:vary_m_est}(a,c), \linelabel{R2-44}\chg{i.e.
-the change in hierarchy of the dispersal schemes Old-New and New-Old,} The most influential factor in these parameter sets is the growth rate of the mutant in old-habitat patches, $a_{\text{old}}$, \chg{as discussed around Fig.~\ref{fig:vary_m_est}.} \chg{The ranking of the symmetric dispersal schemes is mostly determined by the mutant dispersal bias.}
-
-For very large fecundity differences in the old habitat ($\omega^\text{old}_m = 1.1$ and $\omega^\text{old}_m=0$), the ranking of dispersal schemes is the same for all emigration probabilities; from highest to lowest, we find \chg{New-New, Old-New, unbiased dispersal, New-Old, and Old-Old} (cf.
-Fig.~\ref{fig:rescue}(c,d)).
-In this parameter setting, the probability of evolutionary rescue is dominated by the dispersal behavior of the mutant.
-Since the mutant is barely viable in old-habitat patches, it is always preferential for it to disperse towards new-habitat patches.
-Two schemes follow this rule, \chg{the New-New and Old-New dispersal scheme}, but differ in the preference of the wild type.
-Since wild-type individuals also preferentially disperse to new-habitat patches under \chg{the New-New scheme}, more individuals are present in those patches.
-Therefore, the total amount of mutations over the deterioration time is increased which generates more mutants on average.
-This explains the New-New dispersal scheme being consistently higher than the Old-New dispersal scheme.
-%This effect is even stronger for large dispersal rates that increase the wild-type population size in the whole environment.
-Together with the effect of relaxed competition, this explains the strong increase of the \chg{New-Old}, and the \chg{unbiased} dispersal scheme for large emigration probabilities $m$ in Fig.~\ref{fig:rescue}(c).
-
-Lastly, the probability of evolutionary rescue often reaches a local (or global) maximum for intermediate emigration probabilities (Figs.~\ref{fig:rescue}(c,d)).
-This extends previous results \citep{uecker_2014,tomasini_2019} to arbitrary dispersal schemes affecting the immigration process.
-The maximum can be attributed to the interaction of the three regions we identified when analyzing the establishment probability $\varphi_k$ (see also Fig.~\ref{fig:vary_m_est}).
-As such it is a result of the largely positive effect of dispersal (initial increase of the probability of evolutionary rescue) and the negative effect of population mixing \chg{(dispersal to old patches)}.
-\footnote{\francois{This discussion is still difficult.}}
-
-\subsection*{Habitat of origin of the rescue mutant and standing genetic variation}
-Similar to what \chg{we} found for the probability of adaptation, rescue mutants mainly originate from old-habitat patches.
+The ranking of the different dispersal scheme is overall conserved from the previous steps of our analysis (Fig.~\ref{fig:source_sink}).
+Differences between the dispersal schemes are more marked when the fecundity of the mutant in old-habitat patches is lower (Fig.~\ref{fig:rescue_1},\subref{fig:rescue_lethal}), including when the mutant cannot reproduce at all in old-habitat patches ($\omega^\text{old}_m=0$).
+It is comparatively better for rescue that the mutant preferentially disperses into new-habitat patches, where it is fitter, and for the wild-type to also preferentially disperse into new-habitat patches, thereby releasing competition in old-habitat patches.
+
+When mutant fecundity in old-habitat patches is comparatively low (Fig.~\ref{fig:rescue_1},\subref{fig:rescue_lethal}), the probability of evolutionary rescue often reaches a local (or global) maximum at intermediate emigration probabilities.
+This finding extends previous results \citep{uecker_2014, tomasini_2019} to arbitrary dispersal schemes affecting the immigration process.
+
+\subsubsection*{Habitat of origin of the rescue mutant and standing genetic variation}
+Similar to what we found for the probability of adaptation, rescue mutants mainly originate from old-habitat patches (Fig.~\ref{fig:sgv_origin}).
 Mutations are more likely to appear in the more populated patches (old-habitat).
-However, a low mutant fecundity $\omega^\text{old}_m$ decreases the chance of establishment of these mutants that appear in old-habitat patches (compare black and yellow symbols in Fig.~\ref{fig:sgv}(a)).
-Lower fecundity values $\omega^\text{old}_m$ therefore comparatively increase the probability for the rescue mutation to have appeared in the already deteriorated habitat, i.e.
-in new-habitat patches.
-\linelabel{R1-35-2}\chg{The probability of a soft selective sweep for evolutionary rescue from mutants of the different habitat types is very low (circles in Fig.~\ref{fig:sgv}(a)).
-Our choice of a small mutation rate implies a hard selective sweep regime ($\theta K_{\text{old}} M = 0.08<1$) \citep{wilson_2018,hermisson_2017}.
-}
+However, a low mutant fecundity in old-habitat patches $\omega^\text{old}_m$ decreases the chance of establishment of these mutants that appear in old-habitat patches (compare black and yellow symbols in Fig.~\ref{fig:sgv_origin}).
+Here again, we cannot rule out that multiple mutant lineages having appeared in the same habitat type established.
+\linelabel{R1-35-2} Only when mutants from both habitat types are present can we be sure that at least two lineages contributed to establishment (i.e., there was a soft sweep).
+In this case, the probability of rescue with a mix of origins was very low in our simulations (circles in Fig.~\ref{fig:sgv_origin}).
+Note that our choice of a small mutation rate implies a hard selective sweep regime ($\theta K_{\text{old}} M = 0.08<1$) \citep{wilson_2018, hermisson_2017}.
 
@@
  		 \includegraphics[width=\linewidth]{fig6a.pdf}
-  		\caption{}
+  		\caption{Habitat of origin of the rescue mutants}
+  		\label{fig:sgv_origin}
 	\end{subfigure}%
@@
  		 \includegraphics[width=\linewidth]{fig6b.pdf}
-  		\caption{}
+  		\caption{Contribution of de novo mutations compared to standing genetic variation}
+  		\label{fig:sgv_sgv}
 	\end{subfigure}
-	\caption{\textbf{Habitat of origin of the rescue mutation and the impact of standing genetic variation.} \small (a) We compare the source of successful mutations under small (black) and large (yellow) fecundity differences in the old habitat.
-Decreasing the fecundity of the mutant results in more successful mutations emerging in new-habitat patches ($+$) when compared to the contribution from old-habitat patches ($\times$).
-We have chosen $\pi_m=\pi_w=0$.
+	\caption{\textbf{Habitat of origin of the rescue mutation and the impact of standing genetic variation.} \small (a) We compare the source of successful mutations for different mutant fecundity in the old-habitat patches  (black vs.
+yellow).
+Decreasing the fecundity of the mutant results in more successful mutations emerging in new-habitat patches ($+$) compared to the contribution from old-habitat patches ($\times$).
+Dispersal is unbiased ($\pi_m=\pi_w=0$).
 (b) For slower environmental degradation, i.e.
-$\tau=200$, the influence of standing genetic variation (sgv) on the probability of evolutionary rescue decreases.
+$\tau=200$\florence{this value does not appear in the figure?}, the influence of standing genetic variation (sgv) on the probability of evolutionary rescue decreases.
 The simulations are done by letting the system evolve for $1,000$ generations before the first deterioration event happens.
@@
 is there a difference?
-why use one over the other?}}, we ran simulations \linelabel{R1-45}where we let the system evolve for $1,000$ generations before the first degradation event happened.
-Fig.~\ref{fig:sgv}(b) shows the relative contribution of de novo mutations and of standing genetic variation, i.e.
+why use one over the other?}\florence{it is variation}}, we ran simulations \linelabel{R1-45}where we let the system evolve for $1,000$ generations before the first degradation event happened.
+Fig.~\ref{fig:sgv_sgv} shows the relative contribution of de novo mutations and of standing genetic variation, i.e.
 mutations that appeared before the first degradation event happened.
-\linelabel{R2-39}\chg{For a successful rescue event due to standing genetic variation, mutants that were initially present (at time $t=0$) need to survive at least until sufficiently many patches have deteriorated that the probability of adaptation, $P_{\text{adapt}}$, becomes positive, compare to Fig.~\ref{fig:source_sink}(b,d) and Fig.~SX in the SI.} \linelabel{R1-46}\chg{Decreasing the value of $\tau$ reduces the time between two consecutive degradation events and the overall time-span of the entire environmental change.
-Therefore, the impact of standing genetic variation on the probability of evolutionary rescue increases for smaller values of $\tau$.}
-
-Additionally, the relative contribution of standing genetic variation declines as the emigration rate $m$ increases. 
-This holds because \linelabel{R1-49}\chg{the mutant is almost exclusively found in old-habitat patches if the frequency of these patches, $f_{\text{old}}$, and emigration rates $m$ are high} (see Fig.~SX in SI).
+\linelabel{R2-39}For a successful rescue event due to standing genetic variation, mutants that were initially present (at time $t=0$) need to survive at least until sufficiently many patches have deteriorated that the probability of adaptation, $P_{\text{adapt}}$, becomes non zero, compare to Fig.~\ref{fig:source_sink}(b,d) and Fig.~SX\florence{which figure?} in the SI.
+\linelabel{R1-46}Reducing the time between two consecutive degradation events (reducing $\tau$) also reduces the overall time span of the entire environmental change.
+Therefore, the proportion of rescue mutants due to standing genetic variation increases as $\tau$ decreases.
+
+The relative contribution of standing genetic variation declines as the emigration rate $m$ increases.
+With high $m$ and a high proportion of old-habitat patches $f_{\text{old}}$, mutants are \linelabel{R1-49} almost exclusively found in old-habitat patches (see Fig.~SX in SI \florence{which figure?}).
 Thus, mutants that existed prior to the first deterioration event are very unlikely to survive even for a rapidly changing environment.
+\florence{I do not understand what you mean and could not rephrase this paragraph.
+Missing a logical step-- how come is it detrimental for the mutant to be in old-habitat patches (higher estab proba)}
 
@@
 % Brief summary
-We have studied the probabilities of establishment, adaptation and evolutionary rescue under four non-\chg{uniform} dispersal schemes and compared them to \chg{unbiased} dispersal.
-Our analysis builds on the probability of establishment of a single mutant lineage in a heterogeneous environment with a fixed patch configuration.
+We have studied the probabilities of establishment, adaptation and evolutionary rescue under four non-uniform dispersal schemes and compared them to unbiased dispersal.
 In line with previous results, we find that the probabilities of establishment, adaptation and evolutionary rescue can display up to three different phases when varying the dispersal rate $m$.
-The different dispersal schemes, by changing the population dynamics, \chg{alter the parameter range of these regions.}\footnote{\francois{Unclear.
-Suggestion (not great but a bit better maybe): "The dispersal scheme determines the population dynamics and consequently the parameter regions corresponding to the three phases."}}
-%broaden, tighten or shift these regions.
+The dispersal schemes affect population dynamics and consequently the parameter regions corresponding to the three phases.
 
@@
 % General comparison to other models and our resulting formula
-Theoretical studies that investigated the effects of spatial subdivision on the adaptation of a population in a heterogeneous environment can be classified into two types.
-One type of models, classically analyzed in a population genetic context\francois{framework}, assumes constant population sizes in all patches, independent of their local habitat type and of dispersal strength.
+Theoretical studies that investigated the effects of spatial subdivision on the adaptation of a population in a heterogeneous environment can be classified into two types, depending on how they treat demography.
+One type of models, classically analyzed in a population genetic framework, assumes constant population sizes in all patches, independent of their local habitat type and of dispersal strength (which we later call implicit demography).
 Results obtained in this framework show that larger dispersal rates tend to decrease the probability of successful establishment of a rare mutant favored in some part of the environment \citep[e.g.][]{garcia_1997}.
-This \st{inhibiting effect of dispersal on adaptation, also} \francois{effect} termed ``gene swamping'', is a result of an increase in absolute numbers of non-adapted individuals in the habitat where the rare mutant is beneficial \linelabel{R1-51}\chg{before reproduction and density regulation.
-This results in a lower mutant frequency \citep{lenormand_2002,tomasini_2018} and thus in lower reproductive success.} Additionally, for very high dispersal rates, the population homogenizes and individuals encounter an averaged environment.
-Therefore, the type with the largest overall growth rate, averaged over the environment, is favored.
-\francois{In our model, gene swamping was rarely observed; the dominant effects were explained by the impact of dispersal on demography.} \footnote{\francois{Would consider shortening this paragraph which is not directly relevant.
-e.g.
-remove the sentence "this results in a lower mutant frequency...", but keep the references}}.
-
-The second type of models explicitly takes into account demographic effects due to dispersal, often in the context of source-sink systems \citep{holt_1985,pulliam_1988}. 
-Here, the effect of dispersal on adaptation depends on the growth rate differences of the mutant and the wild type in the two habitats \citep{kawecki_2000}. 
-In accordance with this result \francois{these results}, we find that dispersal \francois{monotonically} increases the probability of adaptation if the mutant is just slightly less fit than the wild type in the old habitat %, $\omega^\text{old}_m / \omega^\text{old}_w= 0.99$
-(Fig.~\ref{fig:vary_m_est}(c)). 
-When the mutant's fecundity in old habitats, $\omega^\text{old}_m$, is smaller \francois{than what?}, establishment probabilities are non-monotonic with a local maximum at intermediate dispersal rates \francois{and increases at large dispersal rates thanks to relaxed competition} (Fig.~\ref{fig:vary_m_est}(a)).
-\st{However, if the mutant's disadvantage in old-habitat patches is even greater, i.e.
-$\omega^\text{old}_m$} \francois{When the fecundity of the mutant is even smaller, the local maximum remains but relaxed competition no longer occurs} \st{adaptation is also hindered for large dispersal rates.
-The establishment probability becomes hump-shaped } (cf.
-Fig.~S2 in SI).
-\footnote{\francois{I was confused by this paragraph so rephrased to make it more uniform and better describe the figures and the key differences}}.
-
-\st{Hence, models with and without demographic change differ most at higher emigration rates.
-This corresponds to region (iii) in our analysis; we illustrate this point in} \francois{Compared to models with implicit demography, explicitly modelling the population dynamics causes relaxed competition that increases the probability of establishment at high dispersal rates }Fig.~S5 in SI, the non-demographic version of Fig.~\ref{fig:vary_m_est}.
-
+This ``gene swamping'' effect is due to the dispersal of non-adapted individuals in the habitat where the rare mutant is beneficial \linelabel{R1-51}, decreasing the local frequency of the mutant \citep{lenormand_2002,tomasini_2018}.
+In our model, gene swamping was rarely observed; the dominant effects were explained by the impact of dispersal on demography\footnote{\florence{not sure we should keep this last sentence}}.
+
+The second type of models explicitly takes into account demographic effects due to dispersal, often in the context of source-sink systems \citep{holt_1985, pulliam_1988}. 
+Here, the effect of dispersal on adaptation depends on the growth rate differences of the mutant and the wild type in the two habitats \citep{kawecki_2000}, which we also observe.
+When the mutant is just slightly less fit than the wild type in the old habitat (Fig.~\ref{fig:vary_m_est_old-45}), the probability of adaptation monotonically increases with dispersal.
+When the mutant's fecundity is lower, however, establishment probabilities reach a local maximum at intermediate dispersal rates, and increase again at large dispersal rates thanks to relaxed competition (Fig.~\ref{fig:vary_m_est_old-35}(a)).
+When the fecundity of the mutant is even smaller, the local maximum remains but relaxed competition no longer occurs (cf.
+Fig.~\ref{Sfig:weak_fecundity}).
+
+Compared to models with implicit demography, explicitly modelling the population dynamics allows for relaxed competition at high dispersal rates, which increases the probability of establishment.
+To illustrate this point, Fig.~\ref{Sfig:pop_gen} is the non-demographic version of Fig.~\ref{fig:vary_m_est} (all patches are always at carrying capacity in Fig.~\ref{Sfig:pop_gen}).
 \francois{In figure S5, there does seem to have relaxed competition for curves orange and purple.
@@
 Also I think a more balanced discussion of the three phases: initial rise thanks to exportation of mut in new habitats; decline due to back-migration; final increase due to relaxed competition, and which effect appears or not in the pop gen model, would be nice.
-Right now we focus a lot on relaxed competition.}
-
-\subsection*{\linelabel{R1-52}\chg{Standing genetic variation} and evolutionary rescue}
-
-\st{Besides the general structure of the probability of evolutionary rescue when varying the emigration probability $m$, }we also studied the effect \francois{contribution} of standing genetic variation on this quantity \francois{to evolutionary rescue}.
-We find that the contribution of standing genetic variation on the overall probability of evolutionary rescue increases with the speed of the environmental change, determined by $\tau$ (Fig.~\ref{fig:sgv}(b)).
+Right now we focus a lot on relaxed competition.}\florence{I am also confused by the increases in orange and purple.
+}
+
+\subsection*{\linelabel{R1-52}Standing genetic variation and evolutionary rescue}
+
+We also studied the contribution of standing genetic variation to evolutionary rescue.
+This contribution increases with the speed of environmental change (i.e., with smaller interval between degradations $\tau$), (Fig.~\ref{fig:sgv_sgv}).
 This observation has also been made in a quantitative genetics setting where the adaptive trait is continuous (and not discrete as in our model) \citep{matuszewski_2015}.
-Experimental results with \textit{Caenorhabditis elegans} indicate that under slow environmental change the impact of standing genetic variation is small \citep{guzella_2018}.
-This is because the evolution of the trait is driven by \textit{de novo} mutations with small effects.
-For a sudden or fast environmental change (small $\tau$), standing genetic variation becomes increasingly important for the probability of evolutionary rescue.
-These observations are in line with our findings in Fig.~\ref{fig:sgv}(b).
-%and other theoretical results obtained in this context \citep{matuszewski_2015}.
-
-\subsection*{The effect of \chg{biased} dispersal patterns on adaptation and evolutionary rescue}
-The importance of considering dispersal schemes \chg{other than} \chg{unbiased} dispersal has been highlighted in several papers \citep{edelaar_2008,clobert_2009,edelaar_2012}.
-This has led to a number of simulation studies exploring the effect of various dispersal schemes onto \chg{(local)} adaptation %and niche width
+Experimental results with \textit{Caenorhabditis elegans} also indicate that the impact of standing genetic variation is small under slow environmental change\citep{guzella_2018}.
+
+\subsection*{The effect of biased dispersal patterns on adaptation and evolutionary rescue}
+The importance of considering dispersal schemes other than unbiased dispersal has been highlighted in several papers \citep{edelaar_2008,clobert_2009,edelaar_2012}.
+This has led to a number of simulation studies exploring the effect of various dispersal schemes on (local) adaptation %and niche width
 \citep[e.g.][]{vuilleumier_2010,holt_2015,mortier_2018,pellerin_2018}.
 
-Two of these simulation studies examined the effects of matching habitat choice on adaptation \chg{in} a heterogeneous environment \citep{vuilleumier_2010,holt_2015}.
-Both investigations indicate that matching habitat choice increases the probability of adaptation when compared to \chg{unbiased} dispersal.
-This is in line with our findings: we predict that type-dependent habitat choice \chg{where each type favors patches they are relatively more fit in (Old-New),} generates higher probabilities of establishment \chg{(}and evolutionary rescue\chg{)} than \chg{unbiased} dispersal \chg{(0-0)} (Figs.~\ref{fig:vary_m_est}, \ref{fig:source_sink}, \ref{fig:rescue}).
-
-The dispersal schemes also slightly affect the origin of the successful mutant lineage, cf.
-SI, Fig.~S6.
-Population densities, especially in new-habitat patches, are altered when compared to the \chg{unbiased} dispersal scheme.\footnote{\pete{This paragraph could be dropped in line with a comment of a reviewer.} \francois{agree to drop it}}
-
-To conclude, the effects of dispersal schemes are two-fold.
-By changing population densities in both habitat types, the dispersal schemes \chg{alter} the growth rate of the mutant in old-habitat patches.
-This is the primary reason for the ranking of the dispersal schemes.
-In addition, they also affect the number of mutations arising in either habitat type.
-This has a minor effect on the probability of evolutionary rescue for the explored parameter range but is relevant when studying the origin of the successful mutant\florence{mutation} lineage \chg{(see also SI, Fig.~S6)}.
-\francois{As the genetic background may vary across patches, the origin of a successful mutation will also affect which neutral and deleterious mutations will hitchhike with it}.
-\linelabel{R1-54}\chg{The origin of a successful mutant is of special interest once multiple loci are considered\francois{.
-Rescue by a single mutant implies} \st{which might promote} neutral (or even deleterious) hitchhiking effects.
-Similarly, in the case of polygenic rescue or under recombination \citep[e.g.][]{uecker_2015}, the origin of a mutant is likely to affect its success.}\florence{[mention genetic background]}
+Two of these simulation studies examined the effects of matching habitat choice on adaptation in a heterogeneous environment \citep{vuilleumier_2010,holt_2015}.
+Both investigations indicate that matching habitat choice increases the probability of adaptation compared to unbiased dispersal.
+This is in line with our findings: when each type favors patches where they are relatively fitter (Old-New scheme), probabilities of establishment and evolutionary rescue are higher than under unbiased dispersal (0-0) (Figs.~\ref{fig:vary_m_est}, \ref{fig:source_sink}, \ref{fig:rescue}).
+\florence{I kept the sentence in case I missed something, but this does not seem to be the case??
+orange can be below black}
+
+We identified two types of effects of the different dispersal schemes.
+First, by changing population densities in both habitat types, the dispersal schemes alter the growth rate of the mutant in old-habitat patches $a_{\text{old}}$.
+This is the primary reason for the ranking of the dispersal schemes\footnote{\florence{really need a figure!}}.
+Secondly, the different dispersal schemes affect the number of mutations arising in either habitat type.
+This has a minor effect on the probability of evolutionary rescue for the explored parameter range but is relevant when studying the origin of the successful mutant lineage \chg{(see also Fig.~\label{Sfig:natal_habitat})}.
+As the genetic background may vary across patches, the origin of a successful mutation will also affect which mutations will hitchhike with it.
+\linelabel{R1-54}Similarly, in the case of polygenic rescue or under recombination \citep[e.g.][]{uecker_2015}, the origin of a mutant is likely to affect its success.
 
@@
 \subsection*{Generality of our theoretical analysis and future directions}
-Our mathematical results apply in the case where the mutant offspring numbers can be written in the form ``migration times reproduction (and potential regulation)'' with rates that are constant in time (\linelabel{R1-58-1}\chg{or as in case of evolutionary rescue, constant between two deterioration events}; see the mean reproduction matrix in eq.~\eqref{eq:mean_repro}).
-Biologically, this means that the resident population is stationary and the mutant is either at low numbers or unaffected by its own density.
+
+Our mathematical results rely on the simplifications that the wild type population does not change over time and that the mutant is rare enough that mutants live independently of each other, and do not affect wild-type individuals.
+This allows us to summarize mutant population dynamics with the $\lambda$ terms presented in eq.\eqref{eq:lambda}.
 Furthermore, for our approximation in eq.
-\eqref{eq:estab_approx} to generate accurate predictions, it is essential that growth rate differences between the wild type and the mutant are weak and dispersal is low.
-Formally, just two of these parameters need to be small -- see also the corresponding discussion in \citet{tomasini_2018}.
-
-Since we summarize the population dynamics in our parameters of the reproduction matrix, the approach taken here is quite general and can account for various dispersal schemes and local \chg{type-dependent} population dynamics, \linelabel{AE-3-1}\chg{i.e.
-different reproduction and competitive parameters.} However, it cannot account for \chg{type-dependent carrying capacities, explicit spatial structure %as for example the stepping stone model.
+\eqref{eq:estab_approx} to generate accurate predictions, it is essential that growth rate differences between the wild type and the mutant are weak and dispersal is low -- these conditions are however not needed when we numerically solve system~\eqref{eq:ext_prob}.
+Formally, just two of these parameters need to be small, as discussed also in \citet{tomasini_2018}.
+
+Our approach can account for various dispersal schemes and local type-dependent population dynamics, \linelabel{AE-3-1}i.e.
+different reproduction and competitive parameters.
+However, it cannot account for type-dependent carrying capacities, explicit spatial structure %as for example the stepping stone model.
 It also cannot account for
-or \st{time-inhomogeneous} \francois{rapidly changing} environments.} \linelabel{R1-58-2} The \chg{latter} is the reason for our \chg{poor}\footnote{\pete{any better word here?}} approximation in the context of evolutionary rescue (Fig.~\ref{fig:rescue})\footnote{\francois{I would remove this sentence}}.
-Additionally, in order to obtain analytical solutions, it is important that the stationary population sizes of the wild type have an accessible solution.
-This is not the case if we consider non-linear emigration rates that depend on habitat choice like those incorporated in some simulation studies \citep[e.g][]{holt_2015,mortier_2018}.
-
-In contrast, it \chg{is} possible to \chg{implement} a cost of dispersal \chg{or} a different life cycle.
-In particular, the variation of the life cycle could yield \chg{distinct} results regarding adaptation \citep{holt_2015} and, more generally, in the context of the evolution of dispersal \citep{massol_2015}.
+or rapidly changing environments.
+\linelabel{R1-58-2} The latter is the reason for our less accurate approximate in the context of evolutionary rescue (Fig.~\ref{fig:rescue})\footnote{\francois{I would remove this sentence}\florence{I am ok with the sentence}}.
+Additionally, in order to obtain analytical solutions\florence{but it is still possible if we do things numerically, right?
+if so, I would remove the two following sentences}, it is important that the stationary population sizes of the wild type have an accessible solution.
+This is not the case if we consider non-linear emigration rates that depend on habitat choice like those incorporated in some simulation studies \citep[e.g][]{holt_2015, mortier_2018}.
+
+Our model can readily be extended and include a cost of dispersal or a different life cycle.
+In particular, the variation of the life cycle could yield distinct results regarding adaptation \citep{holt_2015}\footnote{\florence{+ cite Debarre and Gandon hard soft selection please :-)}} and, more generally, in the context of the evolution of dispersal \citep{massol_2015}.
 \\
 
+\subsection*{Conclusion}
 In conclusion, we studied the effect of dispersal and different dispersal schemes on the probability of establishment, adaptation and evolutionary rescue of a mutant under divergent selection in a subdivided population.
-Our quantitative approach disentangles the interaction of dispersal and adaptation.
-We recover previous results on adaptation and provide a general framework for studying evolutionary dynamics of a subdivided population in heterogeneous environments \chg{in discrete time}.
-This unifying approach allows us to identify the forces responsible for the different predictions obtained in the population genetics literature and under source-sink dynamics, respectively.
+We recover previous results on adaptation and provide a general framework for studying evolutionary dynamics of a subdivided population in heterogeneous environments in discrete time.
+This unifying approach allows us to identify the forces responsible for the different predictions obtained in the population genetics literature and under source-sink dynamics.
 We find that including population demography significantly alters the results for high dispersal rates.
-For constant population sizes, high dispersal rates have a negative effect on establishment, while with explicit demography the effect is largely positive.
+For constant population sizes, high dispersal rates have a negative effect on establishment, while with explicit demography the effect is largely positive\footnote{\florence{reconsider these sentences given the results in S5??}}.
 The latter is a result of relaxed competition in old-habitat patches.
@@
 
-\subsection*{Stationary wild-type population sizes}
+\subsection{Stationary wild-type population sizes}
 We denote by $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ the deterministic stationary population sizes of the wild type in old- and new-habitat patches.
@@
 This value cannot be larger than $K_{\text{new}}$, the carrying capacity of new-habitat patches. So, 
-\begin{equation}\label{Seq:Nhatnew}
-\widehat{N}_w^{\text{new}} = \min\left(K_{\text{new}},  \frac{m \omega_w^{\text{new}} f_{\text{old}}  K_{\text{old}}}{
-	1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} - \omega_w^{\text{new}} (1-f_{\text{old}} + \widehat{\pi}_w f_{\text{old}} (1-m))
-} \right).
-\end{equation}
-
-\subsection*{Wild-type population sizes after dispersal}
+\begin{subequations}\label{Seq:Nhat}
+	\begin{align}
+	\widehat{N}_w^{\text{old}} &= K_{\text{old}} \\
+	%
+	\widehat{N}_w^{\text{new}} &= \min\left(K_{\text{new}},  \frac{m \omega_w^{\text{new}} f_{\text{old}}  K_{\text{old}}}{
+	1-f_{\text{old}} +  e^{\pi_w} f_{\text{old}} - \omega_w^{\text{new}} (1-f_{\text{old}} + e^{\pi_w} f_{\text{old}} (1-m))
+} \right).\label{Seq:Nhatnew}
+	\end{align}
+\end{subequations}
+
+\subsection{Wild-type population sizes after dispersal}
 
@@
 
-\subsection*{Wild-type population sizes during the environmental change}
+\subsection{Wild-type population sizes during the environmental change \label{sec:app:WTpopsizeEnv}}
+\florence{I suggest moving this subsection at the end of Appendix A}
 Lastly, compute the (deterministic) wild-type population size over time during the environmental change.
@@
 
-\subsection*{The local per capita growth rate $a_{\text{old}}$}
+\subsection{Local per capita growth rates}
+\subsubsection{The local per capita growth rate $a_{\text{old}}$}
 As stated in eq.~\eqref{eq:sold} in the main text, we define the per capita growth rate of rare mutant in the old habitat by
@@
 %a_{\text{old}} = \left\{ \begin{array}{ll}
-%\dfrac{\omega^\text{old}_m K_{\text{old}}}{\omega^\text{old}_w \left(\frac{(1-f_{\text{old}})m\widehat{\pi}_w K_{\text{new}} + (1-m-f_{\text{old}}(1-m-\widehat{\pi}_w))K_{\text{old}}}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
-%\dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m\widehat{\pi}_w + r(1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}.
+%\dfrac{\omega^\text{old}_m K_{\text{old}}}{\omega^\text{old}_w \left(\frac{(1-f_{\text{old}})m e^{\pi_w} K_{\text{new}} + (1-m-f_{\text{old}}(1-m- e^{\pi_w}))K_{\text{old}}}{1-f_{\text{old}}+ e^{\pi_w} f_{\text{old}}} \right)}\, - 1 , & \text{if } \widehat{N}_w^{\text{new}} = K_{\text{new}};  \\
+%\dfrac{\omega^\text{old}_m}{\omega^\text{old}_w \left(1 - \frac{rm(1-f_{\text{old}})}{f_{\text{old}}m e^{\pi_w} + r(1-f_{\text{old}}+ e^{\pi_w} f_{\text{old}}(1-m))}\right)}\, - 1 , & \text{otherwise}.
 %\end{array}
@@
 
-\subsection*{The local per capita growth rate $a_{\text{new}}$}
+\subsubsection{The local per capita growth rate $a_{\text{new}}$}
 In new-habitat patches, we assume that the carrying capacity $K_{\text{new}}$ is not reached during the establishment phase of the mutant.
@@
 \rho &\approx 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{(\xi -1 + \mu_m^{\text{new}})^2 + 2(1-\xi+\mu_m^{\text{new}})\mu_m^{\text{old}} + (\mu_m^{\text{old}})^2  }\right)\\
-&= 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{\frac{\gamma}{1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}}  }\right),
+&= 1+\frac{1}{2}\varepsilon\left(1+ \xi - \mu + \sqrt{\frac{\gamma}{1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}}  }\right),
 \end{aligned}
@@
 \begin{equation}
-\gamma = (1-f_{\text{old}})(\xi-1+\mu)^2 + \widehat{\pi}_m f_{\text{old}} (\xi-1-\mu)^2\, .
+\gamma = (1-f_{\text{old}})(\xi-1+\mu)^2 +  e^{\pi_m} f_{\text{old}} (\xi-1-\mu)^2\, .
 \end{equation}
@@
 \begin{aligned}
-\varphi_{\text{old}} &= \varepsilon + \frac{\varepsilon\left(1-\xi\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}  + \frac{\varepsilon\left(\mu_m^{\text{old}} -\mu_m^{\text{new}} + 2 \mu_m^{\text{new}}\xi  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}\, ,\\
-\varphi_{\text{new}} &= \varepsilon\xi + \frac{\varepsilon\xi \left(\xi-1\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}} + \frac{\varepsilon\left(\xi \mu_m^{\text{new}} -\xi \mu_m^{\text{old}} + 2 \mu_m^{\text{old}}  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})}}}\, .
+\varphi_{\text{old}} &= \varepsilon + \frac{\varepsilon\left(1-\xi\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}  + \frac{\varepsilon\left(\mu_m^{\text{old}} -\mu_m^{\text{new}} + 2 \mu_m^{\text{new}}\xi  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}\, ,\\
+\varphi_{\text{new}} &= \varepsilon\xi + \frac{\varepsilon\xi \left(\xi-1\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}} + \frac{\varepsilon\left(\xi \mu_m^{\text{new}} -\xi \mu_m^{\text{old}} + 2 \mu_m^{\text{old}}  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}\, .
 \end{aligned}
@@
 \begin{equation}
-C = (1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 + \widehat{\pi}_m f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, ,
+C = (1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 +  e^{\pi_m} f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, ,
 \end{equation}
@@
 \begin{aligned}
-\varphi_{\text{old}} &= a_{\text{old}} + \frac{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})a_{\text{old}} \left(a_{\text{old}}-a_{\text{new}}\right)}{\sqrt{C}}  + \\
-&\qquad \qquad \qquad m \frac{\left(\widehat{\pi}_m f_{\text{old}} a_{\text{old}} - (1-f_{\text{old}}) a_{\text{old}} + 2 (1-f_{\text{old}}) a_{\text{new}}  \right)}{\sqrt{C}}\, ,\\
-\varphi_{\text{new}} &= a_{\text{new}} + \frac{(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}})a_{\text{new}} \left(a_{\text{new}} - a_{\text{old}}\right)}{\sqrt{C}} \\
-&\qquad \qquad \qquad + m \frac{\left(a_{\text{new}} (1-f_{\text{old}}) - a_{\text{new}} \widehat{\pi}_m f_{\text{old}} + 2 a_{\text{old}} \widehat{\pi}_m f_{\text{old}}  \right)}{\sqrt{C}}\, .
+\varphi_{\text{old}} &= a_{\text{old}} + \frac{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})a_{\text{old}} \left(a_{\text{old}}-a_{\text{new}}\right)}{\sqrt{C}}  + \\
+&\qquad \qquad \qquad m \frac{\left( e^{\pi_m} f_{\text{old}} a_{\text{old}} - (1-f_{\text{old}}) a_{\text{old}} + 2 (1-f_{\text{old}}) a_{\text{new}}  \right)}{\sqrt{C}}\, ,\\
+\varphi_{\text{new}} &= a_{\text{new}} + \frac{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})a_{\text{new}} \left(a_{\text{new}} - a_{\text{old}}\right)}{\sqrt{C}} \\
+&\qquad \qquad \qquad + m \frac{\left(a_{\text{new}} (1-f_{\text{old}}) - a_{\text{new}}  e^{\pi_m} f_{\text{old}} + 2 a_{\text{old}}  e^{\pi_m} f_{\text{old}}  \right)}{\sqrt{C}}\, .
 \end{aligned}
@@
 \begin{aligned}
-\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}} \qquad \quad \;  \; +  \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
-& \qquad \qquad + \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}\widehat{\pi}_m f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
-\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} +  \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+\widehat{\pi}_m f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
-& \qquad \quad \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} \widehat{\pi}_m f_{\text{old}} - (a_{\text{new}}-a_{\text{old}}) \widehat{\pi}_m f_{\text{old}} \right)\, .}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
+\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}} \qquad \quad \;  \; +  \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
+& \qquad \qquad + \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} e^{\pi_m} f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
+\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} +  \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
+& \qquad \quad \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}  e^{\pi_m} f_{\text{old}} - (a_{\text{new}}-a_{\text{old}})  e^{\pi_m} f_{\text{old}} \right)\, .}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
 \end{aligned}
@@
 \begin{equation}
-a_{\text{new}} \approx \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}}\left(1+\frac{m f_{\text{old}} (1-\widehat{\pi}_w)}{1-f_{\text{old}}+\widehat{\pi}_w f_{\text{old}}}\right)}.
+a_{\text{new}} \approx \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}}\left(1+\frac{m f_{\text{old}} (1- e^{\pi_w})}{1-f_{\text{old}}+ e^{\pi_w} f_{\text{old}}}\right)}.
 \end{equation}

### 2020-07-08 ms/revision.tex
fin premiere grosse revision main text

@@
 We followed the suggestion and added a sentence explaining this relationship (line~\ref{R1-33}).
-
+\florence{(FD Not exactly true for Nnew)}
 \begin{referee}
@@
 
-However, \florence{we do not provide this formula in the manuscript, because the manuscript is already long.}\st{ since we do not want to go into more details of soft selective sweeps (also due to the length of the manuscript), we do not provide this formula in the manuscript.}
+However, \florence{we do not provide this formula in the manuscript, because the manuscript is already long.}\st{ since we do not want to go into more details of soft selective sweeps (also due to the length of the manuscript), we do not provide this formula in the manuscript.} \florence{not soft sweep}
 

### 2020-07-10 ms/main.tex
DL modifs done on overleaf

@@
 the Old-Old and the Old-New dispersal schemes \florence{It does not make intuitive sense why $\widehat{N}_w^{\text{new}}$ is higher with WT biased towards old...
-maybe then figures with $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ would help}.
+maybe then figures with $\widehat{N}_w^{\text{old}}$ and $\widehat{N}_w^{\text{new}}$ would help}\florence{edit: I plotted $\widehat{N}_w^{\text{new}}$ and it is not largest for WT preferring old..
+\url{https://github.com/flodebarre/PeteRescue/tree/master/code}}.
 
@@
 Secondly, the different dispersal schemes affect the number of mutations arising in either habitat type.
-This has a minor effect on the probability of evolutionary rescue for the explored parameter range but is relevant when studying the origin of the successful mutant lineage \chg{(see also Fig.~\label{Sfig:natal_habitat})}.
+This has a minor effect on the probability of evolutionary rescue for the explored parameter range but is relevant when studying the origin of the successful mutant lineage \chg{(see also Fig.~\ref{Sfig:natal_habitat})}.
 As the genetic background may vary across patches, the origin of a successful mutation will also affect which mutations will hitchhike with it.
@@
 
+\florence{[add here figure of $\widehat{N}_w^k$ as function of $m$]}
+
 \subsection{Wild-type population sizes after dispersal}
@@
 We then replace $\widehat{N}_w^{\text{old}}$ by $K_{\text{old}}$ (since old-habitat patches are assumed to be at carrying capacity after density regulation), and $\widehat{N}_w^{\text{new}}$ by the formula given in eq.~\eqref{Seq:Nhatnew}.
+
+\florence{[add here multi-panel figure of $\tilde{N}_w^k$ as function of $m$]}
 
@@
 
-The mean reproduction matrix $\mathcal{M}$ of a mutant gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides (see also eq.~\eqref{eq:mean_repro} in the main text):
-\begin{equation}\label{eq:repromatrix}
+The mean reproduction matrix $\mathcal{M}$ of a mutant gives the average number of offspring in a certain habitat, dependent on the habitat type in which the mutant resides (see also eq.~\eqref{eq:lambda} in the main text):
+\begin{equation}\label{Seq:mean_repro}
 \mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
@@
 \end{equation}
-\begin{equation}\label{Seq:mean_repro}
-\mathcal{M} = \bordermatrix{ ~ & \text{old patch} & \text{new patch} \cr
-	\text{old patch} & \left(1-m_m^{\text{new}}\right) (1+a_{\text{old}}) & m_m^{\text{new}} (1+a_{\text{new}}) \cr
-	\text{new patch} & m_m^{\text{old}} (1+a_{\text{old}}) & \left(1-m_m^{\text{old}}\right) (1+a_{\text{new}})
-}\ ,
-\end{equation}
 where the rows denote the parent locations, and the columns the patch type of the offspring.
 
-Our goal is to apply Theorem 5.6 from \citet{haccou_book} which states that for a slightly super-critical branching process, \chg{i.e.
-where the survival probability is slightly above zero}, the establishment probability can be expressed in terms of the largest eigenvalue $\rho$ and the corresponding left- and right-eigenvectors of the mean reproduction matrix \chg{$\mathcal{M}$}, denoted by $u$ and $v$, respectively.
+Our goal is to apply Theorem 5.6 from \citet{haccou_book} which states that for a slightly super-critical branching process, i.e.
+where the survival probability is slightly above zero, the establishment probability can be expressed in terms of the largest eigenvalue $\rho$ and the corresponding left- and right-eigenvectors of the mean reproduction matrix $\mathcal{M}$, denoted by $u$ and $v$, respectively.
 The eigenvectors should be normalized in the following way: $u_1+u_2 = 1$ and $\sum_{i=1}^2 u_i v_i = 1$.
@@
 %
-For $\varepsilon\to 0$ we find that $\rho\to 1$ \chg{(eq.~\eqref{eq:eigenvalue}) which means that the branching is slightly super-critical if $\rho>1$ and real.
-A sufficient condition for this to be true is}
+For $\varepsilon\to 0$, we find that $\rho\to 1$ (eq.~\eqref{eq:eigenvalue}) which means that the branching is slightly super-critical if $\rho>1$ and real.
+A sufficient condition for this to be true is that the first term in the parentheses of \eqref{eq:eigenvalue} is positive:
 \begin{equation}
@@
 %
-\chg{In case that the branching process is not super-critical the establishment probability in eq.~\eqref{Seq:theory} becomes negative and as such is not a probability anymore.
-Hence, we can simply reject negative solutions of the establishment probability and by that implicitly justify that our approximation is valid.}%\\
+In case that the branching process is not super-critical the establishment probability in eq.~\eqref{Seq:theory} becomes negative and as such is not a probability anymore.
+Hence, we can simply reject negative solutions of the establishment probability and by that implicitly justify that our approximation is valid.%\\
 
@@
 \subsection{Computing the establishment probability}
-For the solution of eq.~\eqref{Seq:theory} it remains to compute the normalized eigenvectors.
+Eq.~\eqref{Seq:theory} involves normalized eigenvectors.
 Their precise form is of not much insight.
 We therefore omit stating them explicitly but refer to the \textit{Mathematica} notebook.
-Solving eq.~\eqref{Seq:theory} to the first order of $\varepsilon$ we find
-
-\begin{equation}
-\begin{aligned}
-\varphi_{\text{old}} &= \varepsilon + \frac{\varepsilon\left(1-\xi\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}  + \frac{\varepsilon\left(\mu_m^{\text{old}} -\mu_m^{\text{new}} + 2 \mu_m^{\text{new}}\xi  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}\, ,\\
-\varphi_{\text{new}} &= \varepsilon\xi + \frac{\varepsilon\xi \left(\xi-1\right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}} + \frac{\varepsilon\left(\xi \mu_m^{\text{new}} -\xi \mu_m^{\text{old}} + 2 \mu_m^{\text{old}}  \right)}{\sqrt{\frac{\gamma}{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})}}}\, .
-\end{aligned}
-\end{equation}
-%
-Transforming back to the original variables and replacing $\gamma$ by the constant $C$ from the main text (eq.~\eqref{eq:normalization})
-\begin{equation}
-C = (1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}) \left((1-f_{\text{old}})(a_{\text{new}}-a_{\text{old}}+m)^2 +  e^{\pi_m} f_{\text{old}} (a_{\text{new}}-a_{\text{old}}-m)^2\right)\, ,
-\end{equation}
-we obtain
-
-\begin{equation}
-\begin{aligned}
-\varphi_{\text{old}} &= a_{\text{old}} + \frac{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})a_{\text{old}} \left(a_{\text{old}}-a_{\text{new}}\right)}{\sqrt{C}}  + \\
-&\qquad \qquad \qquad m \frac{\left( e^{\pi_m} f_{\text{old}} a_{\text{old}} - (1-f_{\text{old}}) a_{\text{old}} + 2 (1-f_{\text{old}}) a_{\text{new}}  \right)}{\sqrt{C}}\, ,\\
-\varphi_{\text{new}} &= a_{\text{new}} + \frac{(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}})a_{\text{new}} \left(a_{\text{new}} - a_{\text{old}}\right)}{\sqrt{C}} \\
-&\qquad \qquad \qquad + m \frac{\left(a_{\text{new}} (1-f_{\text{old}}) - a_{\text{new}}  e^{\pi_m} f_{\text{old}} + 2 a_{\text{old}}  e^{\pi_m} f_{\text{old}}  \right)}{\sqrt{C}}\, .
-\end{aligned}
-\end{equation}
-%
-Slightly re-ordering the terms, this gives the establishment probability of a single mutant individual, eq.~\eqref{eq:estab_approx} in the main text:
-\begin{equation}\label{Seq:estab_approx}
-\begin{aligned}
-\varphi_{\text{old}} &\approx \qquad \quad \; \; \; \; a_{\text{old}} \qquad \quad \;  \; +  \quad a_{\text{old}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{old}}-a_{\text{new}}) \\
-& \qquad \qquad + \frac{m}{\sqrt{C}} \left(a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}} e^{\pi_m} f_{\text{old}} - (a_{\text{old}}-a_{\text{new}})(1-f_{\text{old}})\right) \, ,\\
-\varphi_{\text{new}} &\approx \underbrace{ a_{\text{new}}}_{\text{(1) local growth parameter}} +  \quad \underbrace{ a_{\text{new}} \frac{\left(1-f_{\text{old}}+ e^{\pi_m} f_{\text{old}}\right)}{\sqrt{C}}(a_{\text{new}}-a_{\text{old}})}_{\text{(2) effect of the heterogeneous environment}} \\
-& \qquad \quad \underbrace{+\; \frac{m}{\sqrt{C}}\left( a_{\text{new}}(1-f_{\text{old}}) + a_{\text{old}}  e^{\pi_m} f_{\text{old}} - (a_{\text{new}}-a_{\text{old}})  e^{\pi_m} f_{\text{old}} \right)\, .}_{\text{(3) effect of dispersal: new patches $+$ old patches $-$ loss to the other patch type }}
-\end{aligned}
-\end{equation}
-
-\chg{For $m=0$ we see that $\varphi_{\text{old}}=0$, i.e.
-terms (1) and (2) cancel out.
-For the establishment probability in the new habitat we recover Haldane's result for the establishment probability of a slightly advantageous mutant: $\varphi_{\text{new}}=2a_{\text{new}}$ \citep{haldane_1927}.}
-
-\subsection{Disentangling the contributions to the establishment probability \label{sec:SI_contributions}}
+Solving eq.~\eqref{Seq:theory} to the first order of $\varepsilon$ and transforming back to the original variables, we obtain after reordering the equation presented in the main text (eq.~\eqref{eq:estab_approx})
+
+\subsection{Disentangling the contributions to the establishment probability\label{sec:SI_contributions}}
+
+\florence{I know you spent a lot on time on this, so I am not lightly making the suggestion but...
+I think this subsection could be removed.
+Intuitive explanations are given in the main text.
+The separation in three terms is not super intuitive, and their interpretation is more complicated that may seem, because $a_{\text{old}}$ is not constant.
+Plotting $a_{\text{old}}$ already helps a lot}
 \chg{We now proceed to explain} the three regions of the establishment probability from Fig.~\ref{fig:vary_m_est}(a) in the main text.
@@
 	\end{subfigure}
-	\caption{\textbf{Contribution of the different terms in eq.~\eqref{Seq:estab_approx} to the establishment probability $\varphi_{\text{old}}$.} Subfigures (a,b) are the same as Figs.~\ref{fig:vary_m_est}(a,c) in the main text.
+	\caption{\textbf{Contribution of the different terms in eq.~\eqref{eq:estab_approx} to the establishment probability $\varphi_{\text{old}}$.} Subfigures (a,b) are the same as Figs.~\ref{fig:vary_m_est}(a,c) in the main text.
 They show the establishment probability for a single mutant individual arising in an old-habitat patch for varying emigration probabilities $m$.
-In subfigures (c,d) we plot the terms from eq.~\eqref{Seq:estab_approx} separately ($\pi_w=0.5,\pi_m=-0.5$).
+In subfigures (c,d) we plot the terms from eq.~\eqref{eq:estab_approx} separately ($\pi_w=0.5,\pi_m=-0.5$).
 Term 1, the mutant growth rate in old habitats (solid), increases with increasing dispersal rates as a consequence of relaxed competition.
@@
 For clarity, we re-plot Figs.~\ref{fig:vary_m_est}(a,c) in Fig.~\ref{Sfig:contribution}(a,b), respectively.
-We try to explain the ongoing processes for each of the regions through the approximations of the establishment probabilities in eq.~\eqref{Seq:estab_approx}, see also Fig.~\ref{Sfig:contribution}(c,d).
+We try to explain the ongoing processes for each of the regions through the approximations of the establishment probabilities in eq.~\eqref{eq:estab_approx}, see also Fig.~\ref{Sfig:contribution}(c,d).
 Note that these explanations are only valid for the establishment probability of a mutant initially in an old-habitat patch, $\varphi_{\text{old}}$.
@@
 Mutants disperse from old- to new-habitat patches where they \chg{have a higher growth rate}.
-This effect is mediated through the third term of the establishment probability in eq.~\eqref{Seq:estab_approx}.
+This effect is mediated through the third term of the establishment probability in eq.~\eqref{eq:estab_approx}.
 \chg{Note that the first two terms of the approximation cancel out for small emigration probabilities $m$.}
-While the third term increases with increasing emigration rate $m$, the second term in eq.~\eqref{Seq:estab_approx} decreases, cf. Fig.~\ref{Sfig:contribution}. 
+While the third term increases with increasing emigration rate $m$, the second term in eq.~\eqref{eq:estab_approx} decreases, cf. Fig.~\ref{Sfig:contribution}. 
 In the formula this is mediated through \chg{the increase of the local growth rate in the old-habitat, $a_{\text{old}}$, i.e.\ it becomes less negative.
@@
 This results in even less competitive pressure in old-habitat patches.
-Eventually, this yields a positive growth rate $a_{\text{old}}$ (first term in eq.~\eqref{Seq:estab_approx}).
+Eventually, this yields a positive growth rate $a_{\text{old}}$ (first term in eq.~\eqref{eq:estab_approx}).
 Therefore, this region is driven by the local growth rate in old habitats.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
+\clearpage
 \renewcommand{\theequation}{C\arabic{equation}}
@@
 
+\florence{I don't understand this section, which does not do much besides announcing a plot... Remove? Unless AmNat wants figures in appendix sections (in which case all figures need to be moved)?}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Probability of establishment for large frequencies of old-habitat patches}
+\florence{The figure is not referred to in the main text. Is this because we can remove it...?}
 \chg{We plot the probability of establishment for a large frequency of old-habitat patches, $f_{\text{old}}$.
@@
 
+\florence{y axis label in panel b needs to be corrected}
 \begin{figure}[h!]
@@
 The dispersal process and the dynamics in old-habitat patches remain as studied before.
-In new-habitat patches we now assume that the population remains at carrying capacity, i.e.
-there is no longer a declining wild-type population.
-This means that the local growth rate $a_{\text{old}}$ in eq.~\eqref{Seq:s_old} takes the form for $\widehat{N}_w^{\text{new}}=K_{\text{new}}$.
-For simplicity we will also assume that $K_{\text{new}}=K_{\text{old}}=K$.
-In order to maintain the divergent selection assumption, we assume that the fecundity of the wild-type in the new habitat is below the fecundity of the mutant.
-%For simplicity, we assume that the fecundity values are given as
-%\begin{equation}
-%    \omega^\text{old}_w^{\text{old}} = \omega^\text{old}_m^{\text{new}} \quad \text{and} \quad \omega^\text{old}_w^{\text{new}} = \omega^\text{old}_m^{\text{old}}.
-%\end{equation}
-Therefore, we need to adjust the local growth rate of a single mutant, $a_{\text{new}}$. The local growth rate in old habitats, $a_{\text{old}}$, remains as outlined in eq.~\eqref{Seq:s_old}. We have
+We now assume that the population remains at carrying capacity in new-habitat patches, i.e.
+there is no longer a declining wild-type population, so that the stationary wild-type population size is $\widehat{N}_w^{\text{new}}=K_{\text{new}}$.
+For simplicity we will also assume that $K_{\text{new}}=K_{\text{old}}=K$\florence{maybe there would not be the final increase in purple and orange if you used the original K values?}.
+Since there is now density regulation in new habitat patches, we need to define again the per capita growth rate of the mutant in new-habitat patches:
 \begin{equation}
@@
 \end{equation}
-which with the help of eq.~\eqref{Seq:wt_eq1} yields
-\begin{equation}
-a_{\text{new}} \approx \frac{\omega_m^{\text{new}}}{\omega_w^{\text{new}}\left(1+\frac{m f_{\text{old}} (1- e^{\pi_w})}{1-f_{\text{old}}+ e^{\pi_w} f_{\text{old}}}\right)}.
-\end{equation}
-
-\chg{Note, that we again used that during the establishment phase the wild type is much more abundant than the mutant which explains the approximation in the two equations.}
-Plugging this in the approximation of the establishment probability from eq.~\eqref{Seq:estab_approx} we find the curves in Fig.~\ref{Sfig:pop_gen}.
-
-We see that, as briefly mentioned in the main text, region (iii) of the establishment probability disappears in these type of models except for the Old-New and the New-Old dispersal schemes.
+where we replace $\widetilde{N}_w^{\text{new}}$ by the formula provided in eq.~\eqref{Seq:Ntildenew}, with $\widehat{N}_w^{k}=K_{k}$.
+
+Once again, we assume that the wild type is much more abundant than the mutant during the establishment phase.
+We finally plug this new formula of $a_{\text{new}}$ in the formulas for the establishment probabilities.
+
+\florence{I stopped here}We see that, as briefly mentioned in the main text, region (iii) of the establishment probability disappears in these type of models except for the Old-New and the New-Old dispersal schemes.
 The reason for the disappearance of the region is that relaxed competition only plays a subordinate role for the symmetric dispersal schemes (Old-Old, New-New and 0-0).
@@
 
-\begin{figure}[h!]
+\begin{figure}[h]
 	\centering
@@
 The analytical approximation (dashed lines) however deviates strongly from the data due to large growth rates ($a_{\text{new}}\approx 0.2$) so that the conditions for the approximation to hold are violated.
-In this case, in eq.~\eqref{Seq:estab_approx} higher order corrections would need to be taken into account.
+In this case, in eq.~\eqref{eq:estab_approx} higher order corrections would need to be taken into account.
 \chg{The fecundity values in the new habitat are given by $\omega_w^{\text{new}} = \omega_m^{\text{old}}$ and $\omega_m^{\text{new}} = \omega_w^{\text{old}}$ and the carrying capacity is $K_{\text{new}}=K_{\text{old}}=K=500$.} Missing data points (mostly for the negative density-dependent dispersal scheme -- green triangles) are explained by too large computation times.
@@
 The habitat type of the origin of the rescue mutation is largely independent of the considered dispersal scheme.
-For $\omega^\text{old}_m=1.35$ we have plotted the relative contribution of each natal habitat type to the probability of evolutionary rescue, Fig.~\ref{Sfig:natal_habitat}.
-We do not see large differences between the three symmetric dispersal schemes (0-0, Old-Old, and New-New) and the Old-New dispersal scheme.
-\chg{Only for the New-Old dispersal schemes we see an increase in the contribution of mutants emerging from old-habitat patches for very high emigration probabilities $m$.
+Fig.~\ref{Sfig:natal_habitat} shows the relative contribution of each natal habitat type to the probability of evolutionary rescue (with $\omega^\text{old}_m=1.35$).
+The establishment probabilities are close among the three symmetric dispersal schemes (0-0, Old-Old, and New-New) and the Old-New dispersal scheme.
+The New-Old dispersal scheme differs a bit: the contribution of mutants emerging from old-habitat patches increases for very high emigration probabilities $m$.
 A possible explanation is the large probability of establishment for a mutant emerging in old-habitat patches for this parameter set (cf.
-Fig.~\ref{Sfig:contribution}(a)).}
+Fig.~\ref{Sfig:contribution}(a)).
 
@@
 	\includegraphics[width=0.6\textwidth]{figS7.pdf}
-	\caption{\textbf{Habitat type of the origin of the rescue mutant dependent on the dispersal scheme.} Varying the emigration probability $m$ we plot the relative contributions of each habitat type to the probability of evolutionary rescue.
+	\caption{\textbf{Habitat type of the origin of the rescue mutant depending on the dispersal scheme.} Relative contributions of each habitat type to the probability of evolutionary rescue, as a function of the emigration probability $m$.
 The color-coding is as in the main text: black for 0-0, blue for the Old-Old, green for the New-New, orange for the Old-New, and purple for the New-Old dispersal scheme.}
